### 2017-11-07 vipassana-for-hackers.tex
switching to latex...fuck pages

\documentclass{article}

\usepackage{hyperref}
\usepackage{graphicx}

\begin{document}

\begin{titlepage}
   \vspace*{\stretch{1.0}}
   \begin{center}
     \Huge\textbf{Vipassana for Hackers}\\
     \Large\textbf{Instigating the Curious}\\
     \vspace{5cm}
     \large\textit{Steven Deobald}\\
     \large\textit{Version 1.1}\\
     \large\textit\today
   \end{center}
   \vspace*{\stretch{2.0}}
\end{titlepage}

\begin{center}
  \Huge{Target Audience}
\end{center}

Should you bother reading this paper? Not everyone associates themselves with the label ``Hacker''. It comes with a few well-known negative connotations so my first task is to quickly identify what is meant by ``Hacker'', since this paper is for these people.

My first exposure to the term ``hacking,'' outside of the late-80's imagery which became fodder for the 1995 film, was through my cousin Jennie. She is an Ecologist, a self-described ``Mud Scientist'', who most people would not associate with the term ``Hacker''. She visited MIT during her PhD to find the term extended to ``roof and tunnel hacking,'' an idea and practice which delighted her to no end. MIT or not, Jennie is a Hacker. Every Science is about exploration and discovery, and these are the concepts at the root of ``hacking.''

Carpentry, painting, gardening, the entire field of Genetics, baking bread, Philosophy, electronics, fashion, Psychology, health, writing, and bicycles. Almost any discipline can fall under the banner of ``hacking'' I'm using for this paper, since almost any discipline can (but does not necessarily) include the potential for two essential aspects of the Hacker Mentality: Rationality and Curiosity.

This is all that is required.

If you have ever enjoyed taking apart a computer or car, experimenting with bread recipes, researching and writing, pulling apart the semantics of your work, building something, rebuilding something, exploring a cave or abandoned construction site... I wrote this paper for you. If you enjoy any exploratory vocation or hobby you will almost certainly find Vipassana utterly fascinating. The reason is this: As acutely as a human being can disassemble a stereo, examine the slide of a microscope, or explore a cave, human beings have the capacity to disassemble, examine, and explore ourselves.

What may be even more fascinating, for the hacker, is the fact that by practicing Vipassana one can seemingly increase the capacity for rational thought and behaviour.

I have written this paper because, thus far, I have found the written and recorded material on the subject of Vipassana to fall into three broad categories:

Well-meaning Neuroscience papers which completely miss the mechanics of nervous-system-based meditation techniques and their core outcomes.
Descriptions of a ``silent meditation retreat,'' usually given by someone who has not attended a course. [1]
Descriptions of Vipassana from an individual and subjective perspective, usually given by someone who has attended one course. (``Oh my god! It was sooooo intense!'')

By using phrases like ``disassemble, examine, and explore ourselves'' I'm at the risk of sounding spiritual. This is the opposite of my intention. My intention is to pique curiosity in fellow hackers by explaining the biological mechanics of Vipassana, to the best of my ability and as I understand them. To curb the impression that this paper falls into the category of New Age Philosophy, I would like to take a moment to address what this paper is not.

\pagebreak

\begin{center}
  \Huge{What This Paper is Not}
\end{center}

This not an instruction manual. You will not learn Vipassana from reading this paper.

This is not a white paper. I am not a scientist. This is not science.

This is not an expert's treatise on the subject of nervous-system-based meditations, nor even an expert Vipassana meditator's description of the technique. I am not an expert. I have only attended four 10-day courses thus far. Each has been substantially different from the others.

This paper will endeavour to avoid falling into the trap of ``Buddhist Guys'' (a mildly derogatory term coined by a serious meditator friend of mine). You might be familiar with phrases commonly attributed to this archetype such as ``When the student is ready, the teacher will appear.'' Mysticism, mystery, and mystique are to be avoided. If I fall into this trap, email me: steven@nilenso.com

This is not a discussion on ``spirituality'', for any definition of the term.

This is not a dissection of the administrative details of the course, though I will provide a brief overview of the course schedule.

This is not a description of past experiences, though I will use a few to highlight core effects of the technique.

This is not a comparison with other meditation techniques, though I will address other techniques for context. It is worth noting that there are a few meditation techniques labelled “Vipassana” and this paper deals specifically with Vipassana as taught by S.N. Goenka.

This is not an examination of Buddhist Philosophy and will exclude commentary or mention of the following concepts:

\begin{itemize}
  \item Siddhartha Gautama (The Buddha)
  \item Enlightenment
  \item Religion / Buddhism
  \item Prophecy / Prediction of the future
  \item Telepathy / ESP
  \item Reincarnation
  \item ``Energies''
  \item Kalapas
  \item Metta / Morality
  \item Sankharas / ``Impurity''
\end{itemize}

These last two bullet points are subject to one exception each.

\pagebreak

\begin{center}
  \LARGE{Morality's Exception: Lying}
\end{center}

A 10-day Vipassana course requires that all students take the standard ``5 Precepts'' [2], as per Theravada. I do not recommend getting drunk, raping and murdering your fellow students, and stealing their belongings. However, the four precepts which forbid those behaviours are quite clearly external, making them components of course administration and therefore not applicable to this document.

Lying, however, is perhaps least interesting in its external form. As an educated adult, it is difficult to say anything at all without lying to some small degree. As one begins to systematically observe the nervous system, which is the basic mechanic of Vipassana, one finds that the brain and nervous system are acutely aware of even the tiniest untruth. If I tell myself I am feeling something which I am not (``my body is exploding into a billion tiny particles! my muscles are raging rivers of lava inside my body!'') the process of empirical and objective observation ends, and the process stops.

Therefore, the exception given to Lying (or the precept governing Lying) is because it actually prevents the technique from working, not because I have some particular crusade against lying.

\begin{center}
  \LARGE{Habit's Exception: Sankhara}
\end{center}

The notion of Sankhara can be split into two camps, in terms of its definition and use. The first, the notion that the body and mind contain ``impurities'' and that the objective of meditation is to rid the body-mind of such impurities, will be excluded from all discussion here. The second, more common use of the term ``Sankhara'' refers to ``conditioned phenomena'' or ``mental dispositions''. The Hindi word, संस्कार (``sanskaar'') roughly translates in this context to ``impression''. This second definition will apply to this paper.

Here, we might choose a more specific term from Psychology: habits. Sankharas, as far as this document is concerned, can be thought of as the ingrained habits of the brain and nervous system. Some concrete examples include liking (or disliking) the taste of olives, my body's ability to type at this QWERTY keyboard without consciously considering each keystroke, and the physical and emotional response I have to news of violence in my city of residence.

\pagebreak

\begin{center}
  \Huge{A Finger Pointing at the Moon}
\end{center}

The fundamental difficulty in any description of Vipassana is this: It is impossible to convey a description of Vipassana in terms known to someone who has not experienced it. During a Vipassana course this paradox is addressed using two metaphors: First, a story of a sighted brother explaining the concept of ``white'' to his blind brother [3]. Second, the classic Indian story of blind men each describing an elephant and each producing a partial truth.

The first metaphor is a better framework for explaining the difficulty of writing about nervous system meditations. It is fundamentally impossible to express the experience of observing ``whiteness'' to an individual who is blind from birth. This person requires sight to understand. Similarly, it is fundamentally impossible to describe the effects of a nervous system meditation which alters perceptions and the interactions of the nervous system itself, since the nervous system (including the brain) ultimately form our entire experience.

The hacker community has been on the verge of this exploration for quite some time. The easiest and most recent examples of this are no doubt the works of Douglas Hofstadter. ``Gödel, Escher, Bach: An Eternal Golden Braid'' captured the imaginations of the hacker community and grew from esoteric pub conversation piece to dinner party staple, around the same time as the term ``blogging'' evolved from portmanteau to an accessible part-time career for stay-at-home parents. The term ``strange loop'', coined in this book, has since become a mild obsession of philosophical and artistic computer scientists, linguists, and mathematicians, even inspiring the namesake software conference held in St. Louis. Hofstadter's more cohesive ``I Am a Strange Loop'' tackles the problem of consciousness more succinctly. However, other than directing the reader to Zen \footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana. The reason I do not discuss Zazen in detail or compare it to Vipassana in this paper is because Zazen, in my experience, is not systematic. The subject of inspection — the nervous system — remains, but Zen teachers tend to avoid explicit instruction, leaving the student to explore on her own. The outcomes, therefore, appear similar but watered down.} at multiple points, these books do not invite the reader to explore these concepts directly. There is no hacking. A thorough read of GEB or IAaSL is akin to reading all four volumes of Knuth's ``The Art of Computer Programming'' without having ever touched a computer or to reading ``The Joy of Cooking'' cover to cover without having ever set foot inside a kitchen.

\pagebreak

\begin{center}
  \Huge{Computing Sidebar: LISP}
\end{center}

I will attempt to frame the self-exploration of nervous system meditations like Vipassana using one tool familiar to the Computer Science discipline: LISP. If you are neither a Computer Scientist nor a Linguist and have no interest in computer languages, feel free to skip to the next section.

Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' [4] and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely. LISP has some very interesting properties: Homoiconicity, metaprogramming, self-mutation and self-replication, recursion and trampolining, and — in the case of modern LISPs and other languages with strong concurrency models — hardened notions of immutability and the reification of time.

Combined, these properties tend to blur the lines of compile time and run time, of code and data, of execution and interpretation. Decades ago, thanks to these ``strange loop'' properties of LISP, it was thought to herald the dawn of the Age of Artificial Intelligence. While that did not happen it left us with a fascinating building material: LISP appears to have no upper-bound. It is ``turtles all the way up.'' LISP is not ``turtles all the way down'', as some would describe it. Even if written on a LISP Machine, there are still paths of wires and electrons which need to be dealt with. These are not written in LISP.

Vipassana would appear to be ``turtles all the way down'' in precisely this sense. Vipassana is the process of digging through your consciousness using consciousness itself. At no point do you reach the machine beneath the machine: ``Aha! I found it! The Consciousness Machine. Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' The machine at the bottom of your consciousness is a strange loop. Having a look at this strange loop is a lot more fun than reading about it.

\pagebreak

\begin{center}
  \Huge{Course Structure}
\end{center}

For those unfamiliar, I will briefly overview the rules of the course. During the course you are not permitted books, phones, talking, looking people in the eye, touching, making sounds, snacks, music, or writing implements.

Put more succinctly, your activities will include: Eat, sleep, poop, bath, meditate.

That's it.

Meditation will consume approximately 10 to 11 hours per day.

\pagebreak

\begin{center}
  \Huge{Suspending Disbelief}
\end{center}

Before explaining the mechanics and outcomes of Vipassana, I must request that you suspend disbelief about three items for the duration of the paper.

\begin{center}
  \LARGE{1. Meditation is a thing}
\end{center}

For the purposes of this document, understand ``meditation'' to mean a single point of awareness with unwavering focus. ``Unwavering focus'' means that we can hold sustained attention on this single point (whatever it may be) without getting distracted by thoughts of our taxes, what's for lunch, or that really persistent itch on our ear.

What I mean by this is that mental exercises exist which can take consciousness beyond ``flow'', ``calm'', or ``elated'' states of mind caused by activities one might describe as ``meditative'' rather than ``meditation''. Biochemical responses to hallucinogens fall on the same side of the spectrum, though they may appear momentarily insightful.

I have drawn my personal spectrum for sample activities below.

\begin{figure}[h]
  \includegraphics[width=\linewidth]{meditation-spectrum-1.jpeg}
  \caption{Sample meditation spectrum.}
  \label{fig:meditation-spectrum}
\end{figure}

For the purposes of this paper, we will keep all notions of “mindfulness” in the “meditative” camp, on the right. Mindfulness is not meditation. This paper will be discussing only two practices from this list: Anapana and Vipassana. An illustrated definition of “awareness” and “attention” follows, in the section entitled “On Attention”.

2. Your brain and nervous system have a tight feedback loop

When we say “tight feedback loop”, this means the brain and nervous system are tied together in a feedback loop faster and tighter than you could ever imagine without experiencing it directly.

We all know the classic examples of connectivity between the brain and nervous system: A child touches a hot stove and retracts her hand. A pubescent boy sees a boy or girl he is attracted to and feels ``butterflies in his stomach'' so intense he vomits. Despite healthy diet and exercise, a high-strung individual suffers from stomach ulcers.

These examples all exist on a gross level, coarse-grained enough that we can identify them externally and even give them names like ``butterflies''. However, your nervous system is interacting with your brain at every conceivable level. No matter how small, you do not have a thought or emotion emerge in your consciousness (or unconscious) without a physical sensation accompanying it. You do not feel physical sensation, consciously or subconsciously, without an accompanying thought or emotion. This interaction is continuous.

Unless you have tried a meditation system with the nervous system as its meditation object, you have almost undoubtedly not experienced this.

3. Dissolution of the body
is a thing

This will be the most difficult of the three. ``Dissolution of the body,'' though strange-sounding at first, does not fall into the categories where we would normally place magic, levitation, or ESP. The term simply refers to our capacity to see through our entire body.

As we investigate this concept, it seems increasingly reasonable in conjunction with points \#1 and \#2. If we can narrow our attention down to anything we can perceive and what we choose to perceive with our brain and nervous system in tandem is a narrow part of the body, surely if we sustain attention on that part for long enough we will start to see it clearly.

Normally, our attention on our body is almost entirely external: my hands on this keyboard, the itch on my ear, my butt in this chair. I don't tend to spend a lot of energy focused on the sensations provided to my brain by my gallbladder or my stomach unless I have a painful gallstone or I am experiencing acute hunger. Physical sensations provided to our awareness by the nervous system for the internals of the body are largely hidden from us — but they can be unearthed and explored.

I should note here that in four courses I have not felt the precise shape of my gallbladder or stomach at any point. I have ``seen'' the shape of my bones in my fingers, shoulders, chest, and pelvis. I have witnessed the rush of blood squooshing through my cardiovascular system (near the surface) with terrifying clarity. I have enough anecdotal evidence to say that it appears this process does work and can be extended to our deepest internal organs and the spinal cord.

To help visualize what I mean, a sample spectrum of sensations from the gross or obvious to the curious and subtle might look like this:

As the field of sensation progresses to the left, the sensations felt become increasingly difficult to describe. “Surreal hot / cold” refers to sensations of temperature we wouldn’t normally consider possible: our muscles either on fire or freezing solid. “Spatial distortion” refers to complete alterations of spacial presence — the partial or complete replacement of one’s normal sense of proprioception. This may include reorientation in space (“I was upside-down!”) or the feeling that one is gigantic or tiny or that any one body part has one of these or other distorted qualities.

“Throbbing” refers to thick, heavy, slower vibrations; these are often felt rhythmically and this rhythm accelerates until the sensation is described as a “vibration”. Beyond vibration is the “invisible”, where a particular body part is no longer seen and a deeper part of the body is experienced directly. The first such example in most cases tends to be the skin, since the external surface of the body is the first part of the nervous system the course instructions ask the meditator to focus on. Beyond this the meditator may find muscle tissue, bones, the cardiovascular system, or something else entirely. Usually by the point in the course when a meditator has dissolved her own skin to her attention the sensations become extremely difficult to identify and label so “something else entirely” is likely to be the most common case.

The sample spectrum above is missing many, many sensations. I also feel it necessary to call out that the spectrum is not serial; although I have felt “vibrations” and seen deeper than skin level, I have certainly not experienced all the sensations possible in the spectrum. However, the diagram hopefully provides a reference point for the type of readily-accessible sensations which Vipassana begins with (on the right) and the types of sensations (or absence of sensation) which will convince a person that complete dissolution of the body not only possible but an entirely natural process.

Virtual Reality
and
The Matrix Paradox

The following section will require some visual assistance, for which I will use the same slides I used in the talk which corresponds to this paper.

They look like this:

This image is taken from a talk by Michael Abrash, Chief Scientist of the Oculus VR team, at a talk he gave at Facebook F8 in 2015 [5]. I'm using this slide because it's nicely laid out and because it is incomplete.

Abrash discusses the nature of the ``Matrix Paradox'' [6] as simply the right neurons firing to simulate any particular environment. ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input. The following might be a slightly better model of reality:

Though a classic sense diagram such as Mr. Abrash’s might describe proprioception [7, 8], nocioception [9], and hunger [10] as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain. Touch is strictly an external sense dealing with skin. These are internal senses.

A brain diagram describing The Matrix Paradox is complete only when we map all senses to the brain. I have not, since I haven't named them all. In fact, neurobiology does not yet have names for all sensory input, making such an exercise exceedingly difficult. Proprioception is clearly associated with the cerebellum and hunger is clearly associated with the hypothalamus. Nocioception, however, is a bit more nuanced with regard to the brain's reception of thermal, mechanical, and chemical information — and whether or not the activity is inbound, outbound, conscious, or unconscious. However, assuming all nervous system activity does have corresponding brain activity, one could theoretically map out the Matrix Paradox on the brain as Abrash has started. It seems likely such a diagram would highlight all parts of the entire brain.

On that note, I feel I should introduce an absolutely fantastic RadioLab episode (``Where Am I?'' [11]) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience. On the topic of lacking proprioception: “We have words like ‘deaf’ and ‘blind’. We do not have words for being deaf or blind to one’s own body.” In the Matrix, perhaps this is either a bug or a feature flag.

Now that we have our diagram for visualizing the use of our senses, we can address them in terms of various altered mental states. Before we get there, however, we will try to come up with a meaningful visual definition of “attention” or “awareness” with respect to our senses.

On Attention

The funny thing about Vipassana is that you are doing it already... you're just terrible at it. Vipassana is the inspection of one's own nervous system with narrowly-focused attention but the precision of this focus is not prescribed by the instruction given for the technique. “As narrow as possible”, perhaps, but in day to day life a normal person cannot possibly narrow her attention to the size of a fingertip. Your attention is, in part, already on your nervous system, with some degree of focus. To the extent that it is, you are doing Vipassana.

If your attention is not directed at your nervous system, where is it? Let's look at three environmental examples which can help us highlight our senses and our awareness within them.

Example 1:
A bustling restaurant

...and a corresponding sense map something like this:

Assuming we've just placed food in our mouths, our sense of taste is going off like crazy, we're hearing a lot of scattered sound around us, and we're also sufficiently distracted by most of our other senses. Note that we are probably paying very little attention to the internal senses.
Example 2:
A sensory deprivation tank

The corresponding sense map is on the right.

Here, the classic 5-sense model fails us, since sensory deprivation only deprives us of these 5 senses. When Neo is first exposed to the reality of The Matrix, the pods which imprison the humans loosely resemble such sensory deprivation tanks, which is reason enough to believe that we are not in that matrix. As anyone who has ever entered a sensory deprivation tank knows, it is not like being in a coma. The mind is still very active and it may actually help in causing someone to pay attention to their internal senses.

Example 3:
An anechoic chamber

The corresponding sense map is again on the right.

The map for an anechoic chamber may seem surprising. However, just because the room you are in has a negative decibel rating does not mean you cannot hear. In fact, in this environment, your sense of hearing is likely to be amplified — and turned inward. You will hear your heart beating, your blood pumping, your digestive system sloshing around. The better the anechoic chamber at swallowing sound, the more intense this experience is going to be.

For this reason, the above anechoic chamber has a record of 45 minutes for the longest period anyone has sat inside it. As it turns out, hearing one's own bodily functions at deafening internal volumes is likely to cause someone to go a bit crazy, hence the highlighted bar for thought activity.

The above three examples serve as templates with external input guiding one's understanding of what is happening neurologically.Anapana Meditation:
Practicing Focus

The first 3.5 days of a 10-day Vipassana course are spend practicing ``Anapana'' (breath) meditation. This period is very important as it is time spent practicing acute, one-pointed attention. If we imagine our attention as a pink dot, the goal is the following clown effigy:

If you have ever practiced a single-pointed meditation (particularly breath meditations) before, you know all too well how easily your mind can be distracted. Taxes, Mom's birthday, an upcoming meeting, a fun math problem, what's for lunch, the itch on my ear... anything can become a distraction because everything which is not ``my breath'' is by definition a distraction. And we are not accustomed to spending long stretches of time with unwavering, singular focus. We are especially unaccustomed to doing so with all of our attention focused on our breath.

As you practice Anapana for over 30 hours, you will find your attention wandering less and less, while (paradoxically) the distractions become more and more interesting. The itch on your ear becomes an itch on your nose which seems to move with the breath. Your brain will spend less time worrying about taxes and more time trying to solve interesting math problems (if that's your thing).

On the first day, your attention might look something like this:
…where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent “outside” the body on thoughts and emotions. By the fourth day, your attention will often take a shape like this instead:

If we revisit the notion that ``meditation is a thing'', we can see a specific example here in Anapana:

As much as possible, we are forcing our attention to the sense of touch. In the above diagram, our attention is (realistically) not perfectly tuned and narrowed to a tiny spot under our nose. We are still likely to hear sounds around us on the 3rd day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall.

Vipassana Meditation:
Mutually-Recursive
Consciousness I/O

Oddly enough, Vipassana mostly seems to be about remaining calm. If you're thinking ``you're sitting there, silent, for 10 hours a day... what is there to get upset about?'' you are in for a surprise. But first, a quick overview of the early mechanics of Vipassana:
Attention is moved from the nostril area to the top of the head, then systematically moved across every body part, down to the toes. This process then repeats. More complex variations are added as the bodily sensation experienced through this probing of the nervous system becomes increasingly subtle. In this order:

Add toes-to-head to head-to-toes
Dual/symmetrical observation (both arms, both legs)
Observation of a ``ring'' around the entire body
Observation of as much of the body as possible (while moving attention up and down)
Penetrating attention of muscles, tissues, organs
Penetrating attention of the spinal cord
Randomized attention in fingertip-sized dots throughout the body, inside and out
Sweeping attention through the entire body, up and down, in a single breath

Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like this:

I should note that I have never successfully penetrated my muscles or inner tissues much. The average new meditator is likely to feel a great many fascinating sensations on her fingertips, palms, arms, toes, and legs. The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way.

As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, the following diagram roughly summarizes most of my meditation experiences:
...where the purple dots are intense pain and the pink dot continues to be my awareness. At the peak of my meditation practice toward the end of a course, I would find that if my attention were somewhere away (say, my neck) from the painful area of my body (say, my knees), I may not feel any pain at all. As I moved my attention around, the pain in my knees would come back as I approached them, but every time I could sustain focus on another part of my body, the pain would appear to decrease with every return visit to my knees. If I can remain calm and focused while away from my knees, my knees' pain seems to take care of itself. If I can remain calm and focused while my attention moves across my knees, the situation continues to improve.

If you can’t feel pain in your knees despite the fact that the pain is “real”, you have effectively eliminated (some) sensory input. This is possible for all sensory input. The 5 classic sense-doors will quite clearly shut down in deep meditative states. Just because your eyes are closed does not make you blind. Light sensitivity remains, and as a side bonus of an optic nerve injury from a few years ago, I can ``see'' something (the damaged spot of the nerve) even when my eyes are shut and I'm in complete darkness. This blob which I ``see'' disappears in deep meditative states. Similarly, these deep states of meditation will eliminate audio input as well, which is easier for anyone to detect.
Blue Bars:
The New Distractions

At this point, we can revisit distraction. Distraction is really the essence of meditation. Whether it's taxes, math problems, hunger, itching, or pain... our job as meditators is to accept the presence of a distraction and return our attention to our meditation object.

However, once the pink bars have reduced far enough, once the mind is quiet enough, an entirely new form of distraction emerges. This diagram highlights the first likely candidate for the New Distraction: thought.

These ``thoughts'' are not normal thoughts in any sense of the word you are currently familiar. These ``thoughts'' are an entirely new form of brain activity caused by your constant prodding of the nervous system through Vipassana. Everything in blue, unfortunately, falls into the “explaining the concept of ‘white’ to a birth-blind man” category of unexplainable sensory input.

What is meant by ``prodding our nervous system'' and what does this have to do with brain activity? Our second suspension of disbelief was of the deep and intense connection between the brain the nervous system. Alternatively, we can think of the pair as the brain and the body. Alternatively again, we can think of them as a single system.

To think of the brain and body as one system is not incorrect but defeats the model we're drawing out at present. For the brain and nervous system to be ``mutually recursive'', they must be separate processes or entities. What is meant then by stating that the brain and nervous system are mutually recursive?

This statement is meant operationally, not existentially. We do not mean to say that the brain could not live without the nervous system (or vice-versa) but that every event — every operation — occurring in the nervous system has a corresponding event in the brain, and so forth. This mutually recursive process will eventually wear itself out. If it did not, the advent of slight anger would mean we would, from that moment onward, just get angrier and angrier for the rest of our lives. However, during the process of mutual recursion we can be feeling pain in a knee, to which our emotional response is “argh, I hate pain”, which makes us angry, which clenches up our physical systems, which makes the pain worse, and so on.

Vipassana is the process of unwinding the stack: If I can pay attention to my aching knee, remain calm and continue to observe it, slowly my body unclenches, the pain subsides a little, my emotional response is calmer, and so on.

Blue Bars 2:
The New-New Distractions

Let us quickly revisit our third suspension of disbelief: ``dissolution of the body is a thing.'' Neuroscience tells us some interesting things about the nature of our sense of self and how it relates to our body. From ``Where Am I?'' [11]:

“The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.”

“Our being is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.”

...which would seem in line with experiences described by mastered meditators of having ``no I'' or ``no self'' as their fundamental insightful discovery. It also seems necessary to project our experiences from one 10-day Vipassana course into the future, with respect to the extent to which this practice can take us. If it were stated that we could explore the entirety of the nervous system and its relationship to the brain — but not the gallbladder or spinal cord because those are off-limits — we would have reason to believe that there is a portion of the body and/or nervous system which is exempt from the rules governing these interactions.

Assuming we can explore to the very depths of our nervous system by repeatedly and systematically penetrating it with our attention, given the extent to which this observably occurs on a 10-day course, we can then examine ``mutually-recursive consciousness I/O'' in its totality. What occurs?

This I/O feedback between the brain and body appears to create a completely new kind of distraction. I will attempt to highlight this in the following diagram:
Here, all of the sense doors which we have quiesced return to life, as did ``thought'' before them. However, this new sensory input (which remains a distraction... just a very sophisticated distraction) is unlike most of regular experience. The nearest approximation I can come up with would be the experience of taking salvia, an awful drug which is legally permitted in most countries because it is simply not that enjoyable. While I do not at all recommend smoking salvia, the Wikipedia article [9] is worth a glance. The to relevant items from the list of ``immediate effects'' are:
Past memories, such as revisiting places from childhood memory
Overlapping realities, such as the perception of being in several locations at once

The effects of deep Vipassana meditation are akin to this, in the sense that one may experience, in its entirety, a completely different time or reality. That is where similarities end but the comparison may help those who have experience with hallucinogens.

It is worth reiterating that the purpose of Vipassana is not to experience altered mental states or its potentially hallucinogenic effects. New thought patterns and ``new realities'' are, again, simply distractions from the meditation itself. The meditation remains the systematic observation of the nervous system.Examples

While I would like to avoid descriptions of personal experiences, a few may help highlight the above description of new distracted states of mind.

The non-example:
the resolution to paradox

It would seem that one very common effect of Vipassana during deep meditation is to witness the resolution of paradox. A logical, mathematical, or linguistic paradox might emerge in the meditator's mind and just as quickly, it is resolved.

These would appear to mirror Zen koan practice, where the paradox is presented to the meditator verbally, in advance. No such practice occurs in Vipassana. Paradoxes and their solutions appear to be intrinsic to the human makeup, somehow.

Holding a baby

A friend once had a very salvia-like experience, feeling as though she was teleported to her friend's kitchen, holding her friend's 2-year-old while he cried. The experience was so raw, realistic, and unexpected that she actually opened her eyes and stood up in her meditation cell.

Writing a poem

Often, in early stage meditation, a person will find him or herself distracted by an upcoming task: write this email, call this person, do this chore.

I had a strangely parallel experience toward the end of my second 10-day course. My brain appeared to have written a poem: a visceral, all-encompassing work of art which touched my every sense and filled my mind. The temptation was not to read the poem (since ``I'' had written it) but to write the last line... to take credit for the poem and make it complete. Given that this genuinely felt like the most beautiful thing I had ever created, it was extremely difficult to fight the temptation to finish it and instead continue scanning my lower back with my awareness.

Time slowing down

It will become quite clear halfway through a 10-day course that this effect of Vipassana is also quite universal, like paradox and resolution to paradox. The moment-to-moment awareness seems to take longer and longer as the course goes on and each one-hour meditation period can also seem to extend beyond reasonable experiential limit.

However, this is not what I am referring to. The experience of time slowing down in very deep meditative states (which I have not experienced a great deal, granted) is that of time slowing down a great deal ...10s or 100s of thousands of years pass before your awareness in a single meditation period.

Time speeding up

The flip side of this, of course, is time speeding up. One sits down to meditate, takes a comfortable position, begins the process in earnest, and BOOM, the gong is going off. The meditation period is over.

Time going backward

Much harder to describe than the experience of time stretching out or time compressing into a single instant is the experience of time moving backward. It would appear, to one's awareness, that cause and effect have been effectively reversed. Each moment takes you a moment (or many) backward in time, and the outcomes of your thoughts or feelings are the source material to their original causes, now future events.

This experience is not necessarily isolated, either. One may experience this as a sort of ``scrobbling'' of time back and forth. This can be quite disorienting and, as distractions go, is extremely difficult to avoid while trying to keep your attention moving over your right nipple.

But that's still the game.

Outcomes

So, great. Vipassana can trick our bodies and brains to emerge an entirely new set of distractions from the practice of... Vipassana. Why, exactly, would anyone do this?

There is a sketch in ``Where Am I?'' [11] about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:

“Honey, don’t forget what the half-life is on the autonomic nervous system.”

The autonomic nervous system is what causes the ``winding stack'' of thought and emotion described earlier. The autonomic nervous system is what makes -- and keeps -- you angry. And afraid. And sad. And anxious. And everything else.

Vipassana is a systematic practice to calm the nervous system (and brain?) so that every sensory input can be analyzed rationally. You won't become a perfectly rational machine after one 10-day Vipassana course, but you are likely to come out with a much clearer head and a much better capacity for decision-making in general.

I would describe the positive effects of nervous system meditations to live on a spectrum. On one end of the spectrum are the positive, rationality-enhancing qualities of calming one's nervous system with quiet attention. On the other end of the spectrum are the subtler qualities: Meditation seems to make a person more generous, more compassionate, more empathetic. Unlike the trick to healthy rationality — intentional pacifying of the autonomic nervous system — I do not have a pop neuroscience answer as to why this is the case.

I do find the entire endeavour not just fascinating, but endlessly worthwhile. Vipassana is simultaneously the hardest thing I have ever done — and the most valuable.Further Reading

My write-up of my first 10-Day Vipassana course:
http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html

My comparison of meditation techniques and various drugs:
http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html

Appendix A:
Can we ``hack'' Vipassana?

If Vipassana is hacking the mind, is there an opportunity for hacking Vipassana? Can we hack the hacking process? Let's ask Martin Thompson. The following slide is from his 2013 talk ``Performance Testing Java Applications'' [12] and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is appropriate.

He suggests that any system we intend to observe should be observed through a 3rd party observer by dumping network traffic. For computer systems, this makes a great deal of sense. Were the system under test or the load agents (the Death Star and X-Wings, respectively) to perform the measurements, the measurements and the systems would likely interfere with each other.

Unfortunately, human biology is not so keen on this setup. Even if we had the technology to tap into the interactions of the brain and nervous system (we don't), there is no way to observe the interactions of the brain and nervous system without immediately falling into the Matrix Paradox. How do I know the tool I used is trustworthy? How do I know the observations I made with my eyes are trustworthy? This is all once again external.

It is possible to ``jump out'' of Vipassana momentarily to take stock of the effects of the meditation. Did my hearing just cut out? Did I just force my brain to become consciously aware of my normally-unconscious proprioception pertaining to my right arm? Did I just feel my cardiovascular system in exquisite detail or did I imagine it? …but it is not possible to do so without breaking the meditation for the period of this observation.

I performed this exercise a great deal toward the end of my second 10-day course because I was quite confused about everything that I was experiencing. It framed my understanding but it did little for the practice itself. I don't recommend doing this. Particularly for your first course, try your hardest to follow the instructions precisely and give all of your energy to the moment-to-moment attention demanded of you. Otherwise you will find yourself spending 10 days thinking about meditation, rather than meditating.
Observation of the process itself, rather than observation of your mind via your nervous system, is actually just another clever form of distraction. Every distraction you indulge will limit the positive effects of the meditation.

This particular distraction may feel justified as the reification of your skepticism. While it pays to be skeptical, try your hardest to avoid skepticism during the course. Approach the practice skeptically until you decide to take a course and then save your analysis of whatever newfound skepticism the course generates for you until after the course is completed. There will be plenty of time to discuss the meditation and the outcomes of the meditation on the 10th day when talking is permitted; explore your skepticism here. Other meditators are likely to be the only ones who will quite comprehend the concepts you are trying to discuss.

[1] ``The Newsroom'':
\url{https://www.youtube.com/watch?v=2C6h-Yyx9Yk}
[2] Five Precepts:
\url{https://en.wikipedia.org/wiki/Five_Precepts}
[3] S.N. Goenka -- Experience the Truth
\url{https://www.youtube.com/watch?v=mcbGbN8pFyk&list=PLlb2aTuor6J9j8hACeX6GUmd4WF7QJsbu&index=2}
[4] Eric S. Raymond -- How To Become A Hacker
\url{http://www.catb.org/esr/faqs/hacker-howto.html}
[5] Michael Abrash -- Why Virtual Reality Will Matter to You
\url{https://www.youtube.com/watch?v=UDu-cnXI8E8}
[6] Simulation Hypothesis
\url{https://en.wikipedia.org/wiki/Simulation_hypothesis}
[7] The use of visual feedback in restoring brain function:
\url{https://academic.oup.com/brain/article/132/7/1693/328686}
[8] Proprioception:
\url{https://en.wikipedia.org/wiki/Proprioception}
[9] Nocioception:
\url{https://en.wikipedia.org/wiki/Nociception}
[10] Brain activity in hunger and satiety: an exploratory visually stimulated FMRI study
\url{https://www.ncbi.nlm.nih.gov/pubmed/18292747}
[11] RadioLab -- Where Am I?
\url{http://www.radiolab.org/story/91524-where-am-i/}
[12] Martin Thompson -- Performance Testing Java Applications
\url{https://www.youtube.com/watch?v=1DuMvpwWHH4}

\end{document}

### 2017-11-07 build.sh

@@ -1,0 +1,1 @@
+pdflatex vipassana-for-hackers.tex

### 2017-11-09 vipassana-for-hackers.tex
cleaned images, finished references

@@
 Well-meaning Neuroscience papers which completely miss the mechanics of nervous-system-based meditation techniques and their core outcomes.
-Descriptions of a ``silent meditation retreat,'' usually given by someone who has not attended a course. [1]
+Descriptions of a ``silent meditation retreat,'' usually given by someone who has not attended a course. \cite{thenewsroom}
 Descriptions of Vipassana from an individual and subjective perspective, usually given by someone who has attended one course. (``Oh my god! It was sooooo intense!'')
@@
 This is not a comparison with other meditation techniques, though I will address other techniques for context.
-It is worth noting that there are a few meditation techniques labelled “Vipassana” and this paper deals specifically with Vipassana as taught by S.N.
+It is worth noting that there are a few meditation techniques labelled ``Vipassana'' and this paper deals specifically with Vipassana as taught by S.N.
 Goenka.
@@
 
-A 10-day Vipassana course requires that all students take the standard ``5 Precepts'' [2], as per Theravada.
+A 10-day Vipassana course requires that all students take the standard ``5 Precepts'' \cite{fiveprecepts}, as per Theravada.
 I do not recommend getting drunk, raping and murdering your fellow students, and stealing their belongings.
@@
 The fundamental difficulty in any description of Vipassana is this: It is impossible to convey a description of Vipassana in terms known to someone who has not experienced it.
-During a Vipassana course this paradox is addressed using two metaphors: First, a story of a sighted brother explaining the concept of ``white'' to his blind brother [3].
-Second, the classic Indian story of blind men each describing an elephant and each producing a partial truth.
+During a Vipassana course this paradox is addressed using two metaphors: First, a story of a sighted brother explaining the concept of ``white'' to his blind brother.
+\cite{experiencethetruth} Second, the classic Indian story of blind men each describing an elephant and each producing a partial truth.
 
@@
 
-Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' [4] and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely.
+Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' \cite{hackerhowto} and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely.
 LISP has some very interesting properties: Homoiconicity, metaprogramming, self-mutation and self-replication, recursion and trampolining, and — in the case of modern LISPs and other languages with strong concurrency models — hardened notions of immutability and the reification of time.
@@
 
+\vspace{1cm}
 \begin{center}
@@
 \begin{figure}[h]
-  \includegraphics[width=\linewidth]{meditation-spectrum-1.jpeg}
-  \caption{Sample meditation spectrum.}
-  \label{fig:meditation-spectrum}
+  \centering
+  \includegraphics[width=11cm]{images/meditation-spectrum-1.jpeg}
+  \caption{Sample meditation/meditative spectrum.}
+  \label{fig:meditation-spectrum-1}
 \end{figure}
 
-For the purposes of this paper, we will keep all notions of “mindfulness” in the “meditative” camp, on the right.
+For the purposes of this paper, we will keep all notions of ``mindfulness'' in the ``meditative'' camp, on the right.
 Mindfulness is not meditation.
 This paper will be discussing only two practices from this list: Anapana and Vipassana.
-An illustrated definition of “awareness” and “attention” follows, in the section entitled “On Attention”.
+An illustrated definition of ``awareness'' and ``attention'' follows, in the section entitled ``On Attention''.
 
-2. Your brain and nervous system have a tight feedback loop
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=11cm]{images/meditation-spectrum-3.jpeg}
+  \caption{Sample meditation spectrum.}
+  \label{fig:meditation-spectrum-3}
+\end{figure}
 
-When we say “tight feedback loop”, this means the brain and nervous system are tied together in a feedback loop faster and tighter than you could ever imagine without experiencing it directly.
+\vspace{1cm}
+\begin{center}
+  \LARGE{2. Your brain and nervous system have}\\
+  \LARGE{a tight feedback loop}
+\end{center}
+
+When we say ``tight feedback loop'', this means the brain and nervous system are tied together in a feedback loop faster and tighter than you could ever imagine without experiencing it directly.
 
@@
 
-3. Dissolution of the body
-is a thing
+\vspace{1cm}
+\begin{center}
+  \LARGE{3. Dissolution of the body is a thing}
+\end{center}
 
@@
 
-To help visualize what I mean, a sample spectrum of sensations from the gross or obvious to the curious and subtle might look like this:
+To help visualize what I mean, a sample spectrum of sensations from the gross or obvious to the curious and subtle might look like the figure on the following page.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/sensation-spectrum.jpeg}
+  \caption{Sample sensation spectrum.}
+  \label{fig:sensation-spectrum}
+\end{figure}
 
 As the field of sensation progresses to the left, the sensations felt become increasingly difficult to describe.
-“Surreal hot / cold” refers to sensations of temperature we wouldn’t normally consider possible: our muscles either on fire or freezing solid.
-“Spatial distortion” refers to complete alterations of spacial presence — the partial or complete replacement of one’s normal sense of proprioception.
-This may include reorientation in space (“I was upside-down!”) or the feeling that one is gigantic or tiny or that any one body part has one of these or other distorted qualities.
+``Surreal hot / cold'' refers to sensations of temperature we wouldn’t normally consider possible: our muscles either on fire or freezing solid.
+``Spatial distortion'' refers to complete alterations of spacial presence — the partial or complete replacement of one’s normal sense of proprioception.
+This may include reorientation in space (``I was upside-down!'') or the feeling that one is gigantic or tiny or that any one body part has one of these or other distorted qualities.
 
-“Throbbing” refers to thick, heavy, slower vibrations; these are often felt rhythmically and this rhythm accelerates until the sensation is described as a “vibration”.
-Beyond vibration is the “invisible”, where a particular body part is no longer seen and a deeper part of the body is experienced directly.
+``Throbbing'' refers to thick, heavy, slower vibrations; these are often felt rhythmically and this rhythm accelerates until the sensation is described as a ``vibration''.
+Beyond vibration is the ``invisible'', where a particular body part is no longer seen and a deeper part of the body is experienced directly.
 The first such example in most cases tends to be the skin, since the external surface of the body is the first part of the nervous system the course instructions ask the meditator to focus on.
 Beyond this the meditator may find muscle tissue, bones, the cardiovascular system, or something else entirely.
-Usually by the point in the course when a meditator has dissolved her own skin to her attention the sensations become extremely difficult to identify and label so “something else entirely” is likely to be the most common case.
+Usually by the point in the course when a meditator has dissolved her own skin to her attention the sensations become extremely difficult to identify and label so ``something else entirely'' is likely to be the most common case.
 
 The sample spectrum above is missing many, many sensations.
-I also feel it necessary to call out that the spectrum is not serial; although I have felt “vibrations” and seen deeper than skin level, I have certainly not experienced all the sensations possible in the spectrum.
+I also feel it necessary to call out that the spectrum is not serial; although I have felt ``vibrations'' and seen deeper than skin level, I have certainly not experienced all the sensations possible in the spectrum.
 However, the diagram hopefully provides a reference point for the type of readily-accessible sensations which Vipassana begins with (on the right) and the types of sensations (or absence of sensation) which will convince a person that complete dissolution of the body not only possible but an entirely natural process.
 
-Virtual Reality
-and
-The Matrix Paradox
+\pagebreak
+
+\begin{center}
+  \Huge{Virtual Reality}\\
+  \Huge{and}\\
+  \Huge{The Matrix Paradox}
+\end{center}
 
@@
 
-This image is taken from a talk by Michael Abrash, Chief Scientist of the Oculus VR team, at a talk he gave at Facebook F8 in 2015 [5].
-I'm using this slide because it's nicely laid out and because it is incomplete.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/michael-abrash-senses.png}
+  \caption{Traditional 5-sense brain mapping.}
+  \label{fig:michael-abrash-senses}
+\end{figure}
 
-Abrash discusses the nature of the ``Matrix Paradox'' [6] as simply the right neurons firing to simulate any particular environment.
+This image is taken from a talk by Michael Abrash, Chief Scientist of the Oculus VR team, at a talk he gave at the Facebook F8 Summit in 2015.
+\cite{abrashvr} I'm using this slide because it's nicely laid out and because it is incomplete.
+
+Abrash discusses the nature of the ``Matrix Paradox'' \cite{simulationhypothesis} as simply the right neurons firing to simulate any particular environment.
 ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input.
@@
 Though a classic sense diagram such as Mr.
-Abrash’s might describe proprioception [7, 8], nocioception [9], and hunger [10] as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain.
+Abrash’s might describe proprioception \cite{ramachandranbrainfunction, proprioception}, nocioception \cite{nocioception}, and hunger \cite{fmrihunger} as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain.
 Touch is strictly an external sense dealing with skin.
@@
 
-On that note, I feel I should introduce an absolutely fantastic RadioLab episode (``Where Am I?'' [11]) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience.
-On the topic of lacking proprioception: “We have words like ‘deaf’ and ‘blind’.
-We do not have words for being deaf or blind to one’s own body.” In the Matrix, perhaps this is either a bug or a feature flag.
+On that note, I feel I should introduce an absolutely fantastic RadioLab episode (``Where Am I?'' \cite{whereami}) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience.
+On the topic of lacking proprioception: ``We have words like `deaf' and `blind'.
+We do not have words for being deaf or blind to one's own body.'' In the Matrix, perhaps this is either a bug or a feature flag.
 
 Now that we have our diagram for visualizing the use of our senses, we can address them in terms of various altered mental states.
-Before we get there, however, we will try to come up with a meaningful visual definition of “attention” or “awareness” with respect to our senses.
+Before we get there, however, we will try to come up with a meaningful visual definition of ``attention'' or ``awareness'' with respect to our senses.
 
@@
 Vipassana is the inspection of one's own nervous system with narrowly-focused attention but the precision of this focus is not prescribed by the instruction given for the technique.
-“As narrow as possible”, perhaps, but in day to day life a normal person cannot possibly narrow her attention to the size of a fingertip.
+``As narrow as possible'', perhaps, but in day to day life a normal person cannot possibly narrow her attention to the size of a fingertip.
 Your attention is, in part, already on your nervous system, with some degree of focus.
@@
 On the first day, your attention might look something like this:
-…where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent “outside” the body on thoughts and emotions.
+…where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body on thoughts and emotions.
 By the fourth day, your attention will often take a shape like this instead:
@@
 
-If you can’t feel pain in your knees despite the fact that the pain is “real”, you have effectively eliminated (some) sensory input.
+If you can’t feel pain in your knees despite the fact that the pain is ``real'', you have effectively eliminated (some) sensory input.
 This is possible for all sensory input.
@@
 These ``thoughts'' are an entirely new form of brain activity caused by your constant prodding of the nervous system through Vipassana.
-Everything in blue, unfortunately, falls into the “explaining the concept of ‘white’ to a birth-blind man” category of unexplainable sensory input.
+Everything in blue, unfortunately, falls into the ``explaining the concept of `white' to a birth-blind man'' category of unexplainable sensory input.
 
@@
 If it did not, the advent of slight anger would mean we would, from that moment onward, just get angrier and angrier for the rest of our lives.
-However, during the process of mutual recursion we can be feeling pain in a knee, to which our emotional response is “argh, I hate pain”, which makes us angry, which clenches up our physical systems, which makes the pain worse, and so on.
+However, during the process of mutual recursion we can be feeling pain in a knee, to which our emotional response is ``argh, I hate pain'', which makes us angry, which clenches up our physical systems, which makes the pain worse, and so on.
 
@@
 Let us quickly revisit our third suspension of disbelief: ``dissolution of the body is a thing.'' Neuroscience tells us some interesting things about the nature of our sense of self and how it relates to our body.
-From ``Where Am I?'' [11]:
+From ``Where Am I?'' \cite{whereami}:
 
-“The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.”
+``The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.''
 
-“Our being is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.”
+``Our being is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.''
 
@@
 
-There is a sketch in ``Where Am I?'' [11] about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:
+There is a sketch in ``Where Am I?'' \cite{whereami} about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:
 
-“Honey, don’t forget what the half-life is on the autonomic nervous system.”
+``Honey, don’t forget what the half-life is on the autonomic nervous system.''
 
@@
 Let's ask Martin Thompson.
-The following slide is from his 2013 talk ``Performance Testing Java Applications'' [12] and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is appropriate.
+The following slide is from his 2013 talk ``Performance Testing Java Applications'' \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is appropriate.
 
@@
 
-[1] ``The Newsroom'':
+\pagebreak
+
+\begin{thebibliography}{9}
+\bibitem{thenewsroom}
+  Aaron Sorkin.
   \url{https://www.youtube.com/watch?v=2C6h-Yyx9Yk}
-[2] Five Precepts:
-\url{https://en.wikipedia.org/wiki/Five_Precepts}
-[3] S.N. Goenka -- Experience the Truth
+  \textit{The Newsroom}.
+  HBO, 2012.
+
+\bibitem{experiencethetruth}
+  S.N. Goenka.
   \url{https://www.youtube.com/watch?v=mcbGbN8pFyk&list=PLlb2aTuor6J9j8hACeX6GUmd4WF7QJsbu&index=2}
-[4] Eric S. Raymond -- How To Become A Hacker
+  \textit{Experience the Truth}.
+  Vipassana Research Institute, 1992.
+
+\bibitem{hackerhowto}
+  Eric S. Raymond: How To Become A Hacker,
   \url{http://www.catb.org/esr/faqs/hacker-howto.html}
-[5] Michael Abrash -- Why Virtual Reality Will Matter to You
+
+\bibitem{abrashvr}
+  Michael Abrash.
   \url{https://www.youtube.com/watch?v=UDu-cnXI8E8}
-[6] Simulation Hypothesis
+  \textit{Why Virtual Reality Will Matter to You}.
+  Facebook F8 Summit, Fort Mason, 2015.
+
+\bibitem{simulationhypothesis}
   \url{https://en.wikipedia.org/wiki/Simulation_hypothesis}
-[7] The use of visual feedback in restoring brain function:
+  \textit{Simulation Hypothesis}.
+  Wikipedia.
+
+\bibitem{ramachandranbrainfunction}
+  V. S. Ramachandran, Eric L. Altschuler.
   \url{https://academic.oup.com/brain/article/132/7/1693/328686}
-[8] Proprioception:
+  \textit{The use of visual feedback in restoring brain function}.
+  Brain, 132(7):1693–1710, July 2009.
+
+\bibitem{proprioception}
   \url{https://en.wikipedia.org/wiki/Proprioception}
-[9] Nocioception:
+  \textit{Proprioception}.
+  Wikipedia.
+
+\bibitem{nocioception}
   \url{https://en.wikipedia.org/wiki/Nociception}
-[10] Brain activity in hunger and satiety: an exploratory visually stimulated FMRI study
+  \textit{Nocioception}.
+  Wikipedia.
+
+\bibitem{fmrihunger}
+  Dagmar Führer, Stefan Zysset, Michael Stumvoll.
   \url{https://www.ncbi.nlm.nih.gov/pubmed/18292747}
-[11] RadioLab -- Where Am I?
+  \textit{Brain activity in hunger and satiety: an exploratory visually stimulated fMRI study}
+  Obesity, 16(5):945-50, May 2008.
+
+\bibitem{whereami}
+  Jad Abumrad, Robert Krulwich.
   \url{http://www.radiolab.org/story/91524-where-am-i/}
-[12] Martin Thompson -- Performance Testing Java Applications
+  \textit{Where Am I?}
+  RadioLab, May 2006.
+
+\bibitem{testingjava}
+  Martin Thompson.
   \url{https://www.youtube.com/watch?v=1DuMvpwWHH4}
+  \textit{Performance Testing Java Applications}.
+  Devoxx UK, 2013.
+
+\end{thebibliography}
 

### 2017-11-10 vipassana-for-hackers.tex
done up to blue bars 1

@@
 \usepackage{graphicx}
+\usepackage{subcaption}
+\usepackage{float}
+\usepackage{enumitem}
+\setlist{noitemsep}
 
@@
 
-Put more succinctly, your activities will include: Eat, sleep, poop, bath, meditate.
+Put more succinctly, your activities will include: Eat, sleep, poop, bath, meditate.\\
 
-That's it.
+That's it.\\
 
-Meditation will consume approximately 10 to 11 hours per day.
+Meditation will consume approximately ten to eleven hours per day. All courses the world over are free and volunteer-run. Material goods are provided for by donations from previous students.
 
@@
 The following section will require some visual assistance, for which I will use the same slides I used in the talk which corresponds to this paper.
-
-They look like this:
+An unmodified slide is seen here in Figure~\ref{fig:michael-abrash-senses}.
 
@@
 
-Abrash discusses the nature of the ``Matrix Paradox'' \cite{simulationhypothesis} as simply the right neurons firing to simulate any particular environment.
+Abrash discusses the nature of the ``Matrix Paradox'' (more commonly known as the ``Simulation Hypothesis'' \cite{simulationhypothesis}) as simply the right neurons firing to simulate any particular environment.
 ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input.
-The following might be a slightly better model of reality:
+Figure~\ref{fig:michael-abrash-senses-extended} is a slightly better model of reality.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/michael-abrash-senses-extended-attached.jpeg}
+  \caption{A more complete, 7-sense brain mapping.}
+  \label{fig:michael-abrash-senses-extended}
+\end{figure}
 
 Though a classic sense diagram such as Mr.
-Abrash’s might describe proprioception \cite{ramachandranbrainfunction, proprioception}, nocioception \cite{nocioception}, and hunger \cite{fmrihunger} as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain.
+Abrash's might describe proprioception \cite{ramachandranbrainfunction, proprioception}, nocioception \cite{nocioception}, and hunger \cite{fmrihunger} as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain.
 Touch is strictly an external sense dealing with skin.
@@
 
-On that note, I feel I should introduce an absolutely fantastic RadioLab episode (``Where Am I?'' \cite{whereami}) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience.
+On that note, I feel I should introduce an absolutely fantastic RadioLab episode (\textit{Where Am I?} \cite{whereami}) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience.
 On the topic of lacking proprioception: ``We have words like `deaf' and `blind'.
@@
 
-On Attention
+\pagebreak
+
+\begin{center}
+  \Huge{On Attention}
+\end{center}
 
@@
 
-Example 1:
-A bustling restaurant
+\pagebreak
+
+\begin{center}
+  \LARGE{Example 1:}\\
+  \LARGE{A bustling restaurant}
+\end{center}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/attention1-restaurant.jpg}
+  \caption{A bustling restaurant.}
+  \label{fig:bustling-restaurant}
+\end{figure}
 
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=10cm]{images/ma-noisy-restaurant.png}
+  \caption{Sense map of a bustling restaurant.}
+  \label{fig:sense-map-bustling-restaurant}
+\end{figure}
+
 Assuming we've just placed food in our mouths, our sense of taste is going off like crazy, we're hearing a lot of scattered sound around us, and we're also sufficiently distracted by most of our other senses.
 Note that we are probably paying very little attention to the internal senses.
-Example 2:
-A sensory deprivation tank
 
-The corresponding sense map is on the right.
+\pagebreak
+
+\begin{center}
+  \LARGE{Example 2:}\\
+  \LARGE{A sensory deprivation tank}
+\end{center}
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/attention2-tank.jpg}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/ma-sensory-deprivation-tank.png}
+  \end{subfigure}
+  \caption{A sensory deprivation tank and corresponding sense map.}
+  \label{fig:sensory-deprivation-tank}
+\end{figure}
 
@@
 
-Example 3:
-An anechoic chamber
+\pagebreak
 
-The corresponding sense map is again on the right.
+\begin{center}
+  \LARGE{Example 3:}\\
+  \LARGE{An anechoic chamber}
+\end{center}
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/attention3-chamber.png}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/ma-anechoic-chamber.png}
+  \end{subfigure}
+  \caption{An anechoic chamber and corresponding sense map.}
+  \label{fig:anechoic-chamber}
+\end{figure}
 
@@
 
-The above three examples serve as templates with external input guiding one's understanding of what is happening neurologically.Anapana Meditation:
-Practicing Focus
+The above three examples serve as templates with external input guiding one's understanding of what is happening neurologically.
+
+\pagebreak
+
+\begin{center}
+  \Huge{Anapana Meditation:}\\
+  \Huge{Practicing Focus}
+\end{center}
 
@@
 This period is very important as it is time spent practicing acute, one-pointed attention.
-If we imagine our attention as a pink dot, the goal is the following clown effigy:
+If we visualize our attention as any skin (internal or external; lips, nose, nostrils, etc.) covered by the pink triangle in Figure~\ref{fig:burmese-anapana-triangle}, we will get an idea of exactly where one's attention is directed at the beginning of Anapana practice.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/burmese-anapana-triangle.png}
+  \caption{Large, triangular Anapana meditation target.}
+  \label{fig:burmese-anapana-triangle}
+\end{figure}
 
@@
 
-On the first day, your attention might look something like this:
-…where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body on thoughts and emotions.
-By the fourth day, your attention will often take a shape like this instead:
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/face-attention.jpg}
+  \caption{Wandering attention.}
+  \label{fig:wandering-attention}
+\end{figure}
 
-If we revisit the notion that ``meditation is a thing'', we can see a specific example here in Anapana:
+On the first day, your attention might look something like Figure~\ref{fig:wandering-attention}, where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body on thoughts and emotions.
+By the fourth day, your attention will often take a shape something like Figure~\ref{fig:narrow-attention}.
+Over the course of the first three days, the instructions narrow down one's attention from the nose-and-lips triangle from Figure~\ref{fig:burmese-anapana-triangle} to a finger-tip-sized area wherever the breath can be felt beneath the nostrils.
 
-As much as possible, we are forcing our attention to the sense of touch.
-In the above diagram, our attention is (realistically) not perfectly tuned and narrowed to a tiny spot under our nose.
-We are still likely to hear sounds around us on the 3rd day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall.
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{images/face-attention-narrow.jpg}
+  \caption{Narrow attention.}
+  \label{fig:narrow-attention}
+\end{figure}
 
-Vipassana Meditation:
-Mutually-Recursive
-Consciousness I/O
+If we revisit the notion that ``meditation is a thing'', we can see a specific example in Anapana with the help of the sense map in Figure~\ref{fig:anapana-sense-map}.
+Where mindfulness and violin, albeit useful physical and mental practices, are not meditation, we can see from the sense map that Anapana \textit{is} meditation.
+It is the practice of narrowing one's attention, repeatedly, toward a specific sense (in this case, touch) at a specific time (now) in a specific place (just beneath the nostrils).
+As our skill in Anapana improves, we come nearer and nearer to the focus-oriented meditation ideal of a single point of awareness with unwavering focus.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=10cm]{images/ma-anapana.png}
+  \caption{Anapana sense map.}
+  \label{fig:anapana-sense-map}
+\end{figure}
+
+In Figure~\ref{fig:anapana-sense-map}, we can see our attention is (realistically) not perfectly tuned and narrowed to a tiny spot under our nose.
+We are still likely to hear sounds around us on the third day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall.
+
+\pagebreak
+
+\begin{center}
+  \Huge{Vipassana Meditation:}\\
+  \Huge{Mutually-Recursive}\\
+  \Huge{Consciousness I/O}
+\end{center}
 
@@
 what is there to get upset about?'' you are in for a surprise.
-But first, a quick overview of the early mechanics of Vipassana:
-Attention is moved from the nostril area to the top of the head, then systematically moved across every body part, down to the toes.
-This process then repeats.
-More complex variations are added as the bodily sensation experienced through this probing of the nervous system becomes increasingly subtle.
-In this order:
+But first, a quick overview of the early mechanics of Vipassana.
+(See Figure~\ref{fig:vipassana-mechanics}.)
+Attention is moved from the nostril area to the top of the head, then systematically moved across every body part, down to the toes. This process then repeats.
 
-Add toes-to-head to head-to-toes
-Dual/symmetrical observation (both arms, both legs)
-Observation of a ``ring'' around the entire body
-Observation of as much of the body as possible (while moving attention up and down)
-Penetrating attention of muscles, tissues, organs
-Penetrating attention of the spinal cord
-Randomized attention in fingertip-sized dots throughout the body, inside and out
-Sweeping attention through the entire body, up and down, in a single breath
+This visualization and description is to act as an \textit{example only}: this is not the only way to begin Vipassana meditation.
+Though Vipassana instruction is systematic, it is far from mechanical or prescriptive.
+Even in the initial instructions, alternative approaches are offered including alternate orders of observation, alternate areas of observation, and alternate durations of observation.
+The key aspect of any approach to Vipassana is that the \textbf{entire body} is repeatedly observed within each meditation period.
 
-Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like this:
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/burmese-vipassana-mechanics.png}
+  \caption{Simplified example of initial Vipassana mechanics.}
+  \label{fig:vipassana-mechanics}
+\end{figure}
+
+I will take this opportunity to echo a caveat from the beginning of this paper: This section is in no way a form of instruction.
+Do not attempt full-body Vipassana meditation outside of a proper 10-day course.
+Practicing Vipassana is likely the most difficult activity you will ever attempt, should you choose to.
+To try so alone, you will have no teacher for guidance and the experience is bound to be frustrating.
+
+Consider this warning to also extend to any literature which \textbf{does} attempt to convey Vipassana meditation instruction.
+Specifically, Glickman's \textit{Beyond the Breath: Extraordinary Mindfulness Through Whole-Body Vipassana Meditation} \cite{glickman} includes a facsimile of instruction provided on a 10-day introductory Vipassana course.
+Instruction on a course includes 11+ hours of video instruction, at least twice as many hours of audio instruction, and multiple interviews per day with the teacher.
+Trying to replace this with a book is both inadequate and dangerous.\\
+
+More complex variations are added as the bodily sensation experienced through this probing of the nervous system becomes increasingly subtle. In this order:
+
+\begin{enumerate}
+  \item Add toes-to-head to head-to-toes
+  \item Dual/symmetrical observation (both arms, both legs)
+  \item Observation of a ``ring'' around the entire body
+  \item Observation of as much of the body as possible (while moving attention up and down)
+  \item Penetrating attention of muscles, tissues, organs
+  \item Penetrating attention of the spinal cord
+  \item Randomized attention in fingertip-sized dots throughout the body, inside and out
+  \item Sweeping attention through the entire body, up and down, in a single breath
+\end{enumerate}
+
+Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like Figure~\ref{fig:vipassana-sense-map}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=8cm]{images/ma-vipassana1.png}
+  \caption{Vipassana sense map.}
+  \label{fig:vipassana-sense-map}
+\end{figure}
 
@@
 The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way.
+\textit{[citation needed]}
 
-As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, the following diagram roughly summarizes most of my meditation experiences:
-...where the purple dots are intense pain and the pink dot continues to be my awareness.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/burmese-vipassana-pain.png}
+  \caption{Common pain for Westerners in meditative postures.}
+  \label{fig:burmese-vipassana-pain}
+\end{figure}
+
+As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, Figure~\ref{fig:burmese-vipassana-pain} roughly summarizes most of my early meditation experiences, where the purple dots are intense pain and the pink dot continues to be my awareness.
 At the peak of my meditation practice toward the end of a course, I would find that if my attention were somewhere away (say, my neck) from the painful area of my body (say, my knees), I may not feel any pain at all.
@@
 
-If you can’t feel pain in your knees despite the fact that the pain is ``real'', you have effectively eliminated (some) sensory input.
+If an individual cannot feel pain in one's knees despite the fact that the pain is ``real'', she has effectively eliminated (some) sensory input.
 This is possible for all sensory input.
-The 5 classic sense-doors will quite clearly shut down in deep meditative states.
+The five classic sense-doors will quite clearly shut down in deep meditative states.
 Just because your eyes are closed does not make you blind.
@@
 Similarly, these deep states of meditation will eliminate audio input as well, which is easier for anyone to detect.
-Blue Bars:
-The New Distractions
+
+\pagebreak
+
+\begin{center}
+  \Huge{Blue Bars:}\\
+  \Huge{The New Distractions}
+\end{center}
 
@@
 
+\bibitem{fiveprecepts}
+  Edited by Access to Insight.
+  \url{https://www.accesstoinsight.org/ptf/dhamma/sila/pancasila.html}
+  \textit{The Five Precepts: pañca-sila}
+  Access to Insight (BCBS Edition), 30 November 2013.
+
 \bibitem{experiencethetruth}
@@
 
+\bibitem{glickman}
+  Marshall Glickman.
+  \textit{Beyond the Breath: Extraordinary Mindfulness Through Whole-Body Vipassana Meditation}.
+  Tuttle Publishing, 2002.
+  ISBN: 1582900434
+
 \bibitem{testingjava}

### 2017-11-10 vipassana-for-hackers.tex
v1.2 draft complete

@@
 \usepackage{float}
+\usepackage[bottom]{footmisc}
 \usepackage{enumitem}
@@
      \large\textit{Steven Deobald}\\
-     \large\textit{Version 1.1}\\
+     \large\textit{Version 1.2}\\
      \large\textit\today
@@
 You might be familiar with phrases commonly attributed to this archetype such as ``When the student is ready, the teacher will appear.'' Mysticism, mystery, and mystique are to be avoided.
-If I fall into this trap, email me: steven@nilenso.com
+If I fall into this trap, email me: \href{mailto:steven@nilenso.com}{steven@nilenso.com}
 
@@
 
+\vspace{1cm}
+
 \begin{center}
@@
 Hofstadter's more cohesive ``I Am a Strange Loop'' tackles the problem of consciousness more succinctly.
-However, other than directing the reader to Zen \footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana.
+However, other than directing the reader to Zen\footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana.
 The reason I do not discuss Zazen in detail or compare it to Vipassana in this paper is because Zazen, in my experience, is not systematic.
@@
 Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' \cite{hackerhowto} and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely.
-LISP has some very interesting properties: Homoiconicity, metaprogramming, self-mutation and self-replication, recursion and trampolining, and — in the case of modern LISPs and other languages with strong concurrency models — hardened notions of immutability and the reification of time.
+LISP has some very interesting properties:
+
+\begin{itemize}
+\item Homoiconicity: the language is represented in itself and code written in the language is understood by the language
+\item Metaprogramming: the language can program itself and understand itself
+\item Self-mutation and self-replication: a LISP program can inherently alter its own internal behaviour and easily write entirely new programs
+\item Recursion and trampolining: mathematical functional recursion and mutual recursion between two functions are natural operations in LISP
+\end{itemize}
 
@@
 It is ``turtles all the way up.'' LISP is not ``turtles all the way down'', as some would describe it.
-Even if written on a LISP Machine, there are still paths of wires and electrons which need to be dealt with.
+Even if written on a LISP Machine (a computer which executes LISP directly rather than transforming LISP into machine-level instructions), there are still paths of wires and electrons which need to be dealt with.
 These are not written in LISP.
@@
 The Consciousness Machine.
-Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' The machine at the bottom of your consciousness is a strange loop.
+Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' The machine at the bottom of your consciousness \textbf{is} a strange loop.
 Having a look at this strange loop is a lot more fun than reading about it.
@@
 
-On the first day, your attention might look something like Figure~\ref{fig:wandering-attention}, where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body on thoughts and emotions.
+On the first day, your attention might look something like Figure~\ref{fig:wandering-attention}, where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body: on thoughts and emotions.
 By the fourth day, your attention will often take a shape something like Figure~\ref{fig:narrow-attention}.
@@
 It is the practice of narrowing one's attention, repeatedly, toward a specific sense (in this case, touch) at a specific time (now) in a specific place (just beneath the nostrils).
-As our skill in Anapana improves, we come nearer and nearer to the focus-oriented meditation ideal of a single point of awareness with unwavering focus.
+As our skill in Anapana improves, we come nearer and nearer to the focus-oriented meditation ideal of an unwavering single point of awareness.
 
@@
 
-In Figure~\ref{fig:anapana-sense-map}, we can see our attention is (realistically) not perfectly tuned and narrowed to a tiny spot under our nose.
+In Figure~\ref{fig:anapana-sense-map}, we can see that in reality our attention is not perfectly tuned and narrowed to a tiny spot under our nose.
 We are still likely to hear sounds around us on the third day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall.
+By the third day of the course, however, one's attention to sound is likely to be greatly reduced.
+
+For the rest of this paper, make note of the fact that \textit{distraction} is at this stage, and will forever remain, the most important aspect of meditation; it is everything meditation is not, and therefore a significant consideration in all the techniques which follow.
+Meditation is essentially comprised of only two components: (a) the meditation object and (b) distractions.
+The activity of meditation is the repeatedly directing of one's attention from the latter to the former.
 
@@
 
-Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like Figure~\ref{fig:vipassana-sense-map}.
+Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like Figure~\ref{fig:vipassana-sense-map-1}.
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \includegraphics[width=8cm]{images/ma-vipassana1.png}
+  \includegraphics[width=\linewidth]{images/ma-vipassana1.png}
   \caption{Vipassana sense map.}
-  \label{fig:vipassana-sense-map}
+  \label{fig:vipassana-sense-map-1}
 \end{figure}
@@
 The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way.
-\textit{[citation needed]}
+\textit{[citation needed]}\\
 
@@
 
-As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, Figure~\ref{fig:burmese-vipassana-pain} roughly summarizes most of my early meditation experiences, where the purple dots are intense pain and the pink dot continues to be my awareness.
+As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, Figure~\ref{fig:burmese-vipassana-pain} roughly summarizes most of my early meditation experiences, where the purple dots represent intense pain and the pink dot continues to represent my awareness.
 At the peak of my meditation practice toward the end of a course, I would find that if my attention were somewhere away (say, my neck) from the painful area of my body (say, my knees), I may not feel any pain at all.
@@
 
-These ``thoughts'' are not normal thoughts in any sense of the word you are currently familiar.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/ma-vipassana3.png}
+  \caption{Attention on a new distraction.}
+  \label{fig:vipassana-sense-map-2}
+\end{figure}
+
+These distracting new ``thoughts'' are not normal thoughts in any sense of the word you are currently familiar.
 These ``thoughts'' are an entirely new form of brain activity caused by your constant prodding of the nervous system through Vipassana.
 Everything in blue, unfortunately, falls into the ``explaining the concept of `white' to a birth-blind man'' category of unexplainable sensory input.
+Though this brain activity cannot be explained it \textbf{can} be experienced directly and even a first-time meditator is likely to experience such brain activity on her first 10-day course.
+The blue bar in Figure~\ref{fig:vipassana-sense-map-2} does not represent a goal or form of meditation, nor even a binary state.
+Rather, if we imagine brain activity on a continuum, worries about taxes might appear on the far left, interesting math and logic problems in the centre, and the ``blue bars'' of unexplainable thought on the right.
+However fascinating a journey toward this brain activity may be, the thought processes themselves remain nothing more but a distraction from our goal.
 
@@
 Our second suspension of disbelief was of the deep and intense connection between the brain the nervous system.
-Alternatively, we can think of the pair as the brain and the body.
-Alternatively again, we can think of them as a single system.
+We cannot feel the brain directly --- or at all.
+Thanks to the deep interconnection between the brain and the nervous system, however, the body acts as a surface to which we have ready access --- and which in turn provides us access to the brain.
+This brain/body connection can be thought of as \textit{mutual recursion}, as the term is used in the previous section.
 
-To think of the brain and body as one system is not incorrect but defeats the model we're drawing out at present.
-For the brain and nervous system to be ``mutually recursive'', they must be separate processes or entities.
-What is meant then by stating that the brain and nervous system are mutually recursive?
-
-This statement is meant operationally, not existentially.
-We do not mean to say that the brain could not live without the nervous system (or vice-versa) but that every event — every operation — occurring in the nervous system has a corresponding event in the brain, and so forth.
+That the brain/body connection is \textit{mutually recursive} is meant operationally, not existentially.
+We do not mean to say that the brain could not live without the nervous system (or vice-versa) but that every event --- every operation --- occurring in the nervous system has a corresponding event in the brain, and so forth.
 This mutually recursive process will eventually wear itself out.
@@
 
-Vipassana is the process of unwinding the stack: If I can pay attention to my aching knee, remain calm and continue to observe it, slowly my body unclenches, the pain subsides a little, my emotional response is calmer, and so on.
+This stack of mutually-recursive operations is what produces all thought, emotion, and sensation.
+Vipassana is the process of unwinding the stack: If I can pay attention to my aching knee, remain calm and continue to observe it, slowly my body unclenches, the pain subsides a little, my emotional response is calmer, I am able to remain calm longer, and so on.
 
-Blue Bars 2:
-The New-New Distractions
+\pagebreak
+
+\begin{center}
+  \Huge{Blue Bars 2:}\\
+  \Huge{The New-New Distractions}
+\end{center}
 
 Let us quickly revisit our third suspension of disbelief: ``dissolution of the body is a thing.'' Neuroscience tells us some interesting things about the nature of our sense of self and how it relates to our body.
-From ``Where Am I?'' \cite{whereami}:
+From ``Where Am I?'' \cite{whereami}:\\
 
-``The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.''
+\textbf{``The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.''}\\
 
-``Our being is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.''
+\textbf{``Our \textit{being} is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.''}\\
 
@@
 It also seems necessary to project our experiences from one 10-day Vipassana course into the future, with respect to the extent to which this practice can take us.
-If it were stated that we could explore the entirety of the nervous system and its relationship to the brain — but not the gallbladder or spinal cord because those are off-limits — we would have reason to believe that there is a portion of the body and/or nervous system which is exempt from the rules governing these interactions.
+If it were stated that we could explore the entirety of the nervous system and its relationship to the brain --- but not the gallbladder or spinal cord because those are off-limits --- we would have reason to believe that there is a portion of the body and/or nervous system which is exempt from the rules governing these interactions.
 
@@
 
-This I/O feedback between the brain and body appears to create a completely new kind of distraction. I will attempt to highlight this in the following diagram:
+This I/O feedback between the brain and body appears to create \textit{a completely new kind of distraction, which goes well beyond new thought patterns}.
+An attempt to highlight this form of distraction is given in Figure~\ref{fig:vipassana-sense-map-4}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/ma-vipassana4.png}
+  \caption{Attention distracted by all-new senses.}
+  \label{fig:vipassana-sense-map-4}
+\end{figure}
+
 Here, all of the sense doors which we have quiesced return to life, as did ``thought'' before them.
@@
 The nearest approximation I can come up with would be the experience of taking salvia, an awful drug which is legally permitted in most countries because it is simply not that enjoyable.
-While I do not at all recommend smoking salvia, the Wikipedia article [9] is worth a glance.
-The to relevant items from the list of ``immediate effects'' are:
-Past memories, such as revisiting places from childhood memory
-Overlapping realities, such as the perception of being in several locations at once
+While I do not at all recommend smoking salvia, the Wikipedia article \cite{salvia} is worth a glance, since the description of hallucinogens is approachable for anyone.
+The two relevant items from the list of ``immediate effects'' are:
+
+\begin{itemize}
+  \item Past memories, such as revisiting places from childhood memory
+  \item Overlapping realities, such as the perception of being in several locations at once
+\end{itemize}
 
 The effects of deep Vipassana meditation are akin to this, in the sense that one may experience, in its entirety, a completely different time or reality.
-That is where similarities end but the comparison may help those who have experience with hallucinogens.
+A comparable, sober experience might be that of a car crash: participants often feel time slow down or stop, which is sensory input completely detached from the normal flow of consciousness.
+That is where similarities end but the comparisons may help provide us with tangible, and therefore approachable, examples.
 
-It is worth reiterating that the purpose of Vipassana is not to experience altered mental states or its potentially hallucinogenic effects.
-New thought patterns and ``new realities'' are, again, simply distractions from the meditation itself.
-The meditation remains the systematic observation of the nervous system.Examples
+It is worth reiterating that the purpose of Vipassana is not to experience altered mental states.
+New thought patterns and new realities are, again, simply distractions from the meditation itself.
+The meditation remains the systematic observation of the nervous system.
 
-While I would like to avoid descriptions of personal experiences, a few may help highlight the above description of new distracted states of mind.
+\pagebreak
 
-The non-example:
-the resolution to paradox
+\begin{center}
+  \Huge{Examples}
+\end{center}
+
+While I would like to avoid descriptions of personal experiences, a few may help highlight the above description of new and unfamiliar distracted states of mind.
+
+These examples also serve a secondary purpose.
+If one searches for ``vipassana'' on YouTube, one will find Goenka's instructional videos, critiques from individuals who could not complete their course, TED talks, and experience reports from individuals who did complete their first course.
+These experience reports (often titled ``My experience...'') tend to confuse these distracting experiences with the meditation itself.
+Framing a few examples within this paper should help the reader with this differentiation.
+
+\vspace{1cm}
+\begin{center}
+  \LARGE{The non-example:}\\
+  \LARGE{The Resolution to Paradox}
+\end{center}
 
@@
 Paradoxes and their solutions appear to be intrinsic to the human makeup, somehow.
+While the resolution of paradox may be of acute interest to many hackers, indulging in the consideration of a manifest paradox is as much a tangent and distraction to Vipassana as it is to Zazen.
 
-Holding a baby
+\vspace{1cm}
+\begin{center}
+  \LARGE{Holding a Baby}
+\end{center}
 
-A friend once had a very salvia-like experience, feeling as though she was teleported to her friend's kitchen, holding her friend's 2-year-old while he cried.
+A friend once had a hallucination-like experience, feeling as though she was teleported to her friend's kitchen, holding her friend's 2-year-old while the child cried.
 The experience was so raw, realistic, and unexpected that she actually opened her eyes and stood up in her meditation cell.
 
-Writing a poem
+Such distractions are difficult to overcome, since they appear to replace consciousness and reality itself.\\
 
-Often, in early stage meditation, a person will find him or herself distracted by an upcoming task: write this email, call this person, do this chore.
-
-I had a strangely parallel experience toward the end of my second 10-day course.
-My brain appeared to have written a poem: a visceral, all-encompassing work of art which touched my every sense and filled my mind.
-The temptation was not to read the poem (since ``I'' had written it) but to write the last line...
-to take credit for the poem and make it complete.
-Given that this genuinely felt like the most beautiful thing I had ever created, it was extremely difficult to fight the temptation to finish it and instead continue scanning my lower back with my awareness.
-
-Time slowing down
+\vspace{1cm}
+\begin{center}
+  \LARGE{Time Slowing Down}
+\end{center}
 
@@
 However, this is not what I am referring to.
-The experience of time slowing down in very deep meditative states (which I have not experienced a great deal, granted) is that of time slowing down a great deal ...10s or 100s of thousands of years pass before your awareness in a single meditation period.
+The experience of time slowing down in very deep meditative states is that of time slowing down a \textbf{great deal} ...tens or hundreds of thousands of years pass before your awareness in a single meditation period.
 
-Time speeding up
+It is difficult to describe how this effect remains nothing but a distraction.
+However, if meditation remains subdivided into two categories, the meditation object and distractions therefrom, we must accept that even experiences which alter our fundamental perceptions are little more than our brain's creative attempts to distract us from the meditation itself.
 
-The flip side of this, of course, is time speeding up.
-One sits down to meditate, takes a comfortable position, begins the process in earnest, and BOOM, the gong is going off.
-The meditation period is over.
+\pagebreak
 
-Time going backward
-
-Much harder to describe than the experience of time stretching out or time compressing into a single instant is the experience of time moving backward.
-It would appear, to one's awareness, that cause and effect have been effectively reversed.
-Each moment takes you a moment (or many) backward in time, and the outcomes of your thoughts or feelings are the source material to their original causes, now future events.
-
-This experience is not necessarily isolated, either.
-One may experience this as a sort of ``scrobbling'' of time back and forth.
-This can be quite disorienting and, as distractions go, is extremely difficult to avoid while trying to keep your attention moving over your right nipple.
-
-But that's still the game.
-
-Outcomes
+\begin{center}
+  \Huge{Outcomes}
+\end{center}
 
@@
 
-There is a sketch in ``Where Am I?'' \cite{whereami} about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:
+There is a sketch in \textit{Where Am I?} \cite{whereami} about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:\\
 
-``Honey, don’t forget what the half-life is on the autonomic nervous system.''
+\textbf{``Honey, don’t forget what the half-life is on the autonomic nervous system.''}\\
 
 The autonomic nervous system is what causes the ``winding stack'' of thought and emotion described earlier.
-The autonomic nervous system is what makes -- and keeps -- you angry.
+The autonomic nervous system is what makes --- and keeps --- you angry.
 And afraid.
@@
 And anxious.
-And everything else.
+And everything else you feel.
 
@@
 On the other end of the spectrum are the subtler qualities: Meditation seems to make a person more generous, more compassionate, more empathetic.
-Unlike the trick to healthy rationality — intentional pacifying of the autonomic nervous system — I do not have a pop neuroscience answer as to why this is the case.
+Why would systematic analysis of the autonomic nervous system have these effects?
+A wide body of research suggests that many forms of meditation, including Vipassana, exhibit these positive effects.
+Many often identify specifically enhanced activity in the prefrontal cortex \cite{differentialengagement, meditationbrainstructure}, which is identified as the centre of the will to live and altruistic tendencies in the human brain.
+\cite{personalityneuroscience} However, literature explaining why or how such effects may occur is limited.
 
-I do find the entire endeavour not just fascinating, but endlessly worthwhile. Vipassana is simultaneously the hardest thing I have ever done — and the most valuable.Further Reading
+Regardless of the mechanism, I do find the entire endeavour not just fascinating, but endlessly worthwhile. Vipassana is simultaneously the hardest thing I have ever done --- and the most valuable.
 
-My write-up of my first 10-Day Vipassana course:
-http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html
+\pagebreak
 
-My comparison of meditation techniques and various drugs:
-http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html
+\begin{center}
+  \Huge{Further Reading}
+\end{center}
 
-Appendix A:
-Can we ``hack'' Vipassana?
+\vspace{1cm}
+
+\begin{itemize}
+\item \textit{The Art of Living: Vipassana Meditation As Taught by S. N. Goenka} (Hart, 1987)
+\item \textit{An Atheist's Glimpse of God}, my write-up of my first 10-Day Vipassana course:
+  \begin{itemize}
+  \item \url{http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html}
+  \end{itemize}
+\item \textit{Drugs, meditation, warnings}, my comparison of meditation techniques and various drugs:
+  \begin{itemize}
+  \item \url{http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html}
+  \end{itemize}
+\end{itemize}
+
+\pagebreak
+
+\begin{center}
+  \Huge{Appendix A:}\\
+  \Huge{Can we ``hack'' Vipassana?}
+\end{center}
 
@@
 Let's ask Martin Thompson.
-The following slide is from his 2013 talk ``Performance Testing Java Applications'' \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is appropriate.
+The following slide in Figure~\ref{fig:load-test} is from his 2013 talk \textit{Performance Testing Java Applications} \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example not only appropriate but reasonably accessible.
 
-He suggests that any system we intend to observe should be observed through a 3rd party observer by dumping network traffic.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/deathstar1.png}
+  \caption{A computer system under load test.}
+  \label{fig:load-test}
+\end{figure}
+
+He suggests that any system we intend to observe should be observed through a 3rd party observer by dumping network traffic (see Figure~\ref{fig:objective-load-test}).
 For computer systems, this makes a great deal of sense.
@@
 
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/deathstar2.png}
+  \caption{A load test with an objective, 3rd-party observer.}
+  \label{fig:objective-load-test}
+\end{figure}
+
 Unfortunately, human biology is not so keen on this setup.
-Even if we had the technology to tap into the interactions of the brain and nervous system (we don't), there is no way to observe the interactions of the brain and nervous system without immediately falling into the Matrix Paradox.
+Even if we had the technology to tap into the interactions of the brain and nervous system (we don't), there is no way to observe the interactions of the brain and nervous system without immediately falling into the Matrix Paradox (see Figure~\ref{fig:nervous-system-observer}).
 How do I know the tool I used is trustworthy?
@@
 This is all once again external.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{images/deathstar3.png}
+  \caption{The nervous system cannot have an objective observer.}
+  \label{fig:nervous-system-observer}
+\end{figure}
 
@@
 Did I just feel my cardiovascular system in exquisite detail or did I imagine it?
-…but it is not possible to do so without breaking the meditation for the period of this observation.
+...but it is not possible to do so without breaking the meditation for the period of this observation.
 
@@
 
+\bibitem{salvia}
+  \url{https://en.wikipedia.org/wiki/Salvia_divinorum}
+  \textit{Salvia divinorum}.
+  Wikipedia.
+
+\bibitem{differentialengagement}
+  Holzel B.K., Ott U., Hempel H., Hackl A., Wolf K., Stark R., Vaitl D.
+  \textit{Differential engagement of anterior cingulate and adjacent medial frontal cortex in adept meditators and non-meditators}
+  Neuroscience Letters, 421:16–21, 2007.
+
+\bibitem{meditationbrainstructure}
+  Fox, Kieran C.R.; Nijeboer, Savannah; Dixon, Matthew L.; Floman, James L.; Ellamil, Melissa; Rumak, Samuel P.; Sedlmeier, Peter; Christoff, Kalina
+  \textit{Is meditation associated with altered brain structure? A systematic review and meta-analysis of morphometric neuroimaging in meditation practitioners}
+  Neuroscience \& Biobehavioral Reviews, 43:48–73, June 2014.
+
+\bibitem{personalityneuroscience}
+  Colin G. DeYoung, Jacob B. Hirsh, Matthew S. Shane, Xenophon Papademetris, Nallakkandi Rajeevan, Jeremy R. Gray.
+  \textit{Testing Predictions From Personality Neuroscience: Brain Structure and the Big Five}
+  Psychological Science, 21(6):820–828, June 2010.
+
 \bibitem{testingjava}

### 2017-11-11 vipassana-for-hackers.tex
fixing warnings... have not fixed hindi rendering (yet)

@@
 
-\usepackage{hyperref}
 \usepackage{graphicx}
@@
 \usepackage{enumitem}
+\usepackage{hyperref}
+
+\setlength{\parindent}{2em}
+\setlength{\parskip}{0.7em}
 \setlist{noitemsep}
@@
 
-I have drawn my personal spectrum for sample activities below.
+I have drawn my personal spectrum for sample activities in Figure~\ref{fig:meditation-spectrum-1}.
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
@@
 On that note, I feel I should introduce an absolutely fantastic RadioLab episode (\textit{Where Am I?} \cite{whereami}) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience.
-On the topic of lacking proprioception: ``We have words like `deaf' and `blind'.
-We do not have words for being deaf or blind to one's own body.'' In the Matrix, perhaps this is either a bug or a feature flag.
+On the topic of lacking proprioception:
+
+\begin{quote}
+  \textbf{We have words like `deaf' and `blind'. We do not have words for being deaf or blind to one's own body.}
+\end{quote}
+
+In the Matrix, perhaps this is either a bug or a feature flag.
 
@@
 Let us quickly revisit our third suspension of disbelief: ``dissolution of the body is a thing.'' Neuroscience tells us some interesting things about the nature of our sense of self and how it relates to our body.
-From ``Where Am I?'' \cite{whereami}:\\
+From \textit{Where Am I?} \cite{whereami}:
 
-\textbf{``The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.''}\\
+\begin{quote}
+  \textbf{The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.}
+\end{quote}
 
-\textbf{``Our \textit{being} is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.''}\\
+\begin{quote}
+  \textbf{Our \textit{being} is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.}
+\end{quote}
 
@@
 
-Such distractions are difficult to overcome, since they appear to replace consciousness and reality itself.\\
+Such distractions are difficult to overcome, since they appear to replace consciousness and reality itself.
 
@@
 
-There is a sketch in \textit{Where Am I?} \cite{whereami} about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:\\
+There is a sketch in \textit{Where Am I?} \cite{whereami} about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:
 
-\textbf{``Honey, don’t forget what the half-life is on the autonomic nervous system.''}\\
+\begin{quote}
+  \textbf{Honey, don’t forget what the half-life is on the autonomic nervous system.}
+\end{quote}
 

### 2017-11-11 vipassana-for-hackers.tex
switching from pdflatex to xelatex... hopefully this does not ruin too many things

@@
 \documentclass{article}
+
+\usepackage{polyglossia}
+\setdefaultlanguage{english}
+\setotherlanguage{sanskrit}
+\usepackage{fontspec}
+\newfontfamily\devanagarifont[Scale=MatchUppercase]{Kohinoor Devanagari}
 
@@
 \begin{center}
-  \LARGE{Habit's Exception: Sankhara}
+  \LARGE{Habit's Exception: Saṅkhāra}
 \end{center}
 
-The notion of Sankhara can be split into two camps, in terms of its definition and use.
+The notion represented by the Pali word \textit{Saṅkhāra} can be split into two camps, in terms of its definition and use.
 The first, the notion that the body and mind contain ``impurities'' and that the objective of meditation is to rid the body-mind of such impurities, will be excluded from all discussion here.
-The second, more common use of the term ``Sankhara'' refers to ``conditioned phenomena'' or ``mental dispositions''.
-The Hindi word, संस्कार (``sanskaar'') roughly translates in this context to ``impression''.
+The second, more common use of the term ``Sankhara'' refers to ``conditioned phenomena'' or ``psychological dispositions''.
+The equivalent Hindi/Sanskrit word, \foreignlanguage{sanskrit}{संस्कार} (sanskaar) roughly translates in this context to ``mental impression''.
 This second definition will apply to this paper.
@@
 Here, we might choose a more specific term from Psychology: habits.
-Sankharas, as far as this document is concerned, can be thought of as the ingrained habits of the brain and nervous system.
+Saṅkhāras, as far as this document is concerned, can be thought of as the ingrained habits of the brain and nervous system.
 Some concrete examples include liking (or disliking) the taste of olives, my body's ability to type at this QWERTY keyboard without consciously considering each keystroke, and the physical and emotional response I have to news of violence in my city of residence.
@@
 
-Put more succinctly, your activities will include: Eat, sleep, poop, bath, meditate.\\
+Put more succinctly, your activities will include: Eat, sleep, poop, bath, meditate.
 
-That's it.\\
+That's it.
 

### 2017-11-11 build.sh

@@ -1,1 +1,10 @@
-pdflatex vipassana-for-hackers.tex
+#!/bin/sh
+
+# pdflatex was used originally but cannot render Sanskrit without a lot of hoop-jumping
+# pdflatex vipassana-for-hackers.tex
+
+# xelatex is apparently less capable than pdflatex at font-rendering in general, but
+# is capable of rendering non-Latin fonts (like Hindi/Sanskrit) by default
+
+xelatex vipassana-for-hackers.tex
+

### 2017-11-11 vipassana-for-hackers.tex
fixing warnings in bibliography

@@
 
-\setlength{\parindent}{2em}
+\setlength{\parindent}{1.3em}
 \setlength{\parskip}{0.7em}
@@
 Instruction on a course includes 11+ hours of video instruction, at least twice as many hours of audio instruction, and multiple interviews per day with the teacher.
-Trying to replace this with a book is both inadequate and dangerous.\\
+Trying to replace this with a book is both inadequate and dangerous.
 
@@
 The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way.
-\textit{[citation needed]}\\
+\textit{[citation needed]}
 
@@
 
-This I/O feedback between the brain and body appears to create \textit{a completely new kind of distraction, which goes well beyond new thought patterns}.
-An attempt to highlight this form of distraction is given in Figure~\ref{fig:vipassana-sense-map-4}.
-
 \begin{figure}[h]
@@
 \end{figure}
+
+This I/O feedback between the brain and body appears to create \textit{a completely new kind of distraction, which goes well beyond new thought patterns}.
+An attempt to highlight this form of distraction is given in Figure~\ref{fig:vipassana-sense-map-4}.
 
@@
 The autonomic nervous system is what causes the ``winding stack'' of thought and emotion described earlier.
-The autonomic nervous system is what makes --- and keeps --- you angry.
+The autonomic nervous system is what makes (and keeps) you angry.
 And afraid.
@@
   \begin{itemize}
-  \item \url{http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html}
+  \item \href{http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html}
+             {\nolinkurl{http://blog.deobald.ca/2013/02/}}\\
+        \href{http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html}
+             {\nolinkurl{an-atheists-glimpse-of-god.html}}
   \end{itemize}
@@
   \begin{itemize}
-  \item \url{http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html}
+  \item \href{http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html}
+             {\nolinkurl{http://blog.deobald.ca/2015/01/}}\\
+        \href{http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html}
+             {\nolinkurl{drugs-meditation-warnings.html}}
   \end{itemize}
@@
 \begin{thebibliography}{9}
+\raggedright
+
 \bibitem{thenewsroom}

### 2017-11-13 vipassana-for-hackers.tex
version bump to 1.3
tweaking final remaining hbox overflow

@@
      \large\textit{Steven Deobald}\\
-     \large\textit{Version 1.2}\\
+     \large\textit{Version 1.3}\\
      \large\textit\today
@@
 
-Here, we might choose a more specific term from Psychology: habits.
+Here, we might choose a more specific term from Psychology: mental habits.
 Saṅkhāras, as far as this document is concerned, can be thought of as the ingrained habits of the brain and nervous system.
@@
 
-More complex variations are added as the bodily sensation experienced through this probing of the nervous system becomes increasingly subtle. In this order:
+More complex variations are added to these mechanics as the bodily sensation experienced through this probing of the nervous system becomes increasingly subtle. In this order:
 

### 2017-11-13 vipassana-for-hackers.tex
fixing inline references from quote to italics

@@
 
-A 10-day Vipassana course requires that all students take the standard ``5 Precepts'' \cite{fiveprecepts}, as per Theravada.
+A 10-day Vipassana course requires that all students take the standard \textit{5 Precepts} \cite{fiveprecepts}, as per Theravada.
 I do not recommend getting drunk, raping and murdering your fellow students, and stealing their belongings.
@@
 The easiest and most recent examples of this are no doubt the works of Douglas Hofstadter.
-``Gödel, Escher, Bach: An Eternal Golden Braid'' captured the imaginations of the hacker community and grew from esoteric pub conversation piece to dinner party staple, around the same time as the term ``blogging'' evolved from portmanteau to an accessible part-time career for stay-at-home parents.
+\textit{Gödel, Escher, Bach: An Eternal Golden Braid} captured the imaginations of the hacker community and grew from esoteric pub conversation piece to dinner party staple, around the same time as the term ``blogging'' evolved from portmanteau to an accessible part-time career for stay-at-home parents.
 The term ``strange loop'', coined in this book, has since become a mild obsession of philosophical and artistic computer scientists, linguists, and mathematicians, even inspiring the namesake software conference held in St.
 Louis.
-Hofstadter's more cohesive ``I Am a Strange Loop'' tackles the problem of consciousness more succinctly.
+Hofstadter's more cohesive \textit{I Am a Strange Loop} tackles the problem of consciousness more succinctly.
 However, other than directing the reader to Zen\footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana.
@@
 There is no hacking.
-A thorough read of GEB or IAaSL is akin to reading all four volumes of Knuth's ``The Art of Computer Programming'' without having ever touched a computer or to reading ``The Joy of Cooking'' cover to cover without having ever set foot inside a kitchen.
+A thorough read of GEB or IAaSL is akin to reading all four volumes of Knuth's \textit{The Art of Computer Programming} \cite{taocp} without having ever touched a computer or to reading \textit{The Joy of Cooking} \cite{joyofcooking} cover to cover without having ever set foot inside a kitchen.
 
@@
 \begin{itemize}
-\item Homoiconicity: the language is represented in itself and code written in the language is understood by the language
-\item Metaprogramming: the language can program itself and understand itself
-\item Self-mutation and self-replication: a LISP program can inherently alter its own internal behaviour and easily write entirely new programs
-\item Recursion and trampolining: mathematical functional recursion and mutual recursion between two functions are natural operations in LISP
+\item \textbf{Homoiconicity:} the language is represented in itself and code written in the language is understood by the language
+\item \textbf{Metaprogramming:} the language can program itself and understand itself
+\item \textbf{Self-mutation and self-replication:} a LISP program can inherently alter its own internal behaviour and easily write entirely new programs
+\item \textbf{Recursion and trampolining:} mathematical functional recursion and mutual recursion between two functions are natural operations in LISP
 \end{itemize}
@@
 The Consciousness Machine.
-Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' The machine at the bottom of your consciousness \textbf{is} a strange loop.
+Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' \textit{The machine at the bottom of your consciousness is a strange loop}.
 Having a look at this strange loop is a lot more fun than reading about it.
@@
 
-Abrash discusses the nature of the ``Matrix Paradox'' (more commonly known as the ``Simulation Hypothesis'' \cite{simulationhypothesis}) as simply the right neurons firing to simulate any particular environment.
+Abrash discusses the nature of the ``Matrix Paradox'' (more commonly known as the \textit{Simulation Hypothesis} \cite{simulationhypothesis}) as simply the right neurons firing to simulate any particular environment.
 ``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input.
@@
 However, just because the room you are in has a negative decibel rating does not mean you cannot hear.
-In fact, in this environment, your sense of hearing is likely to be amplified — and turned inward.
+In fact, in this environment, your sense of hearing is likely to be amplified --- and turned inward.
 You will hear your heart beating, your blood pumping, your digestive system sloshing around.
@@
 
-For this reason, the above anechoic chamber has a record of 45 minutes for the longest period anyone has sat inside it.
+For this reason, the above anechoic chamber at Orfield Laboratories \cite{orfield} has an unofficial record of 45 minutes for the longest period anyone has sat inside it.
 As it turns out, hearing one's own bodily functions at deafening internal volumes is likely to cause someone to go a bit crazy, hence the highlighted bar for thought activity.
@@
 A wide body of research suggests that many forms of meditation, including Vipassana, exhibit these positive effects.
-Many often identify specifically enhanced activity in the prefrontal cortex \cite{differentialengagement, meditationbrainstructure}, which is identified as the centre of the will to live and altruistic tendencies in the human brain.
-\cite{personalityneuroscience} However, literature explaining why or how such effects may occur is limited.
+Many often identify specifically enhanced activity in the prefrontal cortex \cite{differentialengagement, meditationbrainstructure}, which is identified as the centre of the will to live and altruistic tendencies in the human brain.\cite{personalityneuroscience} However, literature explaining why or how such effects may occur is limited.
 
@@
 Let's ask Martin Thompson.
-The following slide in Figure~\ref{fig:load-test} is from his 2013 talk \textit{Performance Testing Java Applications} \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example not only appropriate but reasonably accessible.
+The following slide in Figure~\ref{fig:load-test} is from his 2013 talk \textit{Performance Testing Java Applications} \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is not only appropriate but reasonably accessible.
 
@@
 Approach the practice skeptically until you decide to take a course and then save your analysis of whatever newfound skepticism the course generates for you until after the course is completed.
-There will be plenty of time to discuss the meditation and the outcomes of the meditation on the 10th day when talking is permitted; explore your skepticism here.
+There will be plenty of time to discuss the meditation and the outcomes of the meditation on the tenth day when talking is permitted; explore your skepticism here.
 Other meditators are likely to be the only ones who will quite comprehend the concepts you are trying to discuss.
@@
 
+\bibitem{taocp}
+  Donald Knuth.
+  \textit{The Art of Computer Programming}.
+  Addison-Wesley, 1962, 1968, 1969, 1973, 2005, 2011, 2015, 2017.
+  ISBN: 0-201-03801-3
+
+\bibitem{joyofcooking}
+  Irma S. Rombauer, Marion Rombauer Becker.
+  \textit{The Joy of Cooking, Sixth Edition}.
+  Bobbs-Merrill, 1975.
+  ISBN: 0-02-604570-2
+
 \bibitem{hackerhowto}
@@
 
+\bibitem{orfield}
+  Lindsey Galloway.
+  \url{http://www.bbc.com/travel/story/20121022-the-quietest-place-on-earth}
+  \textit{The quietest place on Earth}
+  BBC, 23 October 2012.
+
 \bibitem{glickman}

### 2018-04-02 vipassana-for-hackers.tex
v1.4
added bridge/wall imagery

@@
      \Huge\textbf{Vipassana for Hackers}\\
-     \Large\textbf{Instigating the Curious}\\
      \vspace{5cm}
      \large\textit{Steven Deobald}\\
-     \large\textit{Version 1.3}\\
+     \large\textit{Version 1.4}\\
      \large\textit\today
@@
 \begin{center}
+  \LARGE{Anapana: The Bridge}
+\end{center}
+
+At the end of the first day of a 10-day Vipassana course, the breath is described as a bridge ``from the known to the unknown, because respiration is one function of the body that can be either conscious or unconscious, intentional or automatic.'' During the 3-day Anapana session of a 10-day course, students are instructed to \textit{consciously} direct attention toward their breath -- and even gently control the breath for brief periods to maintain attention when the mind won't stop wandering -- until awareness of breath, without force, lasts for extended periods.
+Given that the breath is the one function of the body which operates both consciously and unconsciously, it makes sense that it acts as the starting point for many meditation practies, including Vipassana.
+The intention of Vipassana is to train the mind out of habits and tendencies which have embedded themselves into the brain and nervous system at an unconscious level.
+To do this, one must first learn to access one's unconscious.
+Through the breath, Anapana meditation fulfills this purpose.
+
+Vipassana requires a 10-day course to learn.
+Anapana does not.
+Before his death, S.N.
+Goenka released both 70-minute audio instructions as MP3s for the Anapana meditation technique.\cite{anapana} The 70-minute instructions can be found at \url{http://www.vipassana.co/What-is-Anapana} under ``Anapana Meditation for All'' and come as four 15-minute sessions plus a 10-minute ``daily practice'' session.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=8cm]{images/bridge.png}
+  \label{fig:bridge}
+\end{figure}
+
+\pagebreak
+
+\begin{center}
   \Huge{Vipassana Meditation:}\\
@@
 Similarly, these deep states of meditation will eliminate audio input as well, which is easier for anyone to detect.
+
+\begin{center}
+  \LARGE{Vipassana: The Sledgehammer}
+\end{center}
+
+During the Day 4 discourse of a 10-day course, the imagery of the bridge is extended by what is on the far side of that bridge: a giant wall.
+The wall represents the remaining barrier between the conscious and unconscious mind -- and Vipassana is the tool used to break this barrier.
+By systematically examining the nervous system, outside-in, one digs through all the layers of unconscious habit as it manifests in bodily sensation.
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/wall.jpg}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/bridge.png}
+  \end{subfigure}
+  \label{fig:wall}
+\end{figure}
+
+And what comes of this examination?
+Sensations in the body are exploded into view and what we normally experience as a narrow band of sensation from the roughly unpleasant to the roughly pleasant: intense sensations like abdominal pain and annoying sensations like uncontrollable itching are gross sensations we normally consider unpleasant; intense sensations like sexual contact and light sensations like touching fur are gross sensations we normally consider pleasant.
+
+This explosion of sensation occurs on an entirely new axis.
+The range can still be considered pleasant to unpleasant but ``pleasant'' begins to cover new, previously unexperienced territory.
+These experiences have little or no parallel in mundane life and are best described as ``subtle'': fine oscillations and vibration.
+This new axis of experience is the sledgehammer used to break through the wall separating conscious from unconscious.
+Unpleasant sensations and pleasant sensations, both, undergo change.
+As one observes this change, the change itself comes faster and faster and becomes easier and easier to observe.
+The goal is to avoid reacting to unpleasant sensations with aversion and to avoid reacting to these newfound pleasant sensations with craving.
+The task is harder than it sounds.
+However, the calmer a meditator remains while exploring these new depths, the further she explores.
+
+Unfortunately, unlike Anapana meditation, there is no way to learn Vipassana from an MP3.
+This practice of observing subtler and subtler sensation requires quite a bit of time, assistance from a live teacher, and repeated instruction.
+The only way one can learn the technique of Vipassana is to take a 10-day course.
 
@@
 
+\bibitem{anapana}
+  S.N. Goenka
+  \url{http://www.vipassana.co/What-is-Anapana}
+  \textit{What is Anapana?}
+  Vipassana Research Institute, 21 October 2016.
+
 \bibitem{glickman}

### 2018-04-16 vipassana-for-hackers.tex
bump version
spelling
practices spelling
s/practies/practices/

@@
      \large\textit{Steven Deobald}\\
-     \large\textit{Version 1.4}\\
+     \large\textit{Version 1.5}\\
      \large\textit\today
@@
 At the end of the first day of a 10-day Vipassana course, the breath is described as a bridge ``from the known to the unknown, because respiration is one function of the body that can be either conscious or unconscious, intentional or automatic.'' During the 3-day Anapana session of a 10-day course, students are instructed to \textit{consciously} direct attention toward their breath -- and even gently control the breath for brief periods to maintain attention when the mind won't stop wandering -- until awareness of breath, without force, lasts for extended periods.
-Given that the breath is the one function of the body which operates both consciously and unconsciously, it makes sense that it acts as the starting point for many meditation practies, including Vipassana.
+Given that the breath is the one function of the body which operates both consciously and unconsciously, it makes sense that it acts as the starting point for many meditation practices, including Vipassana.
 The intention of Vipassana is to train the mind out of habits and tendencies which have embedded themselves into the brain and nervous system at an unconscious level.

### 2018-05-10 vipassana-for-hackers.tex
Apply fixes by @sycrat and @liammonahan in the base repo
Co-authored-by: Aaron Blair <aaron@aaronpb.me>
Co-authored-by: Liam Monahan <liam@liammonahan.com>
typo fix

--- 2018-04-02 vipassana-for-hackers.tex
@@
 
-The first 3.5 days of a 10-day Vipassana course are spend practicing ``Anapana'' (breath) meditation.
+The first 3.5 days of a 10-day Vipassana course are spent practicing ``Anapana'' (breath) meditation.
 This period is very important as it is time spent practicing acute, one-pointed attention.
@@
 What is meant by ``prodding our nervous system'' and what does this have to do with brain activity?
-Our second suspension of disbelief was of the deep and intense connection between the brain the nervous system.
+Our second suspension of disbelief was of the deep and intense connection between the brain and the nervous system.
 We cannot feel the brain directly --- or at all.

### 2019-05-04 paper-one/build-epub.sh
epub build cleanup

@@ -6,4 +6,4 @@ latexml --dest=paper-one.xml paper-one.tex
 latexmlpost -dest=paper-one.html paper-one.xml
 which ebook-convert
 /usr/local/bin/ebook-convert
-ebook-convert paper-one.html paper-one.epub --language en --output-profile=kindle --authors="Steven Deobald" --title="Vipassana for Hackers"
+ebook-convert paper-one.html ../output/paper-one.epub --language en --output-profile=kindle --authors="Steven Deobald" --title="Vipassana for Hackers"

### 2019-05-06 paper-one/build-epub.sh
move sty files to lib dir

@@ -4,6 +4,6 @@ OUT=../output
 
 latexml --dest=paper-one.xml paper-one.tex
 latexmlpost -dest=paper-one.html paper-one.xml
-which ebook-convert
-/usr/local/bin/ebook-convert
+# which ebook-convert
+# /usr/local/bin/ebook-convert
 ebook-convert paper-one.html ../output/paper-one.epub --language en --output-profile=kindle --authors="Steven Deobald" --title="Vipassana for Hackers"

### 2019-05-06 paper-one/build-pdf.sh

@@ -8,9 +8,16 @@
 
 # Compile twice to resolve references.
 
+SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
+LIBPATH="$SCRIPTPATH/../lib"
+
+PATH=$PATH:$LIBPATH
+
 OUT=../output
 
-xelatex -output-directory=$OUT paper-one.tex
+echo "Running pre-build. This takes a while..."
+xelatex -output-directory=$OUT paper-one.tex > /dev/null 2>&1
+echo "...done. Now running the build."
 xelatex -output-directory=$OUT paper-one.tex
 
 rm $OUT/*.aux $OUT/*.log $OUT/*.out

### 2019-05-06 paper-one/build-epub.sh
css to /lib, clean up build artifacts

@@ -1,9 +1,17 @@
 #!/bin/sh
 
+SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
+LIBPATH="$SCRIPTPATH/../lib"
+PATH=$PATH:$LIBPATH
+
 OUT=../output
 
 latexml --dest=paper-one.xml paper-one.tex
 latexmlpost -dest=paper-one.html paper-one.xml
+
 # which ebook-convert
 # /usr/local/bin/ebook-convert
+
 ebook-convert paper-one.html ../output/paper-one.epub --language en --output-profile=kindle --authors="Steven Deobald" --title="Vipassana for Hackers"
+
+rm paper-one.xml paper-one.html LaTeXML.cache LaTeXML.css ltx-article.css

### 2019-05-06 paper-one/build-pdf.sh

@@ -10,7 +10,6 @@
 
 SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
 LIBPATH="$SCRIPTPATH/../lib"
-
 PATH=$PATH:$LIBPATH
 
 OUT=../output
@@ -20,4 +19,4 @@ xelatex -output-directory=$OUT paper-one.tex > /dev/null 2>&1
 echo "...done. Now running the build."
 xelatex -output-directory=$OUT paper-one.tex
 
-rm $OUT/*.aux $OUT/*.log $OUT/*.out
+rm texput.log $OUT/*.aux $OUT/*.log $OUT/*.out

### 2019-05-06 paper-four/paper-four.tex
paper four first post

--- 2018-04-16 vipassana-for-hackers.tex
@@
 
-\usepackage{polyglossia}
-\setdefaultlanguage{english}
-\setotherlanguage{sanskrit}
-\usepackage{fontspec}
-\newfontfamily\devanagarifont[Scale=MatchUppercase]{Kohinoor Devanagari}
-
-\usepackage{graphicx}
-\usepackage{subcaption}
-\usepackage{float}
-\usepackage[bottom]{footmisc}
 \usepackage{enumitem}
-\usepackage{hyperref}
 
@@
      \Huge\textbf{Vipassana for Hackers}\\
+     \Huge{Paper Four: Mastering Spacetime}\\
      \vspace{5cm}
      \large\textit{Steven Deobald}\\
-     \large\textit{Version 1.5}\\
-     \large\textit\today
+     \large\textit{Version 0.1 DRAFT}\\
+     \large\textit\today\\
+     \vspace{5cm}
+     \large\textit{}\\
    \end{center}
@@
 
-Should you bother reading this paper?
-Not everyone associates themselves with the label ``Hacker''.
-It comes with a few well-known negative connotations so my first task is to quickly identify what is meant by ``Hacker'', since this paper is for these people.
-
-My first exposure to the term ``hacking,'' outside of the late-80's imagery which became fodder for the 1995 film, was through my cousin Jennie.
-She is an Ecologist, a self-described ``Mud Scientist'', who most people would not associate with the term ``Hacker''.
-She visited MIT during her PhD to find the term extended to ``roof and tunnel hacking,'' an idea and practice which delighted her to no end.
-MIT or not, Jennie is a Hacker.
-Every Science is about exploration and discovery, and these are the concepts at the root of ``hacking.''
-
-Carpentry, painting, gardening, the entire field of Genetics, baking bread, Philosophy, electronics, fashion, Psychology, health, writing, and bicycles.
-Almost any discipline can fall under the banner of ``hacking'' I'm using for this paper, since almost any discipline can (but does not necessarily) include the potential for two essential aspects of the Hacker Mentality: Rationality and Curiosity.
-
-This is all that is required.
-
-If you have ever enjoyed taking apart a computer or car, experimenting with bread recipes, researching and writing, pulling apart the semantics of your work, building something, rebuilding something, exploring a cave or abandoned construction site...
-I wrote this paper for you.
-If you enjoy any exploratory vocation or hobby you will almost certainly find Vipassana utterly fascinating.
-The reason is this: As acutely as a human being can disassemble a stereo, examine the slide of a microscope, or explore a cave, human beings have the capacity to disassemble, examine, and explore ourselves.
-
-What may be even more fascinating, for the hacker, is the fact that by practicing Vipassana one can seemingly increase the capacity for rational thought and behaviour.
-
-I have written this paper because, thus far, I have found the written and recorded material on the subject of Vipassana to fall into three broad categories:
-
-Well-meaning Neuroscience papers which completely miss the mechanics of nervous-system-based meditation techniques and their core outcomes.
-Descriptions of a ``silent meditation retreat,'' usually given by someone who has not attended a course. \cite{thenewsroom}
-Descriptions of Vipassana from an individual and subjective perspective, usually given by someone who has attended one course. (``Oh my god! It was sooooo intense!'')
-
-By using phrases like ``disassemble, examine, and explore ourselves'' I'm at the risk of sounding spiritual.
-This is the opposite of my intention.
-My intention is to pique curiosity in fellow hackers by explaining the biological mechanics of Vipassana, to the best of my ability and as I understand them.
-To curb the impression that this paper falls into the category of New Age Philosophy, I would like to take a moment to address what this paper is not.
-
-\pagebreak
-
-\begin{center}
-  \Huge{What This Paper is Not}
-\end{center}
-
-This not an instruction manual. You will not learn Vipassana from reading this paper.
-
-This is not a white paper. I am not a scientist. This is not science.
-
-This is not an expert's treatise on the subject of nervous-system-based meditations, nor even an expert Vipassana meditator's description of the technique.
-I am not an expert.
-I have only attended four 10-day courses thus far.
-Each has been substantially different from the others.
-
-This paper will endeavour to avoid falling into the trap of ``Buddhist Guys'' (a mildly derogatory term coined by a serious meditator friend of mine).
-You might be familiar with phrases commonly attributed to this archetype such as ``When the student is ready, the teacher will appear.'' Mysticism, mystery, and mystique are to be avoided.
-If I fall into this trap, email me: \href{mailto:steven@nilenso.com}{steven@nilenso.com}
-
-This is not a discussion on ``spirituality'', for any definition of the term.
-
-This is not a dissection of the administrative details of the course, though I will provide a brief overview of the course schedule.
-
-This is not a description of past experiences, though I will use a few to highlight core effects of the technique.
-
-This is not a comparison with other meditation techniques, though I will address other techniques for context.
-It is worth noting that there are a few meditation techniques labelled ``Vipassana'' and this paper deals specifically with Vipassana as taught by S.N.
-Goenka.
-
-This is not an examination of Buddhist Philosophy and will exclude commentary or mention of the following concepts:
-
-\begin{itemize}
-  \item Siddhartha Gautama (The Buddha)
-  \item Enlightenment
-  \item Religion / Buddhism
-  \item Prophecy / Prediction of the future
-  \item Telepathy / ESP
-  \item Reincarnation
-  \item ``Energies''
-  \item Kalapas
-  \item Metta / Morality
-  \item Sankharas / ``Impurity''
-\end{itemize}
-
-These last two bullet points are subject to one exception each.
-
-\pagebreak
-
-\begin{center}
-  \LARGE{Morality's Exception: Lying}
-\end{center}
-
-A 10-day Vipassana course requires that all students take the standard \textit{5 Precepts} \cite{fiveprecepts}, as per Theravada.
-I do not recommend getting drunk, raping and murdering your fellow students, and stealing their belongings.
-However, the four precepts which forbid those behaviours are quite clearly external, making them components of course administration and therefore not applicable to this document.
-
-Lying, however, is perhaps least interesting in its external form.
-As an educated adult, it is difficult to say anything at all without lying to some small degree.
-As one begins to systematically observe the nervous system, which is the basic mechanic of Vipassana, one finds that the brain and nervous system are acutely aware of even the tiniest untruth.
-If I tell myself I am feeling something which I am not (``my body is exploding into a billion tiny particles!
-my muscles are raging rivers of lava inside my body!'') the process of empirical and objective observation ends, and the process stops.
-
-Therefore, the exception given to Lying (or the precept governing Lying) is because it actually prevents the technique from working, not because I have some particular crusade against lying.
-
-\vspace{1cm}
-
-\begin{center}
-  \LARGE{Habit's Exception: Saṅkhāra}
-\end{center}
-
-The notion represented by the Pali word \textit{Saṅkhāra} can be split into two camps, in terms of its definition and use.
-The first, the notion that the body and mind contain ``impurities'' and that the objective of meditation is to rid the body-mind of such impurities, will be excluded from all discussion here.
-The second, more common use of the term ``Sankhara'' refers to ``conditioned phenomena'' or ``psychological dispositions''.
-The equivalent Hindi/Sanskrit word, \foreignlanguage{sanskrit}{संस्कार} (sanskaar) roughly translates in this context to ``mental impression''.
-This second definition will apply to this paper.
-
-Here, we might choose a more specific term from Psychology: mental habits.
-Saṅkhāras, as far as this document is concerned, can be thought of as the ingrained habits of the brain and nervous system.
-Some concrete examples include liking (or disliking) the taste of olives, my body's ability to type at this QWERTY keyboard without consciously considering each keystroke, and the physical and emotional response I have to news of violence in my city of residence.
-
-\pagebreak
-
-\begin{center}
-  \Huge{A Finger Pointing at the Moon}
-\end{center}
-
-The fundamental difficulty in any description of Vipassana is this: It is impossible to convey a description of Vipassana in terms known to someone who has not experienced it.
-During a Vipassana course this paradox is addressed using two metaphors: First, a story of a sighted brother explaining the concept of ``white'' to his blind brother.
-\cite{experiencethetruth} Second, the classic Indian story of blind men each describing an elephant and each producing a partial truth.
-
-The first metaphor is a better framework for explaining the difficulty of writing about nervous system meditations.
-It is fundamentally impossible to express the experience of observing ``whiteness'' to an individual who is blind from birth.
-This person requires sight to understand.
-Similarly, it is fundamentally impossible to describe the effects of a nervous system meditation which alters perceptions and the interactions of the nervous system itself, since the nervous system (including the brain) ultimately form our entire experience.
-
-The hacker community has been on the verge of this exploration for quite some time.
-The easiest and most recent examples of this are no doubt the works of Douglas Hofstadter.
-\textit{Gödel, Escher, Bach: An Eternal Golden Braid} captured the imaginations of the hacker community and grew from esoteric pub conversation piece to dinner party staple, around the same time as the term ``blogging'' evolved from portmanteau to an accessible part-time career for stay-at-home parents.
-The term ``strange loop'', coined in this book, has since become a mild obsession of philosophical and artistic computer scientists, linguists, and mathematicians, even inspiring the namesake software conference held in St.
-Louis.
-Hofstadter's more cohesive \textit{I Am a Strange Loop} tackles the problem of consciousness more succinctly.
-However, other than directing the reader to Zen\footnote{Zen is relevant because the practice of Zazen (Zen meditation) is a nervous-system-based meditation similar to Vipassana.
-The reason I do not discuss Zazen in detail or compare it to Vipassana in this paper is because Zazen, in my experience, is not systematic.
-The subject of inspection — the nervous system — remains, but Zen teachers tend to avoid explicit instruction, leaving the student to explore on her own.
-The outcomes, therefore, appear similar but watered down.} at multiple points, these books do not invite the reader to explore these concepts directly.
-There is no hacking.
-A thorough read of GEB or IAaSL is akin to reading all four volumes of Knuth's \textit{The Art of Computer Programming} \cite{taocp} without having ever touched a computer or to reading \textit{The Joy of Cooking} \cite{joyofcooking} cover to cover without having ever set foot inside a kitchen.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Computing Sidebar: LISP}
-\end{center}
-
-I will attempt to frame the self-exploration of nervous system meditations like Vipassana using one tool familiar to the Computer Science discipline: LISP.
-If you are neither a Computer Scientist nor a Linguist and have no interest in computer languages, feel free to skip to the next section.
-
-Eric Raymond described the exercise of LISP mastery to be ``profound enlightenment'' \cite{hackerhowto} and while we can probably agree to ignore the hyperbole, there is something to be said about a language which begins as a calculus and nearly manages to escape formal grammar entirely.
-LISP has some very interesting properties:
-
-\begin{itemize}
-\item \textbf{Homoiconicity:} the language is represented in itself and code written in the language is understood by the language
-\item \textbf{Metaprogramming:} the language can program itself and understand itself
-\item \textbf{Self-mutation and self-replication:} a LISP program can inherently alter its own internal behaviour and easily write entirely new programs
-\item \textbf{Recursion and trampolining:} mathematical functional recursion and mutual recursion between two functions are natural operations in LISP
-\end{itemize}
-
-Combined, these properties tend to blur the lines of compile time and run time, of code and data, of execution and interpretation.
-Decades ago, thanks to these ``strange loop'' properties of LISP, it was thought to herald the dawn of the Age of Artificial Intelligence.
-While that did not happen it left us with a fascinating building material: LISP appears to have no upper-bound.
-It is ``turtles all the way up.'' LISP is not ``turtles all the way down'', as some would describe it.
-Even if written on a LISP Machine (a computer which executes LISP directly rather than transforming LISP into machine-level instructions), there are still paths of wires and electrons which need to be dealt with.
-These are not written in LISP.
-
-Vipassana would appear to be ``turtles all the way down'' in precisely this sense.
-Vipassana is the process of digging through your consciousness using consciousness itself.
-At no point do you reach the machine beneath the machine: ``Aha!
-I found it!
-The Consciousness Machine.
-Now I can do away with the meditation and just read the manufacturer's open specification and I'm sorted.'' \textit{The machine at the bottom of your consciousness is a strange loop}.
-Having a look at this strange loop is a lot more fun than reading about it.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Course Structure}
-\end{center}
-
-For those unfamiliar, I will briefly overview the rules of the course.
-During the course you are not permitted books, phones, talking, looking people in the eye, touching, making sounds, snacks, music, or writing implements.
-
-Put more succinctly, your activities will include: Eat, sleep, poop, bath, meditate.
-
-That's it.
-
-Meditation will consume approximately ten to eleven hours per day. All courses the world over are free and volunteer-run. Material goods are provided for by donations from previous students.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Suspending Disbelief}
-\end{center}
-
-Before explaining the mechanics and outcomes of Vipassana, I must request that you suspend disbelief about three items for the duration of the paper.
-
-\vspace{1cm}
-\begin{center}
-  \LARGE{1. Meditation is a thing}
-\end{center}
-
-For the purposes of this document, understand ``meditation'' to mean a single point of awareness with unwavering focus.
-``Unwavering focus'' means that we can hold sustained attention on this single point (whatever it may be) without getting distracted by thoughts of our taxes, what's for lunch, or that really persistent itch on our ear.
-
-What I mean by this is that mental exercises exist which can take consciousness beyond ``flow'', ``calm'', or ``elated'' states of mind caused by activities one might describe as ``meditative'' rather than ``meditation''.
-Biochemical responses to hallucinogens fall on the same side of the spectrum, though they may appear momentarily insightful.
-
-I have drawn my personal spectrum for sample activities in Figure~\ref{fig:meditation-spectrum-1}.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=11cm]{images/meditation-spectrum-1.jpeg}
-  \caption{Sample meditation/meditative spectrum.}
-  \label{fig:meditation-spectrum-1}
-\end{figure}
-
-For the purposes of this paper, we will keep all notions of ``mindfulness'' in the ``meditative'' camp, on the right.
-Mindfulness is not meditation.
-This paper will be discussing only two practices from this list: Anapana and Vipassana.
-An illustrated definition of ``awareness'' and ``attention'' follows, in the section entitled ``On Attention''.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=11cm]{images/meditation-spectrum-3.jpeg}
-  \caption{Sample meditation spectrum.}
-  \label{fig:meditation-spectrum-3}
-\end{figure}
-
-\vspace{1cm}
-\begin{center}
-  \LARGE{2. Your brain and nervous system have}\\
-  \LARGE{a tight feedback loop}
-\end{center}
-
-When we say ``tight feedback loop'', this means the brain and nervous system are tied together in a feedback loop faster and tighter than you could ever imagine without experiencing it directly.
-
-We all know the classic examples of connectivity between the brain and nervous system: A child touches a hot stove and retracts her hand.
-A pubescent boy sees a boy or girl he is attracted to and feels ``butterflies in his stomach'' so intense he vomits.
-Despite healthy diet and exercise, a high-strung individual suffers from stomach ulcers.
-
-These examples all exist on a gross level, coarse-grained enough that we can identify them externally and even give them names like ``butterflies''.
-However, your nervous system is interacting with your brain at every conceivable level.
-No matter how small, you do not have a thought or emotion emerge in your consciousness (or unconscious) without a physical sensation accompanying it.
-You do not feel physical sensation, consciously or subconsciously, without an accompanying thought or emotion.
-This interaction is continuous.
-
-Unless you have tried a meditation system with the nervous system as its meditation object, you have almost undoubtedly not experienced this.
-
-\vspace{1cm}
-\begin{center}
-  \LARGE{3. Dissolution of the body is a thing}
-\end{center}
-
-This will be the most difficult of the three.
-``Dissolution of the body,'' though strange-sounding at first, does not fall into the categories where we would normally place magic, levitation, or ESP.
-The term simply refers to our capacity to see through our entire body.
-
-As we investigate this concept, it seems increasingly reasonable in conjunction with points \#1 and \#2.
-If we can narrow our attention down to anything we can perceive and what we choose to perceive with our brain and nervous system in tandem is a narrow part of the body, surely if we sustain attention on that part for long enough we will start to see it clearly.
-
-Normally, our attention on our body is almost entirely external: my hands on this keyboard, the itch on my ear, my butt in this chair.
-I don't tend to spend a lot of energy focused on the sensations provided to my brain by my gallbladder or my stomach unless I have a painful gallstone or I am experiencing acute hunger.
-Physical sensations provided to our awareness by the nervous system for the internals of the body are largely hidden from us — but they can be unearthed and explored.
-
-I should note here that in four courses I have not felt the precise shape of my gallbladder or stomach at any point.
-I have ``seen'' the shape of my bones in my fingers, shoulders, chest, and pelvis.
-I have witnessed the rush of blood squooshing through my cardiovascular system (near the surface) with terrifying clarity.
-I have enough anecdotal evidence to say that it appears this process does work and can be extended to our deepest internal organs and the spinal cord.
-
-To help visualize what I mean, a sample spectrum of sensations from the gross or obvious to the curious and subtle might look like the figure on the following page.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/sensation-spectrum.jpeg}
-  \caption{Sample sensation spectrum.}
-  \label{fig:sensation-spectrum}
-\end{figure}
-
-As the field of sensation progresses to the left, the sensations felt become increasingly difficult to describe.
-``Surreal hot / cold'' refers to sensations of temperature we wouldn’t normally consider possible: our muscles either on fire or freezing solid.
-``Spatial distortion'' refers to complete alterations of spacial presence — the partial or complete replacement of one’s normal sense of proprioception.
-This may include reorientation in space (``I was upside-down!'') or the feeling that one is gigantic or tiny or that any one body part has one of these or other distorted qualities.
-
-``Throbbing'' refers to thick, heavy, slower vibrations; these are often felt rhythmically and this rhythm accelerates until the sensation is described as a ``vibration''.
-Beyond vibration is the ``invisible'', where a particular body part is no longer seen and a deeper part of the body is experienced directly.
-The first such example in most cases tends to be the skin, since the external surface of the body is the first part of the nervous system the course instructions ask the meditator to focus on.
-Beyond this the meditator may find muscle tissue, bones, the cardiovascular system, or something else entirely.
-Usually by the point in the course when a meditator has dissolved her own skin to her attention the sensations become extremely difficult to identify and label so ``something else entirely'' is likely to be the most common case.
-
-The sample spectrum above is missing many, many sensations.
-I also feel it necessary to call out that the spectrum is not serial; although I have felt ``vibrations'' and seen deeper than skin level, I have certainly not experienced all the sensations possible in the spectrum.
-However, the diagram hopefully provides a reference point for the type of readily-accessible sensations which Vipassana begins with (on the right) and the types of sensations (or absence of sensation) which will convince a person that complete dissolution of the body not only possible but an entirely natural process.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Virtual Reality}\\
-  \Huge{and}\\
-  \Huge{The Matrix Paradox}
-\end{center}
-
-The following section will require some visual assistance, for which I will use the same slides I used in the talk which corresponds to this paper.
-An unmodified slide is seen here in Figure~\ref{fig:michael-abrash-senses}.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/michael-abrash-senses.png}
-  \caption{Traditional 5-sense brain mapping.}
-  \label{fig:michael-abrash-senses}
-\end{figure}
-
-This image is taken from a talk by Michael Abrash, Chief Scientist of the Oculus VR team, at a talk he gave at the Facebook F8 Summit in 2015.
-\cite{abrashvr} I'm using this slide because it's nicely laid out and because it is incomplete.
-
-Abrash discusses the nature of the ``Matrix Paradox'' (more commonly known as the \textit{Simulation Hypothesis} \cite{simulationhypothesis}) as simply the right neurons firing to simulate any particular environment.
-``How do we know we are not in the matrix?'' The question is valid, but the diagram is missing a great deal of sensory input.
-Figure~\ref{fig:michael-abrash-senses-extended} is a slightly better model of reality.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/michael-abrash-senses-extended-attached.jpeg}
-  \caption{A more complete, 7-sense brain mapping.}
-  \label{fig:michael-abrash-senses-extended}
-\end{figure}
-
-Though a classic sense diagram such as Mr.
-Abrash's might describe proprioception \cite{ramachandranbrainfunction, proprioception}, nocioception \cite{nocioception}, and hunger \cite{fmrihunger} as physical ``feeling'', they are definitively not a subset of ``touch'', as they correspond to entirely different activity in the brain.
-Touch is strictly an external sense dealing with skin.
-These are internal senses.
-
-A brain diagram describing The Matrix Paradox is complete only when we map all senses to the brain.
-I have not, since I haven't named them all.
-In fact, neurobiology does not yet have names for all sensory input, making such an exercise exceedingly difficult.
-Proprioception is clearly associated with the cerebellum and hunger is clearly associated with the hypothalamus.
-Nocioception, however, is a bit more nuanced with regard to the brain's reception of thermal, mechanical, and chemical information — and whether or not the activity is inbound, outbound, conscious, or unconscious.
-However, assuming all nervous system activity does have corresponding brain activity, one could theoretically map out the Matrix Paradox on the brain as Abrash has started.
-It seems likely such a diagram would highlight all parts of the entire brain.
-
-On that note, I feel I should introduce an absolutely fantastic RadioLab episode (\textit{Where Am I?} \cite{whereami}) which addresses this topic of ``other senses'' from the perspectives brought by modern Neuroscience.
-On the topic of lacking proprioception:
-
-\begin{quote}
-  \textbf{We have words like `deaf' and `blind'. We do not have words for being deaf or blind to one's own body.}
-\end{quote}
-
-In the Matrix, perhaps this is either a bug or a feature flag.
-
-Now that we have our diagram for visualizing the use of our senses, we can address them in terms of various altered mental states.
-Before we get there, however, we will try to come up with a meaningful visual definition of ``attention'' or ``awareness'' with respect to our senses.
-
-\pagebreak
-
-\begin{center}
-  \Huge{On Attention}
-\end{center}
-
-The funny thing about Vipassana is that you are doing it already...
-you're just terrible at it.
-Vipassana is the inspection of one's own nervous system with narrowly-focused attention but the precision of this focus is not prescribed by the instruction given for the technique.
-``As narrow as possible'', perhaps, but in day to day life a normal person cannot possibly narrow her attention to the size of a fingertip.
-Your attention is, in part, already on your nervous system, with some degree of focus.
-To the extent that it is, you are doing Vipassana.
-
-If your attention is not directed at your nervous system, where is it? Let's look at three environmental examples which can help us highlight our senses and our awareness within them.
-
-\pagebreak
-
-\begin{center}
-  \LARGE{Example 1:}\\
-  \LARGE{A bustling restaurant}
-\end{center}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/attention1-restaurant.jpg}
-  \caption{A bustling restaurant.}
-  \label{fig:bustling-restaurant}
-\end{figure}
-
-...and a corresponding sense map something like this:
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=10cm]{images/ma-noisy-restaurant.png}
-  \caption{Sense map of a bustling restaurant.}
-  \label{fig:sense-map-bustling-restaurant}
-\end{figure}
-
-Assuming we've just placed food in our mouths, our sense of taste is going off like crazy, we're hearing a lot of scattered sound around us, and we're also sufficiently distracted by most of our other senses.
-Note that we are probably paying very little attention to the internal senses.
-
-\pagebreak
-
-\begin{center}
-  \LARGE{Example 2:}\\
-  \LARGE{A sensory deprivation tank}
-\end{center}
-
-\begin{figure}[h!]
-  \centering
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/attention2-tank.jpg}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/ma-sensory-deprivation-tank.png}
-  \end{subfigure}
-  \caption{A sensory deprivation tank and corresponding sense map.}
-  \label{fig:sensory-deprivation-tank}
-\end{figure}
-
-Here, the classic 5-sense model fails us, since sensory deprivation only deprives us of these 5 senses.
-When Neo is first exposed to the reality of The Matrix, the pods which imprison the humans loosely resemble such sensory deprivation tanks, which is reason enough to believe that we are not in that matrix.
-As anyone who has ever entered a sensory deprivation tank knows, it is not like being in a coma.
-The mind is still very active and it may actually help in causing someone to pay attention to their internal senses.
-
-\pagebreak
-
-\begin{center}
-  \LARGE{Example 3:}\\
-  \LARGE{An anechoic chamber}
-\end{center}
-
-\begin{figure}[h!]
-  \centering
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/attention3-chamber.png}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/ma-anechoic-chamber.png}
-  \end{subfigure}
-  \caption{An anechoic chamber and corresponding sense map.}
-  \label{fig:anechoic-chamber}
-\end{figure}
-
-The map for an anechoic chamber may seem surprising.
-However, just because the room you are in has a negative decibel rating does not mean you cannot hear.
-In fact, in this environment, your sense of hearing is likely to be amplified --- and turned inward.
-You will hear your heart beating, your blood pumping, your digestive system sloshing around.
-The better the anechoic chamber at swallowing sound, the more intense this experience is going to be.
-
-For this reason, the above anechoic chamber at Orfield Laboratories \cite{orfield} has an unofficial record of 45 minutes for the longest period anyone has sat inside it.
-As it turns out, hearing one's own bodily functions at deafening internal volumes is likely to cause someone to go a bit crazy, hence the highlighted bar for thought activity.
-
-The above three examples serve as templates with external input guiding one's understanding of what is happening neurologically.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Anapana Meditation:}\\
-  \Huge{Practicing Focus}
-\end{center}
-
-The first 3.5 days of a 10-day Vipassana course are spend practicing ``Anapana'' (breath) meditation.
-This period is very important as it is time spent practicing acute, one-pointed attention.
-If we visualize our attention as any skin (internal or external; lips, nose, nostrils, etc.) covered by the pink triangle in Figure~\ref{fig:burmese-anapana-triangle}, we will get an idea of exactly where one's attention is directed at the beginning of Anapana practice.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/burmese-anapana-triangle.png}
-  \caption{Large, triangular Anapana meditation target.}
-  \label{fig:burmese-anapana-triangle}
-\end{figure}
-
-If you have ever practiced a single-pointed meditation (particularly breath meditations) before, you know all too well how easily your mind can be distracted.
-Taxes, Mom's birthday, an upcoming meeting, a fun math problem, what's for lunch, the itch on my ear...
-anything can become a distraction because everything which is not ``my breath'' is by definition a distraction.
-And we are not accustomed to spending long stretches of time with unwavering, singular focus.
-We are especially unaccustomed to doing so with all of our attention focused on our breath.
-
-As you practice Anapana for over 30 hours, you will find your attention wandering less and less, while (paradoxically) the distractions become more and more interesting.
-The itch on your ear becomes an itch on your nose which seems to move with the breath.
-Your brain will spend less time worrying about taxes and more time trying to solve interesting math problems (if that's your thing).
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/face-attention.jpg}
-  \caption{Wandering attention.}
-  \label{fig:wandering-attention}
-\end{figure}
-
-On the first day, your attention might look something like Figure~\ref{fig:wandering-attention}, where the wavy pink cloud represents any attention spent on any part of the body other than the nostrils or spent ``outside'' the body: on thoughts and emotions.
-By the fourth day, your attention will often take a shape something like Figure~\ref{fig:narrow-attention}.
-Over the course of the first three days, the instructions narrow down one's attention from the nose-and-lips triangle from Figure~\ref{fig:burmese-anapana-triangle} to a finger-tip-sized area wherever the breath can be felt beneath the nostrils.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{images/face-attention-narrow.jpg}
-  \caption{Narrow attention.}
-  \label{fig:narrow-attention}
-\end{figure}
-
-If we revisit the notion that ``meditation is a thing'', we can see a specific example in Anapana with the help of the sense map in Figure~\ref{fig:anapana-sense-map}.
-Where mindfulness and violin, albeit useful physical and mental practices, are not meditation, we can see from the sense map that Anapana \textit{is} meditation.
-It is the practice of narrowing one's attention, repeatedly, toward a specific sense (in this case, touch) at a specific time (now) in a specific place (just beneath the nostrils).
-As our skill in Anapana improves, we come nearer and nearer to the focus-oriented meditation ideal of an unwavering single point of awareness.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=10cm]{images/ma-anapana.png}
-  \caption{Anapana sense map.}
-  \label{fig:anapana-sense-map}
-\end{figure}
-
-In Figure~\ref{fig:anapana-sense-map}, we can see that in reality our attention is not perfectly tuned and narrowed to a tiny spot under our nose.
-We are still likely to hear sounds around us on the third day of the course, especially if the sound is monkeys clambering over the roof of the meditation hall.
-By the third day of the course, however, one's attention to sound is likely to be greatly reduced.
-
-For the rest of this paper, make note of the fact that \textit{distraction} is at this stage, and will forever remain, the most important aspect of meditation; it is everything meditation is not, and therefore a significant consideration in all the techniques which follow.
-Meditation is essentially comprised of only two components: (a) the meditation object and (b) distractions.
-The activity of meditation is the repeatedly directing of one's attention from the latter to the former.
-
-\pagebreak
-
-\begin{center}
-  \LARGE{Anapana: The Bridge}
-\end{center}
-
-At the end of the first day of a 10-day Vipassana course, the breath is described as a bridge ``from the known to the unknown, because respiration is one function of the body that can be either conscious or unconscious, intentional or automatic.'' During the 3-day Anapana session of a 10-day course, students are instructed to \textit{consciously} direct attention toward their breath -- and even gently control the breath for brief periods to maintain attention when the mind won't stop wandering -- until awareness of breath, without force, lasts for extended periods.
-Given that the breath is the one function of the body which operates both consciously and unconsciously, it makes sense that it acts as the starting point for many meditation practices, including Vipassana.
-The intention of Vipassana is to train the mind out of habits and tendencies which have embedded themselves into the brain and nervous system at an unconscious level.
-To do this, one must first learn to access one's unconscious.
-Through the breath, Anapana meditation fulfills this purpose.
-
-Vipassana requires a 10-day course to learn.
-Anapana does not.
-Before his death, S.N.
-Goenka released both 70-minute audio instructions as MP3s for the Anapana meditation technique.\cite{anapana} The 70-minute instructions can be found at \url{http://www.vipassana.co/What-is-Anapana} under ``Anapana Meditation for All'' and come as four 15-minute sessions plus a 10-minute ``daily practice'' session.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=8cm]{images/bridge.png}
-  \label{fig:bridge}
-\end{figure}
-
-\pagebreak
-
-\begin{center}
-  \Huge{Vipassana Meditation:}\\
-  \Huge{Mutually-Recursive}\\
-  \Huge{Consciousness I/O}
-\end{center}
-
-Oddly enough, Vipassana mostly seems to be about remaining calm.
-If you're thinking ``you're sitting there, silent, for 10 hours a day...
-what is there to get upset about?'' you are in for a surprise.
-But first, a quick overview of the early mechanics of Vipassana.
-(See Figure~\ref{fig:vipassana-mechanics}.)
-Attention is moved from the nostril area to the top of the head, then systematically moved across every body part, down to the toes. This process then repeats.
-
-This visualization and description is to act as an \textit{example only}: this is not the only way to begin Vipassana meditation.
-Though Vipassana instruction is systematic, it is far from mechanical or prescriptive.
-Even in the initial instructions, alternative approaches are offered including alternate orders of observation, alternate areas of observation, and alternate durations of observation.
-The key aspect of any approach to Vipassana is that the \textbf{entire body} is repeatedly observed within each meditation period.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/burmese-vipassana-mechanics.png}
-  \caption{Simplified example of initial Vipassana mechanics.}
-  \label{fig:vipassana-mechanics}
-\end{figure}
-
-I will take this opportunity to echo a caveat from the beginning of this paper: This section is in no way a form of instruction.
-Do not attempt full-body Vipassana meditation outside of a proper 10-day course.
-Practicing Vipassana is likely the most difficult activity you will ever attempt, should you choose to.
-To try so alone, you will have no teacher for guidance and the experience is bound to be frustrating.
-
-Consider this warning to also extend to any literature which \textbf{does} attempt to convey Vipassana meditation instruction.
-Specifically, Glickman's \textit{Beyond the Breath: Extraordinary Mindfulness Through Whole-Body Vipassana Meditation} \cite{glickman} includes a facsimile of instruction provided on a 10-day introductory Vipassana course.
-Instruction on a course includes 11+ hours of video instruction, at least twice as many hours of audio instruction, and multiple interviews per day with the teacher.
-Trying to replace this with a book is both inadequate and dangerous.
-
-More complex variations are added to these mechanics as the bodily sensation experienced through this probing of the nervous system becomes increasingly subtle. In this order:
-
-\begin{enumerate}
-  \item Add toes-to-head to head-to-toes
-  \item Dual/symmetrical observation (both arms, both legs)
-  \item Observation of a ``ring'' around the entire body
-  \item Observation of as much of the body as possible (while moving attention up and down)
-  \item Penetrating attention of muscles, tissues, organs
-  \item Penetrating attention of the spinal cord
-  \item Randomized attention in fingertip-sized dots throughout the body, inside and out
-  \item Sweeping attention through the entire body, up and down, in a single breath
-\end{enumerate}
-
-Through this process, attention will move from the sense of touch to all the internal senses and the sense map will look more like Figure~\ref{fig:vipassana-sense-map-1}.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=\linewidth]{images/ma-vipassana1.png}
-  \caption{Vipassana sense map.}
-  \label{fig:vipassana-sense-map-1}
-\end{figure}
-
-I should note that I have never successfully penetrated my muscles or inner tissues much.
-The average new meditator is likely to feel a great many fascinating sensations on her fingertips, palms, arms, toes, and legs.
-The torso, since it is surrounded in longer nerve fibres, will take longer to observe in the same way.
-\textit{[citation needed]}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/burmese-vipassana-pain.png}
-  \caption{Common pain for Westerners in meditative postures.}
-  \label{fig:burmese-vipassana-pain}
-\end{figure}
-
-As a Canadian who spent little (if any) of my childhood sitting cross-legged on the floor, Figure~\ref{fig:burmese-vipassana-pain} roughly summarizes most of my early meditation experiences, where the purple dots represent intense pain and the pink dot continues to represent my awareness.
-At the peak of my meditation practice toward the end of a course, I would find that if my attention were somewhere away (say, my neck) from the painful area of my body (say, my knees), I may not feel any pain at all.
-As I moved my attention around, the pain in my knees would come back as I approached them, but every time I could sustain focus on another part of my body, the pain would appear to decrease with every return visit to my knees.
-If I can remain calm and focused while away from my knees, my knees' pain seems to take care of itself.
-If I can remain calm and focused while my attention moves across my knees, the situation continues to improve.
-
-If an individual cannot feel pain in one's knees despite the fact that the pain is ``real'', she has effectively eliminated (some) sensory input.
-This is possible for all sensory input.
-The five classic sense-doors will quite clearly shut down in deep meditative states.
-Just because your eyes are closed does not make you blind.
-Light sensitivity remains, and as a side bonus of an optic nerve injury from a few years ago, I can ``see'' something (the damaged spot of the nerve) even when my eyes are shut and I'm in complete darkness.
-This blob which I ``see'' disappears in deep meditative states.
-Similarly, these deep states of meditation will eliminate audio input as well, which is easier for anyone to detect.
-
-\begin{center}
-  \LARGE{Vipassana: The Sledgehammer}
-\end{center}
-
-During the Day 4 discourse of a 10-day course, the imagery of the bridge is extended by what is on the far side of that bridge: a giant wall.
-The wall represents the remaining barrier between the conscious and unconscious mind -- and Vipassana is the tool used to break this barrier.
-By systematically examining the nervous system, outside-in, one digs through all the layers of unconscious habit as it manifests in bodily sensation.
-
-\begin{figure}[h!]
-  \centering
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/wall.jpg}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/bridge.png}
-  \end{subfigure}
-  \label{fig:wall}
-\end{figure}
-
-And what comes of this examination?
-Sensations in the body are exploded into view and what we normally experience as a narrow band of sensation from the roughly unpleasant to the roughly pleasant: intense sensations like abdominal pain and annoying sensations like uncontrollable itching are gross sensations we normally consider unpleasant; intense sensations like sexual contact and light sensations like touching fur are gross sensations we normally consider pleasant.
-
-This explosion of sensation occurs on an entirely new axis.
-The range can still be considered pleasant to unpleasant but ``pleasant'' begins to cover new, previously unexperienced territory.
-These experiences have little or no parallel in mundane life and are best described as ``subtle'': fine oscillations and vibration.
-This new axis of experience is the sledgehammer used to break through the wall separating conscious from unconscious.
-Unpleasant sensations and pleasant sensations, both, undergo change.
-As one observes this change, the change itself comes faster and faster and becomes easier and easier to observe.
-The goal is to avoid reacting to unpleasant sensations with aversion and to avoid reacting to these newfound pleasant sensations with craving.
-The task is harder than it sounds.
-However, the calmer a meditator remains while exploring these new depths, the further she explores.
-
-Unfortunately, unlike Anapana meditation, there is no way to learn Vipassana from an MP3.
-This practice of observing subtler and subtler sensation requires quite a bit of time, assistance from a live teacher, and repeated instruction.
-The only way one can learn the technique of Vipassana is to take a 10-day course.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Blue Bars:}\\
-  \Huge{The New Distractions}
-\end{center}
-
-At this point, we can revisit distraction.
-Distraction is really the essence of meditation.
-Whether it's taxes, math problems, hunger, itching, or pain...
-our job as meditators is to accept the presence of a distraction and return our attention to our meditation object.
-
-However, once the pink bars have reduced far enough, once the mind is quiet enough, an entirely new form of distraction emerges.
-This diagram highlights the first likely candidate for the New Distraction: thought.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/ma-vipassana3.png}
-  \caption{Attention on a new distraction.}
-  \label{fig:vipassana-sense-map-2}
-\end{figure}
-
-These distracting new ``thoughts'' are not normal thoughts in any sense of the word you are currently familiar.
-These ``thoughts'' are an entirely new form of brain activity caused by your constant prodding of the nervous system through Vipassana.
-Everything in blue, unfortunately, falls into the ``explaining the concept of `white' to a birth-blind man'' category of unexplainable sensory input.
-Though this brain activity cannot be explained it \textbf{can} be experienced directly and even a first-time meditator is likely to experience such brain activity on her first 10-day course.
-The blue bar in Figure~\ref{fig:vipassana-sense-map-2} does not represent a goal or form of meditation, nor even a binary state.
-Rather, if we imagine brain activity on a continuum, worries about taxes might appear on the far left, interesting math and logic problems in the centre, and the ``blue bars'' of unexplainable thought on the right.
-However fascinating a journey toward this brain activity may be, the thought processes themselves remain nothing more but a distraction from our goal.
-
-What is meant by ``prodding our nervous system'' and what does this have to do with brain activity?
-Our second suspension of disbelief was of the deep and intense connection between the brain the nervous system.
-We cannot feel the brain directly --- or at all.
-Thanks to the deep interconnection between the brain and the nervous system, however, the body acts as a surface to which we have ready access --- and which in turn provides us access to the brain.
-This brain/body connection can be thought of as \textit{mutual recursion}, as the term is used in the previous section.
-
-That the brain/body connection is \textit{mutually recursive} is meant operationally, not existentially.
-We do not mean to say that the brain could not live without the nervous system (or vice-versa) but that every event --- every operation --- occurring in the nervous system has a corresponding event in the brain, and so forth.
-This mutually recursive process will eventually wear itself out.
-If it did not, the advent of slight anger would mean we would, from that moment onward, just get angrier and angrier for the rest of our lives.
-However, during the process of mutual recursion we can be feeling pain in a knee, to which our emotional response is ``argh, I hate pain'', which makes us angry, which clenches up our physical systems, which makes the pain worse, and so on.
-
-This stack of mutually-recursive operations is what produces all thought, emotion, and sensation.
-Vipassana is the process of unwinding the stack: If I can pay attention to my aching knee, remain calm and continue to observe it, slowly my body unclenches, the pain subsides a little, my emotional response is calmer, I am able to remain calm longer, and so on.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Blue Bars 2:}\\
-  \Huge{The New-New Distractions}
-\end{center}
-
-Let us quickly revisit our third suspension of disbelief: ``dissolution of the body is a thing.'' Neuroscience tells us some interesting things about the nature of our sense of self and how it relates to our body.
-From \textit{Where Am I?} \cite{whereami}:
-
-\begin{quote}
-  \textbf{The studies reported that people who had once been able-bodied and then became paralyzed felt… less. Less happy than able-bodied people, less sad than able-bodied people. Just less.}
-\end{quote}
-
-\begin{quote}
-  \textbf{Our \textit{being} is rooted in a body-state. If I were able to remove from your brain, the representation of your body, you would not know that you were you.}
-\end{quote}
-
-...which would seem in line with experiences described by mastered meditators of having ``no I'' or ``no self'' as their fundamental insightful discovery.
-It also seems necessary to project our experiences from one 10-day Vipassana course into the future, with respect to the extent to which this practice can take us.
-If it were stated that we could explore the entirety of the nervous system and its relationship to the brain --- but not the gallbladder or spinal cord because those are off-limits --- we would have reason to believe that there is a portion of the body and/or nervous system which is exempt from the rules governing these interactions.
-
-Assuming we can explore to the very depths of our nervous system by repeatedly and systematically penetrating it with our attention, given the extent to which this observably occurs on a 10-day course, we can then examine ``mutually-recursive consciousness I/O'' in its totality.
-What occurs?
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/ma-vipassana4.png}
-  \caption{Attention distracted by all-new senses.}
-  \label{fig:vipassana-sense-map-4}
-\end{figure}
-
-This I/O feedback between the brain and body appears to create \textit{a completely new kind of distraction, which goes well beyond new thought patterns}.
-An attempt to highlight this form of distraction is given in Figure~\ref{fig:vipassana-sense-map-4}.
-
-Here, all of the sense doors which we have quiesced return to life, as did ``thought'' before them.
-However, this new sensory input (which remains a distraction...
-just a very sophisticated distraction) is unlike most of regular experience.
-The nearest approximation I can come up with would be the experience of taking salvia, an awful drug which is legally permitted in most countries because it is simply not that enjoyable.
-While I do not at all recommend smoking salvia, the Wikipedia article \cite{salvia} is worth a glance, since the description of hallucinogens is approachable for anyone.
-The two relevant items from the list of ``immediate effects'' are:
-
-\begin{itemize}
-  \item Past memories, such as revisiting places from childhood memory
-  \item Overlapping realities, such as the perception of being in several locations at once
-\end{itemize}
-
-The effects of deep Vipassana meditation are akin to this, in the sense that one may experience, in its entirety, a completely different time or reality.
-A comparable, sober experience might be that of a car crash: participants often feel time slow down or stop, which is sensory input completely detached from the normal flow of consciousness.
-That is where similarities end but the comparisons may help provide us with tangible, and therefore approachable, examples.
-
-It is worth reiterating that the purpose of Vipassana is not to experience altered mental states.
-New thought patterns and new realities are, again, simply distractions from the meditation itself.
-The meditation remains the systematic observation of the nervous system.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Examples}
-\end{center}
-
-While I would like to avoid descriptions of personal experiences, a few may help highlight the above description of new and unfamiliar distracted states of mind.
-
-These examples also serve a secondary purpose.
-If one searches for ``vipassana'' on YouTube, one will find Goenka's instructional videos, critiques from individuals who could not complete their course, TED talks, and experience reports from individuals who did complete their first course.
-These experience reports (often titled ``My experience...'') tend to confuse these distracting experiences with the meditation itself.
-Framing a few examples within this paper should help the reader with this differentiation.
-
-\vspace{1cm}
-\begin{center}
-  \LARGE{The non-example:}\\
-  \LARGE{The Resolution to Paradox}
-\end{center}
-
-It would seem that one very common effect of Vipassana during deep meditation is to witness the resolution of paradox.
-A logical, mathematical, or linguistic paradox might emerge in the meditator's mind and just as quickly, it is resolved.
-
-These would appear to mirror Zen koan practice, where the paradox is presented to the meditator verbally, in advance.
-No such practice occurs in Vipassana.
-Paradoxes and their solutions appear to be intrinsic to the human makeup, somehow.
-While the resolution of paradox may be of acute interest to many hackers, indulging in the consideration of a manifest paradox is as much a tangent and distraction to Vipassana as it is to Zazen.
-
-\vspace{1cm}
-\begin{center}
-  \LARGE{Holding a Baby}
-\end{center}
-
-A friend once had a hallucination-like experience, feeling as though she was teleported to her friend's kitchen, holding her friend's 2-year-old while the child cried.
-The experience was so raw, realistic, and unexpected that she actually opened her eyes and stood up in her meditation cell.
-
-Such distractions are difficult to overcome, since they appear to replace consciousness and reality itself.
-
-\vspace{1cm}
-\begin{center}
-  \LARGE{Time Slowing Down}
-\end{center}
-
-It will become quite clear halfway through a 10-day course that this effect of Vipassana is also quite universal, like paradox and resolution to paradox.
-The moment-to-moment awareness seems to take longer and longer as the course goes on and each one-hour meditation period can also seem to extend beyond reasonable experiential limit.
-
-However, this is not what I am referring to.
-The experience of time slowing down in very deep meditative states is that of time slowing down a \textbf{great deal} ...tens or hundreds of thousands of years pass before your awareness in a single meditation period.
-
-It is difficult to describe how this effect remains nothing but a distraction.
-However, if meditation remains subdivided into two categories, the meditation object and distractions therefrom, we must accept that even experiences which alter our fundamental perceptions are little more than our brain's creative attempts to distract us from the meditation itself.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Outcomes}
-\end{center}
-
-So, great. Vipassana can trick our bodies and brains to emerge an entirely new set of distractions from the practice of... Vipassana. Why, exactly, would anyone do this?
-
-There is a sketch in \textit{Where Am I?} \cite{whereami} about an arguing couple who must overcome their emotional temptation to get wound up by anger, ending in this wonderful quote:
-
-\begin{quote}
-  \textbf{Honey, don’t forget what the half-life is on the autonomic nervous system.}
-\end{quote}
-
-The autonomic nervous system is what causes the ``winding stack'' of thought and emotion described earlier.
-The autonomic nervous system is what makes (and keeps) you angry.
-And afraid.
-And sad.
-And anxious.
-And everything else you feel.
-
-Vipassana is a systematic practice to calm the nervous system (and brain?) so that every sensory input can be analyzed rationally.
-You won't become a perfectly rational machine after one 10-day Vipassana course, but you are likely to come out with a much clearer head and a much better capacity for decision-making in general.
-
-I would describe the positive effects of nervous system meditations to live on a spectrum.
-On one end of the spectrum are the positive, rationality-enhancing qualities of calming one's nervous system with quiet attention.
-On the other end of the spectrum are the subtler qualities: Meditation seems to make a person more generous, more compassionate, more empathetic.
-Why would systematic analysis of the autonomic nervous system have these effects?
-A wide body of research suggests that many forms of meditation, including Vipassana, exhibit these positive effects.
-Many often identify specifically enhanced activity in the prefrontal cortex \cite{differentialengagement, meditationbrainstructure}, which is identified as the centre of the will to live and altruistic tendencies in the human brain.\cite{personalityneuroscience} However, literature explaining why or how such effects may occur is limited.
-
-Regardless of the mechanism, I do find the entire endeavour not just fascinating, but endlessly worthwhile. Vipassana is simultaneously the hardest thing I have ever done --- and the most valuable.
-
-\pagebreak
-
-\begin{center}
-  \Huge{Further Reading}
-\end{center}
-
-\vspace{1cm}
-
-\begin{itemize}
-\item \textit{The Art of Living: Vipassana Meditation As Taught by S. N. Goenka} (Hart, 1987)
-\item \textit{An Atheist's Glimpse of God}, my write-up of my first 10-Day Vipassana course:
-  \begin{itemize}
-  \item \href{http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html}
-             {\nolinkurl{http://blog.deobald.ca/2013/02/}}\\
-        \href{http://blog.deobald.ca/2013/02/an-atheists-glimpse-of-god.html}
-             {\nolinkurl{an-atheists-glimpse-of-god.html}}
-  \end{itemize}
-\item \textit{Drugs, meditation, warnings}, my comparison of meditation techniques and various drugs:
-  \begin{itemize}
-  \item \href{http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html}
-             {\nolinkurl{http://blog.deobald.ca/2015/01/}}\\
-        \href{http://blog.deobald.ca/2015/01/drugs-meditation-warnings.html}
-             {\nolinkurl{drugs-meditation-warnings.html}}
-  \end{itemize}
-\end{itemize}
-
-\pagebreak
-
-\begin{center}
-  \Huge{Appendix A:}\\
-  \Huge{Can we ``hack'' Vipassana?}
-\end{center}
-
-If Vipassana is hacking the mind, is there an opportunity for hacking Vipassana?
-Can we hack the hacking process?
-Let's ask Martin Thompson.
-The following slide in Figure~\ref{fig:load-test} is from his 2013 talk \textit{Performance Testing Java Applications} \cite{testingjava} and while I have no intention of ostracizing non-computer hackers, software was my career once upon a time and the example is not only appropriate but reasonably accessible.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/deathstar1.png}
-  \caption{A computer system under load test.}
-  \label{fig:load-test}
-\end{figure}
-
-He suggests that any system we intend to observe should be observed through a 3rd party observer by dumping network traffic (see Figure~\ref{fig:objective-load-test}).
-For computer systems, this makes a great deal of sense.
-Were the system under test or the load agents (the Death Star and X-Wings, respectively) to perform the measurements, the measurements and the systems would likely interfere with each other.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/deathstar2.png}
-  \caption{A load test with an objective, 3rd-party observer.}
-  \label{fig:objective-load-test}
-\end{figure}
-
-Unfortunately, human biology is not so keen on this setup.
-Even if we had the technology to tap into the interactions of the brain and nervous system (we don't), there is no way to observe the interactions of the brain and nervous system without immediately falling into the Matrix Paradox (see Figure~\ref{fig:nervous-system-observer}).
-How do I know the tool I used is trustworthy?
-How do I know the observations I made with my eyes are trustworthy?
-This is all once again external.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{images/deathstar3.png}
-  \caption{The nervous system cannot have an objective observer.}
-  \label{fig:nervous-system-observer}
-\end{figure}
-
-It is possible to ``jump out'' of Vipassana momentarily to take stock of the effects of the meditation.
-Did my hearing just cut out?
-Did I just force my brain to become consciously aware of my normally-unconscious proprioception pertaining to my right arm?
-Did I just feel my cardiovascular system in exquisite detail or did I imagine it?
-...but it is not possible to do so without breaking the meditation for the period of this observation.
-
-I performed this exercise a great deal toward the end of my second 10-day course because I was quite confused about everything that I was experiencing.
-It framed my understanding but it did little for the practice itself.
-I don't recommend doing this.
-Particularly for your first course, try your hardest to follow the instructions precisely and give all of your energy to the moment-to-moment attention demanded of you.
-Otherwise you will find yourself spending 10 days thinking about meditation, rather than meditating.
-Observation of the process itself, rather than observation of your mind via your nervous system, is actually just another clever form of distraction.
-Every distraction you indulge will limit the positive effects of the meditation.
-
-This particular distraction may feel justified as the reification of your skepticism.
-While it pays to be skeptical, try your hardest to avoid skepticism during the course.
-Approach the practice skeptically until you decide to take a course and then save your analysis of whatever newfound skepticism the course generates for you until after the course is completed.
-There will be plenty of time to discuss the meditation and the outcomes of the meditation on the tenth day when talking is permitted; explore your skepticism here.
-Other meditators are likely to be the only ones who will quite comprehend the concepts you are trying to discuss.
-
-\pagebreak
-
-\begin{thebibliography}{9}
-\raggedright
-
-\bibitem{thenewsroom}
-  Aaron Sorkin.
-  \url{https://www.youtube.com/watch?v=2C6h-Yyx9Yk}
-  \textit{The Newsroom}.
-  HBO, 2012.
-
-\bibitem{fiveprecepts}
-  Edited by Access to Insight.
-  \url{https://www.accesstoinsight.org/ptf/dhamma/sila/pancasila.html}
-  \textit{The Five Precepts: pañca-sila}
-  Access to Insight (BCBS Edition), 30 November 2013.
-
-\bibitem{experiencethetruth}
-  S.N. Goenka.
-  \url{https://www.youtube.com/watch?v=mcbGbN8pFyk&list=PLlb2aTuor6J9j8hACeX6GUmd4WF7QJsbu&index=2}
-  \textit{Experience the Truth}.
-  Vipassana Research Institute, 1992.
-
-\bibitem{taocp}
-  Donald Knuth.
-  \textit{The Art of Computer Programming}.
-  Addison-Wesley, 1962, 1968, 1969, 1973, 2005, 2011, 2015, 2017.
-  ISBN: 0-201-03801-3
-
-\bibitem{joyofcooking}
-  Irma S. Rombauer, Marion Rombauer Becker.
-  \textit{The Joy of Cooking, Sixth Edition}.
-  Bobbs-Merrill, 1975.
-  ISBN: 0-02-604570-2
-
-\bibitem{hackerhowto}
-  Eric S. Raymond: How To Become A Hacker,
-  \url{http://www.catb.org/esr/faqs/hacker-howto.html}
-
-\bibitem{abrashvr}
-  Michael Abrash.
-  \url{https://www.youtube.com/watch?v=UDu-cnXI8E8}
-  \textit{Why Virtual Reality Will Matter to You}.
-  Facebook F8 Summit, Fort Mason, 2015.
-
-\bibitem{simulationhypothesis}
-  \url{https://en.wikipedia.org/wiki/Simulation_hypothesis}
-  \textit{Simulation Hypothesis}.
-  Wikipedia.
-
-\bibitem{ramachandranbrainfunction}
-  V. S. Ramachandran, Eric L. Altschuler.
-  \url{https://academic.oup.com/brain/article/132/7/1693/328686}
-  \textit{The use of visual feedback in restoring brain function}.
-  Brain, 132(7):1693–1710, July 2009.
-
-\bibitem{proprioception}
-  \url{https://en.wikipedia.org/wiki/Proprioception}
-  \textit{Proprioception}.
-  Wikipedia.
-
-\bibitem{nocioception}
-  \url{https://en.wikipedia.org/wiki/Nociception}
-  \textit{Nocioception}.
-  Wikipedia.
-
-\bibitem{fmrihunger}
-  Dagmar Führer, Stefan Zysset, Michael Stumvoll.
-  \url{https://www.ncbi.nlm.nih.gov/pubmed/18292747}
-  \textit{Brain activity in hunger and satiety: an exploratory visually stimulated fMRI study}
-  Obesity, 16(5):945-50, May 2008.
-
-\bibitem{whereami}
-  Jad Abumrad, Robert Krulwich.
-  \url{http://www.radiolab.org/story/91524-where-am-i/}
-  \textit{Where Am I?}
-  RadioLab, May 2006.
-
-\bibitem{orfield}
-  Lindsey Galloway.
-  \url{http://www.bbc.com/travel/story/20121022-the-quietest-place-on-earth}
-  \textit{The quietest place on Earth}
-  BBC, 23 October 2012.
-
-\bibitem{anapana}
-  S.N. Goenka
-  \url{http://www.vipassana.co/What-is-Anapana}
-  \textit{What is Anapana?}
-  Vipassana Research Institute, 21 October 2016.
-
-\bibitem{glickman}
-  Marshall Glickman.
-  \textit{Beyond the Breath: Extraordinary Mindfulness Through Whole-Body Vipassana Meditation}.
-  Tuttle Publishing, 2002.
-  ISBN: 1582900434
-
-\bibitem{salvia}
-  \url{https://en.wikipedia.org/wiki/Salvia_divinorum}
-  \textit{Salvia divinorum}.
-  Wikipedia.
-
-\bibitem{differentialengagement}
-  Holzel B.K., Ott U., Hempel H., Hackl A., Wolf K., Stark R., Vaitl D.
-  \textit{Differential engagement of anterior cingulate and adjacent medial frontal cortex in adept meditators and non-meditators}
-  Neuroscience Letters, 421:16–21, 2007.
-
-\bibitem{meditationbrainstructure}
-  Fox, Kieran C.R.; Nijeboer, Savannah; Dixon, Matthew L.; Floman, James L.; Ellamil, Melissa; Rumak, Samuel P.; Sedlmeier, Peter; Christoff, Kalina
-  \textit{Is meditation associated with altered brain structure? A systematic review and meta-analysis of morphometric neuroimaging in meditation practitioners}
-  Neuroscience \& Biobehavioral Reviews, 43:48–73, June 2014.
-
-\bibitem{personalityneuroscience}
-  Colin G. DeYoung, Jacob B. Hirsh, Matthew S. Shane, Xenophon Papademetris, Nallakkandi Rajeevan, Jeremy R. Gray.
-  \textit{Testing Predictions From Personality Neuroscience: Brain Structure and the Big Five}
-  Psychological Science, 21(6):820–828, June 2010.
-
-\bibitem{testingjava}
-  Martin Thompson.
-  \url{https://www.youtube.com/watch?v=1DuMvpwWHH4}
-  \textit{Performance Testing Java Applications}.
-  Devoxx UK, 2013.
-
-\end{thebibliography}
-
 \end{document}

### 2019-05-06 paper-four/build-pdf.sh

@@ -1,0 +1,14 @@
+#!/bin/sh
+
+SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
+LIBPATH="$SCRIPTPATH/../lib"
+PATH=$PATH:$LIBPATH
+
+OUT=../output
+
+echo "Running pre-build. This takes a while..."
+xelatex -output-directory=$OUT paper-four.tex # > /dev/null 2>&1
+echo "...done. Now running the build."
+xelatex -output-directory=$OUT paper-four.tex
+
+rm texput.log $OUT/*.aux $OUT/*.log $OUT/*.out

### 2019-05-06 paper-four/paper-four.tex
notes recorded for paper four

@@
 
+\begin{itemize}
+  \item Old students (Goenka; forward ref other methods of learning)
+  \item People looking to deepen practice
+  \item Meditators who haven't chosen a technique yet
+\end{itemize}
+
+\pagebreak
+
+\begin{center}
+  \Huge{Eka Maggo? Learning Vipassana Elsewhere}
+\end{center}
+
+\begin{itemize}
+  \item Ledi Sayadaw: 5 hours a day anapana, ``hairs of the head and the body''
+  \item Other students of U Ba Khin: ``The Quiet Mind'', ``Beyond the Breath''
+  \item ``Vipassana'': sensations matter more than terminology (3-day course, 10-day course), other forms of ``Vipassana'' not of interest, Zazen
+\end{itemize}
+
+\pagebreak
+
+\begin{center}
+  \Huge{Concepts, Ideas, Help}
+\end{center}
+
+\begin{itemize}
+  \item The Cheat Code: moving attention => sati vs. anicca
+  \item Circle Awareness Drawing: how the 5 sense doors shut down
+  \item Learn to sit cross-legged
+\end{itemize}
+
+\pagebreak
+
+\begin{center}
+  \Huge{Warnings}
+\end{center}
+
+\begin{itemize}
+  \item This document itself is a hindrance
+  \item Don't get bogged down by ``perfect objectivity'' or ``perfect equanimity'' - yes muscles and blood flow follow your attention
+  \item Do not use meditation to get high
+  \item Do not mix meditations (1. dangers, 2. meditation ``doing itself'' - anatta)
+\end{itemize}
+
 \end{document}

### 2019-05-06 paper-one/paper-one.tex
paper four first post
notes recorded for paper four
registration info for paper one, makefile build

--- 2018-05-10 vipassana-for-hackers.tex
@@
      \Huge\textbf{Vipassana for Hackers}\\
+     \Huge{Paper One: Curious Mechanics}\\
      \vspace{5cm}
      \large\textit{Steven Deobald}\\
-     \large\textit{Version 1.5}\\
-     \large\textit\today
+     \large\textit{Version 1.6}\\
+     \large\textit\today\\
+     \vspace{5cm}
+     \large\textit{Contributions by: Preethi Govindarajan, Mat Kelly, Liam Monahan}\\
    \end{center}
@@
 You might be familiar with phrases commonly attributed to this archetype such as ``When the student is ready, the teacher will appear.'' Mysticism, mystery, and mystique are to be avoided.
-If I fall into this trap, email me: \href{mailto:steven@nilenso.com}{steven@nilenso.com}
+If I fall into this trap, email me: \href{mailto:steven@deobald.ca}{steven@deobald.ca}
 
@@
 
-Vipassana requires a 10-day course to learn.
-Anapana does not.
-Before his death, S.N.
-Goenka released both 70-minute audio instructions as MP3s for the Anapana meditation technique.\cite{anapana} The 70-minute instructions can be found at \url{http://www.vipassana.co/What-is-Anapana} under ``Anapana Meditation for All'' and come as four 15-minute sessions plus a 10-minute ``daily practice'' session.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=8cm]{images/bridge.png}
-  \label{fig:bridge}
-\end{figure}
+Vipassana requires a 10-day course to learn. Anapana does not. Before his death,
+S.N. Goenka released 70-minute audio instructions (for children) and 20-minute audio
+instructions (for adults) as MP3s for the Anapana meditation technique.\cite{anapana}
+Both 70-minute and 20-minute instructions can be found at
+\url{https://www.vridhamma.org/what-is-anapana/} under ``Anapana Meditation for
+All''. Under the videos are links to a 10-minute daily practice session, labelled
+``Group Sitting'' under each respective language.
 
@@
 
-\begin{figure}[h!]
-  \centering
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/wall.jpg}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/bridge.png}
-  \end{subfigure}
-  \label{fig:wall}
-\end{figure}
-
 And what comes of this examination?
@@
 
-Regardless of the mechanism, I do find the entire endeavour not just fascinating, but endlessly worthwhile. Vipassana is simultaneously the hardest thing I have ever done --- and the most valuable.
+Regardless of the mechanism, I do find the entire endeavour not just fascinating, but
+endlessly worthwhile. Vipassana is simultaneously the hardest thing I have ever done
+--- and the most valuable.
+
+\pagebreak
+
+\begin{center}
+  \Huge{Try it Yourself}
+\end{center}
+
+If you have not tried a 10-day Vipassana course, I have tried to repeatedly warn you
+within the bounds of this paper that reading this paper doesn't really teach you
+anything about Vipassana. The 10-day course serves two purposes: First, to try the
+technique yourself to see if it suits you. Second, to learn how to meditate so you
+can practice on your own. Both of these are possible only with direct experience. You
+can't learn how to meditate by reading a paper like this. You have to try it.
+
+If you would like to try it, Vipassana centres exist in many countries and the
+instructions have been translated into most languages. You can apply for a course
+near you online at \url{https://www.dhamma.org/en-US/courses/search} \cite{dhamma}.
+
+\vspace{1cm}
+
+\begin{center}
+  \LARGE{A Word of Warning}
+\end{center}
+
+This paper exists solely to pique your interest in trying a serious meditation
+practice yourself. Unfortunately, in detailing the mechanics of the technique, it is
+likely to become a hindrance for some people. A reader of this paper may be tempted
+to return to the contents of the paper for assistance during meditation. Please do
+not do that. If you choose to take a 10-day Vipassana course, mentally set this paper
+aside and follow the instructions given very carefully and attentively. The 10-day
+course is very detailed and thoughtfully designed --- a very systematic way of
+learning meditation in a relatively short period of time. This paper is only intended
+to inspire you to try it.
 
@@
   S.N. Goenka
-  \url{http://www.vipassana.co/What-is-Anapana}
+  \url{https://www.vridhamma.org/what-is-anapana/}
+  \url{https://www.vridhamma.org/Mini-Anapana-English}
   \textit{What is Anapana?}
@@
 
+\bibitem{dhamma}
+  Vipassana International Academy
+  \url{https://www.dhamma.org}
+  \textit{Vipassana Meditation Website}
+
 \end{thebibliography}

### 2019-05-06 paper-four/build-pdf.sh

@@ -11,4 +11,4 @@ xelatex -output-directory=$OUT paper-four.tex # > /dev/null 2>&1
 echo "...done. Now running the build."
 xelatex -output-directory=$OUT paper-four.tex
 
-rm texput.log $OUT/*.aux $OUT/*.log $OUT/*.out
+rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out

### 2019-05-06 paper-one/build-epub.sh

@@ -14,4 +14,4 @@ latexmlpost -dest=paper-one.html paper-one.xml
 
 ebook-convert paper-one.html ../output/paper-one.epub --language en --output-profile=kindle --authors="Steven Deobald" --title="Vipassana for Hackers"
 
-rm paper-one.xml paper-one.html LaTeXML.cache LaTeXML.css ltx-article.css
+rm -f paper-one.xml paper-one.html LaTeXML.cache LaTeXML.css ltx-article.css

### 2019-05-06 paper-one/build-pdf.sh

@@ -19,4 +19,4 @@ xelatex -output-directory=$OUT paper-one.tex > /dev/null 2>&1
 echo "...done. Now running the build."
 xelatex -output-directory=$OUT paper-one.tex
 
-rm texput.log $OUT/*.aux $OUT/*.log $OUT/*.out
+rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out

### 2019-05-06 paper-three/paper-three.tex
adding paper 3

\documentclass{article}

\usepackage{enumitem}

\setlength{\parindent}{1.3em}
\setlength{\parskip}{0.7em}
\setlist{noitemsep}

\begin{document}

\begin{titlepage}
   \vspace*{\stretch{1.0}}
   \begin{center}
     \Huge\textbf{Vipassana for Hackers}\\
     \Huge{Paper Three: Why Bother?}\\
     \vspace{5cm}
     \large\textit{Steven Deobald}\\
     \large\textit{Version 0.1 DRAFT}\\
     \large\textit\today\\
     \vspace{5cm}
     \large\textit{}\\
   \end{center}
   \vspace*{\stretch{2.0}}
\end{titlepage}

\begin{center}
  \Huge{Target Audience}
\end{center}

Anyone curious about Vipassana

Anyone who has never meditated before but would like to learn

Anyone who is wondering why people keep talking about Vipassana

\begin{itemize}
  \item posture
  \item sleep
  \item health (activation / motivation)
  \item ethics (activation / motivation)
  \item your children: a. knowing how to meditate, b. cross-legged posture
  \item emotion (i.e. anger)
  \item mundane sphere / productivity (21 lessons, seinfeld)
  \item reset frame of reference outside oneself, outside one's own lifetime: ``trees for god'' and obvious karma (sidu/booga smoking)
  \item unlearning obsessive / repetitive thought, enhancing creativity
  \item controlling unbounded sexuality without repression
  \item clarity: in thought, work, planning
  \item clarification: ``isn't that what makes us human?'' (emotions) --- rather, what makes us animal
  \item Die Standing Up

\end{itemize}

\end{document}

### 2019-05-06 paper-three/build-pdf.sh

@@ -1,0 +1,22 @@
+#!/bin/sh
+
+# pdflatex was used originally but cannot render Sanskrit without a lot of hoop-jumping
+# pdflatex vipassana-for-hackers.tex
+
+# xelatex is apparently less capable than pdflatex at font-rendering in general, but
+# is capable of rendering non-Latin fonts (like Hindi/Sanskrit) by default
+
+# Compile twice to resolve references.
+
+SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
+LIBPATH="$SCRIPTPATH/../lib"
+PATH=$PATH:$LIBPATH
+
+OUT=../output
+
+echo "Running pre-build. This takes a while..."
+xelatex -output-directory=$OUT paper-three.tex # > /dev/null 2>&1
+echo "...done. Now running the build."
+xelatex -output-directory=$OUT paper-three.tex
+
+rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out

### 2019-05-08 paper-three/paper-three.tex
beginning paper 3

@@
      \Huge\textbf{Vipassana for Hackers}\\
-     \Huge{Paper Three: Why Bother?}\\
+     \Huge{Paper Three: Why Meditate?}\\
      \vspace{5cm}
@@
 
-Anyone curious about Vipassana
+\textit{Vipassana for Hackers, Paper One: Curious Mechanics} was written to avoid discussing the
+outcomes or consequences of meditation in detail. The focus of that paper was only
+the internal mechanics of Vipassana meditation, to pique the interest of potential
+meditators who had heard of Vipassana elsewhere. Outcomes are discussed only so far
+as they assist the reader in understanding what is written earlier in the paper
+regarding the senses. \textit{Paper Two: The Brain} goes further into the mechanics as they pertain to the
+nervous system. Here, outcomes are discussed as they pertain to
+neuroplasticity. Neither paper directly discusses why an individual might choose to
+try this particular technique of meditation.
 
-Anyone who has never meditated before but would like to learn
+As before, the ``Hacker'' of \textit{Vipassana for Hackers} is not meant to identify
+computer programmers. Instead, it is meant as a label for a culture of curious and
+creative people who enjoy exploring, learning, and creating.
 
-Anyone who is wondering why people keep talking about Vipassana
+\textit{Paper Three: Why Meditate?} is written for anyone who has ever asked
+themselves that very question or asked that question of their friends who
+meditate. It is for both those who are curious about the practice of Vipassana
+specifically and those who are curious about meditation in general. It is for people
+who have meditated in other traditions and are curious about the benefits of
+Vipassana. It is also for people who have never meditated in their entire lives. It
+is intended for anyone who keeps hearing about Vipassana meditation --- in the media,
+in books, and from friends --- and wants to learn what all the fuss is about.
+
+The reader need not have read \textit{Paper One} or \textit{Paper Two}. In fact, it
+is the intention of this paper to be the most accessible of the series and readers
+with only a faint interest in the topic of meditation should start here.
+
+\pagebreak
+
+\begin{center}
+  \Huge{Vipassana Basics}
+\end{center}
+
+Before we get to a discussion about why meditation is valuable, some basic
+understanding of what meditation is (and isn't) is required.
+
+The technique of Vipassana is based on a single underlying principle:
+
+\Huge\textbf{Every experience which emerges in the mind, whether a thought or emotion, always surfaces
+with a corresponding sensation on the body.}
+
+It is important to understand this point as it underpins all other aspects of the
+technique of Vipassana. Someone who is learning Vipassana need not accept this
+principle as fact. Rather, a 10-day Vipassana course is a sort of laboratory where the
+principle can be tested and experienced for oneself.
+
+TODO: sense circle diagram
+
+5 Senses by Daniel Falk from the Noun Project: https://thenounproject.com/daniel2021/collection/human-body-senses/
+Thought by Nociconist from the Noun Project: https://thenounproject.com/search/?q=thought&i=2025873
+Heart by Rafael Garcia Motta from the Noun Project: https://thenounproject.com/search/?q=heart&i=807960
+Body by Makarenko Andrey from the Noun Project: https://thenounproject.com/search/?q=body&i=789989
 

### 2019-05-09 paper-three/paper-three.tex
added sense door diagram

@@
 
+\usepackage{graphicx}
+\usepackage{subcaption}
+\usepackage{float}
 \usepackage{enumitem}
+\usepackage{hyperref}
 
@@
 
-\Huge\textbf{Every experience which emerges in the mind, whether a thought or emotion, always surfaces
-with a corresponding sensation on the body.}
+\vspace{1cm}
+\textbf{Every experience which emerges in the mind, whether a thought, emotion, or contact of the five senses, always surfaces with a corresponding sensation on the body.}
+\vspace{1cm}
 
@@
 
-TODO: sense circle diagram
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=11cm]{images/sense-doors.png}
+  \caption{The sense doors and bodily sensation.}
+  \label{fig:sense-doors}
+\end{figure}
 
-5 Senses by Daniel Falk from the Noun Project: https://thenounproject.com/daniel2021/collection/human-body-senses/
-Thought by Nociconist from the Noun Project: https://thenounproject.com/search/?q=thought&i=2025873
-Heart by Rafael Garcia Motta from the Noun Project: https://thenounproject.com/search/?q=heart&i=807960
-Body by Makarenko Andrey from the Noun Project: https://thenounproject.com/search/?q=body&i=789989
+The totality of human experience can be categorized according to the ``sense doors''
+listed in Figure~\ref{fig:sense-doors} \cite{sense-icons}: The five external sense doors of sight, sound,
+taste, smell, and touch are listed at the top. The internal sense door of ``mind'' is
+broken down into thought and emotion, second to the bottom. At the very bottom of the
+diagram is bodily sensation, the object of meditation in Vipassana.
+
+Once these eight experiences are listed, there is no experience left undescribed. All
+human experience from the mundane (imagination, daydreaming, physical pleasures,
+physical discomforts, etc.) to the supramundane (out-of-body experiences,
+hallucinations, etc.) are subsets of these seven sense doors and their reflection in
+bodily sensation, the eighth.
+
+TODO: Attention in one place only
 
@@
 
+\pagebreak
+
+\begin{thebibliography}{9}
+\raggedright
+
+\bibitem{sense-icons}
+  5 Senses by Daniel Falk from the Noun Project
+  \url{https://thenounproject.com/daniel2021/collection/human-body-senses/}
+  Thought by Nociconist from the Noun Project
+  \url{https://thenounproject.com/search/?q=thought&i=2025873}
+  Heart by Rafael Garcia Motta from the Noun Project
+  \url{https://thenounproject.com/search/?q=heart&i=807960}
+  Body by Makarenko Andrey from the Noun Project
+  \url{https://thenounproject.com/search/?q=body&i=789989}
+  \textit{The Noun Project}.
+
+\end{thebibliography}
+
 \end{document}

### 2019-06-19 paper-three/paper-three.tex
initial notes on paper three

@@
 
-TODO: Attention in one place only
+Mapping all of sensory experience to these eight categories begs the question of
+attention, of awareness: Where does the meditator try to fasten her awareness? Where
+is awareness normally? For the average person, awareness jumps around across these
+eight categories. Even when one tries to focus on a difficult intellectual problem,
+the discomforts of back pain and hunger or the distraction of an irritating sound
+would draw attention away from thought, the desired object of attention. Vipassana
+meditation asks the meditator to use bodily sensation as a gateway to the other seven
+sense experiences. Rather than focusing on sound, focus on the sensation generated in
+the body by the ear sense door. Rather than focusing on a thought or emotion, focus
+on the sensation in the body generated by that thought or emotion. This is extremely
+difficult to do, which is why (for lay people, in most cases) a 10-day silent
+residential course \cite{dhamma} is necessary to learn the technique.
+
+\begin{center}
+  \Huge{The Mundane and The Supramundane}
+\end{center}
+
+Reasons for meditating will be broken down into two categories. \textit{The Mudane}
+in this context refers not to the tedious but to the earthly, the material. Most people will begin
+meditating for reasons in the mundane field simply because most people have never
+experienced the supramundane. \textit{The Supramundane} in this context refers not to
+spirituality or religion but experiences which transcend the material, physical
+world.
+
+Because supramundane experiences tend to occur only in deep meditative states,
+the reasons for meditating listed will predominantly fall in the mundane
+category. Whether or not the meditator experiences supramundane states in deep
+meditation, these altered states of mind are never the goal of meditation. The goal
+of meditation is to change the meditator's mental habits, to move away from unhealthy
+mental patterns which cause harmful behaviours toward healthy mental patterns which
+encourage productive behaviour. Obviously this change is only visible in the mundane
+world, outside of meditation.
+
+\begin{center}
+  \Huge{Meditation vs. Naps}
+\end{center}
+
+While staying at a friend's house, I excused myself in the evening to meditate. He
+sincerely asked, ``Is meditating for an hour really more valuable than using that
+time for a good nap?''
 

### 2019-09-28 paper-two/paper-two.tex
adding paper 2 draft

\documentclass{article}

\usepackage{polyglossia}
\setdefaultlanguage{english}
\setotherlanguage{sanskrit}
\usepackage{fontspec}
\newfontfamily\devanagarifont[Scale=MatchUppercase]{Kohinoor Devanagari}

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage[bottom]{footmisc}
\usepackage{enumitem}
\usepackage{hyperref}

\setlength{\parindent}{1.3em}
\setlength{\parskip}{0.7em}
\setlist{noitemsep}

\begin{document}

\begin{titlepage}
   \vspace*{\stretch{1.0}}
   \begin{center}
     \Huge\textbf{Vipassana for Hackers}\\
     \Huge{Paper Two: The Brain}\\
     \vspace{5cm}
     \large\textit{Preethi Govindarajan}\\
     \large\textit{Version 0.1}\\
     \large\textit\today\\
     \vspace{5cm}
     \large\textit{Contributions by: Steven Deobald}\\
   \end{center}
   \vspace*{\stretch{2.0}}
\end{titlepage}

\begin{center}
  \Huge{Context}
\end{center}

I took my first Vipassana course in 2017 and since then I have been trying to wade through the research that has been published in this field, especially white papers dealing with the effects of a 10-day course on the brain. There is not much in terms of this research and the research that does exist usually uses a sample of experienced meditators and/or self-reports as a outcome. I have tried to write down what I experienced during my 10-day courses over the past year and tried to corroborate it with research that does exist regarding meditation and the brain.
Disclaimer: My primary field of research is biology but I am far removed from neuroscience. I have tried to simplify the work out there so as to better understand it. So, if there are any corrections to make or editing in terms of the content, please feel free to get in touch with me.
Before going any further, I will start by describing the different parts of the brain that have shown up in scientific literature as correlated to meditative practices.
Parts of the brain and their functions
Default Mode Network:
These are a set of structures in the mid-line of the brain (which is a evolutionarily more primitive area related to memory and emotion) which connects to structures in the cortex (which is evolutionarily the most recent part of the brain and contains the executive and higher order functions). It is the constellation of regions that light up when the brain is not engaged in any external/goal-oriented tasks (1). These regions are functionally connected and are involved in
a) The neurological basis of the self —
The regions involved in this are the Posterior Cingulate Cortex (PCC), venteromedial Prefrontal Cortex (vmPFC) and the Inferiour Parietal Lobule (IPL) (1).
Cerebral blood flow and metabolic rate is 40\% higher in the PCC compared to the average across the brain making it one of the most active regions of the brain. (Leech R, Sharp DJ (July 2013). “The role of the posterior cingulate cortex in cognition and disease”. Brain. 137 (Pt 1): 12–32.)
The PCC has been thought to mediate interactions between emotion and memory. It consistently lights up when a person recollects something from their life and depending on the emotional importance of the memory the strength of the lighting up changes. It is also activated by emotional stimuli both positive and negative and acts as an interface between the world and us by gauging the importance of different stimuli(2).

Magnetic resonance Image of the areas of the brain in the Default Mode Network. http://www.frontiersin.org/Neurotrauma/10.3389/fneur.2013.00016/full
The mPFC is divided into venteromedial PreFrontal Cortex (vmPFC) and dorsomedial Prefrontal Cortex (dmPFC). The vmPFC is involved in creating a conceptual self. It is involved in self-related processing and assigns personal significance to self-related information (3).
It is hypothesized that the activation of these three regions of the brain are responsible for providing a sense of self — as a subjective agent in space and time. It functions as a network in which phenomena such as self representations are accessible via the PCC and are gated into conscious awareness by activity in the MPFC which is influenced by changing internal and external demands. The PCC which acts as a brain-wide connectivity hub through which a unitary notion of self is created by a large scale integration of DMN activity (4).
b) Thinking about others —
The Dorsal Medial Prefrontal Cortex (dmPFC) is involved in Theory of the mind(considering the mental states of others), empathy, moral reasoning and altruistic behavior (1,5,6)
c) Autobiographical memory and future simulations —
The DMN is also involved in remembering the past and imagining the future and story comprehension(1).
Salience Network (SN):
Salience: The perceptual quality by which an observable thing stands out relative to its environment
The SN is an intrinsically connected large-scale network anchored in the Anterior Insula (AI) and dorsal Anterior Cingulate Cortex. Both these regions have reached a high degree of specialization in the great apes. It is the collection of the regions in the brain that help decide which stimuli deserves our attention. It acts as a switch between the internally directed DMN and the externally directed Central Executive Network (involved in high level cognitive functions) (7).

The Salience Network consists of the Anterior Cingulate Cortex and the Insula. It helps to switch between the Default Mode Network and the Central Executive Network
a) Anterior Cingulate Cortex (ACC): The ACC is the front end of the cingulate cortex and collars around the corpus callosum (the band connecting the two hemispheres of the brain). It is the connector between the emotional (limbic system) and the cognitive part of the brain. It is involved in functions such as attention allocation, reward anticipation, decision making, morality, impulse control, emotional awareness and registering pain (8–12). It also appears to play a role in the regulation of Autonomic functions such as blood pressure and heart rate (13).
b) Anterior Insular Cortex (AIC): The insula is a part of the cerebral cortex that is located deep within the sulcus (which is the fissure separating the four lobes of the brain). The AIC projects itself into the amygdala. It is involved in multimodal sensory processing such as audio-visual integration tasks, interoceptive awareness (so its activity is directly related to ones sense of internal body states), empathy and conscious awareness (14). It also plays a role in the regulation of autonomic functions such as bodily sensations (as well as where the degree of pain is judged), taste and control of the immune system (15).
The AIC and ACC together give rise to our interoceptive and conscious self-awareness.
Central Executive Network:
This network is activated when a high-level cognitive tasks or external goal-oriented tasks are being performed. Executive functions are cognitive processes involved in cognitive control of behavior.
The regions of the brain involved in this are Dorsolateral Prefrontal Cortex, Orbito Frontal Cortex and the Posterior Parietal Cortex.
Dorsolateral Prefrontal Cortex (DL-PFC): It is a part of the pre-frontal cortex and is found in primates and humans. It is a part of the Central Executive Network. The DL-PFC is involved in higher cognitive processes including working memory (holding different pieces of information and manipulating and using them for tasks), selective attention, cognitive flexibility (switching between tasks) and planning. It also seems to be involved in social cognition and lying and increases dopamine levels in the brain.
Barbey AK, Koenigs M, Grafman J (May 2013). “Dorsolateral prefrontal contributions to human working memory” (PDF). Cortex. 49 (5): 1195–1205.
Monsell S (2003). “Task switching”. Trends in Cognitive Sciences. 7 (3): 134–140. doi:10.1016/S1364–6613(03)00028–7. PMID 12639695.
Ito, Ayahito; Abe, Nobuhito; Fujii, Toshikatsu; Hayashi, Akiko; Ueno, Aya; Mugikura, Shunji; Takahashi, Shoki; Mori, Etsuro (2012). “The contribution of the dorsolateral prefrontal cortex to the preparation for deception and truth-telling”. Brain Research. 1464: 43–52. doi:10.1016/j.brainres.2012.05.004
Orbito Frontal Cortex (OFC): This is an area found in front of both the hemispheres and just above the eye. It is a part of the Pre-Frontal Cortex and it is thought to be involved in the cognitive process of decision making. It is thought to represent emotion and reward in decision making. It also receives input from multiple sensory modalities and in turn activates the amygdala and the hypothalamus
Fuster, J.M. The Prefrontal Cortex, (Raven Press, New York, 1997
Rolls, ET (November 2004). “Convergence of sensory systems in the orbitofrontal cortex in primates and brain design for emotion”. The Anatomical Record Part A: Discoveries in Molecular, Cellular, and Evolutionary Biology. 281 (1): 1212–25. doi:10.1002/ar.a.20126
Price, Joseph L. (2006). “Chapter 3: Connections of the orbital cortex”. In Rauch, Scott L.; Zald, David H. (eds.). The Orbitofrontal Cortex. New York: Oxford University Press. p. 45.
Posterior Parietal Cortex (PPC): It receives information from the Auditory, Visual and Somatosensory systems and in turn activates the motor cortex or DL-PFC.
Martin, R. E. (n.d.). Let’s Get to Know the Parietal Lobes! [PDF]. Retrieved from \url{http://gablab.mit.edu/downloads/Parietal_Primer.pd}

Approximate location of the OFC on an MRI
(Maybe remove)
The limbic system: The limbic system is a complex set of structures that lies on both sides of the thalamus, just under the cerebrum. It includes the hypothalamus, the hippocampus, the amygdala, and several other nearby areas. It appears to be primarily responsible for our emotional life, and has a lot to do with the formation of memories.

The course
I applied for this ten day course after I quit my job. I was very nervous as soon as the acceptance came through until the first day. I had never done anything like this before and I was very comfortably placed within the idea that meditation with its spiritual connotation was a waste of my time. It was for people who had time to waste and I was not one of them. I went for my first course in Chennai and once I got there I had to fill all the forms all over again keep all my luggage in a locker and go to my room. I had to share the room with one other person and I was so out of my element and knew everyone could see me and tell that I was the odd one out, the one that was not supposed to be there and I was too scared to actually make conversation with anyone before silence was enforced. So with my brain filled with thoughts about myself, about the people around me, I started this meditation business. Within the first two days, I had made up elaborate stories about my fellow meditators, all superheroes tirelessly working to save humanity together. My default mode network was in overdrive.
For the first 3.5 days you are asked to focus your attention on the area below your nostrils and above your upper lip. You start with observing your breath coming in and going out. As the teacher mentions in the late evening discourses, the reason for starting with your breath is because this activity is both conscious and unconscious and acts as a bridge.
At least during the meditation hours, I was trying hard, I wanted to give this an honest shot.I focused on the small patch beneath my nostrils above my upper lip. To begin with when you are asked to observe your breath and your mind wanders, that is the default mode network working, as you bring your attention back to your breath, your salience network is activated. Slowly at first, with your mind wandering for many many minutes before you realize you are not with your breath but within the 3.5 days, it pretty quickly learns to bring it back and almost does it by habit. As I gave in to this activity, tried to focused more, tried to sit still, there were longer periods of awareness on that patch of skin and stranger sensations arising and passing. By the third day, it felt like entire ecosystems writhing and flopping and crashing were alive in the area below my nostrils above my upper lip. There have been studies that have shown that once you bring your attention back to breath, the neural structures involved in the control of the Autonomic Nervous System and attention start firing more actively (DL-PFC-a part of the CEN, the Hippocampus and the ACC — part of the SN). There are also global dampening changes seen in the brain especially within the DMN and the Amygdala (which is involved in the flight-or-fight response/stress response). These changes have antagonistically been termed the relaxation response. This relaxation response can be thought of as a gateway to altered states of mind. (16)
There has also been research in mice, showing a cluster of nerves called the pre-botzinger complex found in the brain stem of most mammals which fire with every breath taken. This breathing pacemaker seems to work for all kinds of breathing such as yawns, sighs and gasps. They also appear to play a role in calming and arousal (17). It would be interesting to see the activation in these neuron clusters during the first three days of the course even amongst novice meditators.
After lunch on the fourth day. The instructions for Vipassana are given, where you transfer your attention from the patch below your nostril to the top of your head. From there, for over two hours, the instructions are given to slowly move your attention till your feet and then repeat. The two criteria for observation here are focused attention and open monitoring (observing sensations objectively).
The Salience Network (ACC and AIC) and the Central Executive Network (DL-PFC, PPC and OFC) which are involved in the control of attention start firing rapidly. Usually as the person is engaged in an external task, the Central executive network (Orbito Frontal Cortex and Posterior Parietal Cortex), receiving signals from all the sensory networks (Visual, Audio, Somatosensory etc..) continuously and in response sends signals to the different motor cortices, amygdala, hypothalamus to perform a task, feel an emotion, etc..
Now imagine a situation where once the Salience Network is activated and in turn activates the Central Executive Network and the task that it is supposed to perform is not external but to observe sensation and do nothing. So as you sit there with your eyes closed and not moving and scanning your body, you are doing what the central executive network has always been doing except attempting to not react.
This is hard at first, with the body involuntarily reacting by jerking and writhing in response to your observation. Your awareness itself is jerky and it is hard to observe sensations consciously.
There is also the pain. This eclipses all other sensations. It is difficult trying to be aware of any other sensation with the pain that was emanating from my legs and back. The Orbito Prefrontal Cortex (OFC), is involved in assigning emotion to sensation (18) and the ACC plays a vital role in perception of sensation. These two parts of the brain are usually functionally connected. In the beginning, as I sat trying to observe sensations and I was only aware of the pain, the ACC was firing rapidly and the OFC then assigned these signals to the limbic system (including the amygdala and hippocampus) which then put me in a less than ideal emotional state.
But as your awareness gets smoother, this functional connectivity between the ACC and the OFC slowly starts changing and as you continue to pay attention to the different sensations and attempt to not assign emotion and observe with a non-judgmental attitude particularly toward unpleasant stimuli, the OFC still fires but the limbic system is quietened, so you react less. It is like both your conscious and unconscious awarenesses are slowly aligning until some points when it feels like it is pure awareness moving through your body or some parts of your body. In addition the insula (part of the Salience Network) and the somatosensory cortex which are involved in sensory awareness were firing more actively.
(The somatosensory cortex is located in the gyrus and the parietal lobes: The sensory homunculus which is a part of the somatosensory cortex is a cortical representation of the body based on the degree of sensory innervation)

Somatosensory and Motor cortices
Another strange thing that starts happening is as your mind wandering slows down and you sit with your eyes closed, your sense of your self as a subjective agent in space and time changes, first just very slightly, as you close your eyes and meditate, it feels like you have shifted in space or turned in different directions (during these times the vmPFC and PCC which are a part of the DMN become less active), and then as the meditations get deeper and sometimes, the DMN quietened down completely for a very short duration and in addition to this, the insula is firing more rapidly, it could lead to other alternative networks of consciousness arising. It is as if losing the influence of one’s narrative leads to new insights that are usually kept from consciousness. Experiencing this even for a few seconds can lead to lasting changes even after this activity wears off. These experiences can be quite similar to being on certain kinds of drugs or even in some cases of ecstatic seizures where the abnormal activity of the anterior insula leads to heightened self awareness and feelings of bliss and lack of ambiguity. (Fabienne Picard, State of belief, subjective certainty and bliss as a product of cortical dysfunction, Cortex, Volume 49, Issue 9, 2013,
Pages 2494–2500ISSN 0010 9452, https://doi.org/10.1016/j.cortex.2013.01.006.)
So to conclude, I believe attending a ten day course literally did change the habit pattern of my mind at least for a while.
It quietens the Default Mode Network and causes the Salience Network and certain parts of the Central Executive Network to fire more rapidly (the parts that are aware of the sensations while quietening the reacting part of this network). This causes changes to a persons behavior at least for a short while and this is usually seen as soon as a person leaves the course. But repeated practice is essential for any sustained neuroplasticity since rewriting many years of habit formation requires more than a ten-day course. However, the ten day course is well designed to give the minimum observational change neccesary for people interested to keep practicing.
There have also been studies looking at the structural changes in the brain and functional changes especially to the default mode network and the salience network but these are usually measured after one has clocked over thousands of hours of practice. Therefore I am not delving into those right now.
How is this different from just concentrating on tasks hard? It is different because when concentrating on any task, your default mode network is quetened and the salience network and CEN are firing entering you into a state of flow but the reacting part of the brain is also working and working hard.
There are aspects to this technique that I have not understood or experienced that I have not delved into right now and may do once i know it better.
I was really intrigued as I left the center as to what had occurred over those ten days that changed my brain, I very quickly noticed though that as soon as talking was allowed, these changes slipped away really quickly, but with each hour long sitting a small tiny bit sticks just a little more.
Reference
Andrews-Hanna, Jessica R. (2012–06–01). “The brain’s default network and its adaptive role in internal mentation”. The Neuroscientist: A Review Journal Bringing Neurobiology, Neurology and Psychiatry. 18 (3): 251–270. doi:10.1177/1073858411403316. ISSN 1089–4098. PMC 3553600. PMID 21677128.
Maddock, Richard J.; Garrett, Amy S.; Buonocore, Michael H. (January 2003). “Posterior cingulate cortex activation by emotional words: fMRI evidence from a valence decision task”. Human Brain Mapping. 18 (1): 30–41.
Andrews-Hanna, Jessica R.; Smallwood, Jonathan; Spreng, R. Nathan (2014–05–01). “The default network and self-generated thought: component processes, dynamic control, and clinical relevance”. Annals of the New York Academy of Sciences. 1316(1): 29–52. doi:10.1111/nyas.12360. ISSN 1749–6632. PMC 4039623. PMID 24502540.
Davey CG, Pujol J, Harrison BJ. Mapping the self in the brain’s default mode network. Neuroimage. 2016 May 15;132:390–397.
Isoda, M., \& Noritake, A. (2015). What makes the dorsomedial frontal cortex active during reading the mental states of others?. Neural basis of social learning, social deciding, and other-regarding preferences, 51.
Waytz, A., Zaki, J., \& Mitchell, J. P. (2012). Response of dorsomedial prefrontal cortex predicts altruistic behavior. The Journal of Neuroscience, 32(22), 7646–7650.
Menon, V; Toga, A (2015). Salience Network. Elsevier. pp. 597–611. ISBN 978–0–12–397316–0.
Pardo JV, Pardo PJ, Janer KW, Raichle ME (January 1990). “The anterior cingulate cortex mediates processing selection in the Stroop attentional conflict paradigm”. Proceedings of the National Academy of Sciences of the United States of America. 87 (1): 256–9. doi:10.1073/pnas.87.1.256. PMID 2296583.
Bush G, Vogt BA, Holmes J, Dale AM, Greve D, Jenike MA, Rosen BR (January 2002). “Dorsal anterior cingulate cortex: a role in reward-based decision making”. Proceedings of the National Academy of Sciences of the United States of America. 99 (1): 523–8. doi:10.1073/pnas.012470999. PMC 117593. PMID 11756669.
Sevinc G, Gurvit H, Spreng RN (July 2017). “Salience network engagement with the detection of morally laden information”. Social Cognitive and Affective Neuroscience. 12 (7): 1118–1127. doi:10.1093/scan/nsx035. PMID 28338944.
Jackson PL, Brunet E, Meltzoff AN, Decety J (2006). “Empathy examined through the neural mechanisms involved in imagining how I feel versus how you feel pain”. Neuropsychologia. 44 (5): 752–61. doi:10.1016/j.neuropsychologia.2005.07.015. PMID 16140345.
Bush G, Luu P, Posner MI (June 2000). “Cognitive and emotional influences in anterior cingulate cortex”. Trends in Cognitive Sciences. 4 (6): 215–222. doi:10.1016/S1364–6613(00)01483–2. PMID 10827444
Gianaros PJ, Derbyshire SW, May JC, Siegle GJ, Gamalo MA, Jennings JR. Anterior cingulate activity correlates with blood pressure during stress. Psychophysiology. 2005;42(6):627–35.
Xiaosi Gu, Patrick R. Hof, Karl J. Friston, Jin Fan. J Comp Neurol. 2013 Oct 15; 521(15): 3371–3388.
Rolls ET. Functions of the anterior insula in taste, autonomic, and related functions. Brain Cogn. 2016 Dec;110:4–19.
Lazar SW; Bush George; Gollub RL.; Fricchione, GL.; Khalsa G; Benson H. Functional brain mapping of the relaxation response and meditation.
Neuroreport. 11(7):1581–1585, May 15, 2000.
J. Muñoz-Ortiz, E. Muñoz-Ortiz, L. López-Meraz, L. Beltran-Parrazal, C. Morgado-Valle. The pre-Bötzinger complex: Generation and modulation of respiratory rhythm, Neurología (English Edition), 2018, ISSN 2173–5808,
https://doi.org/10.1016/j.nrleng.2018.05.006.
REMPEL‐CLOWER, N. L. (2007), Role of Orbitofrontal Cortex Connections in
Emotion. Annals of the New York Academy of Sciences, 1121:
72–86. doi:10.1196/annals.1401.026

\pagebreak

\begin{thebibliography}{9}
\raggedright

\bibitem{thenewsroom}
  Aaron Sorkin.
  \url{https://www.youtube.com/watch?v=2C6h-Yyx9Yk}
  \textit{The Newsroom}.
  HBO, 2012.

\end{thebibliography}

\end{document}

### 2019-09-28 paper-two/build-epub.sh

@@ -1,0 +1,17 @@
+#!/bin/sh
+
+SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
+LIBPATH="$SCRIPTPATH/../lib"
+PATH=$PATH:$LIBPATH
+
+OUT=../output
+
+latexml --dest=paper-two.xml paper-two.tex
+latexmlpost -dest=paper-two.html paper-two.xml
+
+# which ebook-convert
+# /usr/local/bin/ebook-convert
+
+ebook-convert paper-two.html ../output/paper-two.epub --language en --output-profile=kindle --authors="Steven Deobald" --title="Vipassana for Hackers"
+
+rm -f paper-two.xml paper-two.html LaTeXML.cache LaTeXML.css ltx-article.css

### 2019-09-28 paper-two/build-pdf.sh

@@ -1,0 +1,22 @@
+#!/bin/sh
+
+# pdflatex was used originally but cannot render Sanskrit without a lot of hoop-jumping
+# pdflatex vipassana-for-hackers.tex
+
+# xelatex is apparently less capable than pdflatex at font-rendering in general, but
+# is capable of rendering non-Latin fonts (like Hindi/Sanskrit) by default
+
+# Compile twice to resolve references.
+
+SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
+LIBPATH="$SCRIPTPATH/../lib"
+PATH=$PATH:$LIBPATH
+
+OUT=../output
+
+echo "Running pre-build. This takes a while..."
+xelatex -output-directory=$OUT paper-two.tex > /dev/null 2>&1
+echo "...done. Now running the build."
+xelatex -output-directory=$OUT paper-two.tex
+
+rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out

### 2019-09-28 paper-two/paper-two.tex
p2: brain functions started

@@
-\documentclass{article}
+\documentclass[twocolumn]{article}
 
@@
 
-I took my first Vipassana course in 2017 and since then I have been trying to wade through the research that has been published in this field, especially white papers dealing with the effects of a 10-day course on the brain.
-There is not much in terms of this research and the research that does exist usually uses a sample of experienced meditators and/or self-reports as a outcome.
-I have tried to write down what I experienced during my 10-day courses over the past year and tried to corroborate it with research that does exist regarding meditation and the brain.
-Disclaimer: My primary field of research is biology but I am far removed from neuroscience.
-I have tried to simplify the work out there so as to better understand it.
-So, if there are any corrections to make or editing in terms of the content, please feel free to get in touch with me.
-Before going any further, I will start by describing the different parts of the brain that have shown up in scientific literature as correlated to meditative practices.
-Parts of the brain and their functions
+I took my first Vipassana course in 2017. Vipassana meditation was so unlike anything
+I had ever experienced before, after the course I was left extremely curious
+about what exactly had happened to me during those ten days. For months afterward, I
+spent my mornings and evenings wading through the research in the field of
+meditation, with a specific focus on white papers dealing with the effects a
+10-day Vipassana course has on the brains of participants. The research in this area
+is limited. The quality research which does exist usually uses a sample of highly
+experienced meditators rather than beginners and/or self-reports rather than
+objective measures.
+
+Over the past year, I have tried to write down what I experienced during that first
+10-day course and the 10-day courses I have taken since, corroborating my experience
+with research that does exist regarding meditation and the brain.
+
+\textbf{Proviso:} S.N. Goenka, the principal teacher of Vipassana meditation,
+actively dissuades students from precisely the sort of brain-centred biological
+inquiry presented in this paper.
+
+\begin{quote}
+  \textbf{The brain itself is just a physical organ. As you deal with other parts of
+    the body, you deal with the brain in the same way, that's all. Nothing special to
+    do with the brain. But the mind is totally different. In the West, all importance
+    is given to the brain as if the mind is located here. Nothing doing, it is
+    everywhere. The mind is in the whole body. So give attention to the whole
+    body.} \cite{goenkabrain}
+\end{quote}
+
+This paper does not contradict Mr. Goenka's sentiment, but instead acts as a starting
+point for those who view the brain as the quintessential physical component of mind.
+
+\textbf{Disclaimer:} Although my primary field of research is biology, my work is far removed
+from neuroscience. I have tried to simplify the research available so as to better
+understand it. If there are any corrections to make or editing in terms of the
+content, please feel free to get in touch with me.
+
+\begin{center}
+  \Huge{Brain Function and Anatomy}
+\end{center}
+
+Before dissecting the experience of meditation as it pertains to the brain, this
+section describes the different parts of the brain that have shown up in
+scientific literature as correlated to meditative practices. Neuroscience divides the
+brain into the reified geography of the brain, its anatomy, and the abstract concepts
+governed by brain activity, its function. Between concrete physiology and abstract
+outcomes exist networks of cooperative structures which correspond to general
+high-level activities of the brain.
+
 Default Mode Network:
@@
 I was really intrigued as I left the center as to what had occurred over those ten days that changed my brain, I very quickly noticed though that as soon as talking was allowed, these changes slipped away really quickly, but with each hour long sitting a small tiny bit sticks just a little more.
-Reference
+
+\pagebreak
+
+References
+
 Andrews-Hanna, Jessica R.
@@
 
+\bibitem{goenkabrain}
+  S.N. Goenka.
+  \url{https://www.vridhamma.org/A-store-house-of-answers-by-Shri-S-N-Goenka}
+  \textit{Answers by Mr. S. N. Goenka}
+
 \bibitem{thenewsroom}

### 2019-09-28 paper-two/paper-two.tex
switching explained, dmn started

@@
 
-\begin{center}
-  \Huge{Context}
-\end{center}
+\section{Introduction}
 
@@
 spent my mornings and evenings wading through the research in the field of
-meditation, with a specific focus on white papers dealing with the effects a
-10-day Vipassana course has on the brains of participants. The research in this area
+meditation. I was specifically focused on white papers dealing with the effects of a
+10-day Vipassana course on the brains of participants. The research in this area
 is limited. The quality research which does exist usually uses a sample of highly
@@
 This paper does not contradict Mr. Goenka's sentiment, but instead acts as a starting
-point for those who view the brain as the quintessential physical component of mind.
+point for readers who view the function of the brain as central to the activity of
+the mind.
 
-\textbf{Disclaimer:} Although my primary field of research is biology, my work is far removed
+\textbf{Disclaimer:} Although my primary field of research is within the field of
+biology, my work is far removed
 from neuroscience. I have tried to simplify the research available so as to better
@@
 
-\begin{center}
-  \Huge{Brain Function and Anatomy}
-\end{center}
+\section{Brain Function and Anatomy}
 
@@
 section describes the different parts of the brain that have shown up in
-scientific literature as correlated to meditative practices. Neuroscience divides the
+scientific literature as correlated to meditative practices. Recent neuroscience divides the
 brain into the reified geography of the brain, its anatomy, and the abstract concepts
 governed by brain activity, its function. Between concrete physiology and abstract
-outcomes exist networks of cooperative structures which correspond to general
+functional outcomes exist networks of cooperative structures which correspond to general
 high-level activities of the brain.
 
-Default Mode Network:
-These are a set of structures in the mid-line of the brain (which is a evolutionarily more primitive area related to memory and emotion) which connects to structures in the cortex (which is evolutionarily the most recent part of the brain and contains the executive and higher order functions).
-It is the constellation of regions that light up when the brain is not engaged in any external/goal-oriented tasks (1).
-These regions are functionally connected and are involved in
+The Default Mode Network (DMN) is the constellation of regions which fire when
+the brain is not engaged in any external or goal-oriented
+tasks. As the name suggests, the brain is in its ``default
+mode''. \cite{defaultnetworkadaptive} The DMN is anticorrelated to the Central Executive
+Network (CEN), which is active during externally-directed, high-level cognitive
+functions. \cite{saliencenetwork} This anticorrelation between the DMN and the CEN is
+governed by the Salience Network (SN), the collection of regions in the brain which
+help decide which stimuli deserve our attention. The SN acts as a switch between the
+internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}
+
+[ TODO: insert switching diagram ]
+
+\subsection{Default Mode Network}
+
+The DMN is comprised of specific anatomy including portions of the mid-line of the
+brain, an evolutionarily primitive area related to memory and emotion, and
+structures in the cortex, an evolutionarily recent part of the brain containting the
+executive and higher order functions. \cite{defaultnetworkanatomy}
+
+These regions are functionally connected and are involved in the neurological basis
+of the self, considering the mental states of others, remembering the past, and
+imagining the future.
+
 a) The neurological basis of the self —
@@
 
-Andrews-Hanna, Jessica R.
-(2012–06–01).
-“The brain’s default network and its adaptive role in internal mentation”.
-The Neuroscientist: A Review Journal Bringing Neurobiology, Neurology and Psychiatry.
-18 (3): 251–270.
-doi:10.1177/1073858411403316.
-ISSN 1089–4098.
-PMC 3553600.
-PMID 21677128.
 Maddock, Richard J.; Garrett, Amy S.; Buonocore, Michael H.
@@
 Waytz, A., Zaki, J., \& Mitchell, J. P. (2012). Response of dorsomedial prefrontal cortex predicts altruistic behavior. The Journal of Neuroscience, 32(22), 7646–7650.
-Menon, V; Toga, A (2015). Salience Network. Elsevier. pp. 597–611. ISBN 978–0–12–397316–0.
+
 Pardo JV, Pardo PJ, Janer KW, Raichle ME (January 1990).
@@
 
+\bibitem{defaultnetworkadaptive}
+  Andrews-Hanna, Jessica R.
+  \textit{The brain’s default network and its adaptive role in internal mentation.}
+  The Neuroscientist: A Review Journal Bringing Neurobiology, Neurology and
+  Psychiatry. 18 (3): 251–270. doi:10.1177/1073858411403316. ISSN 1089–4098. PMC
+  3553600. PMID 21677128. (2012–06–01).
+
+\bibitem{defaultnetworkanatomy}
+  Buckner, R. L.; Andrews-Hanna, Jessica R.; Schacter, D. L.
+  \textit{The Brain's Default Network: Anatomy, Function, and Relevance to Disease.}
+  Annals of the New York Academy of Sciences. 1124 (1): 1–38. CiteSeerX
+  10.1.1.689.6903. doi:10.1196/annals.1440.011. PMID 18400922, 2008
+
+\bibitem{saliencenetwork}
+  Menon, V; Toga, A.
+  \textit{Salience Network.}
+  Elsevier. pp. 597–611, 2015
+  ISBN: 978–0–12–397316–0.
+
 \bibitem{thenewsroom}

### 2019-09-28 paper-two/paper-two.tex
dmn done, sn next

@@
 structures in the cortex, an evolutionarily recent part of the brain containting the
-executive and higher order functions. \cite{defaultnetworkanatomy}
-
-These regions are functionally connected and are involved in the neurological basis
+executive and higher order functions. \cite{defaultnetworkanatomy} These functionally
+connected regions are involved in the neurological basis
 of the self, considering the mental states of others, remembering the past, and
@@
 
-a) The neurological basis of the self —
-The regions involved in this are the Posterior Cingulate Cortex (PCC), venteromedial Prefrontal Cortex (vmPFC) and the Inferiour Parietal Lobule (IPL) (1).
-Cerebral blood flow and metabolic rate is 40\% higher in the PCC compared to the average across the brain making it one of the most active regions of the brain.
-(Leech R, Sharp DJ (July 2013).
-“The role of the posterior cingulate cortex in cognition and disease”.
-Brain.
-137 (Pt 1): 12–32.)
-The PCC has been thought to mediate interactions between emotion and memory.
-It consistently lights up when a person recollects something from their life and depending on the emotional importance of the memory the strength of the lighting up changes.
-It is also activated by emotional stimuli both positive and negative and acts as an interface between the world and us by gauging the importance of different stimuli(2).
+\subsubsection{The neurological basis of the self}
 
-Magnetic resonance Image of the areas of the brain in the Default Mode Network. http://www.frontiersin.org/Neurotrauma/10.3389/fneur.2013.00016/full
-The mPFC is divided into venteromedial PreFrontal Cortex (vmPFC) and dorsomedial Prefrontal Cortex (dmPFC).
-The vmPFC is involved in creating a conceptual self.
-It is involved in self-related processing and assigns personal significance to self-related information (3).
-It is hypothesized that the activation of these three regions of the brain are responsible for providing a sense of self — as a subjective agent in space and time.
-It functions as a network in which phenomena such as self representations are accessible via the PCC and are gated into conscious awareness by activity in the MPFC which is influenced by changing internal and external demands.
-The PCC which acts as a brain-wide connectivity hub through which a unitary notion of self is created by a large scale integration of DMN activity (4).
-b) Thinking about others —
-The Dorsal Medial Prefrontal Cortex (dmPFC) is involved in Theory of the mind(considering the mental states of others), empathy, moral reasoning and altruistic behavior (1,5,6)
-c) Autobiographical memory and future simulations —
-The DMN is also involved in remembering the past and imagining the future and story comprehension(1).
-Salience Network (SN):
-Salience: The perceptual quality by which an observable thing stands out relative to its environment
+The regions involved in an individual's conceptual ``self'' are the Posterior
+Cingulate Cortex (PCC), Ventromedial Prefrontal Cortex (vmPFC), and the Inferior
+Parietal Lobule (IPL). \cite{defaultnetworkadaptive}
+
+Cerebral blood flow and metabolic rate is 40\%
+higher in the PCC compared to the average across the brain, making it one of the most
+active regions of the brain. \cite{pccrole}
+
+The PCC has been thought to mediate interactions between emotion and memory. Under
+fMRI, it consistently lights up when a person recollects something from their
+life. The strength of PCC brain activity in such an experiment varies depending on
+the emotional importance of the memory recalled. It is also activated by emotional
+stimuli, both positive and negative, and acts as an interface between the external
+world and the individual by gauging the importance of different
+stimuli. \cite{pccemotion}
+
+[ TODO: MRI of DMN image ] \cite{dmnmri}
+
+The vmPFC is involved in creating a conceptual self by self-related processing
+and the assignment of personal significance to self-related
+information. \cite{dmnself} The vmPFC is paired with the Dorsomedial Prefrontal
+Cortex (dmPFC), which assists in the creation of the conceptual self through an
+individual's consideration of ``others'', to construct the Medial Prefrontal Cortex
+(mPFC).
+
+[ TODO: explain IPL? ]
+
+It is hypothesized that the activation of these three regions of the brain (PCC,
+vmPFC, and IPL) are responsible for providing a sense of self --- a subjective agent
+in space and time. It functions as a network in which phenomena such as self
+representations are accessible via the PCC [ TODO: ref ] and are gated into conscious awareness by
+activity in the mPFC, as influenced by changing internal and external demands. The
+PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
+created by a large scale integration of DMN activity. \cite{mappingself}
+
+\subsubsection{Thinking about others}
+
+The \textit{Theory of Mind} is a body of work in philosophy and psychology which
+describes an individual's ability to consider the mental states of others, such as
+appreciating another individual's false beliefs (knowledge based on incorrect or
+outdated information). The Theory of Mind captures not only one's ability to
+attribute beliefs, desires, and emotions to others but also to oneself --- and to
+differentiate the two. \cite{autistictheoryofmind}
+
+In addition to The Theory of Mind, the dmPFC is involved in empathy, moral reasoning,
+and altruistic behavior. \cite{defaultnetworkadaptive,dmpfcothers,dmpfcaltruism}
+
+\subsubsection{Autobiographical memory and future simulations}
+
+The DMN is also involved in remembering the past, imagining the future, and story
+comprehension. \cite{defaultnetworkadaptive}
+
+[ TODO: expand? how? ]
+
+\subsection{Salience Network}
+
+\begin{quote}
+  \textbf{Salience: The perceptual quality by which an observable thing stands out
+    relative to its environment}
+\end{quote}
+
 The SN is an intrinsically connected large-scale network anchored in the Anterior Insula (AI) and dorsal Anterior Cingulate Cortex.
@@
 
-Maddock, Richard J.; Garrett, Amy S.; Buonocore, Michael H.
-(January 2003).
-“Posterior cingulate cortex activation by emotional words: fMRI evidence from a valence decision task”.
-Human Brain Mapping.
-18 (1): 30–41.
-Andrews-Hanna, Jessica R.; Smallwood, Jonathan; Spreng, R.
-Nathan (2014–05–01).
-“The default network and self-generated thought: component processes, dynamic control, and clinical relevance”.
-Annals of the New York Academy of Sciences.
-1316(1): 29–52.
-doi:10.1111/nyas.12360.
-ISSN 1749–6632.
-PMC 4039623.
-PMID 24502540.
-Davey CG, Pujol J, Harrison BJ. Mapping the self in the brain’s default mode network. Neuroimage. 2016 May 15;132:390–397.
-Isoda, M., \& Noritake, A.
-(2015).
-What makes the dorsomedial frontal cortex active during reading the mental states of others?.
-Neural basis of social learning, social deciding, and other-regarding preferences, 51.
-Waytz, A., Zaki, J., \& Mitchell, J. P. (2012). Response of dorsomedial prefrontal cortex predicts altruistic behavior. The Journal of Neuroscience, 32(22), 7646–7650.
-
 Pardo JV, Pardo PJ, Janer KW, Raichle ME (January 1990).
@@
 
-\bibitem{thenewsroom}
-  Aaron Sorkin.
-  \url{https://www.youtube.com/watch?v=2C6h-Yyx9Yk}
-  \textit{The Newsroom}.
-  HBO, 2012.
+\bibitem{dmnmri}
+  \textit{Magnetic Resonance Image of the areas of the brain in the Default Mode Network.}
+  http://www.frontiersin.org/Neurotrauma/10.3389/fneur.2013.00016/full
+
+\bibitem{pccrole}
+  Leech R, Sharp DJ
+  \textit{The role of the posterior cingulate cortex in cognition and disease.}
+  Brain. 137 (Pt 1): 12–32., July 2013.
+
+\bibitem{pccemotion}
+  Maddock, Richard J.; Garrett, Amy S.; Buonocore, Michael H.
+  \textit{Posterior cingulate cortex activation by emotional words: fMRI evidence from a
+    valence decision task.}
+  Human Brain Mapping. 18 (1): 30–41. January 2003.
+
+\bibitem{dmnself}
+  Andrews-Hanna, Jessica R.; Smallwood, Jonathan; Spreng, R. Nathan.
+  \textit{The default network and self-generated thought: component processes,
+    dynamic control, and clinical relevance.}
+  Annals of the New York Academy of Sciences. 1316(1):
+  29–52. doi:10.1111/nyas.12360. ISSN 1749–6632. PMC 4039623. PMID 24502540. 2014-05-01.
+
+\bibitem{mappingself}
+  Davey CG, Pujol J, Harrison BJ.
+  \textit{Mapping the self in the brain’s default mode network.}
+  Neuroimage. 132:390–397., 2016-05-15
+
+\bibitem{autistictheoryofmind}
+  Baron-Cohen, Simon; Leslie, Alan M.; Frith, Uta.
+  \textit{Does the autistic child have a ``theory of mind''?.}
+  Cognition. 21 (1): 37–46. doi:10.1016/0010-0277(85)90022-8. PMID
+  2934210. Pdf. October 1985.
+
+\bibitem{dmpfcothers}
+  Isoda, M., \& Noritake, A.
+  \textit{What makes the dorsomedial frontal cortex active
+    during reading the mental states of others?.}
+  Neural basis of social learning, social deciding, and other-regarding preferences,
+  51. 2015.
+
+\bibitem{dmpfcaltruism}
+  Waytz, A., Zaki, J., \& Mitchell, J. P.
+  \textit{Response of dorsomedial prefrontal cortex predicts altruistic behavior.}
+  The Journal of Neuroscience, 32(22), 7646–7650. 2012.
 

### 2019-09-29 paper-two/paper-two.tex
mostly done anatomy of paper 2

@@
 
+\subsection{Central Executive Network}
+
+The CEN is activated when high-level cognitive tasks or external goal-oriented tasks
+are being performed. Executive functions are cognitive processes involved in
+cognitive control of behavior. The regions of the brain involved in executive
+functions are Dorsolateral Prefrontal Cortex (DL-PFC), Orbito Frontal Cortex (OFC),
+and the Posterior Parietal Cortex (PPC).
+
+\subsubsection{Dorsolateral Prefrontal Cortex}
+
+The DL-PFC is a part of the Prefrontal Cortex found in primates, including
+humans. The DL-PFC is involved in higher cognitive processes including working memory
+(holding different pieces of information, manipulating them, and using them for
+tasks), selective attention, cognitive flexibility (switching between tasks) and
+planning. It also seems to be involved in social cognition and lying. The DL-PFC can
+increase dopamine levels in the brain. \cite{dlpfcmemory,dlpfctasks,dlpfclying}
+
+\subsubsection{Orbito Frontal Cortex}
+
+The OFC is an area found in front of both hemispheres of the brain, just above the
+eye. It is again part of the Prefrontal Cortex and is thought to be involved in
+decision making through emotion and reward. It also receives input from multiple
+sensory modalities and in turn activates the Amygdala and the
+Hypothalamus. \cite{theprefrontalcortex,ofcprimates,theorbitofrontalcortex}
+
+\subsubsection{Posterior Parietal Cortex}
+
+The PPC receives information from the auditory, visual and somatosensory systems and
+in turn activates the Motor Cortex or DL-PFC. \cite{parietallobes}
+
+[ TODO: why? how? expand. ]
+
+[ TODO: image of Approximate location of the OFC on an MRI ]
+
 \subsection{Salience Network}
@@
 
-The SN is an intrinsically connected large-scale network anchored in the Anterior Insula (AI) and dorsal Anterior Cingulate Cortex.
-Both these regions have reached a high degree of specialization in the great apes.
-It is the collection of the regions in the brain that help decide which stimuli deserves our attention.
-It acts as a switch between the internally directed DMN and the externally directed Central Executive Network (involved in high level cognitive functions) (7).
+The SN is an intrinsically connected large-scale network anchored in the Anterior
+Insular Cortex (AIC) and Dorsal Anterior Cingulate Cortex (ACC). Both regions have reached a
+high degree of specialization in the great apes. It is the collection of the regions
+in the brain that help decide which stimuli deserves our attention. It acts as a
+switch between the internally-directed DMN and the externally-directed Central
+Executive Network. \cite{saliencenetwork}
 
-The Salience Network consists of the Anterior Cingulate Cortex and the Insula. It helps to switch between the Default Mode Network and the Central Executive Network
-a) Anterior Cingulate Cortex (ACC): The ACC is the front end of the cingulate cortex and collars around the corpus callosum (the band connecting the two hemispheres of the brain).
-It is the connector between the emotional (limbic system) and the cognitive part of the brain.
-It is involved in functions such as attention allocation, reward anticipation, decision making, morality, impulse control, emotional awareness and registering pain (8–12).
-It also appears to play a role in the regulation of Autonomic functions such as blood pressure and heart rate (13).
-b) Anterior Insular Cortex (AIC): The insula is a part of the cerebral cortex that is located deep within the sulcus (which is the fissure separating the four lobes of the brain).
-The AIC projects itself into the amygdala.
-It is involved in multimodal sensory processing such as audio-visual integration tasks, interoceptive awareness (so its activity is directly related to ones sense of internal body states), empathy and conscious awareness (14).
-It also plays a role in the regulation of autonomic functions such as bodily sensations (as well as where the degree of pain is judged), taste and control of the immune system (15).
-The AIC and ACC together give rise to our interoceptive and conscious self-awareness.
-Central Executive Network:
-This network is activated when a high-level cognitive tasks or external goal-oriented tasks are being performed. Executive functions are cognitive processes involved in cognitive control of behavior.
-The regions of the brain involved in this are Dorsolateral Prefrontal Cortex, Orbito Frontal Cortex and the Posterior Parietal Cortex.
-Dorsolateral Prefrontal Cortex (DL-PFC): It is a part of the pre-frontal cortex and is found in primates and humans.
-It is a part of the Central Executive Network.
-The DL-PFC is involved in higher cognitive processes including working memory (holding different pieces of information and manipulating and using them for tasks), selective attention, cognitive flexibility (switching between tasks) and planning.
-It also seems to be involved in social cognition and lying and increases dopamine levels in the brain.
-Barbey AK, Koenigs M, Grafman J (May 2013). “Dorsolateral prefrontal contributions to human working memory” (PDF). Cortex. 49 (5): 1195–1205.
-Monsell S (2003). “Task switching”. Trends in Cognitive Sciences. 7 (3): 134–140. doi:10.1016/S1364–6613(03)00028–7. PMID 12639695.
-Ito, Ayahito; Abe, Nobuhito; Fujii, Toshikatsu; Hayashi, Akiko; Ueno, Aya; Mugikura, Shunji; Takahashi, Shoki; Mori, Etsuro (2012).
-“The contribution of the dorsolateral prefrontal cortex to the preparation for deception and truth-telling”.
-Brain Research.
-1464: 43–52.
-doi:10.1016/j.brainres.2012.05.004
-Orbito Frontal Cortex (OFC): This is an area found in front of both the hemispheres and just above the eye.
-It is a part of the Pre-Frontal Cortex and it is thought to be involved in the cognitive process of decision making.
-It is thought to represent emotion and reward in decision making.
-It also receives input from multiple sensory modalities and in turn activates the amygdala and the hypothalamus
-Fuster, J.M. The Prefrontal Cortex, (Raven Press, New York, 1997
-Rolls, ET (November 2004).
-“Convergence of sensory systems in the orbitofrontal cortex in primates and brain design for emotion”.
-The Anatomical Record Part A: Discoveries in Molecular, Cellular, and Evolutionary Biology.
-281 (1): 1212–25.
-doi:10.1002/ar.a.20126
-Price, Joseph L. (2006). “Chapter 3: Connections of the orbital cortex”. In Rauch, Scott L.; Zald, David H. (eds.). The Orbitofrontal Cortex. New York: Oxford University Press. p. 45.
-Posterior Parietal Cortex (PPC): It receives information from the Auditory, Visual and Somatosensory systems and in turn activates the motor cortex or DL-PFC.
-Martin, R. E. (n.d.). Let’s Get to Know the Parietal Lobes! [PDF]. Retrieved from \url{http://gablab.mit.edu/downloads/Parietal_Primer.pd}
+\subsubsection{Anterior Cingulate Cortex}
 
-Approximate location of the OFC on an MRI
-(Maybe remove)
-The limbic system: The limbic system is a complex set of structures that lies on both sides of the thalamus, just under the cerebrum.
-It includes the hypothalamus, the hippocampus, the amygdala, and several other nearby areas.
-It appears to be primarily responsible for our emotional life, and has a lot to do with the formation of memories.
+The ACC is the front end of the Cingulate Cortex and collars around the Corpus
+Callosum, the band connecting the two hemispheres of the brain. It is the connector
+between the emotional (Limbic System) and the cognitive (Prefrontal Cortex) part of
+the brain. It is involved in functions such as attention allocation, reward
+anticipation, decision making, morality, impulse control, emotional awareness and
+registering pain. \cite{accstroop,accreward,snmorality,empathypain,acccognitive} It
+also appears to play a role in the regulation of Autonomic functions such as blood
+pressure and heart rate. \cite{accbloodpressure}
 
-The course
+\subsubsection{Anterior Insular Cortex}
+
+The Insula is a part of the Cerebral Cortex located deep within the Sulcus, the
+fissure separating the four lobes of the brain. The AIC physically projects itself
+into the Amygdala. It is involved in multimodal sensory processing such as
+audio-visual integration tasks, interoceptive awareness (so its activity is directly
+related to an individual's sense of internal body states), empathy and conscious
+awareness. \cite{aicemotion}
+
+It also plays a role in the regulation of autonomic functions such as bodily
+sensations (including judgement of the severity of pain), taste, and control of the
+immune system. \cite{aicautonomic}
+
+The AIC and ACC together give rise to our interoceptive and conscious
+self-awareness. [ TODO: reference ]
+
+\subsection{The Limbic System}
+
+The limbic system is a complex set of structures that lies on both sides of the
+Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
+Amygdala, and several other nearby areas. It appears to be primarily responsible for
+our emotional life, and has a lot to do with the formation of memories. [ TODO:
+  reference ]
+
+\section{The Course}
+
 I applied for this ten day course after I quit my job.
@@
 
-Pardo JV, Pardo PJ, Janer KW, Raichle ME (January 1990).
-“The anterior cingulate cortex mediates processing selection in the Stroop attentional conflict paradigm”.
-Proceedings of the National Academy of Sciences of the United States of America.
-87 (1): 256–9.
-doi:10.1073/pnas.87.1.256.
-PMID 2296583.
-Bush G, Vogt BA, Holmes J, Dale AM, Greve D, Jenike MA, Rosen BR (January 2002).
-“Dorsal anterior cingulate cortex: a role in reward-based decision making”.
-Proceedings of the National Academy of Sciences of the United States of America.
-99 (1): 523–8.
-doi:10.1073/pnas.012470999.
-PMC 117593.
-PMID 11756669.
-Sevinc G, Gurvit H, Spreng RN (July 2017).
-“Salience network engagement with the detection of morally laden information”.
-Social Cognitive and Affective Neuroscience.
-12 (7): 1118–1127.
-doi:10.1093/scan/nsx035.
-PMID 28338944.
-Jackson PL, Brunet E, Meltzoff AN, Decety J (2006).
-“Empathy examined through the neural mechanisms involved in imagining how I feel versus how you feel pain”.
-Neuropsychologia.
-44 (5): 752–61.
-doi:10.1016/j.neuropsychologia.2005.07.015.
-PMID 16140345.
-Bush G, Luu P, Posner MI (June 2000). “Cognitive and emotional influences in anterior cingulate cortex”. Trends in Cognitive Sciences. 4 (6): 215–222. doi:10.1016/S1364–6613(00)01483–2. PMID 10827444
-Gianaros PJ, Derbyshire SW, May JC, Siegle GJ, Gamalo MA, Jennings JR. Anterior cingulate activity correlates with blood pressure during stress. Psychophysiology. 2005;42(6):627–35.
-Xiaosi Gu, Patrick R. Hof, Karl J. Friston, Jin Fan. J Comp Neurol. 2013 Oct 15; 521(15): 3371–3388.
-Rolls ET. Functions of the anterior insula in taste, autonomic, and related functions. Brain Cogn. 2016 Dec;110:4–19.
 Lazar SW; Bush George; Gollub RL.; Fricchione, GL.; Khalsa G; Benson H. Functional brain mapping of the relaxation response and meditation.
@@
 
+\bibitem{accstroop}
+  Pardo JV, Pardo PJ, Janer KW, Raichle ME.
+  \textit{The anterior cingulate cortex mediates processing selection in the Stroop
+    attentional conflict paradigm.}
+  Proceedings of the National Academy of Sciences of the United States of
+  America. 87 (1): 256–9. doi:10.1073/pnas.87.1.256. PMID 2296583. January 1990.
+
+\bibitem{accreward}
+  Bush G, Vogt BA, Holmes J, Dale AM, Greve D, Jenike MA, Rosen BR.
+  \textit{Dorsal anterior cingulate cortex: a role in reward-based decision making.}
+  Proceedings of the National Academy of Sciences of the United States of
+  America. 99 (1): 523–8. doi:10.1073/pnas.012470999. PMC 117593. PMID
+  11756669. January 2002.
+
+\bibitem{snmorality}
+  Sevinc G, Gurvit H, Spreng RN.
+  \textit{Salience network engagement with the detection of morally laden
+    information.}
+  Social Cognitive and Affective Neuroscience. 12 (7):
+  1118–1127. doi:10.1093/scan/nsx035. PMID 28338944. July 2017.
+
+\bibitem{empathypain}
+  Jackson PL, Brunet E, Meltzoff AN, Decety J.
+  \textit{Empathy examined through the neural mechanisms involved in imagining how I
+    feel versus how you feel pain.}
+  Neuropsychologia. 44 (5): 752–61. doi:10.1016/j.neuropsychologia.2005.07.015. PMID
+  16140345. 2006.
+
+\bibitem{acccognitive}
+  Bush G, Luu P, Posner MI.
+  \textit{Cognitive and emotional influences in anterior cingulate cortex.}
+  Trends in Cognitive Sciences. 4 (6):
+  215–222. doi:10.1016/S1364–6613(00)01483–2. PMID 10827444. June 2000.
+
+\bibitem{accbloodpressure}
+  Gianaros PJ, Derbyshire SW, May JC, Siegle GJ, Gamalo MA, Jennings JR.
+  \textit{Anterior cingulate activity correlates with blood pressure during
+    stress.}
+  Psychophysiology. 42(6):627–35. 2005.
+
+\bibitem{aicemotion}
+  Xiaosi Gu, Patrick R. Hof, Karl J. Friston, Jin Fan.
+  \textit{Anterior Insular Cortex and Emotional Awareness.}
+  J Comp Neurol. 521(15): 3371–3388. 2013-08-15.
+
+\bibitem{aicautonomic}
+  Rolls ET.
+  \textit{Functions of the anterior insula in taste, autonomic, and related
+    functions.}
+  Brain Cogn. 110:4–19. December 2016.
+
+\bibitem{dlpfcmemory}
+  Barbey AK, Koenigs M, Grafman J.
+  \textit{Dorsolateral prefrontal contributions to human working memory.}
+  Cortex. 49 (5): 1195–1205. May 2013.
+
+\bibitem{dlpfctasks}
+  Monsell S.
+  \textit{Task switching.}
+  Trends in Cognitive Sciences. 7 (3):
+  134–140. doi:10.1016/S1364–6613(03)00028–7. PMID 12639695. 2003.
+
+\bibitem{dlpfclying}
+  Ito, Ayahito; Abe, Nobuhito; Fujii, Toshikatsu; Hayashi, Akiko; Ueno, Aya;
+  Mugikura, Shunji; Takahashi, Shoki; Mori, Etsuro.
+  \textit{The contribution of the dorsolateral prefrontal cortex to the preparation
+    for deception and truth-telling.}
+  Brain Research. 1464: 43–52. doi:10.1016/j.brainres.2012.05.004. 2012.
+
+\bibitem{theprefrontalcortex}
+  Fuster, J.M.
+  \textit{The Prefrontal Cortex}
+  Raven Press, New York, 1997.
+
+\bibitem{ofcprimates}
+  Rolls, ET.
+  \textit{Convergence of sensory systems in the orbitofrontal
+    cortex in primates and brain design for emotion.}
+  The Anatomical Record Part A: Discoveries in Molecular, Cellular, and Evolutionary
+  Biology. 281 (1):1212–25. doi:10.1002/ar.a.20126. November 2004.
+
+\bibitem{theorbitofrontalcortex}
+  Price, Joseph L.
+  \textit{Chapter 3: Connections of the orbital cortex.}
+  Rauch, Scott L.; Zald, David H.
+  The Orbitofrontal Cortex. p. 45.
+  Oxford University Press, New York. 2006.
+
+\bibitem{parietallobes}
+  Martin, R. E.
+  \textit{Let’s Get to Know the Parietal Lobes!}
+  \url{http://gablab.mit.edu/downloads/Parietal_Primer.pdf} [ TODO: 404? ]
+
 \end{thebibliography}

### 2019-09-29 paper-two/paper-two.tex
paper2: all done but conclusion

@@
 
+[ TODO: Summarize DMN? Maybe introduce why it's relevant to meditation? ]
+
 \subsection{Central Executive Network}
@@
 
+[ TODO: image of Approximate location of the OFC on an MRI ]
+
 \subsubsection{Posterior Parietal Cortex}
@@
 
-[ TODO: image of Approximate location of the OFC on an MRI ]
+[ TODO: Summarize CEN? Relevance to meditation? ]
 
@@
   \textbf{Salience: The perceptual quality by which an observable thing stands out
-    relative to its environment}
+    relative to its environment.}
 \end{quote}
@@
 
-I applied for this ten day course after I quit my job.
-I was very nervous as soon as the acceptance came through until the first day.
-I had never done anything like this before and I was very comfortably placed within the idea that meditation with its spiritual connotation was a waste of my time.
-It was for people who had time to waste and I was not one of them.
-I went for my first course in Chennai and once I got there I had to fill all the forms all over again keep all my luggage in a locker and go to my room.
-I had to share the room with one other person and I was so out of my element and knew everyone could see me and tell that I was the odd one out, the one that was not supposed to be there and I was too scared to actually make conversation with anyone before silence was enforced.
-So with my brain filled with thoughts about myself, about the people around me, I started this meditation business.
-Within the first two days, I had made up elaborate stories about my fellow meditators, all superheroes tirelessly working to save humanity together.
-My default mode network was in overdrive.
-For the first 3.5 days you are asked to focus your attention on the area below your nostrils and above your upper lip.
-You start with observing your breath coming in and going out.
-As the teacher mentions in the late evening discourses, the reason for starting with your breath is because this activity is both conscious and unconscious and acts as a bridge.
-At least during the meditation hours, I was trying hard, I wanted to give this an honest shot.I focused on the small patch beneath my nostrils above my upper lip.
-To begin with when you are asked to observe your breath and your mind wanders, that is the default mode network working, as you bring your attention back to your breath, your salience network is activated.
-Slowly at first, with your mind wandering for many many minutes before you realize you are not with your breath but within the 3.5 days, it pretty quickly learns to bring it back and almost does it by habit.
-As I gave in to this activity, tried to focused more, tried to sit still, there were longer periods of awareness on that patch of skin and stranger sensations arising and passing.
-By the third day, it felt like entire ecosystems writhing and flopping and crashing were alive in the area below my nostrils above my upper lip.
-There have been studies that have shown that once you bring your attention back to breath, the neural structures involved in the control of the Autonomic Nervous System and attention start firing more actively (DL-PFC-a part of the CEN, the Hippocampus and the ACC — part of the SN).
-There are also global dampening changes seen in the brain especially within the DMN and the Amygdala (which is involved in the flight-or-fight response/stress response).
-These changes have antagonistically been termed the relaxation response.
-This relaxation response can be thought of as a gateway to altered states of mind.
-(16)
-There has also been research in mice, showing a cluster of nerves called the pre-botzinger complex found in the brain stem of most mammals which fire with every breath taken.
-This breathing pacemaker seems to work for all kinds of breathing such as yawns, sighs and gasps.
-They also appear to play a role in calming and arousal (17).
-It would be interesting to see the activation in these neuron clusters during the first three days of the course even amongst novice meditators.
-After lunch on the fourth day.
-The instructions for Vipassana are given, where you transfer your attention from the patch below your nostril to the top of your head.
-From there, for over two hours, the instructions are given to slowly move your attention till your feet and then repeat.
-The two criteria for observation here are focused attention and open monitoring (observing sensations objectively).
-The Salience Network (ACC and AIC) and the Central Executive Network (DL-PFC, PPC and OFC) which are involved in the control of attention start firing rapidly.
-Usually as the person is engaged in an external task, the Central executive network (Orbito Frontal Cortex and Posterior Parietal Cortex), receiving signals from all the sensory networks (Visual, Audio, Somatosensory etc..) continuously and in response sends signals to the different motor cortices, amygdala, hypothalamus to perform a task, feel an emotion, etc..
-Now imagine a situation where once the Salience Network is activated and in turn activates the Central Executive Network and the task that it is supposed to perform is not external but to observe sensation and do nothing.
-So as you sit there with your eyes closed and not moving and scanning your body, you are doing what the central executive network has always been doing except attempting to not react.
-This is hard at first, with the body involuntarily reacting by jerking and writhing in response to your observation. Your awareness itself is jerky and it is hard to observe sensations consciously.
-There is also the pain.
-This eclipses all other sensations.
-It is difficult trying to be aware of any other sensation with the pain that was emanating from my legs and back.
-The Orbito Prefrontal Cortex (OFC), is involved in assigning emotion to sensation (18) and the ACC plays a vital role in perception of sensation.
-These two parts of the brain are usually functionally connected.
-In the beginning, as I sat trying to observe sensations and I was only aware of the pain, the ACC was firing rapidly and the OFC then assigned these signals to the limbic system (including the amygdala and hippocampus) which then put me in a less than ideal emotional state.
-But as your awareness gets smoother, this functional connectivity between the ACC and the OFC slowly starts changing and as you continue to pay attention to the different sensations and attempt to not assign emotion and observe with a non-judgmental attitude particularly toward unpleasant stimuli, the OFC still fires but the limbic system is quietened, so you react less.
-It is like both your conscious and unconscious awarenesses are slowly aligning until some points when it feels like it is pure awareness moving through your body or some parts of your body.
-In addition the insula (part of the Salience Network) and the somatosensory cortex which are involved in sensory awareness were firing more actively.
-(The somatosensory cortex is located in the gyrus and the parietal lobes: The sensory homunculus which is a part of the somatosensory cortex is a cortical representation of the body based on the degree of sensory innervation)
+I applied for my first 10-day Vipassana course after I quit my job. From the time I
+was accepted into the course until the first day of the course I was very nervous
+about what it might entail. I had never done anything like this (meditation) before
+and I was quite happy with the idea that meditation, with its spiritual connotations
+and religious mumbo-jumbo, was a waste of my time. In my mind, a silent meditation retreat was
+for people who had time to waste --- and I was not one of them. Yet friends had convinced me that I
+should give Vipassana a try and in the interval between workplaces I had
+time to experiment with a course.
 
+I went for my first course in Chennai, my home town. When I arrived, I had to fill
+all the application forms all over again (despite applying online), keep all my
+luggage in a locker, and go to my room. I had to share the room with one other
+person. I felt quite out of my element and I was certain everyone could see in me
+that I was the odd one out, the one that was not supposed to be there. Because I felt
+so out of place, I was too scared to actually make conversation with anyone before
+silence was enforced.
+
+\subsection{Anapana Meditation}
+
+And so it was. With my brain filled with thoughts about myself, about the people
+around me, I started this meditation business. For the first 3.5 days you are asked
+to focus your attention on the area below your nostrils and above your upper
+lip. That's all. Within the first two days, I had made up elaborate stories about my
+fellow meditators, all of them superheroes, tirelessly working to save humanity
+together. My default mode network was in overdrive.
+
+Focusing our attention below the nostrils starts with observing your breath coming in
+and going out. As the teacher mentions in the late evening discourses, the reason for
+starting with your breath is because this the only activity of the body which is both
+conscious and unconscious --- it acts as a bridge to the unconscious mind and the
+involuntary processes of the body.
+
+During the meditation hours I was trying hard. If I was to examine this practice
+empirically, I had to give it an honest shot. I focused on the small patch beneath my
+nostrils above my upper lip. To begin with, while trying to observe your breath and
+your mind wanders, that is the default mode network working. As you realize your mind
+has wandered and bring your attention back to your breath, your salience network is
+activated. This happens slowly at first, with your mind wandering for many, many
+minutes before you realize you are not with your breath. But within the 3.5 days, the
+SN learns quite quickly to bring attention back to the breath. After a few days, this
+refocusing almost happens by habit, almost automatically.
+
+As I gave in to this activity, tried to focus more intently, tried to sit still for
+longer, there were longer periods of awareness on that patch of skin and, with them,
+stranger sensations arising and passing within that physical area. By the third day,
+it felt like entire ecosystems were writhing and flopping and crashing, all of them
+very alive in the area below my nostrils above my upper lip.
+
+There have been studies
+that have shown that once you bring your attention back to breath, the neural
+structures involved in the control of the Autonomic Nervous System and attention
+start firing more actively (in this case, the DL-PFC --- a part of the CEN and the
+Hippocampus and ACC --- both part of the SN). There are also global dampening changes
+seen in the brain, particularly within the DMN and the Amygdala, which is involved in
+the flight-or-fight (stress) response. These changes have been
+termed the ``relaxation response'', as they are antagonistic to the stress
+response. This relaxation response can be thought of as a gateway to altered states
+of mind. \cite{relaxationresponse}
+
+There has also been research in mice, showing a cluster of nerves called the
+pre-Bötzinger Complex (preBötC), found in the brain stem of most mammals, which fires
+with every breath taken. This breathing pacemaker seems to work not only for regular
+conscious and unconscious breathing but for all kinds of breathing --- such as yawns,
+sighs and gasps. The preBötc also appears to play a role in calming and
+arousal. \cite{prebotcgeneration} It would be interesting to see the activation in
+these neuron clusters during the first three days of a Vipassana course, even among
+novice meditators.
+
+\subsection{Vipassana Meditation}
+
+After lunch on the fourth day the instructions for Vipassana are given. During
+Vipassana, you transfer your attention from the patch below your nostril to the top
+of your head. From there, for two hours, the instructions are given to slowly move
+your attention throughout the entire body: ``from the top of the head to the tips of
+the toes'', then repeat. The two criteria for observation of bodily sensations are
+focused attention (making use of the narrow focus practiced during the Anapana
+period) and open monitoring (observing the sensations objectively).
+
+The Salience Network (ACC and AIC) and the Central Executive Network (DL-PFC, OFC,
+and PPC) which are involved in the control of attention start firing rapidly. Usually
+when a person is engaged in an external task, the OFC and PPC are receiving signals
+from all the sensory networks (Visual, Audio, Somatosensory, etc.) and, in response,
+continuously send signals to the different motor cortices, Amygdala, and Hypothalamus to
+respond by performing a task, feeling an emotion, etc.
+
+But Vipassana is an internal task. The Salience Network is activated by focused
+attention, which in turn activates the Central Executive Network. Rather than an
+external task, as the CEN is accustomed to, the conscious instruction is to observe
+bodily sensation --- and do nothing. So as you sit in silence with your eyes closed,
+not moving and scanning your body with your attention, you are doing what the CEN has
+always done except attempting not to react.
+
+This is hard at first, with the body involuntarily reacting by jerking and writhing
+in involuntary response to the act of observation. Initially, your awareness is
+itself jerky and it is hard to observe sensations consciously. There is also the
+pain. This eclipses all other sensations. In my case, it was difficult to maintain
+awareness of any other sensation with the pain that was emanating from my legs and
+back. The OFC is involved in assigning emotion to sensation \cite{ofcemotion} and the
+ACC plays a vital role in perception of sensation. These two parts of the brain are
+usually functionally connected. In the beginning, as I sat trying to observe
+sensations but instead was only aware of the pain, the ACC was firing rapidly and the
+OFC then assigned these signals to the Limbic System (amygdala and
+hippocampus)... which decided to put me in a less than ideal emotional state.
+
+But as your awareness gets smoother and observations get more objective, this
+functional connectivity between the ACC and the OFC slowly starts changing. As you
+continue to pay attention to the different sensations you attempt to avoid assigning
+emotions to them. Without assigning an emotional response, you can observe with a
+non-judgmental attitude, particularly toward unpleasant stimuli. The OFC still fires
+but the Limbic System is quiesced and, in turn, you react less. It is as if both your
+conscious and unconscious awarenesses are slowly aligning until you experience some
+occasions when it feels as if there is nothing but pure, detached awareness moving
+through your body (or some parts of your body). Senses are not dulled, however. In
+addition to a feeling of emotional detachment, the Insula and the Somatosensory
+Cortex, which are involved in sensory awareness, were firing \textbf{more} actively,
+not less.
+
+(The somatosensory cortex is located in the gyrus and the parietal lobes: The sensory
+homunculus which is a part of the somatosensory cortex is a cortical representation
+of the body based on the degree of sensory innervation) [ TODO: Maybe move this to
+  the anatomy section? ]
+
+[ TODO: sensory homunculus diagram ]
 Somatosensory and Motor cortices
-Another strange thing that starts happening is as your mind wandering slows down and you sit with your eyes closed, your sense of your self as a subjective agent in space and time changes, first just very slightly, as you close your eyes and meditate, it feels like you have shifted in space or turned in different directions (during these times the vmPFC and PCC which are a part of the DMN become less active), and then as the meditations get deeper and sometimes, the DMN quietened down completely for a very short duration and in addition to this, the insula is firing more rapidly, it could lead to other alternative networks of consciousness arising.
-It is as if losing the influence of one’s narrative leads to new insights that are usually kept from consciousness.
-Experiencing this even for a few seconds can lead to lasting changes even after this activity wears off.
-These experiences can be quite similar to being on certain kinds of drugs or even in some cases of ecstatic seizures where the abnormal activity of the anterior insula leads to heightened self awareness and feelings of bliss and lack of ambiguity.
-(Fabienne Picard, State of belief, subjective certainty and bliss as a product of cortical dysfunction, Cortex, Volume 49, Issue 9, 2013,
-Pages 2494–2500ISSN 0010 9452, https://doi.org/10.1016/j.cortex.2013.01.006.)
+
+Perhaps the strangest experience during a Vipassana course is that of shifts in your
+sense of self. The ``self'', as a subjective agent in space and time changes, if at
+first only very slightly. As you close your eyes and meditate, it may feel as though
+the body has shifted in space or rotated in different directions. During these times,
+the vmPFC and PCC become less active. As the meditations get deeper the DMN
+occasionally quiets completely (perhaps only for a short duration) while the Insula
+fires more rapidly, leading to alternative networks of consciousness arising. It is
+as if losing the influence of one's own narrative leads to new insights that are
+usually kept from consciousness. It has been found that experiencing this, even for a
+few seconds, can lead to lasting changes long after the activity of meditation itself
+has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
+similar to being on certain halucinogenic drugs or even, in some cases, to ecstatic
+seizures where the abnormal activity of the Anterior Insular Cortex leads to
+heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}
+
+\section{Conclusion}
+
 So to conclude, I believe attending a ten day course literally did change the habit pattern of my mind at least for a while.
@@
 I was really intrigued as I left the center as to what had occurred over those ten days that changed my brain, I very quickly noticed though that as soon as talking was allowed, these changes slipped away really quickly, but with each hour long sitting a small tiny bit sticks just a little more.
-
-\pagebreak
-
-References
-
-Lazar SW; Bush George; Gollub RL.; Fricchione, GL.; Khalsa G; Benson H. Functional brain mapping of the relaxation response and meditation.
-Neuroreport. 11(7):1581–1585, May 15, 2000.
-J.
-Muñoz-Ortiz, E.
-Muñoz-Ortiz, L.
-López-Meraz, L.
-Beltran-Parrazal, C.
-Morgado-Valle.
-The pre-Bötzinger complex: Generation and modulation of respiratory rhythm, Neurología (English Edition), 2018, ISSN 2173–5808,
-https://doi.org/10.1016/j.nrleng.2018.05.006.
-REMPEL‐CLOWER, N. L. (2007), Role of Orbitofrontal Cortex Connections in
-Emotion. Annals of the New York Academy of Sciences, 1121:
-72–86. doi:10.1196/annals.1401.026
 
@@
 
+\bibitem{relaxationresponse}
+  Lazar SW; Bush George; Gollub RL.; Fricchione, GL.; Khalsa G; Benson H.
+  \textit{Functional brain mapping of the relaxation response and meditation.}
+  Neuroreport. 11(7):1581–1585. 2000-05-15.
+
+\bibitem{prebotcgeneration}
+  J. Muñoz-Ortiz, E. Muñoz-Ortiz, L. López-Meraz, L. Beltran-Parrazal,
+  C. Morgado-Valle.
+  \textit{The pre-Bötzinger complex: Generation and modulation of respiratory rhythm}
+  Neurología (English Edition), ISSN 2173–5808,
+  https://doi.org/10.1016/j.nrleng.2018.05.006. 2018.
+
+\bibitem{ofcemotion}
+  Rempel-Clower, N. L.
+  \textit{Role of Orbitofrontal Cortex Connections in Emotion.}
+  Annals of the New York Academy of Sciences,
+  1121:72–86. doi:10.1196/annals.1401.026. 2007.
+
+\bibitem{alteredtraits}
+  Goleman, Daniel; Davidson, Richard J.
+  \textit{Altered Traits}
+  ISBN: 9780399184390. September 2018.
+
+\bibitem{cortexbliss}
+  Picard, Fabienne.
+  \textit{State of belief, subjective certainty and bliss as a product of cortical
+    dysfunction}
+  Cortex, Volume 49, Issue 9, Pages 2494–2500ISSN 0010 9452,
+  https://doi.org/10.1016/j.cortex.2013.01.006. 2013.
+
 \end{thebibliography}

### 2019-09-29 paper-two/paper-two.tex
done first edit of paper2

@@
 
-So to conclude, I believe attending a ten day course literally did change the habit pattern of my mind at least for a while.
-It quietens the Default Mode Network and causes the Salience Network and certain parts of the Central Executive Network to fire more rapidly (the parts that are aware of the sensations while quietening the reacting part of this network).
-This causes changes to a persons behavior at least for a short while and this is usually seen as soon as a person leaves the course.
-But repeated practice is essential for any sustained neuroplasticity since rewriting many years of habit formation requires more than a ten-day course.
-However, the ten day course is well designed to give the minimum observational change neccesary for people interested to keep practicing.
-There have also been studies looking at the structural changes in the brain and functional changes especially to the default mode network and the salience network but these are usually measured after one has clocked over thousands of hours of practice.
-Therefore I am not delving into those right now.
-How is this different from just concentrating on tasks hard?
-It is different because when concentrating on any task, your default mode network is quetened and the salience network and CEN are firing entering you into a state of flow but the reacting part of the brain is also working and working hard.
-There are aspects to this technique that I have not understood or experienced that I have not delved into right now and may do once i know it better.
-I was really intrigued as I left the center as to what had occurred over those ten days that changed my brain, I very quickly noticed though that as soon as talking was allowed, these changes slipped away really quickly, but with each hour long sitting a small tiny bit sticks just a little more.
+I believe attending a 10-day Vipassana course did, in fact, change the habit pattern
+of my mind --- at least for a while.
+
+Vipassana quiets the DMN, causes certain parts of the SN and CEN to fire more
+rapidly. That is, reacting parts of the network and portions of the network which are
+aware of bodily sensation, respectively. This causes changes to a person's behavior
+(in the short term, at least) and these changes can often be seen as soon as someone
+leaves the course on the tenth day.
+
+But repeated practice is essential for any sustained neuroplasticity since rewriting
+many years of habit formation requires more than a ten day course. However, the ten
+day course is well-designed and will give someone experimenting with meditation the
+minimum observational change neccesary to instill an interest, to keep practicing.
+
+There have been studies looking at both the structural changes in the brain and
+functional changes in brain activity, especially to the DMN and the SN. But such
+studies are usually only done on meditators who have clocked thousands of hours of
+practice, which diverges from the purpose of this paper. [ TODO: ref? ]
+
+How is Vipassana different from simply concentrating on a task, as hard as one can?
+When concentrating on any task, your DMN is quieted, the SN and CEN are firing, and you
+enter into a state of flow. But flow or no flow, the reacting part of the brain is
+also working. And it is working hard. The goal of Vipassana is not simply to
+concentrate or two enter a state of flow, but to break old reactionary habit patterns
+altogether.
+
+As I left the Vipassana Center, I was really intrigued as to what exactly had
+occurred over those ten days. What was it that had changed in my brain? I had to
+know. I very quickly noticed, however, that as I left the centre (or even earlier, as
+soon as talking was allowed), those very changes started slipping away. But with each
+hour-long meditation at home, a tiny bit sticks just a little more. And I get closer
+to my answer.
 

### 2019-09-29 paper-two/paper-two.tex
reformatted paper two with physics article template

@@
-\documentclass[twocolumn]{article}
-
-\usepackage{polyglossia}
-\setdefaultlanguage{english}
-\setotherlanguage{sanskrit}
-\usepackage{fontspec}
-\newfontfamily\devanagarifont[Scale=MatchUppercase]{Kohinoor Devanagari}
-
+\documentclass[a4paper, amsfonts, amssymb, amsmath, reprint, showkeys, nofootinbib, twoside]{revtex4-1}
+\usepackage[english]{babel}
+\usepackage[utf8]{inputenc}
+\usepackage[colorinlistoftodos, color=green!40, prependcaption]{todonotes}
 \usepackage{graphicx}
@@
 
-\setlength{\parindent}{1.3em}
-\setlength{\parskip}{0.7em}
+\bibliographystyle{apsrev4-1}
 \setlist{noitemsep}
@@
 
-\begin{titlepage}
-   \vspace*{\stretch{1.0}}
-   \begin{center}
-     \Huge\textbf{Vipassana for Hackers}\\
-     \Huge{Paper Two: The Brain}\\
-     \vspace{5cm}
-     \large\textit{Preethi Govindarajan}\\
-     \large\textit{Version 0.1}\\
-     \large\textit\today\\
-     \vspace{5cm}
-     \large\textit{Contributions by: Steven Deobald}\\
-   \end{center}
-   \vspace*{\stretch{2.0}}
-\end{titlepage}
+\title{%
+  \Huge{Vipassana for Hackers} \\
+  \large{Paper Two: The Brain} \\
+  \large\textit{Version 0.2}
+}
+\author{Preethi Govindarajan}
+\email[Correspondence email address: ]{preethi@deobald.ca}
+\affiliation{Siggu.org}
+\date{\today}
+
+\begin{abstract}
+  \todo{add an abstract} This is the abstract. etc. etc.
+\end{abstract}
+
+\keywords{neuroscience, vipassana, meditation}
+
+\maketitle
+
+\listoftodos
 
@@
     everywhere. The mind is in the whole body. So give attention to the whole
-    body.} \cite{goenkabrain}
+    body.} --- S.N. Goenka \cite{goenkabrain}
 \end{quote}
@@
 \textbf{Disclaimer:} Although my primary field of research is within the field of
-biology, my work is far removed
-from neuroscience. I have tried to simplify the research available so as to better
+biology, my work is far removed from neuroscience. I have tried to simplify the research available so as to better
 understand it. If there are any corrections to make or editing in terms of the
@@
 
-[ TODO: insert switching diagram ]
+\todo{insert switching diagram}
 
@@
 
-[ TODO: MRI of DMN image ] \cite{dmnmri}
+\todo{ insert image of ``MRI of DMN'' }
+\textit{Magnetic Resonance Image of the areas of the brain in the Default Mode Network.}
+http://www.frontiersin.org/Neurotrauma/10.3389/fneur.2013.00016/full
 
@@
 
-[ TODO: explain IPL? ]
+\todo{explain IPL?}
 
@@
 in space and time. It functions as a network in which phenomena such as self
-representations are accessible via the PCC [ TODO: ref ] and are gated into conscious awareness by
+representations are accessible via the PCC \todo{add PCC reference} and are gated into conscious awareness by
 activity in the mPFC, as influenced by changing internal and external demands. The
@@
 
-[ TODO: expand? how? ]
+\todo{expand on DMN involvement in past/future/stories. how does this work?}
 
-[ TODO: Summarize DMN? Maybe introduce why it's relevant to meditation? ]
+\todo{Summarize DMN? Maybe introduce why it's relevant to meditation?}
 
@@
 
-[ TODO: image of Approximate location of the OFC on an MRI ]
+\todo{add image of Approximate location of the OFC on an MRI}
 
@@
 
-[ TODO: why? how? expand. ]
+\todo{why and how does the PCC receive external information and activate the motor cortex/DL-PFC? expand.}
 
-[ TODO: Summarize CEN? Relevance to meditation? ]
+\todo{Summarize CEN? Relevance to meditation?}
 
@@
 The AIC and ACC together give rise to our interoceptive and conscious
-self-awareness. [ TODO: reference ]
+self-awareness. \todo{add reference justifying statement that AIC+ACC create
+  conscious self-awareness}
 
@@
 Amygdala, and several other nearby areas. It appears to be primarily responsible for
-our emotional life, and has a lot to do with the formation of memories. [ TODO:
-  reference ]
+our emotional life, and has a lot to do with the formation of memories. \todo{add
+  reference for limbic system}
 
@@
 homunculus which is a part of the somatosensory cortex is a cortical representation
-of the body based on the degree of sensory innervation) [ TODO: Maybe move this to
-  the anatomy section? ]
+of the body based on the degree of sensory innervation) \todo{Maybe move sensory
+  homunculus diagram and explanation to the anatomy section?}
 
-[ TODO: sensory homunculus diagram ]
+\todo{add sensory homunculus diagram}
 Somatosensory and Motor cortices
@@
 studies are usually only done on meditators who have clocked thousands of hours of
-practice, which diverges from the purpose of this paper. [ TODO: ref? ]
+practice, which diverges from the purpose of this paper. \todo{add references with
+  evidence of structural changes in the brain after meditation}
 
@@
 
-\pagebreak
+\section{Acknowledgements}
 
-\begin{thebibliography}{9}
-\raggedright
+Thank you to Steven Deobald for reviews, edits, and corrections.
+
+\begin{thebibliography}{99}
+
+\section*{References}
 
@@
   ISBN: 978–0–12–397316–0.
-
-\bibitem{dmnmri}
-  \textit{Magnetic Resonance Image of the areas of the brain in the Default Mode Network.}
-  http://www.frontiersin.org/Neurotrauma/10.3389/fneur.2013.00016/full
 
@@
   \textit{Let’s Get to Know the Parietal Lobes!}
-  \url{http://gablab.mit.edu/downloads/Parietal_Primer.pdf} [ TODO: 404? ]
+  \url{http://gablab.mit.edu/downloads/Parietal_Primer.pdf}
+  \todo{parietal lobes article returns a 404. fix.}
 
@@
 
+% \appendix*
+% \input{sections/appendix1.tex}
+
 \end{document}

### 2019-09-29 paper-two/paper-two.tex
Update paper-2
1)added first draft of abstract
2)added tibetean monks reference
3)spelling error

@@
 \begin{abstract}
-  \todo{add an abstract} This is the abstract. etc. etc.
+  Two years ago, I took my first Vipassana course and immediately began researching 
+the effects of meditation on the brain. Written below is a summary of the research 
+available as well as speculations on what goes on in the brain during a 10-day silent 
+Vipassana meditation retreat.
 \end{abstract}
@@
 
-I took my first Vipassana course in 2017. Vipassana meditation was so unlike anything
-I had ever experienced before, after the course I was left extremely curious
-about what exactly had happened to me during those ten days. For months afterward, I
-spent my mornings and evenings wading through the research in the field of
-meditation. I was specifically focused on white papers dealing with the effects of a
-10-day Vipassana course on the brains of participants. The research in this area
-is limited. The quality research which does exist usually uses a sample of highly
-experienced meditators rather than beginners and/or self-reports rather than
+I took my first Vipassana course (as taught by S.N. Goenka) in 2017. 
+Vipassana meditation was so unlike anything I had ever experienced before, 
+after the course I was left extremely curious about what exactly had happened 
+to me during those ten days. For months afterward, I spent my mornings and 
+evenings wading through the research in the field of meditation. I was specifically 
+focused on white papers dealing with the effects of a 10-day Vipassana course on 
+the brains of participants. The research in this area is limited. 
+The quality research which does exist usually uses a sample of highly 
+experienced meditators \cite{tibeteanmonks} rather than beginners and/or self-reports rather than
 objective measures.
@@
 brain, an evolutionarily primitive area related to memory and emotion, and
-structures in the cortex, an evolutionarily recent part of the brain containting the
+structures in the cortex, an evolutionarily recent part of the brain containing the
 executive and higher order functions. \cite{defaultnetworkanatomy} These functionally
@@
 
+\bibitem{tibeteanmonks}
+  Alex Hankey
+  \url:https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1697747/
+  \textit{Studies of Advanced Stages of Meditation in the Tibetan 
+           Buddhist and Vedic Traditions. I: A Comparison of General Changes}
+   Evid Based Complement Alternat Med. 2006 Dec; 3(4): 513–521.
+
 \bibitem{goenkabrain}

### 2019-09-29 paper-two/paper-two.tex
cleaning up network images for paper 2

--- 2019-09-29 paper-two/paper-two.tex
@@
 
-\todo{insert switching diagram}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=7.5cm]{images/top-dmn-sn-cen-switching.jpg}
+  \caption{The Salience Network consists of the Anterior Cingulate Cortex and the
+    Anterior Insular Cortex. It helps switch between the Default Mode Network and the
+  Central Executive Network.}
+  \label{fig:top-dmn-sn-cen-switching}
+\end{figure}
 
@@
 imagining the future.
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/top-dmn.png}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/side-dmn.png}
+  \end{subfigure}
+  \caption{Highlighted: areas of the brain firing when the Default Mode Network is active.}
+  \label{fig:dmn}
+\end{figure}
 
@@
 stimuli. \cite{pccemotion}
-
-\todo{ insert image of ``MRI of DMN'' }
-\textit{Magnetic Resonance Image of the areas of the brain in the Default Mode Network.}
-http://www.frontiersin.org/Neurotrauma/10.3389/fneur.2013.00016/full
 
@@
 
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=7.5cm]{images/top-sn.jpg}
+  \caption{The Salience Network consists of the Anterior Cingulate Cortex and the
+    Anterior Insular Cortex. It helps switch between the Default Mode Network and the
+  Central Executive Network.}
+  \label{fig:top-sn}
+\end{figure}
+
 \subsubsection{Dorsolateral Prefrontal Cortex}
@@
 Hypothalamus. \cite{theprefrontalcortex,ofcprimates,theorbitofrontalcortex}
-
-\todo{add image of Approximate location of the OFC on an MRI}
 
@@
 
+``Brain\_human\_normal\_inferior\_view'' By Patrick J. Lynch, medical illustrator, CC BY
+2.5, https://commons.wikimedia.org/w/index.php?curid=1496648
+
+``Brain\_human\_lateral\_view.svg'' By Patrick J. Lynch, medical illustrator, CC BY
+2.5, https://commons.wikimedia.org/w/index.php?curid=1496651
+
 \begin{thebibliography}{99}

### 2019-09-29 paper-two/paper-two.tex

--- 2019-09-29 paper-two/paper-two.tex
@@
 imagining the future. 
+This network is mostly observed through changes in blood flow to different parts of 
+the brain measured using fMRI (functional Magnetic Resonance Imaging) or 
+PET (Positron Emission Tomography){defaultnetworkadaptive}
 
@@
 
-The regions involved in an individual's conceptual ``self'' are the Posterior
+The regions involved in an individual's conceptual` `self'' are the Posterior
 Cingulate Cortex (PCC), Ventromedial Prefrontal Cortex (vmPFC), and the Inferior
-Parietal Lobule (IPL). \cite{defaultnetworkadaptive}
+Parietal Lobule (IPL).That is to say, these regions are preferentially activated when 
+volunteers were asked to reflect on personal preferences, beliefs, abilities, 
+anything relevant to the self. \cite{defaultnetworkadaptive}
 
-Cerebral blood flow and metabolic rate is 40\%
-higher in the PCC compared to the average across the brain, making it one of the most
-active regions of the brain. \cite{pccrole}
+Cerebral blood flow and metabolic rate is 40\% higher than average in the PCC,
+making it one of the most active regions of the brain. \cite{pccrole}
 
-The PCC has been thought to mediate interactions between emotion and memory. Under
-fMRI, it consistently lights up when a person recollects something from their
-life. The strength of PCC brain activity in such an experiment varies depending on
-the emotional importance of the memory recalled. It is also activated by emotional
-stimuli, both positive and negative, and acts as an interface between the external
-world and the individual by gauging the importance of different
-stimuli. \cite{pccemotion}
+The PCC has been thought to mediate interactions between emotion and memory. 
+Under an fMRI, it consistently lights up when a person recollects something from their
+life. The strength of this activity varies depending on the emotional importance of 
+the memory recalled. It is also activated by emotional stimuli, 
+both positive and negative, and acts as an interface between the external world 
+and the individual by gauging the importance of different stimuli. \cite{pccemotion}
 
@@
 Cortex (dmPFC), which assists in the creation of the conceptual self through an
-individual's consideration of ``others'', to construct the Medial Prefrontal Cortex
-(mPFC).
+individual's consideration of ``others''.
 
-\todo{explain IPL?}
+The Inferior Parietal lobe (IPL) is situated in the back of the brain and it is 
+involved in the perception of facial stimuli.\cite{IPL}
 
@@
 in space and time. It functions as a network in which phenomena such as self
-representations are accessible via the PCC \todo{add PCC reference} and are gated into conscious awareness by
-activity in the mPFC, as influenced by changing internal and external demands. The
-PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
+representations are accessible via the PCC \cite{mappingself} and are gated into 
+conscious awareness by activity in the mPFC, as influenced by changing internal 
+and external demands. The PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
 created by a large scale integration of DMN activity. \cite{mappingself}
@@
 
+\bibitem{IPL}
+ Radua J, Phillips ML, Russell T, Lawrence N, Marshall N, Kalidindi S, El-Hage W, McDonald C, 
+ Giampietro V, Brammer MJ, David AS, Surguladze SA.
+ \textit{Neural response to specific components of fearful faces in healthy 
+  and schizophrenic adults.}
+  Neuroimage. 2010 Jan 1;49(1):939-46. doi: 10.1016/j.neuroimage.2009.08.030. 
+  Epub 2009 Aug 20.
+  \url:https://www.ncbi.nlm.nih.gov/pubmed/19699306
+
 \bibitem{mappingself}

### 2019-09-29 paper-two/paper-two.tex

@@
 PET (Positron Emission Tomography){defaultnetworkadaptive}
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/top-dmn.png}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/side-dmn.png}
+  \end{subfigure}
+  \caption{Highlighted: areas of the brain firing when the Default Mode Network is active.}
+  \label{fig:dmn}
+\end{figure}
 
@@
 and the individual by gauging the importance of different stimuli. \cite{pccemotion}
-
-\todo{ insert image of ``MRI of DMN'' }
-\textit{Magnetic Resonance Image of the areas of the brain in the Default Mode Network.}
-http://www.frontiersin.org/Neurotrauma/10.3389/fneur.2013.00016/full
 

### 2019-09-29 paper-two/paper-two.tex
Added information of theory of mind, completed to do for memory

@@
 The Inferior Parietal lobe (IPL) is situated in the back of the brain and it is 
-involved in the perception of facial stimuli.\cite{IPL}
+involved in the perception of facial stimuli. \cite{ipl}
 
@@
 differentiate the two. \cite{autistictheoryofmind}
+There is a rich body of work pointing to this network (especially the dmPFC) being engaged
+when individuals reflect upon, evaluate, appraise their own thoughts, feelings and desires 
+as well as those of others. \cite{theoryofmind}
 
-In addition to The Theory of Mind, the dmPFC is involved in empathy, moral reasoning,
+In addition to this, the dmPFC is involved in empathy, moral reasoning,
 and altruistic behavior. \cite{defaultnetworkadaptive,dmpfcothers,dmpfcaltruism}
@@
 
-The DMN is also involved in remembering the past, imagining the future, and story
+The critical role of the medial temporal lobe (MTL) which is a part of the DMN 
+in long term memory is well documented. \cite{mtl}
+This network has also been activated during autobiographical memory retrieval, 
+contextual association, and semantic knowledge. \cite{defaultnetworkadaptive}
+
+It has been hypothesized that the adaptive role of memory retrieval is to facilitate 
+construction of episodes to prepare for immediate and distant future scenarios.
+So the DMN is involved in remembering the past, imagining the future, and story
 comprehension. \cite{defaultnetworkadaptive}
-
-\todo{expand on DMN involvement in past/future/stories. how does this work?}
 
@@
 
-\bibitem{IPL}
+\bibitem{ipl}
  Radua J, Phillips ML, Russell T, Lawrence N, Marshall N, Kalidindi S, El-Hage W, McDonald C, 
@@
 
+\bibitem{theoryofmind}
+  Frank Van Overwalle
+ \textit{A dissociation between social mentalizing and general reasoning}
+  NeuroImage. Volume 54, Issue 2, 15 January 2011, Pages 1589-1599
+  \url:https://www.sciencedirect.com/science/article/abs/pii/S1053811910012243
+
 \bibitem{dmpfcothers}
@@
   The Journal of Neuroscience, 32(22), 7646–7650. 2012.
+
+\bibitem{mtl}
+ Scoville WB, Milner B
+ \textit{Loss of recent memory after bilateral hippocampal lesions.}
+  J Neurol Neurosurg Psych,  1957;20:11–21.
 

### 2019-09-29 paper-two/paper-two.tex
Tried to add a summary for DMN.

@@
 It has been hypothesized that the adaptive role of memory retrieval is to facilitate 
-construction of episodes to prepare for immediate and distant future scenarios.
-So the DMN is involved in remembering the past, imagining the future, and story
+construction of episodes to prepare for immediate and distant future scenarios. \cite{defaultnetworkadaptive}
+The DMN is therefore involved in remembering the past, imagining the future, and story
 comprehension. \cite{defaultnetworkadaptive}
+
+Now, that we know a little bit about this default mode network, how does it pertain
+to meditative practices. 
+The DMN is the network that lends itself to excessive rumination. 
+So when I sit down to meditate and my thoughts wander away (and how they wander!),
+it is the DMN that is being activated.
 

### 2019-09-29 paper-two/paper-two.tex
summary of DMN with reference

--- 2019-09-29 paper-two/paper-two.tex
@@
 It has been hypothesized that the adaptive role of memory retrieval is to facilitate 
-construction of episodes to prepare for immediate and distant future scenarios.
-So the DMN is involved in remembering the past, imagining the future, and story
+construction of episodes to prepare for immediate and distant future scenarios. \cite{defaultnetworkadaptive}
+The DMN is therefore involved in remembering the past, imagining the future, and story
 comprehension. \cite{defaultnetworkadaptive}
@@
 \todo{Summarize DMN? Maybe introduce why it's relevant to meditation?}
+Therefore the DMN that plays a role in constructing a sense of self, in memory, 
+in thinking about other also lends itself to excessive rumination and anxiety. \citr{dmndepression} 
+When I sit down to meditate and my thoughts wander away (and how they wander!),
+it is the DMN that is being activated.
 
@@
 
+\bibitem{dmndepression}
+  J. Paul Hamilton, Madison Farmer, Phoebe Fogelman, and Ian H. Gotlib.
+  \textit{Depressive Rumination, the Default-Mode Network, 
+   and the Dark Matter of Clinical Neuroscience}
+   Biol Psychiatry. 2015 Aug 15; 78(4): 224–230. doi: 10.1016/j.biopsych.2015.02.020
+
 \bibitem{accstroop}

### 2019-09-29 paper-two/paper-two.tex
included information about the PPC and reference

@@
 The CEN is activated when high-level cognitive tasks or external goal-oriented tasks
-are being performed. Executive functions are cognitive processes involved in
-cognitive control of behavior. The regions of the brain involved in executive
-functions are Dorsolateral Prefrontal Cortex (DL-PFC), Orbito Frontal Cortex (OFC),
+are being performed. These tasks or executive functions are cognitive processes involved in
+cognitive control of behavior. This depends on three types of brain functions : working memory,
+mental flexibility and self-control. 
+The regions of the brain involved in these executive functions are 
+Dorsolateral Prefrontal Cortex (DL-PFC), Orbito Frontal Cortex (OFC),
 and the Posterior Parietal Cortex (PPC).
@@
 tasks), selective attention, cognitive flexibility (switching between tasks) and
-planning. It also seems to be involved in social cognition and lying. The DL-PFC can
-increase dopamine levels in the brain. \cite{dlpfcmemory,dlpfctasks,dlpfclying}
+planning. It also seems to be involved in social cognition and lying. The DL-PFC has
+also been seen to increase dopamine levels in the brain. \cite{dlpfcmemory,dlpfctasks,dlpfclying}
 
@@
 The OFC is an area found in front of both hemispheres of the brain, just above the
-eye. It is again part of the Prefrontal Cortex and is thought to be involved in
+eyes. It is again a part of the Prefrontal Cortex and is thought to be involved in
 decision making through emotion and reward. It also receives input from multiple
@@
 \subsubsection{Posterior Parietal Cortex}
-
-The PPC receives information from the auditory, visual and somatosensory systems and
-in turn activates the Motor Cortex or DL-PFC. \cite{parietallobes}
-
-\todo{why and how does the PCC receive external information and activate the motor cortex/DL-PFC? expand.}
+The PPC is a region of the parietal cortex that is posterior to the primary somatosensory 
+cortex.
+The posterior parietal cortex receives input from auditory, visual, and somatosensory systems
+as well as a variety of other brain regions, and it integrates this input to facilitate 
+the execution of functions that require diverse information (also known as higher-order functions). 
+It in turn activates the DL-PFC and motor cortex and is involved in the performance of 
+attention related tasks as well as higher-order motor tasks such as grasping and catching. \cite{parietallobesa,parietallobes}
 
@@
 
+\bibitem{parietallobesa}
+  Caspers S, Amunts K, Zilles K. 
+  \textit {Posterior Parietal Cortex: Multimodal Association Cortex.}
+   The Human Nervous System. 3rd ed. New York: Elsevier; 2012.
+
 \bibitem{parietallobes}

### 2019-09-29 paper-two/paper-two.tex
2 comment correction made

@@
 
-The critical role of the medial temporal lobe (MTL) which is a part of the DMN 
+The critical role of the Medial Temporal Lobe (MTL) which is a part of the DMN 
 in long term memory is well documented. \cite{mtl}
@@
 Therefore the DMN that plays a role in constructing a sense of self, in memory, 
-in thinking about other also lends itself to excessive rumination and anxiety. \citr{dmndepression} 
+in thinking about other also lends itself to excessive rumination and anxiety. \cite{dmndepression} 
 When I sit down to meditate and my thoughts wander away (and how they wander!),
@@
 
+So, the Central Executive Network is involved in
+
+cultivate their abilities to
+monitor cognitive processes related to attention and distraction, making them well suited to
+report on these mental events.
 \todo{Summarize CEN? Relevance to meditation?}

### 2019-09-29 paper-two/paper-two.tex
added summary for CEN.

@@
 
-So, the Central Executive Network is involved in
+So, the Central Executive Network is involved in goal-directed behavior and processes 
+related to attention. These goals are usually external and attention is also directed externally(whether 
+voluntary or not). When you meditate, you direct your attention voluntarily
+inside, to the same sensory cues that are directing the PPC.
+When you meditate, you also cultivate your cognitive processes related to attention.
+When your wanders, you notice it and bring it back.
 
-cultivate their abilities to
-monitor cognitive processes related to attention and distraction, making them well suited to
-report on these mental events.
 \todo{Summarize CEN? Relevance to meditation?}

### 2019-09-30 paper-two/paper-two.tex
added information about limbic system and ssc.with references

@@
 structures in the cortex, an evolutionarily recent part of the brain containing the
-executive and higher order functions. \cite{defaultnetworkanatomy} These functionally
-connected regions are involved in the neurological basis
+executive and higher order functions. \cite{defaultnetworkanatomy}
+
+These functionally connected regions are involved in the neurological basis
 of the self, considering the mental states of others, remembering the past, and
 imagining the future.
+
 This network is mostly observed through changes in blood flow to different parts of 
@@
 differentiate the two. \cite{autistictheoryofmind}
-There is a rich body of work pointing to this network (especially the dmPFC) being engaged
-when individuals reflect upon, evaluate, appraise their own thoughts, feelings and desires 
-as well as those of others. \cite{theoryofmind}
+
+There is a rich body of work pointing to this network (especially the dmPFC subsystem) 
+being engaged when individuals reflect upon, evaluate, appraise their own thoughts, 
+feelings and desires as well as those of others. \cite{theoryofmind}
 
@@
 \subsubsection{Autobiographical memory and future simulations}
+
+The Medial Temporal Subsytem is primarily activated during these processes
+This subsystem consists of the Medial Temporal Lobe and parts of the limbic system
+especially the Hippocampus.
+
+The limbic system is a complex set of structures that lies on both sides of the
+Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
+Amygdala, and several other nearby areas. It appears to be primarily responsible for
+our emotional life, and has a lot to do with the formation of memories. \cite{limbic}
 
@@
 
+\subsection {Sidenote: Somatosensory Cortex}
+The primary somatosensory cortex is located in a ridge of cortex found within the parietal lobe.
+It is responsible for processing somatic sensations. These sensations arise from receptors 
+positioned throughout the body. They are responsible for detecting touch, for proprioception 
+(i.e. the position of the body in space), nociception (i.e. pain), and temperature. \cite{somato}
+When such receptors detect one of these sensations,  the information is sent to the thalamus and 
+then to the primary somatosensory cortex.
+The sensory homunculus which is a part of the somatosensory cortex is a cortical representation
+of the body based on the degree of sensory innervation. \cite{sensoryhom}
+
+\todo {include images of the homunculus here}
+
+\todo{Summarize CEN? Relevance to meditation?} 
 So, the Central Executive Network is involved in goal-directed behavior and processes 
@@
 inside, to the same sensory cues that are directing the PPC.
-When you meditate, you also cultivate your cognitive processes related to attention.
-When your wanders, you notice it and bring it back.
-
-\todo{Summarize CEN? Relevance to meditation?}
 
@@
 The AIC and ACC together give rise to our interoceptive and conscious
-self-awareness. \todo{add reference justifying statement that AIC+ACC create
-  conscious self-awareness}
+self-awareness.  \cite{selfaware}
 
-\subsection{The Limbic System}
-
-The limbic system is a complex set of structures that lies on both sides of the
-Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
-Amygdala, and several other nearby areas. It appears to be primarily responsible for
-our emotional life, and has a lot to do with the formation of memories. \todo{add
-  reference for limbic system}
+The salience network is directly involved in switching between the DMN and CN. 
+When you meditate, you cultivate the cognitive processes related to attention.
+When your mind wanders, you notice it and bring it back. 
+The salience network should be directly involved in this process of assigning attention. 
+Meditation should strengthen the salience network and its ability to switch from
+the default.
 
@@
 to focus your attention on the area below your nostrils and above your upper
-lip. That's all. Within the first two days, I had made up elaborate stories about my
+lip. That's all, these were really simple instructions, but so hard to keep up with.
+Within the first two days, I had made up elaborate stories about my
 fellow meditators, all of them superheroes, tirelessly working to save humanity
@@
 
-(The somatosensory cortex is located in the gyrus and the parietal lobes: The sensory
-homunculus which is a part of the somatosensory cortex is a cortical representation
-of the body based on the degree of sensory innervation) \todo{Maybe move sensory
+(The somatosensory cortex is located in the gyrus and the parietal lobes:) \todo{Maybe move sensory
   homunculus diagram and explanation to the anatomy section?}
@@
 
+\bibitem{limbic}
+  Catani M, Dell'Acqua F, Thiebaut De Schotten M
+  \textit {A revised limbic system model for memory, emotion and behaviour}
+  Neuroscience and Biobehavioral Reviews. 37 (8): 1724–37. 
+  doi:10.1016/j.neubiorev.2013.07.001. PMID 23850593.
+
 \bibitem{dmpfcothers}
@@
 
+\bibitem{selfaware}
+  Anil K. Seth, Keisuke Suzuki, Hugo D. Critchley
+  \textit {An Interoceptive Predictive Coding Model of Conscious Presence}
+  Front Psychol. 2011; 2: 395.
+
 \bibitem{dlpfcmemory}
@@
   Oxford University Press, New York. 2006.
+
+\bibitem{somato}
+  Viaene A.N. 
+  \textit {synaptic properties of thalamic input to layers 2/3 and 4 of 
+              primary somatosensory and auditory cortices}
+  Journal of Neurophysiology. 105 (1): 279–292. 
+  doi:10.1152/jn.00747.2010. PMC 3023380.
+
+\bibitem{sensoryhom}
+  Marieb E, Hoehn K. 
+  Human Anatomy and Physiology. 
+  7th Ed. 2007. Pearson Benjamin Cummings: San Francisco.
 

### 2019-09-30 paper-two/paper-two.tex

@@
 
+The following are my observations and speculations about what could be happening 
+in the brain during a 10 course.
+
 I went for my first course in Chennai, my home town. When I arrived, I had to fill
@@
 
-There have been studies
-that have shown that once you bring your attention back to breath, the neural
-structures involved in the control of the Autonomic Nervous System and attention
-start firing more actively (in this case, the DL-PFC --- a part of the CEN and the
-Hippocampus and ACC --- both part of the SN). There are also global dampening changes
-seen in the brain, particularly within the DMN and the Amygdala, which is involved in
+There have been studies that have shown that once you bring your 
+attention back to breath, the neural structures involved in the control
+of the Autonomic Nervous System and attention start firing more actively
+(in this case, the DL-PFC --- a part of the CEN and the Hippocampus a part of the DMN 
+and ACC --- a part of the SN). There are also global dampening changes seen in the brain, 
+particularly within the DMN and the Amygdala, which is involved in
 the flight-or-fight (stress) response. These changes have been
@@
 usually functionally connected. In the beginning, as I sat trying to observe
-sensations but instead was only aware of the pain, the ACC was firing rapidly and the
-OFC then assigned these signals to the Limbic System (amygdala and
-hippocampus)... which decided to put me in a less than ideal emotional state.
+sensations and was only aware of the pain, the ACC was firing rapidly and the
+OFC then assigned these signals to the Limbic System which 
+put me in a less than ideal emotional state.
 
@@
 functional connectivity between the ACC and the OFC slowly starts changing. As you
-continue to pay attention to the different sensations you attempt to avoid assigning
+continue to pay attention to the different sensations and you attempt to avoid assigning
 emotions to them. Without assigning an emotional response, you can observe with a
@@
 conscious and unconscious awarenesses are slowly aligning until you experience some
-occasions when it feels as if there is nothing but pure, detached awareness moving
+occasions when it feels as if there is nothing but detached awareness moving
 through your body (or some parts of your body). Senses are not dulled, however. In
@@
 not less.
-
-(The somatosensory cortex is located in the gyrus and the parietal lobes:) \todo{Maybe move sensory
-  homunculus diagram and explanation to the anatomy section?}
-
-\todo{add sensory homunculus diagram}
-Somatosensory and Motor cortices
 
@@
 has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
-similar to being on certain halucinogenic drugs or even, in some cases, to ecstatic
+similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
 seizures where the abnormal activity of the Anterior Insular Cortex leads to
@@
 
-Vipassana quiets the DMN, causes certain parts of the SN and CEN to fire more
-rapidly. That is, reacting parts of the network and portions of the network which are
-aware of bodily sensation, respectively. This causes changes to a person's behavior
-(in the short term, at least) and these changes can often be seen as soon as someone
-leaves the course on the tenth day.
+Vipassana causes global dampening changes in the DMN, causes certain parts of 
+the SN and CEN to fire more rapidly. That is, reacting parts of the network and 
+portions of the network which are aware of bodily sensation, respectively. 
+This changes a person's behavior (in the short term, at least) and these 
+changes can often be seen as soon as someone leaves the course on the tenth day.
 
@@
 day course is well-designed and will give someone experimenting with meditation the
-minimum observational change neccesary to instill an interest, to keep practicing.
+minimum observational change necessary to instill an interest, to keep practicing.
 
@@
 studies are usually only done on meditators who have clocked thousands of hours of
-practice, which diverges from the purpose of this paper. \todo{add references with
-  evidence of structural changes in the brain after meditation}
+practice, which diverges from the purpose of this paper. \cite {tibeteanmonks
+, experiencemed, experiencemeda}
 
@@
 also working. And it is working hard. The goal of Vipassana is not simply to
-concentrate or two enter a state of flow, but to break old reactionary habit patterns
+concentrate or to enter a state of flow, but to break old reactionary habit patterns
 altogether.
@@
    Cortex, Volume 49, Issue 9, Pages 2494–2500ISSN 0010 9452,
-  https://doi.org/10.1016/j.cortex.2013.01.006. 2013.
+   \url: https://doi.org/10.1016/j.cortex.2013.01.006. 2013.
+
+\bibitem{experiencemed}
+   Brewer JA, Worhunsky PD, Gray JR, Tang YY, Weber J, Kober H.
+   \textit {Meditation experience is associated with differences in default mode 
+          network activity and connectivity.}
+     Proc Natl Acad Sci U S A. 2011 Dec 13;108(50):20254-9. 
+     doi: 10.1073/pnas.1112029108.
+
+\bibitem{experiencemeda}
+  Zoran Josipovic, Ilan Dinstein, Jochen Weber, David J. Heeger
+  \textit {Influence of meditation on anti-correlated networks in the brain}
+   Front. Hum. Neurosci., 03 January 2012,
+   https://doi.org/10.3389/fnhum.2011.00183
 

### 2019-10-01 paper-two/paper-two.tex
editing round 2: finished DMN

@@
 \begin{abstract}
-  Two years ago, I took my first Vipassana course and immediately began researching 
-the effects of meditation on the brain. Written below is a summary of the research 
-available as well as speculations on what goes on in the brain during a 10-day silent 
-Vipassana meditation retreat.
+  Two years ago, immediately following my first Vipassana course, I began researching
+the effects of meditation on the brain. This paper is a summary of the research
+available on the topic and speculations, based on my experience, as to what is
+happening in the brain during a 10-day silent Vipassana meditation retreat.
 \end{abstract}
@@
 I took my first Vipassana course (as taught by S.N. Goenka) in 2017.
-Vipassana meditation was so unlike anything I had ever experienced before, 
-after the course I was left extremely curious about what exactly had happened 
-to me during those ten days. For months afterward, I spent my mornings and 
+Vipassana meditation was so unlike anything I had ever experienced before
+I was left extremely curious about what exactly had happened
+to me during those ten days. For months afterward I spent my mornings and
 evenings wading through the research in the field of meditation. I was specifically
@@
 The quality research which does exist usually uses a sample of highly
-experienced meditators \cite{tibeteanmonks} rather than beginners and/or self-reports rather than
+experienced meditators \cite{tibetanmonks} rather than beginners and/or self-reports rather than
 objective measures.
@@
 This paper does not contradict Mr. Goenka's sentiment, but instead acts as a starting
-point for readers who view the function of the brain as central to the activity of
-the mind.
+point for curious readers who, whether they have taken a Vipassana course or not,
+view the function of the brain as central to the activity of the mind.
 
@@
 understand it. If there are any corrections to make or editing in terms of the
-content, please feel free to get in touch with me.
+content, please feel free to contact me.
 
@@
 scientific literature as correlated to meditative practices. Recent neuroscience divides the
-brain into the reified geography of the brain, its anatomy, and the abstract concepts
-governed by brain activity, its function. Between concrete physiology and abstract
+brain into the reified geography of the brain (anatomy) and the abstract concepts
+governed by brain activity (function). Between concrete physiology and abstract
 functional outcomes exist networks of cooperative structures which correspond to general
@@
 tasks. As the name suggests, the brain is in its ``default
-mode''. \cite{defaultnetworkadaptive} The DMN is anticorrelated to the Central Executive
+mode'' of operation. \cite{defaultnetworkadaptive} The DMN is anticorrelated to the Central Executive
 Network (CEN), which is active during externally-directed, high-level cognitive
@@
 governed by the Salience Network (SN), the collection of regions in the brain which
-help decide which stimuli deserve our attention. The SN acts as a switch between the
+help decide which stimuli deserve our attention. In this way, the SN acts as a switch between the
 internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}
+
+\todo{Introduce both the Limbic System and Somatosensory Cortex here, at a very high level.}
 
@@
 the brain measured using fMRI (functional Magnetic Resonance Imaging) or
-PET (Positron Emission Tomography){defaultnetworkadaptive}
+PET (Positron Emission Tomography). \cite{defaultnetworkadaptive}
 
@@
 
-The regions involved in an individual's conceptual` `self'' are the Posterior
+The regions involved in an individual's conceptual ``self'' are the Posterior
 Cingulate Cortex (PCC), Ventromedial Prefrontal Cortex (vmPFC), and the Inferior
-Parietal Lobule (IPL).That is to say, these regions are preferentially activated when 
-volunteers were asked to reflect on personal preferences, beliefs, abilities, 
-anything relevant to the self. \cite{defaultnetworkadaptive}
-
-Cerebral blood flow and metabolic rate is 40\% higher than average in the PCC,
+Parietal Lobule (IPL). Observationally, these regions are preferentially activated when
+volunteers were asked to reflect on personal preferences, beliefs, abilities, or
+any other qualities relevant to the self. \cite{defaultnetworkadaptive} Cerebral
+blood flow and metabolic rate are 40\% higher than average in the PCC,
 making it one of the most active regions of the brain. \cite{pccrole}
@@
 The PCC has been thought to mediate interactions between emotion and memory.
-Under an fMRI, it consistently lights up when a person recollects something from their
+Under an fMRI, it consistently lights up when a person recollects an event from their
 life. The strength of this activity varies depending on the emotional importance of
 the memory recalled. It is also activated by emotional stimuli,
-both positive and negative, and acts as an interface between the external world 
-and the individual by gauging the importance of different stimuli. \cite{pccemotion}
+both positive and negative, and acts as an navigational interface, between the individual
+and the external world, by gauging the importance of different stimuli. \cite{pccemotion}
 
@@
 
+\todo{Expand on IPL: Why is facial recognition interesting to this paper?}
+
 It is hypothesized that the activation of these three regions of the brain (PCC,
@@
 representations are accessible via the PCC \cite{mappingself} and are gated into
-conscious awareness by activity in the mPFC, as influenced by changing internal 
+conscious awareness by activity in the mPFC \todo{The mPFC needs to be mentioned
+  earlier, defined, and abbreviated.} , as influenced by changing internal
 and external demands. The PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
@@
 
-There is a rich body of work pointing to this network (especially the dmPFC subsystem) 
-being engaged when individuals reflect upon, evaluate, appraise their own thoughts, 
-feelings and desires as well as those of others. \cite{theoryofmind}
-
-In addition to this, the dmPFC is involved in empathy, moral reasoning,
+The DMN (especially the dmPFC subsystem) is engaged when individuals reflect upon,
+evaluate, or appraise their own mental states or the mental states of others. ``Mental
+states'' include thoughts, feelings, and desires. \cite{theoryofmind} In addition to
+consideration of these states, the dmPFC is involved in empathy, moral reasoning,
 and altruistic behavior. \cite{defaultnetworkadaptive,dmpfcothers,dmpfcaltruism}
@@
 
+\todo{Define and describe the Medial Temporal Subsystem: Describe MTL anatomically and MTS relation
+  to the DMN, particularly the partial inclusion of the LS. Use the Anatomical
+  introduction (immediately following the Introduction) to assist in framing this.}
+
 The Medial Temporal Subsytem is primarily activated during these processes
-This subsystem consists of the Medial Temporal Lobe and parts of the limbic system
-especially the Hippocampus.
+This subsystem consists of the Medial Temporal Lobe (MTL) and parts of the Limbic
+System, particularly the Hippocampus.
 
-The limbic system is a complex set of structures that lies on both sides of the
+The Limbic System is a complex set of structures that lies on both sides of the
 Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
 Amygdala, and several other nearby areas. It appears to be primarily responsible for
-our emotional life, and has a lot to do with the formation of memories. \cite{limbic}
+our emotional life and has a specific involvement in the formation of memories. \cite{limbic}
 
-The critical role of the Medial Temporal Lobe (MTL) which is a part of the DMN 
-in long term memory is well documented. \cite{mtl}
-This network has also been activated during autobiographical memory retrieval, 
+The critical role of the MTL in long term memory is well documented. \cite{mtl}
+This network \todo{Which network?} has also been activated during autobiographical memory retrieval,
 contextual association, and semantic knowledge. \cite{defaultnetworkadaptive}
@@
 
-\todo{Summarize DMN? Maybe introduce why it's relevant to meditation?}
-Therefore the DMN that plays a role in constructing a sense of self, in memory, 
-in thinking about other also lends itself to excessive rumination and anxiety. \cite{dmndepression} 
-When I sit down to meditate and my thoughts wander away (and how they wander!),
-it is the DMN that is being activated.
+\subsubsection{Summary}
+
+The DMN plays a role in constructing a sense of self, in memory,
+and in thinking about others. Because of this, it is also a key component in excessive
+rumination and anxiety. \cite{dmndepression}
+
+When a meditator sits down to meditate and thoughts wander away to past experiences
+or anxieties about the future, it is the DMN playing the most active role in this
+distraction.
+
+\todo{Expand on DMN relationship to Monkey Mind. Add papers with supporting evidence?}
 
@@
 studies are usually only done on meditators who have clocked thousands of hours of
-practice, which diverges from the purpose of this paper. \cite {tibeteanmonks
-, experiencemed, experiencemeda}
+practice, which diverges from the purpose of this paper. \cite{tibetanmonks,
+  experiencemed, experiencemeda}
 
@@
 
-\bibitem{tibeteanmonks}
+\bibitem{tibetanmonks}
   Alex Hankey
-  \url:https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1697747/
+  \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1697747/}
   \textit{Studies of Advanced Stages of Meditation in the Tibetan
@@
   NeuroImage. Volume 54, Issue 2, 15 January 2011, Pages 1589-1599
-  \url:https://www.sciencedirect.com/science/article/abs/pii/S1053811910012243
+  \url{https://www.sciencedirect.com/science/article/abs/pii/S1053811910012243}
 
@@
   Catani M, Dell'Acqua F, Thiebaut De Schotten M
-  \textit {A revised limbic system model for memory, emotion and behaviour}
+  \textit{A revised limbic system model for memory, emotion and behaviour}
   Neuroscience and Biobehavioral Reviews. 37 (8): 1724–37.
@@
    Brewer JA, Worhunsky PD, Gray JR, Tang YY, Weber J, Kober H.
-   \textit {Meditation experience is associated with differences in default mode 
+   \textit{Meditation experience is associated with differences in default mode
            network activity and connectivity.}

### 2019-10-03 paper-two/paper-two.tex
round two edits to paper two, done

@@
 
-\todo{Introduce both the Limbic System and Somatosensory Cortex here, at a very high level.}
+\todo{Introduce both the Limbic System and Somatosensory Cortex here, at a very high
+  level.}
+
+Where possible in the following sections these networks are described in terms of
+function. The description of the DMN is also organized by function but the subsystems
+of the CEN and SN offer too many potential functional categorizations and are instead
+organized by anatomy.
 
@@
 are being performed. These tasks or executive functions are cognitive processes involved in
-cognitive control of behavior. This depends on three types of brain functions : working memory,
-mental flexibility and self-control.
-The regions of the brain involved in these executive functions are
-Dorsolateral Prefrontal Cortex (DL-PFC), Orbito Frontal Cortex (OFC),
-and the Posterior Parietal Cortex (PPC).
+cognitive control of behavior. This depends on three types of brain functions:
+working memory, mental flexibility, and self-control. The regions of the brain
+involved in these executive functions are the Dorsolateral Prefrontal Cortex
+(DL-PFC), Orbito Frontal Cortex (OFC), and the Posterior Parietal Cortex (PPC).
 
@@
   \centering
-  \includegraphics[width=7.5cm]{images/top-sn.jpg}
-  \caption{The Salience Network consists of the Anterior Cingulate Cortex and the
-    Anterior Insular Cortex. It helps switch between the Default Mode Network and the
-  Central Executive Network.}
-  \label{fig:top-sn}
+  \includegraphics[width=5cm]{images/top-cen.jpg}
+  \caption{The Central Executive Network.}
+  \label{fig:top-cen}
 \end{figure}
@@
 \subsubsection{Posterior Parietal Cortex}
-The PPC is a region of the parietal cortex that is posterior to the primary somatosensory
-cortex.
-The posterior parietal cortex receives input from auditory, visual, and somatosensory systems
-as well as a variety of other brain regions, and it integrates this input to facilitate
+The PPC is a region of the Parietal Cortex, physically located behind the primary
+Somatosensory Cortex. The PPC receives input from auditory, visual, and somatosensory systems
+as well as a variety of other brain regions \todo{What other brain regions provide
+  input to the PPC?}, and it integrates this input to facilitate
 the execution of functions that require diverse information (also known as higher-order functions).
-It in turn activates the DL-PFC and motor cortex and is involved in the performance of
+In turn, the PPC activates the DL-PFC and motor cortex and is involved in the performance of
 attention related tasks as well as higher-order motor tasks such as grasping and catching. \cite{parietallobesa,parietallobes}
 
-\subsection {Sidenote: Somatosensory Cortex}
-The primary somatosensory cortex is located in a ridge of cortex found within the parietal lobe.
+\todo{PSC section mentions Parietal Lobe but it hasn't been mentioned earlier, nor defined?}
+
+\subsubsection{Sidenote: Somatosensory Cortex}
+The Primary Somatosensory Cortex is located in a ridge of cortex found within the Parietal Lobe.
 It is responsible for processing somatic sensations. These sensations arise from receptors
 positioned throughout the body. They are responsible for detecting touch, for proprioception
-(i.e. the position of the body in space), nociception (i.e. pain), and temperature. \cite{somato}
-When such receptors detect one of these sensations,  the information is sent to the thalamus and
-then to the primary somatosensory cortex.
+(the position of the body in space), nociception (pain), and temperature. \cite{somato}
+When such receptors detect one of these sensations, the information is sent first to
+the Thalamus and then to the Primary Somatosensory Cortex.
+
 The sensory homunculus which is a part of the somatosensory cortex is a cortical representation
@@
 
-\todo {include images of the homunculus here}
+\todo{Include side image of PSC here.}
 
-\todo{Summarize CEN? Relevance to meditation?}
-So, the Central Executive Network is involved in goal-directed behavior and processes
-related to attention. These goals are usually external and attention is also directed externally(whether
-voluntary or not). When you meditate, you direct your attention voluntarily
-inside, to the same sensory cues that are directing the PPC.
+\todo{Include images of the homunculus here}
+
+\subsubsection{Summary}
+
+The Central Executive Network is involved in goal-directed behavior and processes
+related to attention. These goals are usually external and attention is also directed
+externally (regardless of whether attention is voluntary or not). When you meditate,
+attention is directed both internally and voluntarily, to the same sensory cues that
+are normally directing the PPC. \todo{What is interesting about internal/voluntary
+  attention in CEN?}
 
@@
 in the brain that help decide which stimuli deserves our attention. It acts as a
-switch between the internally-directed DMN and the externally-directed Central
-Executive Network. \cite{saliencenetwork}
+switch between the internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}
 
@@
 
-The Insula is a part of the Cerebral Cortex located deep within the Sulcus, the
+The AIC is a part of the Cerebral Cortex located deep within the Sulcus, the
 fissure separating the four lobes of the brain. The AIC physically projects itself
@@
 audio-visual integration tasks, interoceptive awareness (so its activity is directly
-related to an individual's sense of internal body states), empathy and conscious
+related to an individual's sense of internal body states), empathy, and conscious
 awareness. \cite{aicemotion}
@@
 
-The salience network is directly involved in switching between the DMN and CN.
-When you meditate, you cultivate the cognitive processes related to attention.
+\subsubsection{Summary}
+
+The SN is directly involved in switching between the DMN and CEN. When you meditate,
+you cultivate the cognitive processes related to attention.
 When your mind wanders, you notice it and bring it back.
-The salience network should be directly involved in this process of assigning attention.
-Meditation should strengthen the salience network and its ability to switch from
-the default.
+The SN is directly involved in this process of assigning attention, since it is
+responsible for the prominence of any object of attention.
+Meditation, or the act of returning to an object of meditation, can therefore be
+thought of as an exercise in strengthening the SN and its ability to switch from
+the default objects of attention, over time.
 
@@
 was accepted into the course until the first day of the course I was very nervous
-about what it might entail. I had never done anything like this (meditation) before
+about what it might entail. I had never done anything like meditation before
 and I was quite happy with the idea that meditation, with its spiritual connotations
@@
 The following are my observations and speculations about what could be happening
-in the brain during a 10 course.
+in the brain during a 10-day course.
 
 I went for my first course in Chennai, my home town. When I arrived, I had to fill
-all the application forms all over again (despite applying online), keep all my
+the application forms all over again (despite applying online), keep all my
 luggage in a locker, and go to my room. I had to share the room with one other
-person. I felt quite out of my element and I was certain everyone could see in me
-that I was the odd one out, the one that was not supposed to be there. Because I felt
+student. I felt quite out of my element and I was certain everyone else could see
+that I was the odd one out. In my head, I was not supposed to be there. Because I felt
 so out of place, I was too scared to actually make conversation with anyone before
@@
 And so it was. With my brain filled with thoughts about myself, about the people
-around me, I started this meditation business. For the first 3.5 days you are asked
-to focus your attention on the area below your nostrils and above your upper
-lip. That's all, these were really simple instructions, but so hard to keep up with.
-Within the first two days, I had made up elaborate stories about my
-fellow meditators, all of them superheroes, tirelessly working to save humanity
-together. My default mode network was in overdrive.
+around me, I started this meditation business. For the first 3.5 days students are asked
+to focus their attention on the area below their nostrils and above their upper
+lip. That's all. These are undoubtedly very simple instructions but in practice I
+found them very difficult. Maintaining my focus was much harder than I had
+anticipated. Within the first two days, I had made up elaborate stories about my
+fellow meditation students. In these fantasies all of the other students were
+superheroes, tirelessly working to save humanity together. My Default Mode Network
+was in overdrive.
 
-Focusing our attention below the nostrils starts with observing your breath coming in
+Focusing our attention below the nostrils starts with observing breath --- coming in
 and going out. As the teacher mentions in the late evening discourses, the reason for
@@
 
-During the meditation hours I was trying hard. If I was to examine this practice
+\todo{Elaborate on breath/bridge and how it relates to brain function?}
+
+During the meditation hours I was trying hard. I had decided that if I was to examine this practice
 empirically, I had to give it an honest shot. I focused on the small patch beneath my
-nostrils above my upper lip. To begin with, while trying to observe your breath and
-your mind wanders, that is the default mode network working. As you realize your mind
-has wandered and bring your attention back to your breath, your salience network is
-activated. This happens slowly at first, with your mind wandering for many, many
-minutes before you realize you are not with your breath. But within the 3.5 days, the
-SN learns quite quickly to bring attention back to the breath. After a few days, this
+nostrils above my upper lip. Initially, while trying to observe one's breathing, the
+mind will wander --- not just to fantasies of superheroes but to almost any object or
+daydream imaginable --- as the mind wanders the Default Mode Network is working. With
+the realization that the mind has wandered, and the consequent return of attention
+back to the breath, the Salience Network is
+activated. This happens slowly at first. One's mind wanders away for many, many
+minutes before realizing that the object of attention, the breath, has been lost. But
+at some point within the 3.5 days, the SN learns to bring attention back to the breath. After a few days, this
 refocusing almost happens by habit, almost automatically.
 
-As I gave in to this activity, tried to focus more intently, tried to sit still for
-longer, there were longer periods of awareness on that patch of skin and, with them,
+As I gave in to this activity, tried to focus more intently, and tried to sit still for
+longer continuous stretches of time, there were longer periods of awareness on that
+patch of skin. With these longer periods of awareness came
 stranger sensations arising and passing within that physical area. By the third day,
@@
 
-There have been studies that have shown that once you bring your
+\todo{Neurological / physiological basis for these strange sensations?}
+
+There have been studies that have shown that once subjects bring their
 attention back to breath, the neural structures involved in the control
 of the Autonomic Nervous System and attention start firing more actively
-(in this case, the DL-PFC --- a part of the CEN and the Hippocampus a part of the DMN
-and ACC --- a part of the SN). There are also global dampening changes seen in the brain,
-particularly within the DMN and the Amygdala, which is involved in
+(in this case, the DL-PFC --- a part of the CEN, the Hippocampus --- a part of the
+DMN, and the ACC --- a part of the SN). During Anapana there are also global
+dampening changes seen in the brain, particularly within the DMN and the Amygdala, which is involved in
 the flight-or-fight (stress) response. These changes have been
@@
 
+\todo{Relate ``gateway to altered states of mind'' to the Anapana ``bridge''?}
+
 There has also been research in mice, showing a cluster of nerves called the
@@
 conscious and unconscious breathing but for all kinds of breathing --- such as yawns,
-sighs and gasps. The preBötc also appears to play a role in calming and
+sighs and gasps. The preBötC also appears to play a role in calming and
 arousal. \cite{prebotcgeneration} It would be interesting to see the activation in
@@
 After lunch on the fourth day the instructions for Vipassana are given. During
-Vipassana, you transfer your attention from the patch below your nostril to the top
-of your head. From there, for two hours, the instructions are given to slowly move
-your attention throughout the entire body: ``from the top of the head to the tips of
-the toes'', then repeat. The two criteria for observation of bodily sensations are
-focused attention (making use of the narrow focus practiced during the Anapana
-period) and open monitoring (observing the sensations objectively).
+Vipassana, students transfer their attention from the patch below their nostrils to the top
+of the head. From there, for two hours, instructions are given to slowly move
+one's attention throughout the entire body: ``from the top of the head to the tips of
+the toes.'' This movement of attention throughout the body is then repeated, over and
+over. The two objectives while observing bodily sensations are focused attention (making
+use of the narrow focus practiced during the Anapana period) and open monitoring
+(observing the sensations objectively).
 
-The Salience Network (ACC and AIC) and the Central Executive Network (DL-PFC, OFC,
-and PPC) which are involved in the control of attention start firing rapidly. Usually
-when a person is engaged in an external task, the OFC and PPC are receiving signals
+As this process progresses, the Salience Network (ACC and AIC) and the Central Executive Network (DL-PFC, OFC,
+and PPC), which are involved in the control of attention, fire more actively. Usually
+when a person is engaged in an external task, the OFC and PPC receive signals
 from all the sensory networks (Visual, Audio, Somatosensory, etc.) and, in response,
@@
 bodily sensation --- and do nothing. So as you sit in silence with your eyes closed,
-not moving and scanning your body with your attention, you are doing what the CEN has
-always done except attempting not to react.
+not moving your body physically but instead moving throughout your body with your
+attention, the CEN is doing what it has always done. However, the outcomes of
+activity in the CEN is different during Vipassana. Instead of focusing on
+external sensory input the sensory input is internal (bodily sensations) and instead
+of reacting to input by performing tasks or feeling emotions the goal is to not
+react, to simply observe as objectively as possible.
 
 This is hard at first, with the body involuntarily reacting by jerking and writhing
-in involuntary response to the act of observation. Initially, your awareness is
-itself jerky and it is hard to observe sensations consciously. There is also the
+in involuntary response to the very act of observation. Initially, even awareness
+itself has a jerky quality and it is hard to observe sensations consciously. There is also the
 pain. This eclipses all other sensations. In my case, it was difficult to maintain
@@
 sensations and was only aware of the pain, the ACC was firing rapidly and the
-OFC then assigned these signals to the Limbic System which
+OFC then assigned these signals to the Limbic System, which
 put me in a less than ideal emotional state.
 
-But as your awareness gets smoother and observations get more objective, this
-functional connectivity between the ACC and the OFC slowly starts changing. As you
-continue to pay attention to the different sensations and you attempt to avoid assigning
-emotions to them. Without assigning an emotional response, you can observe with a
+Over the final seven days of the course, as awareness gets smoother and observations get more objective, this
+functional connectivity between the ACC and the OFC slowly starts changing. While
+continuing to pay attention to constantly changing bodily sensations the goal remains: avoid
+assigning evaluations (and therefore, emotions) to them. Without assigning an
+emotional response, one learns to observe with a
 non-judgmental attitude, particularly toward unpleasant stimuli. The OFC still fires
-but the Limbic System is quiesced and, in turn, you react less. It is as if both your
-conscious and unconscious awarenesses are slowly aligning until you experience some
-occasions when it feels as if there is nothing but detached awareness moving
-through your body (or some parts of your body). Senses are not dulled, however. In
+but the Limbic System is quiesced and, in turn, reactions become smaller and less
+frequent. This snowball effect almost mirrors the spiral of brain behaviours
+exhibited early in the course. Early in the course, the CEN's normal mode of
+operation is in play, which is to experience unpleasant stimuli and react (desiring to rid the body of
+them), which in turn causes the unpleasant stimuli to appear all the more
+unpleasant. Once a measure of objectivity is reached, however, the brain's function
+snowballs in the other direction. It is as if both conscious and unconscious awarenesses are slowly aligning
+until there are occasions of experience when it feels as if there is nothing but detached awareness moving
+through the body (or some parts of the body). This is not at all akin to the
+detachment of sleep or intoxication, however, and the senses are not dulled. In
 addition to a feeling of emotional detachment, the Insula and the Somatosensory
-Cortex, which are involved in sensory awareness, were firing \textbf{more} actively,
-not less.
+Cortex, which are involved in sensory awareness, were \textbf{more} active, not
+less. As the ACC, OFC, and Limbic System are calmed and attention is narrowed
+further, the sensations become progressively more intense.
 
-Perhaps the strangest experience during a Vipassana course is that of shifts in your
-sense of self. The ``self'', as a subjective agent in space and time changes, if at
-first only very slightly. As you close your eyes and meditate, it may feel as though
+Perhaps the strangest experience during a Vipassana course is that of shifts in the
+sense of self. The ``self'', as a subjective agent in space and time, changes. At
+first, this change is only very slight. As a student closes her eyes and begins to
+meditate, it may feel as though
 the body has shifted in space or rotated in different directions. During these times,
-the vmPFC and PCC become less active. As the meditations get deeper the DMN
+the vmPFC and PCC become less active. \todo{So? What does this mean?} As the meditations get deeper the DMN
 occasionally quiets completely (perhaps only for a short duration) while the Insula
-fires more rapidly, leading to alternative networks of consciousness arising. It is
+fires more rapidly, leading to alternative networks of consciousness
+arising. \todo{What does this mean? What is an example of ``alternative networks of consciousness''?} It is
 as if losing the influence of one's own narrative leads to new insights that are
@@
 Vipassana causes global dampening changes in the DMN, causes certain parts of
-the SN and CEN to fire more rapidly. That is, reacting parts of the network and
+the SN and CEN to activate. That is, reacting parts of the network and
 portions of the network which are aware of bodily sensation, respectively.
-This changes a person's behavior (in the short term, at least) and these
-changes can often be seen as soon as someone leaves the course on the tenth day.
+This changes a person's behavior and these changes can often be seen as soon as
+someone leaves the course on the tenth day.
 
@@
 many years of habit formation requires more than a ten day course. However, the ten
-day course is well-designed and will give someone experimenting with meditation the
+day course is well-designed \todo{How is the course well designed?} and will give someone experimenting with meditation the
 minimum observational change necessary to instill an interest, to keep practicing.
@@
 functional changes in brain activity, especially to the DMN and the SN. But such
-studies are usually only done on meditators who have clocked thousands of hours of
+studies are often only done on meditators who have clocked thousands of hours of
 practice, which diverges from the purpose of this paper. \cite{tibetanmonks,
@@
 enter into a state of flow. But flow or no flow, the reacting part of the brain is
-also working. And it is working hard. The goal of Vipassana is not simply to
-concentrate or to enter a state of flow, but to break old reactionary habit patterns
-altogether.
+also working. The mind may learn to focus intensely but the habit patterns of the
+mind are unlikely to change. The goal of Vipassana is not simply to concentrate or to
+enter a state of flow, but to break old reactionary habit patterns altogether.
 
@@
 occurred over those ten days. What was it that had changed in my brain? I had to
-know. I very quickly noticed, however, that as I left the centre (or even earlier, as
-soon as talking was allowed), those very changes started slipping away. But with each
-hour-long meditation at home, a tiny bit sticks just a little more. And I get closer
-to my answer.
+know. I very quickly noticed, however, that as I left the centre, those very changes
+started slipping away. But with each hour-long meditation at home, a tiny bit sticks
+just a little more. And I get closer to my answer.
 

### 2019-10-04 paper-two/paper-two.tex
reviews of second edit with preethical

@@
 
-\todo{Introduce both the Limbic System and Somatosensory Cortex here, at a very high
-  level.}
-
 Where possible in the following sections these networks are described in terms of
@@
 Parietal Lobule (IPL). Observationally, these regions are preferentially activated when
-volunteers were asked to reflect on personal preferences, beliefs, abilities, or
+subjects were asked to reflect on personal preferences, beliefs, abilities, or
 any other qualities relevant to the self. \cite{defaultnetworkadaptive} Cerebral
@@
 the memory recalled. It is also activated by emotional stimuli,
-both positive and negative, and acts as an navigational interface, between the individual
+both positive and negative, and acts as a navigational interface, between the individual
 and the external world, by gauging the importance of different stimuli. \cite{pccemotion}
@@
 Cortex (dmPFC), which assists in the creation of the conceptual self through an
-individual's consideration of ``others''.
+individual's consideration of ``others''. Together, the vmPFC and dmPFC form the
+Medial Prefrontal Cortex (mPFC).
 
@@
 
-\todo{Expand on IPL: Why is facial recognition interesting to this paper?}
+\todo{Expand on IPL: How does the IPL contribute to our sense of self?}
 
@@
 representations are accessible via the PCC \cite{mappingself} and are gated into
-conscious awareness by activity in the mPFC \todo{The mPFC needs to be mentioned
-  earlier, defined, and abbreviated.} , as influenced by changing internal
+conscious awareness by activity in the mPFC, as influenced by changing internal
 and external demands. The PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
@@
 
-\todo{Define and describe the Medial Temporal Subsystem: Describe MTL anatomically and MTS relation
-  to the DMN, particularly the partial inclusion of the LS. Use the Anatomical
-  introduction (immediately following the Introduction) to assist in framing this.}
+\todo{Define and describe the MTS: Describe MTL anatomically and MTS relation
+  to the DMN, particularly the partial inclusion of the LS.}
 
@@
 The critical role of the MTL in long term memory is well documented. \cite{mtl}
-This network \todo{Which network?} has also been activated during autobiographical memory retrieval,
+The MTS has also been activated during autobiographical memory retrieval,
 contextual association, and semantic knowledge. \cite{defaultnetworkadaptive}
@@
 When a meditator sits down to meditate and thoughts wander away to past experiences
-or anxieties about the future, it is the DMN playing the most active role in this
+or anxieties about the future, classic meditation texts and instructions refer to
+this behaviour as ``Monkey Mind''. It is the DMN playing the most active role in this
 distraction.
-
-\todo{Expand on DMN relationship to Monkey Mind. Add papers with supporting evidence?}
 
@@
 The PPC is a region of the Parietal Cortex, physically located behind the primary
-Somatosensory Cortex. The PPC receives input from auditory, visual, and somatosensory systems
-as well as a variety of other brain regions \todo{What other brain regions provide
-  input to the PPC?}, and it integrates this input to facilitate
-the execution of functions that require diverse information (also known as higher-order functions).
-In turn, the PPC activates the DL-PFC and motor cortex and is involved in the performance of
-attention related tasks as well as higher-order motor tasks such as grasping and catching. \cite{parietallobesa,parietallobes}
+Somatosensory Cortex. The PPC is responsible for facilitating higher-order
+functions. Higher-order functions are defined as those requiring diverse information
+for the purposes of higher evolutionary behaviour: intelligence, memory, planning,
+speech, orientation, decision-making, etc.
 
-\todo{PSC section mentions Parietal Lobe but it hasn't been mentioned earlier, nor defined?}
+To do this, the PPC receives input from auditory, visual, and somatosensory
+systems, integrates this input, and uses the aggregate to activate the DL-PFC and motor cortex and is
+involved in the performance of attention related tasks as well as higher-order motor
+tasks such as grasping and catching. \cite{parietallobesa,parietallobes}
 
 \subsubsection{Sidenote: Somatosensory Cortex}
-The Primary Somatosensory Cortex is located in a ridge of cortex found within the Parietal Lobe.
+The Primary Somatosensory Cortex is located in a ridge of cortex found within the Parietal Cortex.
 It is responsible for processing somatic sensations. These sensations arise from receptors
@@
 
-The Central Executive Network is involved in goal-directed behavior and processes
+The CEN is involved in goal-directed behavior and processes
 related to attention. These goals are usually external and attention is also directed
@@
 attention is directed both internally and voluntarily, to the same sensory cues that
-are normally directing the PPC. \todo{What is interesting about internal/voluntary
-  attention in CEN?}
+are normally directing the PPC.
 
@@
 you cultivate the cognitive processes related to attention.
-When your mind wanders, you notice it and bring it back.
+When the meditator's mind wanders, she notices it and brings it back.
 The SN is directly involved in this process of assigning attention, since it is
@@
 Meditation, or the act of returning to an object of meditation, can therefore be
-thought of as an exercise in strengthening the SN and its ability to switch from
-the default objects of attention, over time.
+thought of as an exercise in strengthening the SN and its ability to perform this switch, over time.
 
@@
 And so it was. With my brain filled with thoughts about myself, about the people
-around me, I started this meditation business. For the first 3.5 days students are asked
-to focus their attention on the area below their nostrils and above their upper
+around me, I started this meditation business. For the first 3.5 days I was instructed
+to focus my attention on the area below my nostrils and above my upper
 lip. That's all. These are undoubtedly very simple instructions but in practice I
@@
 
-\todo{Elaborate on breath/bridge and how it relates to brain function?}
+There has been research in mice, showing a cluster of nerves called the
+pre-Bötzinger Complex (preBötC), found in the brain stem of most mammals, which fires
+with every breath taken. This breathing pacemaker seems to work not only for regular
+conscious and unconscious breathing but for all kinds of breathing --- such as yawns,
+sighs and gasps. The preBötC also appears to play a role in calming and
+arousal. \cite{prebotcgeneration} It would be interesting to see the activation in
+these neuron clusters during the first three days of a Vipassana course, even among
+novice meditators.
 
@@
 it felt like entire ecosystems were writhing and flopping and crashing, all of them
-very alive in the area below my nostrils above my upper lip.
-
-\todo{Neurological / physiological basis for these strange sensations?}
+very alive in the area below my nostrils above my upper lip. Since Anapana meditation
+is very simple and these strange sensations appeared within a few days of my practice
+as an absolute beginner, it would be very interesting (and relatively easy) to study
+the neurological and physiological basis for such sensations.
 
@@
 of mind. \cite{relaxationresponse}
-
-\todo{Relate ``gateway to altered states of mind'' to the Anapana ``bridge''?}
-
-There has also been research in mice, showing a cluster of nerves called the
-pre-Bötzinger Complex (preBötC), found in the brain stem of most mammals, which fires
-with every breath taken. This breathing pacemaker seems to work not only for regular
-conscious and unconscious breathing but for all kinds of breathing --- such as yawns,
-sighs and gasps. The preBötC also appears to play a role in calming and
-arousal. \cite{prebotcgeneration} It would be interesting to see the activation in
-these neuron clusters during the first three days of a Vipassana course, even among
-novice meditators.
 
@@
 attention, the CEN is doing what it has always done. However, the outcomes of
-activity in the CEN is different during Vipassana. Instead of focusing on
+activity in the CEN are different during Vipassana. Instead of focusing on
 external sensory input the sensory input is internal (bodily sensations) and instead
@@
 detachment of sleep or intoxication, however, and the senses are not dulled. In
-addition to a feeling of emotional detachment, the Insula and the Somatosensory
+addition to a feeling of emotional detachment, the AIC and the Somatosensory
 Cortex, which are involved in sensory awareness, were \textbf{more} active, not
@@
 the vmPFC and PCC become less active. \todo{So? What does this mean?} As the meditations get deeper the DMN
-occasionally quiets completely (perhaps only for a short duration) while the Insula
+occasionally quiets completely (perhaps only for a short duration) while the AIC
 fires more rapidly, leading to alternative networks of consciousness
@@
 But repeated practice is essential for any sustained neuroplasticity since rewriting
-many years of habit formation requires more than a ten day course. However, the ten
-day course is well-designed \todo{How is the course well designed?} and will give someone experimenting with meditation the
-minimum observational change necessary to instill an interest, to keep practicing.
+many years of habit formation requires more than a ten day course. However, the
+10-day course is both progressive and systematic. It is designed to give someone
+experimenting with meditation the minimum observational change necessary to instill
+an interest, to keep practicing.
 

### 2019-10-06 paper-two/paper-two.tex
paper 2: adding all new images, with new renders

@@
   \centering
-  \includegraphics[width=7.5cm]{images/top-dmn-sn-cen-switching.jpg}
+  \includegraphics[width=7.5cm]{images/fmri-switching.jpg}
   \caption{The Salience Network consists of the Anterior Cingulate Cortex and the
@@
   Central Executive Network.}
-  \label{fig:top-dmn-sn-cen-switching}
+  \label{fig:fmri-switching}
 \end{figure}
@@
 
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=5cm]{images/fmri-dmn.jpg}
+  \caption{fMRI scan of the Default Mode Network.}
+  \label{fig:fmri-dmn}
+\end{figure}
+
 \subsubsection{Autobiographical memory and future simulations}
@@
 
-\begin{figure}[H]
+\begin{figure}[h!]
   \centering
-  \includegraphics[width=5cm]{images/top-cen.jpg}
-  \caption{The Central Executive Network.}
-  \label{fig:top-cen}
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/top-cen.png}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/side-cen.png}
+  \end{subfigure}
+  \caption{Highlighted: areas of the brain firing when the Central Executive Network is active.}
+  \label{fig:cen}
 \end{figure}
@@
 
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=5cm]{images/fmri-cen.jpg}
+  \caption{fMRI scan of the Central Executive Network.}
+  \label{fig:fmri-cen}
+\end{figure}
+
 \subsubsection{Summary}
@@
 switch between the internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/top-sn.png}
+  \end{subfigure}
+  \begin{subfigure}[b]{0.48\linewidth}
+    \includegraphics[width=\linewidth]{images/side-sn.png}
+  \end{subfigure}
+  \caption{Highlighted: areas of the brain firing when the Salience Network is active.}
+  \label{fig:sn}
+\end{figure}
+
+\todo{Are the labels on either the SVG or the MRI image reversed?}
 
@@
 self-awareness.  \cite{selfaware}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=5cm]{images/fmri-sn.jpg}
+  \caption{fMRI scan of the Salience Network.}
+  \label{fig:fmri-sn}
+\end{figure}
 

### 2019-10-06 paper-two/paper-two.tex
paper two including side homunculus

@@
 
-The sensory homunculus which is a part of the somatosensory cortex is a cortical representation
+The Sensory Homunculus which is a part of the Somatosensory Cortex is a cortical representation
 of the body based on the degree of sensory innervation. \cite{sensoryhom}
 
-\todo{Include side image of PSC here.}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=5cm]{images/side-somatosensory.png}
+  \caption{Location of the Sensory Homunculus.}
+  \label{fig:side-somatosensory}
+\end{figure}
 

### 2019-10-06 paper-two/paper-two.tex
spelling mistake
mentioning approximate locations of brain activity in diagrams

@@
 \title{%
-  \Huge{Vipassana for Hackers} \\
-  \large{Paper Two: The Brain} \\
+  \large{Vipassana for Hackers} \\
+  \Huge{Paper Two: The Brain} \\
   \large\textit{Version 0.2}
@@
   \end{subfigure}
-  \caption{Highlighted: areas of the brain firing when the Default Mode Network is active.}
+  \caption{Highlighted: approximate locations of brain regions firing when the Default Mode Network is active.}
   \label{fig:dmn}
@@
   \end{subfigure}
-  \caption{Highlighted: areas of the brain firing when the Central Executive Network is active.}
+  \caption{Highlighted: approximate locations of brain regions firing when the Central Executive Network is active.}
   \label{fig:cen}
@@
   \includegraphics[width=5cm]{images/side-somatosensory.png}
-  \caption{Location of the Sensory Homunculus.}
+  \caption{Location of the Somatosensory Cortex.}
   \label{fig:side-somatosensory}
@@
   \end{subfigure}
-  \caption{Highlighted: areas of the brain firing when the Salience Network is active.}
+  \caption{Highlighted: approximate locations of brain regions firing when the Salience Network is active.}
   \label{fig:sn}

### 2019-10-07 paper-two/paper-two.tex
completed todo -1 : added information about ipl with 2 references

@@
 The Inferior Parietal lobe (IPL) is situated in the back of the brain and it is
-involved in the perception of facial stimuli. \cite{ipl}
-
-\todo{Expand on IPL: How does the IPL contribute to our sense of self?}
+involved in the perception of facial stimuli. \cite{ipl}.In a study previously conducted where
+subjects were asked to retrieve previous judgments about oneself, their friend and the Dutch queen, 
+the activation of the IPL decreased with decreasing self reference. \cite{iplself} 
+Another study used Transcranial Magnetic Stimulation (TMS) to create a sort of virtual lesion of
+the right-IPL and showed that this disrupts self-other discrimination. \cite{ipltms}
 
@@
 
+\bibitem{iplself}
+ Hans C. Lou, Bruce Luber, Michael Crupain, Julian P. Keenan, Markus Nowak, 
+ Troels W. Kjaer, Harold A. Sackeim, Sarah H. Lisanby
+ \textit{Parietal cortex and representation of the mental Self}
+  Proc Natl Acad Sci U S A. 2004 Apr 27; 101(17): 6827–6832.
+
+\bibitem{ipltms}
+ Lucina Q. Uddin, Istvan Molnar-Szakacs, Eran Zaidel, Marco Iacoboni
+ \textit {rTMS to the right inferior parietal lobule disrupts self–other discrimination}
+  Social Cognitive and Affective Neuroscience, Volume 1, Issue 1, June 2006, 
+  Pages 65–71, https://doi.org/10.1093/scan/nsl003
+
 \bibitem{mappingself}

### 2019-10-07 paper-two/paper-two.tex
more edits to grammar in IPL section

@@
 The Inferior Parietal lobe (IPL) is situated in the back of the brain and it is
-involved in the perception of facial stimuli. \cite{ipl}.In a study previously conducted where
-subjects were asked to retrieve previous judgments about oneself, their friend and the Dutch queen, 
-the activation of the IPL decreased with decreasing self reference. \cite{iplself} 
-Another study used Transcranial Magnetic Stimulation (TMS) to create a sort of virtual lesion of
+involved in the perception of facial stimuli. \cite{ipl}
+The IPL has been hypothesized as being involved in distinguishing self from others, 
+in a study where subjects were asked to retrieve 
+previous judgments about oneself, their friend and the Dutch queen, 
+the activation of the IPL was seen to decrease with decreasing self-reference. \cite{iplself} 
+Another study also used Transcranial Magnetic Stimulation (TMS) to create a sort of "virtual lesion" 
+(transiently disturb the neural circuitry in that area) of
 the right-IPL and showed that this disrupts self-other discrimination. \cite{ipltms}

### 2019-10-07 paper-two/paper-two.tex
updating 2nd todo MTL

@@
 \subsubsection{Autobiographical memory and future simulations}
+The term "autobiographical memory" refers to our memory for specific episodes, 
+and to our conceptual, generic, and schematic knowledge of our lives.
+
+The Medial Temporal Lobe (MTL) which is found on the medial surface of the temporal lobe and plays 
+an important role in memory processing. 
+The MTL consists of the hippocampal formation on top and the parahippocampal gyrus below it as well as
+The entorhinal and perirhinal cortices. 
+The MTL along with the amygdala and the olfactory cortex, hypothalamus and several other 
+anatomically connected areas forms the limbic system. It appears to be primarily responsible for
+our emotional life and has a specific involvement in the formation of memories. \cite{limbic}
 
@@
 Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
-Amygdala, and several other nearby areas. It appears to be primarily responsible for
-our emotional life and has a specific involvement in the formation of memories. \cite{limbic}
-
+Amygdala, a
 The critical role of the MTL in long term memory is well documented. \cite{mtl}

### 2019-10-07 paper-two/paper-two.tex
todo: MTS completed

--- 2019-10-07 paper-two/paper-two.tex
@@
 
-\todo{Define and describe the MTS: Describe MTL anatomically and MTS relation
-  to the DMN, particularly the partial inclusion of the LS.}
+The term "autobiographical memory" refers to our memory for specific episodes, 
+and to our conceptual, generic, and schematic knowledge of our lives.
 
-The Medial Temporal Subsytem is primarily activated during these processes
-This subsystem consists of the Medial Temporal Lobe (MTL) and parts of the Limbic
-System, particularly the Hippocampus.
+The Medial Temporal Lobe (MTL) which is found on the medial surface of the temporal lobe. 
+The critical role of the MTL in long term memory is well documented. \cite{mtl}
 
-The Limbic System is a complex set of structures that lies on both sides of the
-Thalamus, just under the Cerebrum. It includes the Hypothalamus, the Hippocampus, the
-Amygdala, and several other nearby areas. It appears to be primarily responsible for
+The MTL consists of the hippocampal formation on top and the parahippocampal gyrus below it as well as
+the entorhinal and perirhinal cortices. 
+The MTL, amygdala, the olfactory cortex, hypothalamus and several other 
+anatomically connected areas forms the limbic system. It appears to be primarily responsible for
 our emotional life and has a specific involvement in the formation of memories. \cite{limbic}
 
-The critical role of the MTL in long term memory is well documented. \cite{mtl}
-The MTS has also been activated during autobiographical memory retrieval,
+The MTL along with mPFC, IPL and PCC forms the Medial Temporal Subsystem (a part of the DMN). 
+This subsystem is triggered during autobiographical memory retrieval,
 contextual association, and semantic knowledge. \cite{defaultnetworkadaptive}

### 2019-10-07 paper-two/paper-two.tex
cleaning up jpgs

--- 2019-10-07 paper-two/paper-two.tex
@@
 involved in the perception of facial stimuli. \cite{ipl}
-The IPL has been hypothesized as being involved in distinguishing self from others, 
-in a study where subjects were asked to retrieve 
-previous judgments about oneself, their friend and the Dutch queen, 
-the activation of the IPL was seen to decrease with decreasing self-reference. \cite{iplself} 
-Another study also used Transcranial Magnetic Stimulation (TMS) to create a sort of "virtual lesion" 
-(transiently disturb the neural circuitry in that area) of
-the right-IPL and showed that this disrupts self-other discrimination. \cite{ipltms}
+
+\todo{Expand on IPL: How does the IPL contribute to our sense of self?}
 
@@
   \centering
-  \includegraphics[width=5cm]{images/fmri-dmn.jpg}
+  \includegraphics[width=5cm]{images/fmri-dmn.png}
   \caption{fMRI scan of the Default Mode Network.}
@@
   \centering
-  \includegraphics[width=5cm]{images/fmri-cen.jpg}
+  \includegraphics[width=5cm]{images/fmri-cen.png}
   \caption{fMRI scan of the Central Executive Network.}
@@
   \centering
-  \includegraphics[width=5cm]{images/fmri-sn.jpg}
+  \includegraphics[width=5cm]{images/fmri-sn.png}
   \caption{fMRI scan of the Salience Network.}
@@
 
-\bibitem{iplself}
- Hans C. Lou, Bruce Luber, Michael Crupain, Julian P. Keenan, Markus Nowak, 
- Troels W. Kjaer, Harold A. Sackeim, Sarah H. Lisanby
- \textit{Parietal cortex and representation of the mental Self}
-  Proc Natl Acad Sci U S A. 2004 Apr 27; 101(17): 6827–6832.
-
-\bibitem{ipltms}
- Lucina Q. Uddin, Istvan Molnar-Szakacs, Eran Zaidel, Marco Iacoboni
- \textit {rTMS to the right inferior parietal lobule disrupts self–other discrimination}
-  Social Cognitive and Affective Neuroscience, Volume 1, Issue 1, June 2006, 
-  Pages 65–71, https://doi.org/10.1093/scan/nsl003
-
 \bibitem{mappingself}

### 2019-10-08 paper-two/paper-two.tex
homnculus image

--- 2019-10-07 paper-two/paper-two.tex
@@
 
-\todo{Are the labels on either the SVG or the MRI image reversed?}
-
 \subsubsection{Anterior Cingulate Cortex}
@@
 the realization that the mind has wandered, and the consequent return of attention
-back to the breath, the Salience Network is
-activated. This happens slowly at first. One's mind wanders away for many, many
-minutes before realizing that the object of attention, the breath, has been lost. But
+back to the breath, the Salience Network is activated. 
+This happens slowly at first. One's mind wanders away for many, many minutes before 
+realizing that the object of attention, the breath, has been lost. But
 at some point within the 3.5 days, the SN learns to bring attention back to the breath. After a few days, this

### 2019-10-09 paper-two/paper-two.tex
adding homunculus diagram to pdf
update todo what does this mean?

--- 2019-10-07 paper-two/paper-two.tex
@@
 emotional response, one learns to observe with a
-non-judgmental attitude, particularly toward unpleasant stimuli. The OFC still fires
-but the Limbic System is quiesced and, in turn, reactions become smaller and less
+non-judgmental attitude, particularly toward unpleasant stimuli. The ACC is still up-regulated 
+but the OFC and the Limbic System may acquiesce and, in turn, reactions become smaller and less
 frequent. This snowball effect almost mirrors the spiral of brain behaviours
@@
 detachment of sleep or intoxication, however, and the senses are not dulled. In
-addition to a feeling of emotional detachment, the AIC and the Somatosensory
+addition to a feeling of emotional detachment, the ACC, AIC and the Somatosensory
 Cortex, which are involved in sensory awareness, were \textbf{more} active, not
-less. As the ACC, OFC, and Limbic System are calmed and attention is narrowed
+less. As the OFC, and Limbic System are calmed and attention is narrowed
 further, the sensations become progressively more intense.
@@
 first, this change is only very slight. As a student closes her eyes and begins to
-meditate, it may feel as though
-the body has shifted in space or rotated in different directions. During these times,
-the vmPFC and PCC become less active. \todo{So? What does this mean?} As the meditations get deeper the DMN
+meditate, it may feel as though the body has shifted in space or rotated in different direction, 
+due to the temporary down regulation of the prefrontal cortex (don't worry about it, keep observing
+sensations). \todo{So? What does this mean?} As the meditations get deeper the DMN
 occasionally quiets completely (perhaps only for a short duration) while the AIC

### 2019-10-09 paper-two/paper-two.tex
updated info on consciousness

@@
 awareness of any other sensation with the pain that was emanating from my legs and
-back. The OFC is involved in assigning emotion to sensation \cite{ofcemotion} and the
-ACC plays a vital role in perception of sensation. These two parts of the brain are
-usually functionally connected. In the beginning, as I sat trying to observe
-sensations and was only aware of the pain, the ACC was firing rapidly and the
-OFC then assigned these signals to the Limbic System, which
+back. The OFC is involved in assigning emotion to sensation and the shaping of pain by expectations
+\cite{ofcemotion,ofcexpect} and the ACC plays a vital role in perception of sensation. 
+These two parts of the brain are usually functionally connected. 
+In the beginning, as I sat trying to observe sensations and was only aware of the pain, 
+the ACC was firing rapidly and the OFC then assigned these signals to the Limbic System, which
 put me in a less than ideal emotional state.
@@
 first, this change is only very slight. As a student closes her eyes and begins to
-meditate, it may feel as though the body has shifted in space or rotated in different direction, 
-due to the temporary down regulation of the prefrontal cortex (don't worry about it, keep observing
-sensations). \todo{So? What does this mean?} As the meditations get deeper the DMN
+meditate, it may feel as though the body has shifted in space or rotated in different direction, this is
+due to the temporary down regulation of the prefrontal cortex.
+ \todo{So? What does this mean?} As the meditations get deeper the DMN
 occasionally quiets completely (perhaps only for a short duration) while the AIC
 fires more rapidly, leading to alternative networks of consciousness
-arising. \todo{What does this mean? What is an example of ``alternative networks of consciousness''?} It is
-as if losing the influence of one's own narrative leads to new insights that are
+arising. \todo
+Consciousness is defined as our subjective awareness of ourselves and our environment. 
+An altered state of consciousness is defined as alternate patterns or configurations of 
+this experience, which differ qualitatively from a baseline state.
+Altered states of consciousness
+
+{What does this mean? What is an example of ``alternative networks of consciousness''?} 
+It is as if losing the influence of one's own narrative leads to new insights that are
 usually kept from consciousness. It has been found that experiencing this, even for a
@@
 
+\bibitem{ofcexpect}
+ Lauren Y.Atlas, Tor D.Wager
+ \textit{How expectations shape pain.}
+  Neuroscience Letters,  520 (2):140-148, 29 June 2012.
+
 \bibitem{alteredtraits}

### 2019-10-09 paper-two/paper-two.tex
added information about consciousness and altered states

@@
 
-As this process progresses, the Salience Network (ACC and AIC) and the Central Executive Network (DL-PFC, OFC,
-and PPC), which are involved in the control of attention, fire more actively. Usually
-when a person is engaged in an external task, the OFC and PPC receive signals
-from all the sensory networks (Visual, Audio, Somatosensory, etc.) and, in response,
-continuously send signals to the different motor cortices, Amygdala, and Hypothalamus to
+As this process progresses, the Salience Network (ACC and AIC) and some parts of
+the Central Executive Network (DL-PFC, OFC, and PPC), 
+which are involved in the control of attention, fire more actively. The anti-correlation
+between the DMN and CEN reduces. Usually when a person is engaged in an external task, 
+the OFC and PPC receive signals from all the sensory networks (Visual, Audio, Somatosensory, etc.) and, 
+in response, continuously send signals to the different motor cortices, Amygdala, and Hypothalamus to
 respond by performing a task, feeling an emotion, etc.
@@
 unpleasant. Once a measure of objectivity is reached, however, the brain's function
-snowballs in the other direction. It is as if both conscious and unconscious awarenesses are slowly aligning
+snowballs in the other direction. It is as if both conscious and unconscious awareness-es are slowly aligning
 until there are occasions of experience when it feels as if there is nothing but detached awareness moving
@@
 meditate, it may feel as though the body has shifted in space or rotated in different direction, this is
-due to the temporary down regulation of the prefrontal cortex.
- \todo{So? What does this mean?} As the meditations get deeper the DMN
+due to the temporary down regulation of the prefrontal cortex. As the meditations get deeper the DMN
 occasionally quiets completely (perhaps only for a short duration) while the AIC
@@
 arising. \todo
-Consciousness is defined as our subjective awareness of ourselves and our environment. 
+
+[Side note; Consciousness is defined as our subjective awareness of ourselves and our environment. 
 An altered state of consciousness is defined as alternate patterns or configurations of 
 this experience, which differ qualitatively from a baseline state.
-Altered states of consciousness
+Altered states of consciousness include a state of dreaming, epileptic seizures, sensory deprivation, 
+hysterical states of dissociation and depersonalization as well as pharmacologically induced altered states.
 
-{What does this mean? What is an example of ``alternative networks of consciousness''?} 
-It is as if losing the influence of one's own narrative leads to new insights that are
+The entropic brain hypothesis: This was put forth by Robin Carhatt Harris et al, while studying 
+the psychedelic state. Entropy is a measure of the uncertainty in a system and in this case applied 
+to the states of consciousness and their associated neurodynamics. The theory makes a distinction 
+two fundamental modes of consciousness: Primitive consciousness and the normal waking or secondary
+consciousness. 
+Primitive consciousness is characterized by higher entropy and it is hypothesized to have 
+preceded the development of modern, adult, human, normal waking consciousness. 
+It is argued that suppression of this entropy facilitates normal waking consciousness 
+and associated metacognitive functions such as reality-testing and self-awareness.]
+
+Sometimes it feels like loosening the hold of one's own narrative leads to new insights that are
 usually kept from consciousness. It has been found that experiencing this, even for a

### 2019-10-09 paper-two/paper-two.tex

@@
 
-\todo{Include images of the homunculus here}
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=7.5cm]{images/homunculus.jpg}
+  \caption{The Sensory Homunculus mapped across the Somatosensory Cortex.}
+  \label{fig:homunculus}
+\end{figure}
 

### 2019-10-10 paper-two/paper-two.tex
edits to preethis most recent fixes

@@
 fires more rapidly, leading to alternative networks of consciousness
-arising. \todo
+arising.
 
-[Side note; Consciousness is defined as our subjective awareness of ourselves and our environment. 
+Consciousness is defined as our subjective awareness of ourselves and our environment.
 An altered state of consciousness is defined as alternate patterns or configurations of
@@
 
-The entropic brain hypothesis: This was put forth by Robin Carhatt Harris et al, while studying 
-the psychedelic state. Entropy is a measure of the uncertainty in a system and in this case applied 
-to the states of consciousness and their associated neurodynamics. The theory makes a distinction 
-two fundamental modes of consciousness: Primitive consciousness and the normal waking or secondary
-consciousness. 
+The notion of altered states of consciousness has been put forth another way by the
+Entropic Brain Hypothesis, described by Robin Carhart-Harris et al, while studying
+the brain under the influence of psychedelics. \cite{entropic,entropicrevisited} Entropy is a measure of the
+uncertainty in a system and, in this case, applies
+to states of consciousness and their associated neurodynamics. The theory makes a
+distinction between
+two fundamental modes of consciousness: Primitive Consciousness and the normal Secondary
+Consciousness of waking states.
 Primitive consciousness is characterized by higher entropy and it is hypothesized to have
@@
 It is argued that suppression of this entropy facilitates normal waking consciousness
-and associated metacognitive functions such as reality-testing and self-awareness.]
+and associated metacognitive functions such as reality-testing and self-awareness.
+
+Even one ten-day course will expose a student to these altered states of
+consciousness and provide access to Primitive Consciousness. However, unlike
+psychedelics, the Vipassana student is not under the control of an external force. If
+she opens her eyes and stops meditating, these disorienting altered states of
+consciousness instantly disintegrate. Lasting changes to consciousness caused by
+Vipassana take a different shape.
 
@@
 has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
-similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
+similar to the influence of certain hallucinogenic drugs or even, in some cases, to ecstatic
 seizures where the abnormal activity of the Anterior Insular Cortex leads to
-heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}
+heightened self-awareness, feelings of bliss, and a lack of
+ambiguity. \cite{cortexbliss} Outside the activity of meditation, the lasting effects
+(``Altered Traits'') of meditation are not so extreme. They tend to include
+heightened awareness, improved concentration, and increased empathy, among
+others. \cite{alteredtraits}
 
@@
   \textit {Influence of meditation on anti-correlated networks in the brain}
-  Front. Hum. Neurosci., 03 January 2012,
+  Frontiers in Human Neuroscience, 03 January 2012,
   https://doi.org/10.3389/fnhum.2011.00183
+
+\bibitem{entropic}
+  Robin L. Carhart-Harris, Robert Leech, Peter J. Hellyer, Murray Shanahan, Amanda
+  Feilding, Enzo Tagliazucchi, Dante R. Chialvo and David Nutt
+  \textit{The entropic brain: a theory of conscious states informed by neuroimaging
+    research with psychedelic drugs}
+  Frontiers in Human Neuroscience, 03 February 2014,
+  https://doi.org/10.3389/fnhum.2014.00020
+
+\bibitem{entropicrevisited}
+  Carhart-Harris, Robin L.
+  \textit{The entropic brain - revisited}
+  NEUROPHARMACOLOGY, Vol: 142, Pages: 167-178, ISSN: 0028-3908, 2018.
 

### 2019-10-19 paper-two/paper-two.tex
removing 404 pdf, build for 0.3 public release
cleaning up acknowledgements

@@
   \Huge{Paper Two: The Brain} \\
-  \large\textit{Version 0.2}
+  \large\textit{Version 0.3}
 }
@@
 
-\listoftodos
+% \listoftodos
 
@@
 involved in the performance of attention related tasks as well as higher-order motor
-tasks such as grasping and catching. \cite{parietallobesa,parietallobes}
+tasks such as grasping and catching. \cite{parietallobes}
 
@@
 
-Thank you to Steven Deobald for reviews, edits, and corrections.
-
-``Brain\_human\_normal\_inferior\_view'' By Patrick J. Lynch, medical illustrator, CC BY
-2.5, https://commons.wikimedia.org/w/index.php?curid=1496648
-
-``Brain\_human\_lateral\_view.svg'' By Patrick J. Lynch, medical illustrator, CC BY
-2.5, https://commons.wikimedia.org/w/index.php?curid=1496651
+Thank you to Steven Deobald for reviews, edits, and corrections. \texttt{Brain\_human\_normal\_inferior\_view.svg} and
+\texttt{Brain\_human\_lateral\_view.svg} by Patrick J. Lynch, medical illustrator, CC-BY 2.5 \cite{brainsvg}
 
@@
   Epub 2009 Aug 20.
-  \url:https://www.ncbi.nlm.nih.gov/pubmed/19699306
+  \url{https://www.ncbi.nlm.nih.gov/pubmed/19699306}
 
@@
 
-\bibitem{parietallobesa}
+\bibitem{parietallobes}
   Caspers S, Amunts K, Zilles K.
@@
    The Human Nervous System. 3rd ed. New York: Elsevier; 2012.
-
-\bibitem{parietallobes}
-  Martin, R. E.
-  \textit{Let’s Get to Know the Parietal Lobes!}
-  \url{http://gablab.mit.edu/downloads/Parietal_Primer.pdf}
-  \todo{parietal lobes article returns a 404. fix.}
 
@@
   Cortex, Volume 49, Issue 9, Pages 2494–2500ISSN 0010 9452,
-  \url: https://doi.org/10.1016/j.cortex.2013.01.006. 2013.
+  \url{https://doi.org/10.1016/j.cortex.2013.01.006}. 2013.
 
@@
 
+\bibitem{brainsvg}
+  Patrick J. Lynch
+  \texttt{Brain\_human\_normal\_inferior\_view.svg} and \textit{Brain\_human\_lateral\_view.svg}
+  CC-BY 2.5
+  \url{https://commons.wikimedia.org/w/index.php?curid=1496651}
+  \url{https://commons.wikimedia.org/w/index.php?curid=1496648}
+
 \end{thebibliography}

### 2019-10-19 paper-one/build-epub.sh
Fixing epub builds
- fixing epub title and series in both epub scripts
- fixed references section order in paper-two.tex (`section` within bibilography was causing errors)
- fixed svg resize errors by putting SVGs in their own directory... I have no idea why `epub-convert` was looking at the SVGs since the .tex file doesn't even reference them, but whatever.
- added a refrences section title to paper one
- adding graphicsmagick to readme (may not be required but it can't hurt)

@@ -12,6 +12,6 @@ latexmlpost -dest=paper-one.html paper-one.xml
 # which ebook-convert
 # /usr/local/bin/ebook-convert
 
-ebook-convert paper-one.html ../output/paper-one.epub --language en --output-profile=kindle --authors="Steven Deobald" --title="Vipassana for Hackers"
+ebook-convert paper-one.html ../output/paper-one.epub --language en --output-profile=kindle --authors="Steven Deobald" --series="Vipassana for Hackers" --title="Paper One: Curious Mechanics" --no-default-epub-cover
 
 rm -f paper-one.xml paper-one.html LaTeXML.cache LaTeXML.css ltx-article.css

### 2019-10-19 paper-two/build-epub.sh

@@ -12,6 +12,6 @@ latexmlpost -dest=paper-two.html paper-two.xml
 # which ebook-convert
 # /usr/local/bin/ebook-convert
 
-ebook-convert paper-two.html ../output/paper-two.epub --language en --output-profile=kindle --authors="Steven Deobald" --title="Vipassana for Hackers"
+ebook-convert paper-two.html ../output/paper-two.epub --language en --output-profile=kindle --authors="Preethi Govindarajan" --series="Vipassana for Hackers" --title="Paper Two: The Brain" --no-default-epub-cover
 
 rm -f paper-two.xml paper-two.html LaTeXML.cache LaTeXML.css ltx-article.css

### 2019-10-29 paper-three/paper-three.tex
paper three 2-column

@@
-\documentclass{article}
+\documentclass[a4paper, amsfonts, amssymb, amsmath, reprint, showkeys, nofootinbib, twoside]{revtex4-1}
 
+\usepackage[english]{babel}
+\usepackage[utf8]{inputenc}
+\usepackage[colorinlistoftodos, color=green!40, prependcaption]{todonotes}
 \usepackage{graphicx}
@@
 
-\begin{titlepage}
-   \vspace*{\stretch{1.0}}
-   \begin{center}
-     \Huge\textbf{Vipassana for Hackers}\\
-     \Huge{Paper Three: Why Meditate?}\\
-     \vspace{5cm}
-     \large\textit{Steven Deobald}\\
-     \large\textit{Version 0.1 DRAFT}\\
-     \large\textit\today\\
-     \vspace{5cm}
-     \large\textit{}\\
-   \end{center}
-   \vspace*{\stretch{2.0}}
-\end{titlepage}
+\title{%
+  \large{Vipassana for Hackers} \\
+  \Huge{Paper Three: Why Meditate?} \\
+  \large\textit{Version 0.1}
+}
+\author{Steven Deobald}
+\email[Correspondence email address: ]{steven@deobald.ca}
+\affiliation{vipassana-for-hackers.org}
+\date{\today}
 
-\begin{center}
-  \Huge{Target Audience}
-\end{center}
+\begin{abstract}
+ \todo{Write the abstract.}
+\end{abstract}
 
-\textit{Vipassana for Hackers, Paper One: Curious Mechanics} was written to avoid discussing the
+\keywords{neuroplasticity}
+
+\maketitle
+
+\section{Target Audience}
+
+\textit{Vipassana for Hackers, Paper One: Curious Mechanics} was written with the
+explicit intention of avoiding a discussion about specific
 outcomes or consequences of meditation in detail. The focus of that paper was only
@@
 as they assist the reader in understanding what is written earlier in the paper
-regarding the senses. \textit{Paper Two: The Brain} goes further into the mechanics as they pertain to the
-nervous system. Here, outcomes are discussed as they pertain to
+regarding the senses. \textit{Paper Two: The Brain} goes further into the internal
+mechanics as they pertain to the
+hub of the nervous system. Here, outcomes are discussed as they pertain to
 neuroplasticity. Neither paper directly discusses why an individual might choose to
@@
 \textit{Paper Three: Why Meditate?} is written for anyone who has ever asked
-themselves that very question or asked that question of their friends who
+themselves that very question or asked this question of their friends who
 meditate. It is for both those who are curious about the practice of Vipassana
@@
 
-\pagebreak
-
-\begin{center}
-  \Huge{Vipassana Basics}
-\end{center}
+\section{Vipassana Basics}
 
@@
   \centering
-  \includegraphics[width=11cm]{images/sense-doors.png}
-  \caption{The sense doors and bodily sensation.}
+  \includegraphics[width=0.8\linewidth]{images/sense-doors.png}
+  \caption{The sense doors and bodily sensation. \cite{sense-icons}}
   \label{fig:sense-doors}
@@
 The totality of human experience can be categorized according to the ``sense doors''
-listed in Figure~\ref{fig:sense-doors} \cite{sense-icons}: The five external sense doors of sight, sound,
+listed in Figure~\ref{fig:sense-doors}: The five external sense doors of sight, sound,
 taste, smell, and touch are listed at the top. The internal sense door of ``mind'' is
@@
 physical discomforts, etc.) to the supramundane (out-of-body experiences,
-hallucinations, etc.) are subsets of these seven sense doors and their reflection in
-bodily sensation, the eighth.
+hallucinations, pronounced perceptual time dilation, etc.) are subsets of these seven
+sense doors and their reflection in bodily sensation, the eighth.
 
@@
 
-\begin{center}
-  \Huge{The Mundane and The Supramundane}
-\end{center}
+\section{The Mundane and The Supramundane}
 
@@
 meditating for reasons in the mundane field simply because most people have never
-experienced the supramundane. \textit{The Supramundane} in this context refers not to
-spirituality or religion but experiences which transcend the material, physical
-world.
+experienced the supramundane field. \textit{The Supramundane} in this context refers not to
+spirituality or religion but experiences which transcend the material, physical world.
 
@@
 meditation, these altered states of mind are never the goal of meditation. The goal
-of meditation is to change the meditator's mental habits, to move away from unhealthy
-mental patterns which cause harmful behaviours toward healthy mental patterns which
+of meditation is to change the meditator's mental habits. The intention is to move away from unhealthy
+mental patterns, which cause harmful behaviours, toward healthy mental patterns which
 encourage productive behaviour. Obviously this change is only visible in the mundane
@@
 
-\begin{center}
-  \Huge{Meditation vs. Naps}
-\end{center}
+\section{Meditation vs. Naps}
 
@@
   \item mundane sphere / productivity (21 lessons, seinfeld)
-  \item reset frame of reference outside oneself, outside one's own lifetime: ``trees for god'' and obvious karma (sidu/booga smoking)
   \item unlearning obsessive / repetitive thought, enhancing creativity
@@
   \item clarity: in thought, work, planning
-  \item clarification: ``isn't that what makes us human?'' (emotions) --- rather, what makes us animal
   \item Die Standing Up
-
 \end{itemize}
 
-\pagebreak
+\begin{itemize}
+  \item reset frame of reference outside oneself, outside one's own lifetime: ``trees
+    for god'' and obvious karma (sidu/booga smoking)
+  \item Time: Nat Friedman's blog post?
+  \item clarification: ``isn't that what makes us human?'' (emotions) --- rather, what makes us animal
+\end{itemize}
 
-\begin{thebibliography}{9}
-\raggedright
+\section*{Acknowledgements}
+
+Thank you to Preethi Govindarajan for reviewing this paper.
+
+\section*{References}
+
+\begin{thebibliography}{99}
 

### 2019-11-08 paper-four/paper-four.tex
tweaking paper four

@@
-\documentclass{article}
+\documentclass[a4paper, amsfonts, amssymb, amsmath, reprint, showkeys, nofootinbib, twoside]{revtex4-1}
 
+\usepackage[english]{babel}
+\usepackage[utf8]{inputenc}
+\usepackage[colorinlistoftodos, color=green!40, prependcaption]{todonotes}
+\usepackage{graphicx}
+\usepackage{subcaption}
+\usepackage{float}
+\usepackage[bottom]{footmisc}
 \usepackage{enumitem}
+\usepackage{hyperref}
 
@@
 \setlength{\parskip}{0.7em}
+
+\bibliographystyle{apsrev4-1}
 \setlist{noitemsep}
@@
 
-\begin{titlepage}
-   \vspace*{\stretch{1.0}}
-   \begin{center}
-     \Huge\textbf{Vipassana for Hackers}\\
-     \Huge{Paper Four: Mastering Spacetime}\\
-     \vspace{5cm}
-     \large\textit{Steven Deobald}\\
-     \large\textit{Version 0.1 DRAFT}\\
-     \large\textit\today\\
-     \vspace{5cm}
-     \large\textit{}\\
-   \end{center}
-   \vspace*{\stretch{2.0}}
-\end{titlepage}
+\title{%
+  \large{Vipassana for Hackers} \\
+  \Huge{Paper Four: Mastering Spacetime} \\
+  \large\textit{Version 0.1}
+}
+\author{Steven Deobald}
+\email[Correspondence email address: ]{steven@deobald.ca}
+\affiliation{vipassana-for-hackers.org}
+\date{\today}
 
-\begin{center}
-  \Huge{Target Audience}
-\end{center}
+\begin{abstract}
+ \todo{Write the abstract.}
+\end{abstract}
+
+% \keywords{}
+
+\maketitle
+
+\section{Target Audience}
+
+The target audience for this paper remains ``hackers'' for the broadest possible
+definition of that word. \todo{ref: from cherry?}
 
@@
 
-\pagebreak
+\section{Eka Maggo: The One and Only Path}
 
-\begin{center}
-  \Huge{Eka Maggo? Learning Vipassana Elsewhere}
-\end{center}
+\subsection{Learning Vipassana Elsewhere}
 
@@
 
-\pagebreak
+\subsection{How is Vipassana ``Complete''?}
 
-\begin{center}
-  \Huge{Concepts, Ideas, Help}
-\end{center}
+- 7 doors
+- TM says to use your imagination exclusively, with a mantra
+- Anapana variations say to use the breath (sensation) exclusively, with a limitation to a particular area
+- Void-based meditations (yoga, elsewhere) suggest you can divorce yourself of attention from any one of these phenomena… which is pretty unlikely for a beginner
+- “Sound of Silence” focuses on the ear sense door + tinitus
+- other forms of Vipassana focus entirely on thoughts
+- Zazen, generally, says open yourself to all 7 sense doors, all at once.
+... they all choose an object of meditation within these 7 doors.
+
+- mind / body === brain / nervous system
+
+- vipassana the largest superset
+
+- anapana covers time, vipassana covers space
+
+Anapana eliminates space.
+
+It says “focus here, on this tiny area… make the area as tiny as possible, with no
+limit or exception. Now all you have to do is follow your breath in the tiniest
+slices possible, making time stretch out longer and longer with every sub-slice.”
+
+Anapana is mastery over time.
+
+Vipassana is harder.
+It stops ignoring space and actually adds it to the equation.
+You move your attention throughout your entire body (which is the only thing you have direct access to: sensation) with the narrow focus you trained yourself to achieve in the first 3 days of the course.
+As your apparent physical existence begins to break apart, you’ll open up your attention to the whole body, all at once (or as much as possible at once) but even after 9 Vipassana courses, I rarely ever do anything approximating that because I’m quite a beginner.
+🙂
+
+Anyway.
+
+Vipassana is mastery over time + space.
+
+...once you have that, there’s nothing left.
+
+- Day 8: Meditate all the time.
+
+\section{Concepts, Ideas, Help}
 
@@
 
-\pagebreak
-
-\begin{center}
-  \Huge{Warnings}
-\end{center}
+\section{Warnings}
 
@@
 
+\section*{Acknowledgements}
+
+Thank you to Preethi Govindarajan for reviewing this paper.
+
+\section*{References}
+
+\begin{thebibliography}{99}
+
+\bibitem{sense-icons}
+  5 Senses by Daniel Falk from the Noun Project
+  \url{https://thenounproject.com/daniel2021/collection/human-body-senses/}
+  Thought by Nociconist from the Noun Project
+  \url{https://thenounproject.com/search/?q=thought&i=2025873}
+  Heart by Rafael Garcia Motta from the Noun Project
+  \url{https://thenounproject.com/search/?q=heart&i=807960}
+  Body by Makarenko Andrey from the Noun Project
+  \url{https://thenounproject.com/search/?q=body&i=789989}
+  \textit{The Noun Project}.
+
+\bibitem{dhamma}
+  Vipassana International Academy
+  \url{https://www.dhamma.org}
+  \textit{Vipassana Meditation Website}
+
+\end{thebibliography}
+
 \end{document}

### 2019-11-21 the-proposal/the-proposal.tex
first commit on the proposal... here we go.

--- 2019-10-09 paper-two/paper-two.tex
@@
 \documentclass[a4paper, amsfonts, amssymb, amsmath, reprint, showkeys, nofootinbib, twoside]{revtex4-1}
+
 \usepackage[english]{babel}
@@
   \large{Vipassana for Hackers} \\
-  \Huge{Paper Two: The Brain} \\
-  \large\textit{Version 0.2}
+  \Huge{The Proposal} \\
+  \large\textit{Version 0.1}
 }
-\author{Preethi Govindarajan}
-\email[Correspondence email address: ]{preethi@deobald.ca}
+\author{Steven Deobald}
+\email[Correspondence email address: ]{steven@deobald.ca}
 \affiliation{Siggu.org}
@@
 \begin{abstract}
-  Two years ago, immediately following my first Vipassana course, I began researching
-the effects of meditation on the brain. This paper is a summary of the research
-available on the topic and speculations, based on my experience, as to what is
-happening in the brain during a 10-day silent Vipassana meditation retreat.
+I want to turn myself into a human guinea pig.
 \end{abstract}
@@
 
-\listoftodos
+% \listoftodos
 
-\section{Introduction}
+\section{Why Vipassana?}
 
-I took my first Vipassana course (as taught by S.N. Goenka) in 2017.
-Vipassana meditation was so unlike anything I had ever experienced before
-I was left extremely curious about what exactly had happened
-to me during those ten days. For months afterward I spent my mornings and
-evenings wading through the research in the field of meditation. I was specifically
-focused on white papers dealing with the effects of a 10-day Vipassana course on
-the brains of participants. The research in this area is limited.
-The quality research which does exist usually uses a sample of highly
-experienced meditators \cite{tibetanmonks} rather than beginners and/or self-reports rather than
-objective measures.
+zig zig zig
 
-Over the past year, I have tried to write down what I experienced during that first
-10-day course and the 10-day courses I have taken since, corroborating my experience
-with research that does exist regarding meditation and the brain.
+\section*{Acknowledgements}
 
-\textbf{Proviso:} S.N. Goenka, the principal teacher of Vipassana meditation,
-actively dissuades students from precisely the sort of brain-centred biological
-inquiry presented in this paper.
+Thank you to Steven Deobald for reviews, edits, and corrections. \texttt{Brain\_human\_normal\_inferior\_view.svg} and
+\texttt{Brain\_human\_lateral\_view.svg} by Patrick J. Lynch, medical illustrator, CC-BY 2.5 \cite{brainsvg}
 
-\begin{quote}
-  \textbf{The brain itself is just a physical organ. As you deal with other parts of
-    the body, you deal with the brain in the same way, that's all. Nothing special to
-    do with the brain. But the mind is totally different. In the West, all importance
-    is given to the brain as if the mind is located here. Nothing doing, it is
-    everywhere. The mind is in the whole body. So give attention to the whole
-    body.} --- S.N. Goenka \cite{goenkabrain}
-\end{quote}
-
-This paper does not contradict Mr. Goenka's sentiment, but instead acts as a starting
-point for curious readers who, whether they have taken a Vipassana course or not,
-view the function of the brain as central to the activity of the mind.
-
-\textbf{Disclaimer:} Although my primary field of research is within the field of
-biology, my work is far removed from neuroscience. I have tried to simplify the research available so as to better
-understand it. If there are any corrections to make or editing in terms of the
-content, please feel free to contact me.
-
-\section{Brain Function and Anatomy}
-
-Before dissecting the experience of meditation as it pertains to the brain, this
-section describes the different parts of the brain that have shown up in
-scientific literature as correlated to meditative practices. Recent neuroscience divides the
-brain into the reified geography of the brain (anatomy) and the abstract concepts
-governed by brain activity (function). Between concrete physiology and abstract
-functional outcomes exist networks of cooperative structures which correspond to general
-high-level activities of the brain.
-
-The Default Mode Network (DMN) is the constellation of regions which fire when
-the brain is not engaged in any external or goal-oriented
-tasks. As the name suggests, the brain is in its ``default
-mode'' of operation. \cite{defaultnetworkadaptive} The DMN is anticorrelated to the Central Executive
-Network (CEN), which is active during externally-directed, high-level cognitive
-functions. \cite{saliencenetwork} This anticorrelation between the DMN and the CEN is
-governed by the Salience Network (SN), the collection of regions in the brain which
-help decide which stimuli deserve our attention. In this way, the SN acts as a switch between the
-internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}
-
-Where possible in the following sections these networks are described in terms of
-function. The description of the DMN is also organized by function but the subsystems
-of the CEN and SN offer too many potential functional categorizations and are instead
-organized by anatomy.
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=7.5cm]{images/fmri-switching.jpg}
-  \caption{The Salience Network consists of the Anterior Cingulate Cortex and the
-    Anterior Insular Cortex. It helps switch between the Default Mode Network and the
-  Central Executive Network.}
-  \label{fig:fmri-switching}
-\end{figure}
-
-\subsection{Default Mode Network}
-
-The DMN is comprised of specific anatomy including portions of the mid-line of the
-brain, an evolutionarily primitive area related to memory and emotion, and
-structures in the cortex, an evolutionarily recent part of the brain containing the
-executive and higher order functions. \cite{defaultnetworkanatomy}
-
-These functionally connected regions are involved in the neurological basis
-of the self, considering the mental states of others, remembering the past, and
-imagining the future.
-
-This network is mostly observed through changes in blood flow to different parts of
-the brain measured using fMRI (functional Magnetic Resonance Imaging) or
-PET (Positron Emission Tomography). \cite{defaultnetworkadaptive}
-
-\begin{figure}[h!]
-  \centering
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/top-dmn.png}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/side-dmn.png}
-  \end{subfigure}
-  \caption{Highlighted: approximate locations of brain regions firing when the Default Mode Network is active.}
-  \label{fig:dmn}
-\end{figure}
-
-\subsubsection{The neurological basis of the self}
-
-The regions involved in an individual's conceptual ``self'' are the Posterior
-Cingulate Cortex (PCC), Ventromedial Prefrontal Cortex (vmPFC), and the Inferior
-Parietal Lobule (IPL). Observationally, these regions are preferentially activated when
-subjects were asked to reflect on personal preferences, beliefs, abilities, or
-any other qualities relevant to the self. \cite{defaultnetworkadaptive} Cerebral
-blood flow and metabolic rate are 40\% higher than average in the PCC,
-making it one of the most active regions of the brain. \cite{pccrole}
-
-The PCC has been thought to mediate interactions between emotion and memory.
-Under an fMRI, it consistently lights up when a person recollects an event from their
-life. The strength of this activity varies depending on the emotional importance of
-the memory recalled. It is also activated by emotional stimuli,
-both positive and negative, and acts as a navigational interface, between the individual
-and the external world, by gauging the importance of different stimuli. \cite{pccemotion}
-
-The vmPFC is involved in creating a conceptual self by self-related processing
-and the assignment of personal significance to self-related
-information. \cite{dmnself} The vmPFC is paired with the Dorsomedial Prefrontal
-Cortex (dmPFC), which assists in the creation of the conceptual self through an
-individual's consideration of ``others''. Together, the vmPFC and dmPFC form the
-Medial Prefrontal Cortex (mPFC).
-
-The Inferior Parietal lobe (IPL) is situated in the back of the brain and it is
-involved in the perception of facial stimuli. \cite{ipl}
-The IPL has been hypothesized as being involved in distinguishing self from others,
-in a study where subjects were asked to retrieve
-previous judgments about oneself, their friend and the Dutch queen,
-the activation of the IPL was seen to decrease with decreasing self-reference. \cite{iplself}
-Another study also used Transcranial Magnetic Stimulation (TMS) to create a sort of "virtual lesion"
-(transiently disturb the neural circuitry in that area) of
-the right-IPL and showed that this disrupts self-other discrimination. \cite{ipltms}
-
-It is hypothesized that the activation of these three regions of the brain (PCC,
-vmPFC, and IPL) are responsible for providing a sense of self --- a subjective agent
-in space and time. It functions as a network in which phenomena such as self
-representations are accessible via the PCC \cite{mappingself} and are gated into
-conscious awareness by activity in the mPFC, as influenced by changing internal
-and external demands. The PCC acts as a brain-wide connectivity hub, through which a unitary notion of self is
-created by a large scale integration of DMN activity. \cite{mappingself}
-
-\subsubsection{Thinking about others}
-
-The \textit{Theory of Mind} is a body of work in philosophy and psychology which
-describes an individual's ability to consider the mental states of others, such as
-appreciating another individual's false beliefs (knowledge based on incorrect or
-outdated information). The Theory of Mind captures not only one's ability to
-attribute beliefs, desires, and emotions to others but also to oneself --- and to
-differentiate the two. \cite{autistictheoryofmind}
-
-The DMN (especially the dmPFC subsystem) is engaged when individuals reflect upon,
-evaluate, or appraise their own mental states or the mental states of others. ``Mental
-states'' include thoughts, feelings, and desires. \cite{theoryofmind} In addition to
-consideration of these states, the dmPFC is involved in empathy, moral reasoning,
-and altruistic behavior. \cite{defaultnetworkadaptive,dmpfcothers,dmpfcaltruism}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=5cm]{images/fmri-dmn.png}
-  \caption{fMRI scan of the Default Mode Network.}
-  \label{fig:fmri-dmn}
-\end{figure}
-
-\subsubsection{Autobiographical memory and future simulations}
-
-The term "autobiographical memory" refers to our memory for specific episodes,
-and to our conceptual, generic, and schematic knowledge of our lives.
-
-The Medial Temporal Lobe (MTL) which is found on the medial surface of the temporal lobe.
-The critical role of the MTL in long term memory is well documented. \cite{mtl}
-
-The MTL consists of the hippocampal formation on top and the parahippocampal gyrus below it as well as
-the entorhinal and perirhinal cortices.
-The MTL, amygdala, the olfactory cortex, hypothalamus and several other
-anatomically connected areas forms the limbic system. It appears to be primarily responsible for
-our emotional life and has a specific involvement in the formation of memories. \cite{limbic}
-
-The MTL along with mPFC, IPL and PCC forms the Medial Temporal Subsystem (a part of the DMN).
-This subsystem is triggered during autobiographical memory retrieval,
-contextual association, and semantic knowledge. \cite{defaultnetworkadaptive}
-
-It has been hypothesized that the adaptive role of memory retrieval is to facilitate
-construction of episodes to prepare for immediate and distant future scenarios. \cite{defaultnetworkadaptive}
-The DMN is therefore involved in remembering the past, imagining the future, and story
-comprehension. \cite{defaultnetworkadaptive}
-
-\subsubsection{Summary}
-
-The DMN plays a role in constructing a sense of self, in memory,
-and in thinking about others. Because of this, it is also a key component in excessive
-rumination and anxiety. \cite{dmndepression}
-
-When a meditator sits down to meditate and thoughts wander away to past experiences
-or anxieties about the future, classic meditation texts and instructions refer to
-this behaviour as ``Monkey Mind''. It is the DMN playing the most active role in this
-distraction.
-
-\subsection{Central Executive Network}
-
-The CEN is activated when high-level cognitive tasks or external goal-oriented tasks
-are being performed. These tasks or executive functions are cognitive processes involved in
-cognitive control of behavior. This depends on three types of brain functions:
-working memory, mental flexibility, and self-control. The regions of the brain
-involved in these executive functions are the Dorsolateral Prefrontal Cortex
-(DL-PFC), Orbito Frontal Cortex (OFC), and the Posterior Parietal Cortex (PPC).
-
-\begin{figure}[h!]
-  \centering
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/top-cen.png}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/side-cen.png}
-  \end{subfigure}
-  \caption{Highlighted: approximate locations of brain regions firing when the Central Executive Network is active.}
-  \label{fig:cen}
-\end{figure}
-
-\subsubsection{Dorsolateral Prefrontal Cortex}
-
-The DL-PFC is a part of the Prefrontal Cortex found in primates, including
-humans. The DL-PFC is involved in higher cognitive processes including working memory
-(holding different pieces of information, manipulating them, and using them for
-tasks), selective attention, cognitive flexibility (switching between tasks) and
-planning. It also seems to be involved in social cognition and lying. The DL-PFC has
-also been seen to increase dopamine levels in the brain. \cite{dlpfcmemory,dlpfctasks,dlpfclying}
-
-\subsubsection{Orbito Frontal Cortex}
-
-The OFC is an area found in front of both hemispheres of the brain, just above the
-eyes. It is again a part of the Prefrontal Cortex and is thought to be involved in
-decision making through emotion and reward. It also receives input from multiple
-sensory modalities and in turn activates the Amygdala and the
-Hypothalamus. \cite{theprefrontalcortex,ofcprimates,theorbitofrontalcortex}
-
-\subsubsection{Posterior Parietal Cortex}
-The PPC is a region of the Parietal Cortex, physically located behind the primary
-Somatosensory Cortex. The PPC is responsible for facilitating higher-order
-functions. Higher-order functions are defined as those requiring diverse information
-for the purposes of higher evolutionary behaviour: intelligence, memory, planning,
-speech, orientation, decision-making, etc.
-
-To do this, the PPC receives input from auditory, visual, and somatosensory
-systems, integrates this input, and uses the aggregate to activate the DL-PFC and motor cortex and is
-involved in the performance of attention related tasks as well as higher-order motor
-tasks such as grasping and catching. \cite{parietallobesa,parietallobes}
-
-\subsubsection{Sidenote: Somatosensory Cortex}
-The Primary Somatosensory Cortex is located in a ridge of cortex found within the Parietal Cortex.
-It is responsible for processing somatic sensations. These sensations arise from receptors
-positioned throughout the body. They are responsible for detecting touch, for proprioception
-(the position of the body in space), nociception (pain), and temperature. \cite{somato}
-When such receptors detect one of these sensations, the information is sent first to
-the Thalamus and then to the Primary Somatosensory Cortex.
-
-The Sensory Homunculus which is a part of the Somatosensory Cortex is a cortical representation
-of the body based on the degree of sensory innervation. \cite{sensoryhom}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=5cm]{images/side-somatosensory.png}
-  \caption{Location of the Somatosensory Cortex.}
-  \label{fig:side-somatosensory}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=7.5cm]{images/homunculus.jpg}
-  \caption{The Sensory Homunculus mapped across the Somatosensory Cortex.}
-  \label{fig:homunculus}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=5cm]{images/fmri-cen.png}
-  \caption{fMRI scan of the Central Executive Network.}
-  \label{fig:fmri-cen}
-\end{figure}
-
-\subsubsection{Summary}
-
-The CEN is involved in goal-directed behavior and processes
-related to attention. These goals are usually external and attention is also directed
-externally (regardless of whether attention is voluntary or not). When you meditate,
-attention is directed both internally and voluntarily, to the same sensory cues that
-are normally directing the PPC.
-
-\subsection{Salience Network}
-
-\begin{quote}
-  \textbf{Salience: The perceptual quality by which an observable thing stands out
-    relative to its environment.}
-\end{quote}
-
-The SN is an intrinsically connected large-scale network anchored in the Anterior
-Insular Cortex (AIC) and Dorsal Anterior Cingulate Cortex (ACC). Both regions have reached a
-high degree of specialization in the great apes. It is the collection of the regions
-in the brain that help decide which stimuli deserves our attention. It acts as a
-switch between the internally-directed DMN and the externally-directed CEN. \cite{saliencenetwork}
-
-\begin{figure}[h!]
-  \centering
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/top-sn.png}
-  \end{subfigure}
-  \begin{subfigure}[b]{0.48\linewidth}
-    \includegraphics[width=\linewidth]{images/side-sn.png}
-  \end{subfigure}
-  \caption{Highlighted: approximate locations of brain regions firing when the Salience Network is active.}
-  \label{fig:sn}
-\end{figure}
-
-\subsubsection{Anterior Cingulate Cortex}
-
-The ACC is the front end of the Cingulate Cortex and collars around the Corpus
-Callosum, the band connecting the two hemispheres of the brain. It is the connector
-between the emotional (Limbic System) and the cognitive (Prefrontal Cortex) part of
-the brain. It is involved in functions such as attention allocation, reward
-anticipation, decision making, morality, impulse control, emotional awareness and
-registering pain. \cite{accstroop,accreward,snmorality,empathypain,acccognitive} It
-also appears to play a role in the regulation of Autonomic functions such as blood
-pressure and heart rate. \cite{accbloodpressure}
-
-\subsubsection{Anterior Insular Cortex}
-
-The AIC is a part of the Cerebral Cortex located deep within the Sulcus, the
-fissure separating the four lobes of the brain. The AIC physically projects itself
-into the Amygdala. It is involved in multimodal sensory processing such as
-audio-visual integration tasks, interoceptive awareness (so its activity is directly
-related to an individual's sense of internal body states), empathy, and conscious
-awareness. \cite{aicemotion}
-
-It also plays a role in the regulation of autonomic functions such as bodily
-sensations (including judgement of the severity of pain), taste, and control of the
-immune system. \cite{aicautonomic}
-
-The AIC and ACC together give rise to our interoceptive and conscious
-self-awareness.  \cite{selfaware}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=5cm]{images/fmri-sn.png}
-  \caption{fMRI scan of the Salience Network.}
-  \label{fig:fmri-sn}
-\end{figure}
-
-\subsubsection{Summary}
-
-The SN is directly involved in switching between the DMN and CEN. When you meditate,
-you cultivate the cognitive processes related to attention.
-When the meditator's mind wanders, she notices it and brings it back.
-The SN is directly involved in this process of assigning attention, since it is
-responsible for the prominence of any object of attention.
-Meditation, or the act of returning to an object of meditation, can therefore be
-thought of as an exercise in strengthening the SN and its ability to perform this switch, over time.
-
-\section{The Course}
-
-I applied for my first 10-day Vipassana course after I quit my job. From the time I
-was accepted into the course until the first day of the course I was very nervous
-about what it might entail. I had never done anything like meditation before
-and I was quite happy with the idea that meditation, with its spiritual connotations
-and religious mumbo-jumbo, was a waste of my time. In my mind, a silent meditation retreat was
-for people who had time to waste --- and I was not one of them. Yet friends had convinced me that I
-should give Vipassana a try and in the interval between workplaces I had
-time to experiment with a course.
-
-The following are my observations and speculations about what could be happening
-in the brain during a 10-day course.
-
-I went for my first course in Chennai, my home town. When I arrived, I had to fill
-the application forms all over again (despite applying online), keep all my
-luggage in a locker, and go to my room. I had to share the room with one other
-student. I felt quite out of my element and I was certain everyone else could see
-that I was the odd one out. In my head, I was not supposed to be there. Because I felt
-so out of place, I was too scared to actually make conversation with anyone before
-silence was enforced.
-
-\subsection{Anapana Meditation}
-
-And so it was. With my brain filled with thoughts about myself, about the people
-around me, I started this meditation business. For the first 3.5 days I was instructed
-to focus my attention on the area below my nostrils and above my upper
-lip. That's all. These are undoubtedly very simple instructions but in practice I
-found them very difficult. Maintaining my focus was much harder than I had
-anticipated. Within the first two days, I had made up elaborate stories about my
-fellow meditation students. In these fantasies all of the other students were
-superheroes, tirelessly working to save humanity together. My Default Mode Network
-was in overdrive.
-
-Focusing our attention below the nostrils starts with observing breath --- coming in
-and going out. As the teacher mentions in the late evening discourses, the reason for
-starting with your breath is because this the only activity of the body which is both
-conscious and unconscious --- it acts as a bridge to the unconscious mind and the
-involuntary processes of the body.
-
-There has been research in mice, showing a cluster of nerves called the
-pre-Bötzinger Complex (preBötC), found in the brain stem of most mammals, which fires
-with every breath taken. This breathing pacemaker seems to work not only for regular
-conscious and unconscious breathing but for all kinds of breathing --- such as yawns,
-sighs and gasps. The preBötC also appears to play a role in calming and
-arousal. \cite{prebotcgeneration} It would be interesting to see the activation in
-these neuron clusters during the first three days of a Vipassana course, even among
-novice meditators.
-
-During the meditation hours I was trying hard. I had decided that if I was to examine this practice
-empirically, I had to give it an honest shot. I focused on the small patch beneath my
-nostrils above my upper lip. Initially, while trying to observe one's breathing, the
-mind will wander --- not just to fantasies of superheroes but to almost any object or
-daydream imaginable --- as the mind wanders the Default Mode Network is working. With
-the realization that the mind has wandered, and the consequent return of attention
-back to the breath, the Salience Network is activated.
-This happens slowly at first. One's mind wanders away for many, many minutes before
-realizing that the object of attention, the breath, has been lost. But
-at some point within the 3.5 days, the SN learns to bring attention back to the breath. After a few days, this
-refocusing almost happens by habit, almost automatically.
-
-As I gave in to this activity, tried to focus more intently, and tried to sit still for
-longer continuous stretches of time, there were longer periods of awareness on that
-patch of skin. With these longer periods of awareness came
-stranger sensations arising and passing within that physical area. By the third day,
-it felt like entire ecosystems were writhing and flopping and crashing, all of them
-very alive in the area below my nostrils above my upper lip. Since Anapana meditation
-is very simple and these strange sensations appeared within a few days of my practice
-as an absolute beginner, it would be very interesting (and relatively easy) to study
-the neurological and physiological basis for such sensations.
-
-There have been studies that have shown that once subjects bring their
-attention back to breath, the neural structures involved in the control
-of the Autonomic Nervous System and attention start firing more actively
-(in this case, the DL-PFC --- a part of the CEN, the Hippocampus --- a part of the
-DMN, and the ACC --- a part of the SN). During Anapana there are also global
-dampening changes seen in the brain, particularly within the DMN and the Amygdala, which is involved in
-the flight-or-fight (stress) response. These changes have been
-termed the ``relaxation response'', as they are antagonistic to the stress
-response. This relaxation response can be thought of as a gateway to altered states
-of mind. \cite{relaxationresponse}
-
-\subsection{Vipassana Meditation}
-
-After lunch on the fourth day the instructions for Vipassana are given. During
-Vipassana, students transfer their attention from the patch below their nostrils to the top
-of the head. From there, for two hours, instructions are given to slowly move
-one's attention throughout the entire body: ``from the top of the head to the tips of
-the toes.'' This movement of attention throughout the body is then repeated, over and
-over. The two objectives while observing bodily sensations are focused attention (making
-use of the narrow focus practiced during the Anapana period) and open monitoring
-(observing the sensations objectively).
-
-As this process progresses, the Salience Network (ACC and AIC) and some parts of
-the Central Executive Network (DL-PFC, OFC, and PPC), 
-which are involved in the control of attention, fire more actively. The anti-correlation
-between the DMN and CEN reduces. Usually when a person is engaged in an external task, 
-the OFC and PPC receive signals from all the sensory networks (Visual, Audio, Somatosensory, etc.) and, 
-in response, continuously send signals to the different motor cortices, Amygdala, and Hypothalamus to
-respond by performing a task, feeling an emotion, etc.
-
-But Vipassana is an internal task. The Salience Network is activated by focused
-attention, which in turn activates the Central Executive Network. Rather than an
-external task, as the CEN is accustomed to, the conscious instruction is to observe
-bodily sensation --- and do nothing. So as you sit in silence with your eyes closed,
-not moving your body physically but instead moving throughout your body with your
-attention, the CEN is doing what it has always done. However, the outcomes of
-activity in the CEN are different during Vipassana. Instead of focusing on
-external sensory input the sensory input is internal (bodily sensations) and instead
-of reacting to input by performing tasks or feeling emotions the goal is to not
-react, to simply observe as objectively as possible.
-
-This is hard at first, with the body involuntarily reacting by jerking and writhing
-in involuntary response to the very act of observation. Initially, even awareness
-itself has a jerky quality and it is hard to observe sensations consciously. There is also the
-pain. This eclipses all other sensations. In my case, it was difficult to maintain
-awareness of any other sensation with the pain that was emanating from my legs and
-back. The OFC is involved in assigning emotion to sensation and the shaping of pain by expectations
-\cite{ofcemotion,ofcexpect} and the ACC plays a vital role in perception of sensation. 
-These two parts of the brain are usually functionally connected. 
-In the beginning, as I sat trying to observe sensations and was only aware of the pain, 
-the ACC was firing rapidly and the OFC then assigned these signals to the Limbic System, which
-put me in a less than ideal emotional state.
-
-Over the final seven days of the course, as awareness gets smoother and observations get more objective, this
-functional connectivity between the ACC and the OFC slowly starts changing. While
-continuing to pay attention to constantly changing bodily sensations the goal remains: avoid
-assigning evaluations (and therefore, emotions) to them. Without assigning an
-emotional response, one learns to observe with a
-non-judgmental attitude, particularly toward unpleasant stimuli. The ACC is still up-regulated 
-but the OFC and the Limbic System may acquiesce and, in turn, reactions become smaller and less
-frequent. This snowball effect almost mirrors the spiral of brain behaviours
-exhibited early in the course. Early in the course, the CEN's normal mode of
-operation is in play, which is to experience unpleasant stimuli and react (desiring to rid the body of
-them), which in turn causes the unpleasant stimuli to appear all the more
-unpleasant. Once a measure of objectivity is reached, however, the brain's function
-snowballs in the other direction. It is as if both conscious and unconscious awareness-es are slowly aligning
-until there are occasions of experience when it feels as if there is nothing but detached awareness moving
-through the body (or some parts of the body). This is not at all akin to the
-detachment of sleep or intoxication, however, and the senses are not dulled. In
-addition to a feeling of emotional detachment, the ACC, AIC and the Somatosensory
-Cortex, which are involved in sensory awareness, were \textbf{more} active, not
-less. As the OFC, and Limbic System are calmed and attention is narrowed
-further, the sensations become progressively more intense.
-
-Perhaps the strangest experience during a Vipassana course is that of shifts in the
-sense of self. The ``self'', as a subjective agent in space and time, changes. At
-first, this change is only very slight. As a student closes her eyes and begins to
-meditate, it may feel as though the body has shifted in space or rotated in different direction, this is
-due to the temporary down regulation of the prefrontal cortex. As the meditations get deeper the DMN
-occasionally quiets completely (perhaps only for a short duration) while the AIC
-fires more rapidly, leading to alternative networks of consciousness
-arising. \todo
-
-[Side note; Consciousness is defined as our subjective awareness of ourselves and our environment. 
-An altered state of consciousness is defined as alternate patterns or configurations of 
-this experience, which differ qualitatively from a baseline state.
-Altered states of consciousness include a state of dreaming, epileptic seizures, sensory deprivation, 
-hysterical states of dissociation and depersonalization as well as pharmacologically induced altered states.
-
-The entropic brain hypothesis: This was put forth by Robin Carhatt Harris et al, while studying 
-the psychedelic state. Entropy is a measure of the uncertainty in a system and in this case applied 
-to the states of consciousness and their associated neurodynamics. The theory makes a distinction 
-two fundamental modes of consciousness: Primitive consciousness and the normal waking or secondary
-consciousness. 
-Primitive consciousness is characterized by higher entropy and it is hypothesized to have 
-preceded the development of modern, adult, human, normal waking consciousness. 
-It is argued that suppression of this entropy facilitates normal waking consciousness 
-and associated metacognitive functions such as reality-testing and self-awareness.]
-
-Sometimes it feels like loosening the hold of one's own narrative leads to new insights that are
-usually kept from consciousness. It has been found that experiencing this, even for a
-few seconds, can lead to lasting changes long after the activity of meditation itself
-has stopped. \cite{alteredtraits} In the moment, these experiences can be quite
-similar to being on certain hallucinogenic drugs or even, in some cases, to ecstatic
-seizures where the abnormal activity of the Anterior Insular Cortex leads to
-heightened self-awareness, feelings of bliss, and a lack of ambiguity. \cite{cortexbliss}
-
-\section{Conclusion}
-
-I believe attending a 10-day Vipassana course did, in fact, change the habit pattern
-of my mind --- at least for a while.
-
-Vipassana causes global dampening changes in the DMN, causes certain parts of
-the SN and CEN to activate. That is, reacting parts of the network and
-portions of the network which are aware of bodily sensation, respectively.
-This changes a person's behavior and these changes can often be seen as soon as
-someone leaves the course on the tenth day.
-
-But repeated practice is essential for any sustained neuroplasticity since rewriting
-many years of habit formation requires more than a ten day course. However, the
-10-day course is both progressive and systematic. It is designed to give someone
-experimenting with meditation the minimum observational change necessary to instill
-an interest, to keep practicing.
-
-There have been studies looking at both the structural changes in the brain and
-functional changes in brain activity, especially to the DMN and the SN. But such
-studies are often only done on meditators who have clocked thousands of hours of
-practice, which diverges from the purpose of this paper. \cite{tibetanmonks,
-  experiencemed, experiencemeda}
-
-How is Vipassana different from simply concentrating on a task, as hard as one can?
-When concentrating on any task, your DMN is quieted, the SN and CEN are firing, and you
-enter into a state of flow. But flow or no flow, the reacting part of the brain is
-also working. The mind may learn to focus intensely but the habit patterns of the
-mind are unlikely to change. The goal of Vipassana is not simply to concentrate or to
-enter a state of flow, but to break old reactionary habit patterns altogether.
-
-As I left the Vipassana Center, I was really intrigued as to what exactly had
-occurred over those ten days. What was it that had changed in my brain? I had to
-know. I very quickly noticed, however, that as I left the centre, those very changes
-started slipping away. But with each hour-long meditation at home, a tiny bit sticks
-just a little more. And I get closer to my answer.
-
-\section{Acknowledgements}
-
-Thank you to Steven Deobald for reviews, edits, and corrections.
-
-``Brain\_human\_normal\_inferior\_view'' By Patrick J. Lynch, medical illustrator, CC BY
-2.5, https://commons.wikimedia.org/w/index.php?curid=1496648
-
-``Brain\_human\_lateral\_view.svg'' By Patrick J. Lynch, medical illustrator, CC BY
-2.5, https://commons.wikimedia.org/w/index.php?curid=1496651
+\section*{References}
 
@@
 
-\section*{References}
-
-\bibitem{tibetanmonks}
-  Alex Hankey
-  \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1697747/}
-  \textit{Studies of Advanced Stages of Meditation in the Tibetan
-          Buddhist and Vedic Traditions. I: A Comparison of General Changes}
-  Evid Based Complement Alternat Med. 2006 Dec; 3(4): 513–521.
-
-\bibitem{goenkabrain}
-  S.N. Goenka.
-  \url{https://www.vridhamma.org/A-store-house-of-answers-by-Shri-S-N-Goenka}
-  \textit{Answers by Mr. S. N. Goenka}
-
-\bibitem{defaultnetworkadaptive}
-  Andrews-Hanna, Jessica R.
-  \textit{The brain’s default network and its adaptive role in internal mentation.}
-  The Neuroscientist: A Review Journal Bringing Neurobiology, Neurology and
-  Psychiatry. 18 (3): 251–270. doi:10.1177/1073858411403316. ISSN 1089–4098. PMC
-  3553600. PMID 21677128. (2012–06–01).
-
-\bibitem{defaultnetworkanatomy}
-  Buckner, R. L.; Andrews-Hanna, Jessica R.; Schacter, D. L.
-  \textit{The Brain's Default Network: Anatomy, Function, and Relevance to Disease.}
-  Annals of the New York Academy of Sciences. 1124 (1): 1–38. CiteSeerX
-  10.1.1.689.6903. doi:10.1196/annals.1440.011. PMID 18400922, 2008
-
-\bibitem{saliencenetwork}
-  Menon, V; Toga, A.
-  \textit{Salience Network.}
-  Elsevier. pp. 597–611, 2015
-  ISBN: 978–0–12–397316–0.
-
-\bibitem{pccrole}
-  Leech R, Sharp DJ
-  \textit{The role of the posterior cingulate cortex in cognition and disease.}
-  Brain. 137 (Pt 1): 12–32., July 2013.
-
-\bibitem{pccemotion}
-  Maddock, Richard J.; Garrett, Amy S.; Buonocore, Michael H.
-  \textit{Posterior cingulate cortex activation by emotional words: fMRI evidence from a
-    valence decision task.}
-  Human Brain Mapping. 18 (1): 30–41. January 2003.
-
-\bibitem{dmnself}
-  Andrews-Hanna, Jessica R.; Smallwood, Jonathan; Spreng, R. Nathan.
-  \textit{The default network and self-generated thought: component processes,
-    dynamic control, and clinical relevance.}
-  Annals of the New York Academy of Sciences. 1316(1):
-  29–52. doi:10.1111/nyas.12360. ISSN 1749–6632. PMC 4039623. PMID 24502540. 2014-05-01.
-
-\bibitem{ipl}
- Radua J, Phillips ML, Russell T, Lawrence N, Marshall N, Kalidindi S, El-Hage W, McDonald C,
- Giampietro V, Brammer MJ, David AS, Surguladze SA.
- \textit{Neural response to specific components of fearful faces in healthy
-  and schizophrenic adults.}
-  Neuroimage. 2010 Jan 1;49(1):939-46. doi: 10.1016/j.neuroimage.2009.08.030.
-  Epub 2009 Aug 20.
-  \url:https://www.ncbi.nlm.nih.gov/pubmed/19699306
-
-\bibitem{iplself}
- Hans C. Lou, Bruce Luber, Michael Crupain, Julian P. Keenan, Markus Nowak,
- Troels W. Kjaer, Harold A. Sackeim, Sarah H. Lisanby
- \textit{Parietal cortex and representation of the mental Self}
-  Proc Natl Acad Sci U S A. 2004 Apr 27; 101(17): 6827–6832.
-
-\bibitem{ipltms}
- Lucina Q. Uddin, Istvan Molnar-Szakacs, Eran Zaidel, Marco Iacoboni
- \textit {rTMS to the right inferior parietal lobule disrupts self–other discrimination}
-  Social Cognitive and Affective Neuroscience, Volume 1, Issue 1, June 2006,
-  Pages 65–71, https://doi.org/10.1093/scan/nsl003
-
-\bibitem{mappingself}
-  Davey CG, Pujol J, Harrison BJ.
-  \textit{Mapping the self in the brain’s default mode network.}
-  Neuroimage. 132:390–397., 2016-05-15
-
-\bibitem{autistictheoryofmind}
-  Baron-Cohen, Simon; Leslie, Alan M.; Frith, Uta.
-  \textit{Does the autistic child have a ``theory of mind''?.}
-  Cognition. 21 (1): 37–46. doi:10.1016/0010-0277(85)90022-8. PMID
-  2934210. Pdf. October 1985.
-
-\bibitem{theoryofmind}
-  Frank Van Overwalle
-  \textit{A dissociation between social mentalizing and general reasoning}
-  NeuroImage. Volume 54, Issue 2, 15 January 2011, Pages 1589-1599
-  \url{https://www.sciencedirect.com/science/article/abs/pii/S1053811910012243}
-
-\bibitem{limbic}
-  Catani M, Dell'Acqua F, Thiebaut De Schotten M
-  \textit{A revised limbic system model for memory, emotion and behaviour}
-  Neuroscience and Biobehavioral Reviews. 37 (8): 1724–37.
-  doi:10.1016/j.neubiorev.2013.07.001. PMID 23850593.
-
-\bibitem{dmpfcothers}
-  Isoda, M., \& Noritake, A.
-  \textit{What makes the dorsomedial frontal cortex active
-    during reading the mental states of others?.}
-  Neural basis of social learning, social deciding, and other-regarding preferences,
-  51. 2015.
-
-\bibitem{dmpfcaltruism}
-  Waytz, A., Zaki, J., \& Mitchell, J. P.
-  \textit{Response of dorsomedial prefrontal cortex predicts altruistic behavior.}
-  The Journal of Neuroscience, 32(22), 7646–7650. 2012.
-
-\bibitem{mtl}
- Scoville WB, Milner B
- \textit{Loss of recent memory after bilateral hippocampal lesions.}
- J Neurol Neurosurg Psych,  1957;20:11–21.
-
-\bibitem{dmndepression}
-  J. Paul Hamilton, Madison Farmer, Phoebe Fogelman, and Ian H. Gotlib.
-  \textit{Depressive Rumination, the Default-Mode Network,
-          and the Dark Matter of Clinical Neuroscience}
-  Biol Psychiatry. 2015 Aug 15; 78(4): 224–230. doi: 10.1016/j.biopsych.2015.02.020
-
-\bibitem{accstroop}
-  Pardo JV, Pardo PJ, Janer KW, Raichle ME.
-  \textit{The anterior cingulate cortex mediates processing selection in the Stroop
-    attentional conflict paradigm.}
-  Proceedings of the National Academy of Sciences of the United States of
-  America. 87 (1): 256–9. doi:10.1073/pnas.87.1.256. PMID 2296583. January 1990.
-
-\bibitem{accreward}
-  Bush G, Vogt BA, Holmes J, Dale AM, Greve D, Jenike MA, Rosen BR.
-  \textit{Dorsal anterior cingulate cortex: a role in reward-based decision making.}
-  Proceedings of the National Academy of Sciences of the United States of
-  America. 99 (1): 523–8. doi:10.1073/pnas.012470999. PMC 117593. PMID
-  11756669. January 2002.
-
-\bibitem{snmorality}
-  Sevinc G, Gurvit H, Spreng RN.
-  \textit{Salience network engagement with the detection of morally laden
-    information.}
-  Social Cognitive and Affective Neuroscience. 12 (7):
-  1118–1127. doi:10.1093/scan/nsx035. PMID 28338944. July 2017.
-
-\bibitem{empathypain}
-  Jackson PL, Brunet E, Meltzoff AN, Decety J.
-  \textit{Empathy examined through the neural mechanisms involved in imagining how I
-    feel versus how you feel pain.}
-  Neuropsychologia. 44 (5): 752–61. doi:10.1016/j.neuropsychologia.2005.07.015. PMID
-  16140345. 2006.
-
-\bibitem{acccognitive}
-  Bush G, Luu P, Posner MI.
-  \textit{Cognitive and emotional influences in anterior cingulate cortex.}
-  Trends in Cognitive Sciences. 4 (6):
-  215–222. doi:10.1016/S1364–6613(00)01483–2. PMID 10827444. June 2000.
-
-\bibitem{accbloodpressure}
-  Gianaros PJ, Derbyshire SW, May JC, Siegle GJ, Gamalo MA, Jennings JR.
-  \textit{Anterior cingulate activity correlates with blood pressure during
-    stress.}
-  Psychophysiology. 42(6):627–35. 2005.
-
-\bibitem{aicemotion}
-  Xiaosi Gu, Patrick R. Hof, Karl J. Friston, Jin Fan.
-  \textit{Anterior Insular Cortex and Emotional Awareness.}
-  J Comp Neurol. 521(15): 3371–3388. 2013-08-15.
-
-\bibitem{aicautonomic}
-  Rolls ET.
-  \textit{Functions of the anterior insula in taste, autonomic, and related
-    functions.}
-  Brain Cogn. 110:4–19. December 2016.
-
-\bibitem{selfaware}
-  Anil K. Seth, Keisuke Suzuki, Hugo D. Critchley
-  \textit {An Interoceptive Predictive Coding Model of Conscious Presence}
-  Front Psychol. 2011; 2: 395.
-
-\bibitem{dlpfcmemory}
-  Barbey AK, Koenigs M, Grafman J.
-  \textit{Dorsolateral prefrontal contributions to human working memory.}
-  Cortex. 49 (5): 1195–1205. May 2013.
-
-\bibitem{dlpfctasks}
-  Monsell S.
-  \textit{Task switching.}
-  Trends in Cognitive Sciences. 7 (3):
-  134–140. doi:10.1016/S1364–6613(03)00028–7. PMID 12639695. 2003.
-
-\bibitem{dlpfclying}
-  Ito, Ayahito; Abe, Nobuhito; Fujii, Toshikatsu; Hayashi, Akiko; Ueno, Aya;
-  Mugikura, Shunji; Takahashi, Shoki; Mori, Etsuro.
-  \textit{The contribution of the dorsolateral prefrontal cortex to the preparation
-    for deception and truth-telling.}
-  Brain Research. 1464: 43–52. doi:10.1016/j.brainres.2012.05.004. 2012.
-
-\bibitem{theprefrontalcortex}
-  Fuster, J.M.
-  \textit{The Prefrontal Cortex}
-  Raven Press, New York, 1997.
-
-\bibitem{ofcprimates}
-  Rolls, ET.
-  \textit{Convergence of sensory systems in the orbitofrontal
-    cortex in primates and brain design for emotion.}
-  The Anatomical Record Part A: Discoveries in Molecular, Cellular, and Evolutionary
-  Biology. 281 (1):1212–25. doi:10.1002/ar.a.20126. November 2004.
-
-\bibitem{theorbitofrontalcortex}
-  Price, Joseph L.
-  \textit{Chapter 3: Connections of the orbital cortex.}
-  Rauch, Scott L.; Zald, David H.
-  The Orbitofrontal Cortex. p. 45.
-  Oxford University Press, New York. 2006.
-
-\bibitem{somato}
-  Viaene A.N.
-  \textit {synaptic properties of thalamic input to layers 2/3 and 4 of
-              primary somatosensory and auditory cortices}
-  Journal of Neurophysiology. 105 (1): 279–292.
-  doi:10.1152/jn.00747.2010. PMC 3023380.
-
-\bibitem{sensoryhom}
-  Marieb E, Hoehn K.
-  Human Anatomy and Physiology.
-  7th Ed. 2007. Pearson Benjamin Cummings: San Francisco.
-
-\bibitem{parietallobesa}
-  Caspers S, Amunts K, Zilles K.
-  \textit {Posterior Parietal Cortex: Multimodal Association Cortex.}
-   The Human Nervous System. 3rd ed. New York: Elsevier; 2012.
-
-\bibitem{parietallobes}
-  Martin, R. E.
-  \textit{Let’s Get to Know the Parietal Lobes!}
-  \url{http://gablab.mit.edu/downloads/Parietal_Primer.pdf}
-  \todo{parietal lobes article returns a 404. fix.}
-
-\bibitem{relaxationresponse}
-  Lazar SW; Bush George; Gollub RL.; Fricchione, GL.; Khalsa G; Benson H.
-  \textit{Functional brain mapping of the relaxation response and meditation.}
-  Neuroreport. 11(7):1581–1585. 2000-05-15.
-
-\bibitem{prebotcgeneration}
-  J. Muñoz-Ortiz, E. Muñoz-Ortiz, L. López-Meraz, L. Beltran-Parrazal,
-  C. Morgado-Valle.
-  \textit{The pre-Bötzinger complex: Generation and modulation of respiratory rhythm}
-  Neurología (English Edition), ISSN 2173–5808,
-  https://doi.org/10.1016/j.nrleng.2018.05.006. 2018.
-
-\bibitem{ofcemotion}
-  Rempel-Clower, N. L.
-  \textit{Role of Orbitofrontal Cortex Connections in Emotion.}
-  Annals of the New York Academy of Sciences,
-  1121:72–86. doi:10.1196/annals.1401.026. 2007.
-
-\bibitem{ofcexpect}
- Lauren Y.Atlas, Tor D.Wager
- \textit{How expectations shape pain.}
-  Neuroscience Letters,  520 (2):140-148, 29 June 2012.
-
-\bibitem{alteredtraits}
-  Goleman, Daniel; Davidson, Richard J.
-  \textit{Altered Traits}
-  ISBN: 9780399184390. September 2018.
-
-\bibitem{cortexbliss}
-  Picard, Fabienne.
-  \textit{State of belief, subjective certainty and bliss as a product of cortical
-          dysfunction}
-  Cortex, Volume 49, Issue 9, Pages 2494–2500ISSN 0010 9452,
-  \url: https://doi.org/10.1016/j.cortex.2013.01.006. 2013.
-
-\bibitem{experiencemed}
-   Brewer JA, Worhunsky PD, Gray JR, Tang YY, Weber J, Kober H.
-   \textit{Meditation experience is associated with differences in default mode
-           network activity and connectivity.}
-   Proc Natl Acad Sci U S A. 2011 Dec 13;108(50):20254-9.
-   doi: 10.1073/pnas.1112029108.
-
-\bibitem{experiencemeda}
-  Zoran Josipovic, Ilan Dinstein, Jochen Weber, David J. Heeger
-  \textit {Influence of meditation on anti-correlated networks in the brain}
-  Front. Hum. Neurosci., 03 January 2012,
-  https://doi.org/10.3389/fnhum.2011.00183
+\bibitem{entropicrevisited}
+  Carhart-Harris, Robin L.
+  \textit{The entropic brain - revisited}
+  NEUROPHARMACOLOGY, Vol: 142, Pages: 167-178, ISSN: 0028-3908, 2018.
 

### 2019-11-21 paper-four/build-pdf.sh

@@ -11,4 +11,4 @@ xelatex -output-directory=$OUT paper-four.tex # > /dev/null 2>&1
 echo "...done. Now running the build."
 xelatex -output-directory=$OUT paper-four.tex
 
-rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out
+rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out $OUT/*.tdo $OUT/*.bib

### 2019-11-21 paper-one/build-pdf.sh

@@ -19,4 +19,4 @@ xelatex -output-directory=$OUT paper-one.tex > /dev/null 2>&1
 echo "...done. Now running the build."
 xelatex -output-directory=$OUT paper-one.tex
 
-rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out
+rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out $OUT/*.tdo $OUT/*.bib

### 2019-11-21 paper-two/build-pdf.sh

@@ -19,4 +19,4 @@ xelatex -output-directory=$OUT paper-two.tex > /dev/null 2>&1
 echo "...done. Now running the build."
 xelatex -output-directory=$OUT paper-two.tex
 
-rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out
+rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out $OUT/*.tdo $OUT/*.bib

### 2019-11-21 the-proposal/build-pdf.sh

@@ -15,8 +15,8 @@ PATH=$PATH:$LIBPATH
 OUT=../output
 
 echo "Running pre-build. This takes a while..."
-xelatex -output-directory=$OUT paper-three.tex # > /dev/null 2>&1
+xelatex -output-directory=$OUT the-proposal.tex > /dev/null 2>&1
 echo "...done. Now running the build."
-xelatex -output-directory=$OUT paper-three.tex
+xelatex -output-directory=$OUT the-proposal.tex
 
 rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out $OUT/*.tdo $OUT/*.bib

### 2019-11-24 paper-three/paper-three.tex
couple small notes for p3
tweaking paper four
p3: on sleep, p2: some typo fixes

@@
 \email[Correspondence email address: ]{steven@deobald.ca}
-\affiliation{vipassana-for-hackers.org}
+\affiliation{www.vipassana-for-hackers.org}
 \date{\today}
@@
 \begin{abstract}
- \todo{Write the abstract.}
+ \todo{Meditation requires a time commitment. Like other activities we consider
+   worthy of our time for their benefits to our health, such as sufficient sleep, exercise, fresh air, and a
+   healthy diet, it is often the (rational) first step of individuals considering a
+   meditation practice to ask: Why should I bother? What are the outcomes of
+   meditation? And do the benefits of those outcomes outweigh the time meditation
+   costs the practitioner? This paper answers those questions as they pertain to
+   meditation in general, and to Vipassana, specifically.}
 \end{abstract}
@@
 \textit{Vipassana for Hackers, Paper One: Curious Mechanics} was written with the
-explicit intention of avoiding a discussion about specific
-outcomes or consequences of meditation in detail. The focus of that paper was only
+explicit intention of avoiding a discussion about the specific
+outcomes or consequences of meditation in detail. That paper was intentionally,
+though artificially, restricted to
 the internal mechanics of Vipassana meditation, to pique the interest of potential
@@
 as they assist the reader in understanding what is written earlier in the paper
-regarding the senses. \textit{Paper Two: The Brain} goes further into the internal
+regarding human sensory experience. \textit{Paper Two: The Brain} goes further into the internal
 mechanics as they pertain to the
-hub of the nervous system. Here, outcomes are discussed as they pertain to
+hub of the nervous system. In the second paper, outcomes are discussed as they pertain to
 neuroplasticity. Neither paper directly discusses why an individual might choose to
@@
 computer programmers. Instead, it is meant as a label for a culture of curious and
-creative people who enjoy exploring, learning, and creating.
+creative people who enjoy exploring, learning, and creating. Librarians, scientists,
+musicians, architects, medical practitioners, carpenters, artists, lovers of books,
+mechanics, journalists, academics,
+hobbyists of all stripes --- all ``hackers'', of a sort. If you think you belong
+here, you do.
 
 \textit{Paper Three: Why Meditate?} is written for anyone who has ever asked
-themselves that very question or asked this question of their friends who
-meditate. It is for both those who are curious about the practice of Vipassana
+themselves that very question or asked this question of a friend who
+meditates. It is for both those who are curious about the practice of Vipassana
 specifically and those who are curious about meditation in general. It is for people
@@
 The reader need not have read \textit{Paper One} or \textit{Paper Two}. In fact, it
-is the intention of this paper to be the most accessible of the series and readers
+is the intention of this paper to be the most accessible of the series. Readers
 with only a faint interest in the topic of meditation should start here.
 
-\section{Vipassana Basics}
+\section{The Mundane and The Supramundane}
+
+There are two fields of human experience and the following analysis of the value
+provided by different meditation techniques will break down all points into these two
+categories. \textit{The Mudane}
+in this context refers not to the tedious but to the earthly, the material. Most people will begin
+meditating for reasons in the mundane field simply because most people have never
+experienced the supramundane field. \textit{The Supramundane} in this context refers not to
+spirituality or religion but experiences which transcend the material, physical world.
+
+Because supramundane experiences tend to occur only in deep meditative states,
+the reasons for meditating listed will predominantly fall in the mundane
+category. Whether or not the meditator experiences supramundane states in deep
+meditation, these altered states of mind are never the goal of meditation. The goal
+of meditation is to change the meditator's mental habits. The intention is to move away from unhealthy
+mental patterns --- those which cause harmful behaviours --- toward healthy mental patterns which
+encourage productive behaviour. Obviously this change is only visible in the mundane
+world, outside of meditation.
+
+\section{The Mundane Sphere of Experience}
+
+\subsection{Sleep}
+
+Why do we have difficulty sleeping? If one imagines a sleepless night of one's past,
+it often followed an anxiety-inducing event or preceded a stressful event. When we
+fight with a family member or have a difficult day at work, we become anxious and
+sometimes cannot escape from replaying that event over and over in our mind's
+eye in exchange for sleep. When the next day brings a final exam or a job interview,
+we repeatedly imagine the future and its outcomes while we lie awake in bed. Whether
+we are anxious about the past or the future, it seems that anxiety has a great deal
+to do with our inability to sleep. Sometimes this anxiety is apparently disconnected
+from our lives entirely. We may ruminate about anything: long-past childhood
+experiences, politics, global warming, human suffering at scales completely
+unmanageable through any actions of our own. Anxiety is anxiety.
+
+\todo{more anxiety/sleep references}
+
+And, as it turns out, anxiety has a lot to do with our inability to
+sleep. When we are anxious we can't sleep. \cite{mellman2006, staner2003}
+But this relationship is dangerously recursive: when we can't sleep we become
+anxious. This effect occurs both at the narrow and personally-observable level, within a single
+night of poor sleep. But it also occurs on a lifelong scale and there is mounting
+evidence that sleep deprivation in childhood and adulthood has a causal relationship
+with chronic anxiety. \cite{gregory2005, willis2015}
+
+Meditation objectively improves sleep. Black et al. found in a 2015 study that this
+improvement is is not simply over the baseline, however, but also an improvement
+above and beyond what can be achieved through Sleep
+Hygiene Education (SHE). \cite{black2015} The difficulty with this statement, of
+course, is defining the term ``meditation''. In this particular study, the meditation
+in question is \textit{Mindfulness}, as taught in the UCLA Mindfulness
+Course. \cite{uclamaps} Mindfulness is a very accessible form of meditation, varying
+in format. It usually involves multiple techniques, including \textit{Open
+  Monitoring} which is characterized by
+openness to whatever is happening in the present moment in a variety of postures
+(sitting, standing, walking, eating, etc.), \textit{Focused Attention} which is
+usually a narrow breath awareness meditation, and \textit{Loving-Kindness} in which
+meditators actively direct compassion to themselvse and others. All three of these
+practices mirror the three meditation practices taught in Vipassana. However, in Vipassana, Open
+Monitoring finds its parallel in objective observation of sensation internal to the
+body, as we will see later.
+
+\subsection{Meditation vs. Naps}
+
+While staying at a friend's house, I excused myself in the evening to meditate. He
+sincerely asked, ``Is meditating for an hour really more valuable than using that
+time for a good nap?''
+
+\todo{look through leather notebook}
+
+\begin{itemize}
+  \item posture
+  \item sleep
+  \item health (activation / motivation)
+  \item ethics (activation / motivation)
+  \item your children: a. knowing how to meditate, b. cross-legged posture
+  \item emotion (i.e. anger)
+  \item mundane sphere / productivity (21 lessons, seinfeld)
+  \item unlearning obsessive / repetitive thought, enhancing creativity
+  \item controlling unbounded sexuality without repression (louis c.k. joke?)
+  \item clarity: in thought, work, planning
+  \item Die Standing Up
+  \item \url{https://www.pnas.org/content/early/2019/10/18/1909959116}
+\end{itemize}
+
+\begin{itemize}
+  \item reset frame of reference outside oneself, outside one's own lifetime: ``trees
+    for god'' and obvious karma (sidu/booga smoking)
+  \item Time: Nat Friedman's blog post?
+  \item clarification: ``isn't that what makes us human?'' (emotions) --- rather,
+    what makes us animal
+  \item ``How to Change Your Mind'' --- ego dissolution --- specific advantages?
+\end{itemize}
+
+\section{Why Vipassana?}
+
+Material notes: free course, code of ethics, 10 days, etc.
+
+\subsection{Vipassana Basics}
 
@@
 \vspace{1cm}
-\textbf{Every experience which emerges in the mind, whether a thought, emotion, or contact of the five senses, always surfaces with a corresponding sensation on the body.}
+\textbf{Every experience which emerges in the mind, whether a thought, emotion, or
+  contact of the five senses, always surfaces with a corresponding sensation in or on the body.}
 \vspace{1cm}
+
+\cite{hauke2018}
 
@@
 
+This concept, that sensory input is ``reflected'' in internal bodily sensation,
+
 Mapping all of sensory experience to these eight categories begs the question of
@@
 residential course \cite{dhamma} is necessary to learn the technique.
-
-\section{The Mundane and The Supramundane}
-
-Reasons for meditating will be broken down into two categories. \textit{The Mudane}
-in this context refers not to the tedious but to the earthly, the material. Most people will begin
-meditating for reasons in the mundane field simply because most people have never
-experienced the supramundane field. \textit{The Supramundane} in this context refers not to
-spirituality or religion but experiences which transcend the material, physical world.
-
-Because supramundane experiences tend to occur only in deep meditative states,
-the reasons for meditating listed will predominantly fall in the mundane
-category. Whether or not the meditator experiences supramundane states in deep
-meditation, these altered states of mind are never the goal of meditation. The goal
-of meditation is to change the meditator's mental habits. The intention is to move away from unhealthy
-mental patterns, which cause harmful behaviours, toward healthy mental patterns which
-encourage productive behaviour. Obviously this change is only visible in the mundane
-world, outside of meditation.
-
-\section{Meditation vs. Naps}
-
-While staying at a friend's house, I excused myself in the evening to meditate. He
-sincerely asked, ``Is meditating for an hour really more valuable than using that
-time for a good nap?''
-
-\begin{itemize}
-  \item posture
-  \item sleep
-  \item health (activation / motivation)
-  \item ethics (activation / motivation)
-  \item your children: a. knowing how to meditate, b. cross-legged posture
-  \item emotion (i.e. anger)
-  \item mundane sphere / productivity (21 lessons, seinfeld)
-  \item unlearning obsessive / repetitive thought, enhancing creativity
-  \item controlling unbounded sexuality without repression
-  \item clarity: in thought, work, planning
-  \item Die Standing Up
-\end{itemize}
-
-\begin{itemize}
-  \item reset frame of reference outside oneself, outside one's own lifetime: ``trees
-    for god'' and obvious karma (sidu/booga smoking)
-  \item Time: Nat Friedman's blog post?
-  \item clarification: ``isn't that what makes us human?'' (emotions) --- rather, what makes us animal
-\end{itemize}
 
@@
 
+\bibitem{mellman2006}
+  Mellman, Thomas A.
+  \textit{Sleep and Anxiety Disorders}
+  Psychiatric Clinics of North America, Volume 29, Issue 4, Pages 1047-1058, December 2006.
+  \url{https://doi.org/10.1016/j.psc.2006.08.005}
+
+\bibitem{staner2003}
+  Staner, L.
+  \textit{Sleep and anxiety disorders.}
+  Dialogues in Clinical Neuroscience, Volume 5(3), Pages 249–258, September 2003.
+  PMID: 22033804; PMCID: PMC3181635.
+
+\bibitem{gregory2005}
+  Gregory, A.M., Caspi, A., Eley, T.C. et al.
+  \textit{Prospective Longitudinal Associations Between Persistent Sleep Problems in
+    Childhood and Anxiety and Depression Disorders in Adulthood}
+  Journal of Abnormal Child Psychology, Volume 33, Issue 2, pp 157–163, April 2005.
+  \url{https://doi.org/10.1007/s10802-005-1824-0}
+
+\bibitem{willis2015}
+  Willis, T.A., Gregory, A.M.
+  \textit{Anxiety Disorders and Sleep in Children and Adolescents.}
+  Sleep Medicine Clinics, Volume 10, Issue 2, Pages 125-131, June 2015.
+  \url{https://doi.org/10.1016/j.jsmc.2015.02.002}
+
+\bibitem{hauke2018}
+  Hauke G., Kritikos A.
+  \textit{Building a Body of Evidence: From Sensation to Emotion and Psychotherapy.}
+  Embodiment in Psychotherapy. Springer, Cham.
+  ISBN: 978-3-319-92888-3. December 2018.
+  \url{https://doi.org/10.1007/978-3-319-92889-0_1}
+
+\bibitem{black2015}
+  Black D.S., O’Reilly G.A., Olmstead R., Breen E.C., Irwin M.R.
+  \textit{Mindfulness Meditation and Improvement in Sleep Quality and Daytime Impairment
+    Among Older Adults With Sleep Disturbances: A Randomized Clinical Trial.}
+  JAMA Internal Medicine, Volume 175(4), Pages 494–501, 2015.
+
+\bibitem{uclamaps}
+  Mindful Awareness Practices (MAPs) Classes
+  \url{https://www.uclahealth.org/marc/maps-classes}
+
 \bibitem{dhamma}

### 2019-11-24 paper-two/paper-two.tex
Fixing epub builds
- fixing epub title and series in both epub scripts
- fixed references section order in paper-two.tex (`section` within bibilography was causing errors)
- fixed svg resize errors by putting SVGs in their own directory... I have no idea why `epub-convert` was looking at the SVGs since the .tex file doesn't even reference them, but whatever.
- added a refrences section title to paper one
- adding graphicsmagick to readme (may not be required but it can't hurt)
paper three 2-column
p3: on sleep, p2: some typo fixes

@@
 
-\section{Acknowledgements}
+\section*{Acknowledgements}
 
@@
 
+\section*{References}
+
 \begin{thebibliography}{99}
-
-\section*{References}
 
@@
           dysfunction}
-  Cortex, Volume 49, Issue 9, Pages 2494–2500ISSN 0010 9452,
-  \url{https://doi.org/10.1016/j.cortex.2013.01.006}. 2013.
+  Cortex, Volume 49, Issue 9, Pages 2494–2500ISSN 0010 9452. 2013.
+  \url{https://doi.org/10.1016/j.cortex.2013.01.006}.
 
@@
   \textit {Influence of meditation on anti-correlated networks in the brain}
-  Frontiers in Human Neuroscience, 03 January 2012,
-  https://doi.org/10.3389/fnhum.2011.00183
+  Frontiers in Human Neuroscience, 03 January 2012.
+  \url{https://doi.org/10.3389/fnhum.2011.00183}
 
@@
     research with psychedelic drugs}
-  Frontiers in Human Neuroscience, 03 February 2014,
-  https://doi.org/10.3389/fnhum.2014.00020
+  Frontiers in Human Neuroscience, 03 February 2014.
+  \url{https://doi.org/10.3389/fnhum.2014.00020}
 

### 2019-11-25 paper-three/paper-three.tex
meditation planes for paper 3

@@
 \begin{abstract}
- \todo{Meditation requires a time commitment. Like other activities we consider
+ Meditation requires a time commitment. Like other activities we consider
  worthy of our time for their benefits to our health, such as sufficient sleep, exercise, fresh air, and a
@@
  costs the practitioner? This paper answers those questions as they pertain to
-   meditation in general, and to Vipassana, specifically.}
+ meditation in general, and to Vipassana, specifically.
 \end{abstract}
@@
 
-\section{The Mundane and The Supramundane}
+\section{The Mundane vs. The Supramundane}
 
@@
 provided by different meditation techniques will break down all points into these two
-categories. \textit{The Mudane}
+categories. \textit{The Mundane}
 in this context refers not to the tedious but to the earthly, the material. Most people will begin
@@
 experienced the supramundane field. \textit{The Supramundane} in this context refers not to
-spirituality or religion but experiences which transcend the material, physical world.
+spirituality or religion but experiences which transcend the material, physical
+world. Put another way, mundane experiences are those which can be
+described. Supramundane experiences, due to their very nature of transcending the
+material world, are ineffable. Just because an experience is ineffable, however, does
+not mean conversation cannot exist around it. An experience of dimethyltryptamine
+(DMT) may be ineffable but it is almost a guarantee that someone using this drug will
+talk about it afterward. In this way, we will discuss the consequences of
+supramundane experiences (and their value) toward the end of this paper.
 
@@
 world, outside of meditation.
+
+In \textit{Paper One: Curious Mechanics}, two sample spectrums of meditations were
+listed. One included activities which were simply meditative, rather than anything
+which could be considered a formal meditation practice. These included sports,
+playing a musical instrument, or yogic practices such as asanas and pranayams. We
+will ignore such activities here. The following plane attempts to roughly locate
+meditation practices between focus and insight on two axes, rather than one.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=5cm]{images/focus-insight-plane.png}
+  \caption{Meditation Quadrants}
+  \label{fig:meditation-quadrants}
+\end{figure}
+
+The locations of different meditation practices on the graph are only for
+illustrative purposes. There are many other axes available such as a practitioner's
+experience level, intensity of practice, and an individual's physiology and mental
+health. However, this plane helps ground a comparison of different meditation
+techniques for the rest of the paper. The three mindfulness meditation categories are
+Focused Attention Meditation (FAM), Open Monitoring Meditation (OMM), and
+Loving-Kindness Meditation (LKM). These will be explained as mindfulness meditations
+come up in discussion later.
+
+Returning to the topic of the supramundane, the following spectrum illustrates where
+each of these practices approximately falls, across that dichotomy.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=5cm]{images/mundane-supramundane-plane.png}
+  \caption{Mundane vs. Supramundane Meditations}
+  \label{fig:mundane-vs-supramundane}
+\end{figure}
+
+Meditations within the mundane sphere are easier to understand and easier to
+teach. As a consequence, they are easier to submit to rigorous scientific
+study. Where the broad categories of FAM, OMM, and LKM have been studied
+scientifically, we will observer respective parallels in Anapana, Vipassana, and
+Metta meditations.
 
@@
 Course. \cite{uclamaps} Mindfulness is a very accessible form of meditation, varying
-in format. It usually involves multiple techniques, including \textit{Open
-  Monitoring} which is characterized by
+in format according to the instructor. It usually involves multiple techniques,
+including \textit{Open Monitoring}, which is characterized by
 openness to whatever is happening in the present moment in a variety of postures
-(sitting, standing, walking, eating, etc.), \textit{Focused Attention} which is
-usually a narrow breath awareness meditation, and \textit{Loving-Kindness} in which
+(sitting, standing, walking, eating, etc.); \textit{Focused Attention}, which is
+usually a narrow breath awareness meditation; and \textit{Loving-Kindness}, in which
 meditators actively direct compassion to themselvse and others. All three of these
@@
   \item \url{https://www.pnas.org/content/early/2019/10/18/1909959116}
-\end{itemize}
-
-\begin{itemize}
-  \item reset frame of reference outside oneself, outside one's own lifetime: ``trees
-    for god'' and obvious karma (sidu/booga smoking)
-  \item Time: Nat Friedman's blog post?
-  \item clarification: ``isn't that what makes us human?'' (emotions) --- rather,
-    what makes us animal
-  \item ``How to Change Your Mind'' --- ego dissolution --- specific advantages?
 \end{itemize}
@@
 residential course \cite{dhamma} is necessary to learn the technique.
+
+\section{The Supramundane Sphere of Experience}
+
+\begin{itemize}
+  \item reset frame of reference outside oneself, outside one's own lifetime: ``trees
+    for god'' and obvious karma (sidu/booga smoking)
+  \item Time: Nat Friedman's blog post?
+  \item bible; ``a perfect set of rules''; nature itself <=> perception; ``perfect
+    perception (vision)''; inner perception (sensation mirror) --- harris v. peterson
+  \item clarification: ``isn't that what makes us human?'' (emotions) --- rather,
+    what makes us animal
+  \item ``How to Change Your Mind'' --- ego dissolution --- specific advantages?
+  \item seeing oneself as aggregate and the path from singular => aggregate
+    (neuroscience behind hemispheres)
+\end{itemize}
 

### 2019-11-28 paper-three/paper-three.tex
digestion (p3) and sample proposal formats
citation add, format update

@@
 
-Meditation objectively improves sleep. Black et al. found in a 2015 study that this
+Meditation objectively improves sleep across a number of meditation techniques.
+\cite{nagendra2012} Black et al. found in a 2015 study that this
 improvement is is not simply over the baseline, however, but also an improvement
@@
   \item sleep
+  \item digestion (``Make sure to pay attention to your poops!'' first course)
+  \item diet ^
+  \item schedule
   \item health (activation / motivation)
@@
   \item emotion (i.e. anger)
+  \item dealing with death
   \item mundane sphere / productivity (21 lessons, seinfeld)
@@
   Mellman, Thomas A.
-  \textit{Sleep and Anxiety Disorders}
+  \textit{Sleep and Anxiety Disorders.}
   Psychiatric Clinics of North America, Volume 29, Issue 4, Pages 1047-1058, December 2006.
@@
 
+\bibitem{nagendra2012}
+  Nagendra R.P., Maruthai N., Kutty B.M.
+  \textit{Meditation and its regulatory role on sleep.}
+  Frontiers in Neurology, 3:54, April 2012.
+  \url{https://doi.org/10.3389/fneur.2012.00054}
+
 \bibitem{black2015}

### 2019-11-29 the-proposal/the-proposal.tex
draft abstract

@@
 \email[Correspondence email address: ]{steven@deobald.ca}
-\affiliation{Siggu.org}
+\affiliation{www.vipassana-for-hackers.org}
 \date{\today}
@@
 \begin{abstract}
-I want to turn myself into a human guinea pig.
+Vipassana meditation (as taught by S.N. Goenka in the tradition of Sayagyi U Ba Khin)
+is unlike other meditation techniques insofar as its claims to completeness and
+outcomes are concerned. Vipassana claims to ultimately explore the entire field of
+mind and matter, with the goals of total liberation and full enlightenment. Implicit
+within these claims is a complete understanding of human consciousness. These are
+difficult claims to prove because the time commitment required to research the
+technique orders on multiple decades --- if not multiple generations. This is complicated by the
+fact that the time commitment is demanded of both the researcher \textit{and the
+subject}. Until now, research on meditation of all kinds has covered only one of
+two fields: (1) controlled experiments which must inherently rely on superficial data
+gathered from beginners and (2) observational exploratory research of monks --- expert
+meditators who have dedicated their lives to the practice. I propose bridging this
+gap by submitting myself to the middle ground. While remaining a layperson, I will
+commit to a high ratio of waking medition hours for the rest of my life. Internally,
+I will conduct qualitative research into the nature of consciousness while
+simultaneously exposing myself as a subject to long-term quantitative studies.
+
+\todo{add references to (1) and (2)}
+
 \end{abstract}

### 2019-11-29 the-proposal/the-proposal.tex
started intro draft for proposal

@@
 I will conduct qualitative research into the nature of consciousness while
-simultaneously exposing myself as a subject to long-term quantitative studies.
+externally exposing myself as a subject to long-term quantitative studies.
 
@@
 
-\keywords{neuroscience, vipassana, meditation}
+\keywords{neuroscience, psychology, vipassana, meditation}
 
@@
 
-% \listoftodos
+\listoftodos
 
-\section{Why Vipassana?}
+\section{Introduction}
 
-zig zig zig
+Research into the effects of meditation has been conducted in earnest for half a
+century but the quality of research in this field varies wildly. Randomized
+controlled trials were missing from much research conducted during the first few
+decades of meditation study. The importance of active controls was often missed even
+when randomized controlled trials were attempted. Double-blind studies are inherently
+impossible with meditation research; a subject will always know if she is receiving
+meditation instruction or an active control instruction. \cite{alteredtraits}
+
+Add to these difficulties the very nature of meditation research itself. There are
+many techniques of meditation and it is very important to capture the specific
+technique under study to make meaningful assertions about its
+effects. \cite{alteredtraits} However, even within the definition of a single
+meditation technique there exist variations in instruction between
+teachers and each student's comprehension of the instructions
+received. \cite{davidson2015} Even if researchers could cement (or at least
+accurately record) semantics and terminology, the subject of study is often unclear:
+Are we looking for health benefits? Increases in productivity? Increases in
+intelligence? How long these effects persist? How much do we want to explore concrete
+hypotheses versus exploratory analysis of long-term effects? How much can be learned
+about the nature of consciousness? Can these learnings be measured objectively or
+communicated meaningfully?
+
+Because Vipassana is globally standardized, it affords researchers with definitive
+solutions to the difficulties presented by varying teaching methodologies. In
+exchange for this, the difficulty of long-term study is compounded by the intrinsic
+seriousness demanded of a Vipassana practitioner: the concept of ``continuous
+piercing awareness of constantly changing bodily sensation'' is not only a
+requirement of serious Vipassana practice but could very well act as a surrogate
+definition for the practice itself. This seriousness poses obvious difficulties.
+
+\subsection{Why Vipassana?}
+
+1. ``complete'', standardized, global, multilingual, 100\% free
+2. mundane (sleep) vs. supramundane (total eradication of suffering)
+
+3. who is steven deobald
+
+\section{Problem Statement}
+
+A large scale controlled study of beginner-to-intermediate
+students of Vipassana, each practicing a minimum of two hours daily, is
+possible. Randomization will not be possible in such a study as the students
+self-select this
+
+- middle ground: me
+
+(vs. monks)
+
+\subsection{Overview}
+\subsection{Hypothesis}
 
@@
 
-Thank you to Steven Deobald for reviews, edits, and corrections. \texttt{Brain\_human\_normal\_inferior\_view.svg} and
-\texttt{Brain\_human\_lateral\_view.svg} by Patrick J. Lynch, medical illustrator, CC-BY 2.5 \cite{brainsvg}
+Thank you to Preethi Govindarajan for reviews, edits, and corrections.
 
@@
 \begin{thebibliography}{99}
+
+\bibitem{alteredtraits}
+  Goleman D., Davidson R.J.
+  \textit{Altered Traits.}
+  ISBN: 9780399184390. September 2018.
+
+\bibitem{davidson2015}
+  Davidson R.J., Kaszniak A.W.
+  \textit{Conceptual and methodological issues in research on mindfulness and meditation.}
+  American Psychologist, Volume 70(7), Pages 581-92. October 2015.
+  \url{https://doi.org/10.1037/a0039512}
 

### 2019-11-30 the-proposal/the-proposal.tex
intro

@@
 gap by submitting myself to the middle ground. While remaining a layperson, I will
-commit to a high ratio of waking medition hours for the rest of my life. Internally,
-I will conduct qualitative research into the nature of consciousness while
-externally exposing myself as a subject to long-term quantitative studies.
+commit to a high ratio of waking medition hours for the rest of my life. Individually
+and internally, I will conduct qualitative research into the consequences of
+Vipassana meditation and the nature of consciousness while collectively and
+externally pledging myself as a subject for long-term quantitative
+studies with a broader community of researchers.
 
@@
 about the nature of consciousness? Can these learnings be measured objectively or
-communicated meaningfully?
+even communicated meaningfully?
 
@@
 requirement of serious Vipassana practice but could very well act as a surrogate
-definition for the practice itself. This seriousness poses obvious difficulties.
+description of the practice itself. This seriousness poses obvious difficulties and
+it is the intention of my study to begin breaking ground in solutions to those
+difficulties.
 
-\subsection{Why Vipassana?}
+This study will prove significant in three fields of research:
 
-1. ``complete'', standardized, global, multilingual, 100\% free
-2. mundane (sleep) vs. supramundane (total eradication of suffering)
+First, and most accessible, is the continued quantitative research of meditation in
+the broader neuroscience and psychology disciplines, where my participation will be
+more as subject than researcher.
 
-3. who is steven deobald
+Second is the qualitative research into the long-term consequences of
+Vipassana meditation, what it reveals about the nature of human consciousness, and
+reusable techniques for its exploration --- a field currently consisting of a bridge
+between psychology, philosophy, linguistics, and contemplative studies.
+
+Third is the exploratory research intended to objectively define
+consciousness. As such research must pertain to all forms of consciousness it
+therefore includes all non-human forms of consciousness. Findings will inform our
+understanding of the Mind-Body Problem of psychology and philosophy, the entire field
+of bioethics, and --- perhaps most importantly --- the nascent field of Artificial
+Consciousness as a subfield of Artificial Intelligence. As of this writing, this
+overarching integral field of study has yet to emerge and has no name as a discipline.
 
@@
 
-\subsection{Overview}
+\subsection{Overview (?)}
 \subsection{Hypothesis}
+
+\section{Objectives and Aims}
+
+\subsection{Overall Objective}
+\subsection{Specific Aims}
+
+\section{Background and Significance}
+
+\subsection{Preliminary Research Review}
+- summary of related research
+- strengths and weaknesses
+- justification
+--- what hasn't been done by others?
+--- why is this research necessary?
+
+\subsection{Why Vipassana?}
+
+1. ``complete'', standardized, global, multilingual, 100\% free
+2. mundane (sleep) vs. supramundane (total eradication of suffering)
+
+3. who is steven deobald
+
+\subsection{Why now?}
+
+\section{Research Design and Methods}
+\subsection{OVERVIEW}
+
+\subsection{Study Design / Research Method}
+- case study type?
+
+\subsection{POPULATION AND STUDY SAMPLE}
+\subsection{SAMPLE SIZE AND SELECTION OF SAMPLE}
+\subsection{SOURCES OF DATA}
+\subsection{COLLECTION OF DATA}
+\subsection{EXPOSURE ASSESSMENT}
+\subsection{DATA MANAGEMENT}
+\subsection{DATA ANALYSIS STRATEGIES}
+
+\subsection{Participants}
+- people & roles
+
+\subsection{ETHICS AND HUMAN SUBJECTS ISSUES}
+\subsection{TIMEFRAMES}
+
+\section{STRENGTHS AND WEAKNESSES OF THE STUDY}
+\section{PUBLIC HEALTH SIGNIFICANCE}
+\section{BUDGET AND MOTIVATION}
+- resources required
+
+\section{Conflict of Interest Statement?}
+
+\section{REFERENCES}
+\section{APPENDICES}
+\subsection{Appendix 1: Proposed Daily Schedule}
 

### 2019-11-30 the-proposal/the-proposal.tex
proposal: add hildt ref

@@
 of bioethics, and --- perhaps most importantly --- the nascent field of Artificial
-Consciousness as a subfield of Artificial Intelligence. As of this writing, this
-overarching integral field of study has yet to emerge and has no name as a discipline.
+Consciousness as a subfield of Artificial Intelligence. \cite{hildt2019} As of this
+writing, this overarching integral field of study has yet to emerge and has no name
+as a discipline.
 
@@
 
+\bibitem{hildt2019}
+  Artificial Intelligence: Does Consciousness Matter?
+  \todo{stop doing this and use bibtex}
+
 \bibitem{entropicrevisited}

### 2019-11-30 the-proposal/the-proposal.tex
proposal: wrangling the build to work with bibtex, updating the makefile

--- 2019-11-30 the-proposal/the-proposal.tex
@@
 impossible with meditation research; a subject will always know if she is receiving
-meditation instruction or an active control instruction. \cite{alteredtraits}
+meditation instruction or an active control instruction. \cite{goleman2017altered}
 
@@
 technique under study to make meaningful assertions about its
-effects. \cite{alteredtraits} However, even within the definition of a single
+effects. \cite{goleman2017altered} However, even within the definition of a single
 meditation technique there exist variations in instruction between
 teachers and each student's comprehension of the instructions
-received. \cite{davidson2015} Even if researchers could cement (or at least
+received. \cite{davidson2015conceptual} Even if researchers could cement (or at least
 accurately record) semantics and terminology, the subject of study is often unclear:
@@
 of bioethics, and --- perhaps most importantly --- the nascent field of Artificial
-Consciousness as a subfield of Artificial Intelligence. As of this writing, this
-overarching integral field of study has yet to emerge and has no name as a discipline.
+Consciousness as a subfield of Artificial Intelligence. \cite{hildt2019artificial} As of this
+writing, this overarching integral field of study has yet to emerge and has no name
+as a discipline.
 
@@
 \subsection{Participants}
-- people & roles
+- people and roles
 
@@
 
-\begin{thebibliography}{99}
+\bibliography{the-proposal}{}
+\bibliographystyle{chicago}
 
-\bibitem{alteredtraits}
-  Goleman D., Davidson R.J.
-  \textit{Altered Traits.}
-  ISBN: 9780399184390. September 2018.
-
-\bibitem{davidson2015}
-  Davidson R.J., Kaszniak A.W.
-  \textit{Conceptual and methodological issues in research on mindfulness and meditation.}
-  American Psychologist, Volume 70(7), Pages 581-92. October 2015.
-  \url{https://doi.org/10.1037/a0039512}
-
-\bibitem{entropicrevisited}
-  Carhart-Harris, Robin L.
-  \textit{The entropic brain - revisited}
-  NEUROPHARMACOLOGY, Vol: 142, Pages: 167-178, ISSN: 0028-3908, 2018.
-
-\end{thebibliography}
-
+%\begin{thebibliography}{99}
+%
+%\bibitem{alteredtraits}
+%  Goleman D., Davidson R.J.
+%  \textit{Altered Traits.}
+%  ISBN: 9780399184390. September 2018.
+%
+%\bibitem{davidson2015}
+%  Davidson R.J., Kaszniak A.W.
+%  \textit{Conceptual and methodological issues in research on mindfulness and meditation.}
+%  American Psychologist, Volume 70(7), Pages 581-92. October 2015.
+%  \url{https://doi.org/10.1037/a0039512}
+%
+%\bibitem{hildt2019}
+%  Artificial Intelligence: Does Consciousness Matter?
+%  \todo{stop doing this and use bibtex}
+%
+%\bibitem{entropicrevisited}
+%  Carhart-Harris, Robin L.
+%  \textit{The entropic brain - revisited}
+%  NEUROPHARMACOLOGY, Vol: 142, Pages: 167-178, ISSN: 0028-3908, 2018.
+%
+%
+%\end{thebibliography}
+%
 % \appendix*

### 2019-11-30 the-proposal/build-pdf.sh

@@ -12,11 +12,15 @@ SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
 LIBPATH="$SCRIPTPATH/../lib"
 PATH=$PATH:$LIBPATH
 
-OUT=../output
+echo "Running bibliography pre-build. This takes a while..."
+xelatex the-proposal.tex > /dev/null 2>&1
+echo "Generating bibliography..."
+bibtex the-proposal.aux
+read -p "Press enter to continue"
 
-echo "Running pre-build. This takes a while..."
-xelatex -output-directory=$OUT the-proposal.tex > /dev/null 2>&1
-echo "...done. Now running the build."
-xelatex -output-directory=$OUT the-proposal.tex
+echo "Running PDF pre-build. This takes a while..."
+xelatex the-proposal.tex > /dev/null 2>&1
+echo "...done. Now running the PDF build."
+xelatex the-proposal.tex
 
-rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out $OUT/*.tdo $OUT/*.bib
+mv the-proposal.pdf ../output/the-proposal.pdf

### 2019-12-01 the-proposal/the-proposal.tex
proposal: finally done the intro, maybe

@@
 
-\bibliographystyle{apsrev4-1}
 \setlist{noitemsep}
@@
 exchange for this, the difficulty of long-term study is compounded by the intrinsic
-seriousness demanded of a Vipassana practitioner: the concept of ``continuous
-piercing awareness of constantly changing bodily sensation'' is not only a
+seriousness demanded of a Vipassana practitioner: the Pāli concept of \textit{ātāpī
+  sampajāno satimā} (continuous piercing awareness of constantly
+changing bodily sensation \cite{goenka1999discourses}) is not only a
 requirement of serious Vipassana practice but could very well act as a surrogate
@@
 
-%\begin{thebibliography}{99}
-%
-%\bibitem{alteredtraits}
-%  Goleman D., Davidson R.J.
-%  \textit{Altered Traits.}
-%  ISBN: 9780399184390. September 2018.
-%
-%\bibitem{davidson2015}
-%  Davidson R.J., Kaszniak A.W.
-%  \textit{Conceptual and methodological issues in research on mindfulness and meditation.}
-%  American Psychologist, Volume 70(7), Pages 581-92. October 2015.
-%  \url{https://doi.org/10.1037/a0039512}
-%
-%\bibitem{hildt2019}
-%  Artificial Intelligence: Does Consciousness Matter?
-%  \todo{stop doing this and use bibtex}
-%
-%\bibitem{entropicrevisited}
-%  Carhart-Harris, Robin L.
-%  \textit{The entropic brain - revisited}
-%  NEUROPHARMACOLOGY, Vol: 142, Pages: 167-178, ISSN: 0028-3908, 2018.
-%
-%
-%\end{thebibliography}
-%
 % \appendix*

### 2019-12-01 the-proposal/the-proposal.tex
proposal: problem statement-phenomenology

@@
 
+Because Vipassana does not permit any form of imagination, any attempt to analyze it
+through philosophical phenomenology becomes a hindrance to actually practicing
+Vipassana. \cite{patrik1994phenomenological} I have previously discussed this paradox
+in \textit{Vipassana for Hackers, Paper One: Curious Mechanics} \cite{deobald2017curious}
+
 \subsection{Overview (?)}

### 2019-12-02 the-proposal/the-proposal.tex
proposal: i think i just accidentally wrote the design and methods section instead of the problem statement

@@
 
-Because Vipassana is globally standardized, it affords researchers with definitive
+Because Vipassana \footnote{For the remainder of this research proposal,
+  ``Vipassana'' will always refer to the specific technique ``Vipassana as taught by S.N. Goenka in the
+  tradition of Sayagyi U Ba Khin'' and lineage-identical instruction, such as that
+  taught by Ledi Sayadaw (in writing) or other students of U Ba Khin contemporary
+  to S.N. Goenka, unless noted otherwise.} is globally standardized, it affords
+researchers with definitive
 solutions to the difficulties presented by varying teaching methodologies. In
@@
 
+\subsection{Overview}
+
+As the structure of studies conducted on Vipassana is inherently very difficult, due to
+the nature of Vipassana meditation instruction, we must first examine which kinds of
+studies are possible, which are not, and what will make possible studies worthwhile.
+
+\subsubsection{Expertise}
+
 A large scale controlled study of beginner-to-intermediate
@@
 possible. Randomization will not be possible in such a study as the students
-self-select this
+self-select this meditation technique for themselves. The recommendation to practice
+the technique ``a minimum of two hours daily'' is a component of instruction, given
+to students even on the most basic 10-day introductory
+course. \cite{dhammaorg2017guidelines,goenka2001importance} As such, it is possible
+to conduct such a study on students who have dedicated themselves to the practice of
+Vipassana in the manner recommended --- and an increasingly large number of Vipassana
+meditators do so, worldwide.
 
-- middle ground: me
+However, such a study suffers from the very nature of
+the instruction and its target audience. On the lifelong scale of Vipassana
+practice, the spectrum spans from laypeople practicing two hours per day to
+renunciates (monks and nuns) who practice up to twenty-four hours per day once they
+reach the stage where deep sleep no longer occurs due to the practice itself. It is
+also worth noting that a junior renunciate will still have less \textit{experience}
+than an senior lay meditator, which means that the only objective measure of
+experience is total number of hours meditated. \cite{goleman2017altered}
 
-(vs. monks)
+Meditation research often has difficulty defining and capturing \textit{expertise},
+however, and total hours of experience is still confounded by the rate in which those
+hours of experience are accumulated (hours per day). We must capture both, as accurately as
+possible. With a high hours-per-day rate and extended (lifelong) experience, the very
+broad category of \textit{expert} meditators quickly requires upward of 100,000 hours
+of practice, even if we limit practice hours to formal, sitting meditation and
+estimate twelve (12) hours per day. If we take the much more modest rate of six (6)
+hours per day, assume an adult can learn meditation from approximately 20 years of
+age, and is of an age when they may participate in an extended study (say, 50-60
+years of age), we arrive at roughly 60,000 hours of practice. If we use the oft-cited
+``10,000 hours'' measurement for competence in a subject, we might reasonably
+describe our rough spectrum of expertise as such:
+
+\begin{itemize}
+	\item Beginner: 0-10,000 hours
+	\item Intermediate: 10,000-60,000 hours
+	\item Expert: 60,000+ hours
+\end{itemize}
+
+As a consequence, even studies which claim to observe ``expert'' or ``long-term''
+Vipassana meditators are often predominantly observing beginners (7.9-8.6 mean years
+of experience with 2 hours of daily practice).
+
+\subsubsection{Methodological Issues}
+
+This is in addition to a large number of other
+methodological issues with studies of Vipassana, as described by Alberto Chiesa in
+\textit{Vipassana meditation: systematic review of current evidence},
+2010. \cite{chiesa2010vipassana} These include the lack of: study replication,
+randomized trials, active plus inactive controls, and double/single blinding.
+
+Chiesa also notes that it would be beneficial to capture data ``both from a clinical
+and from a neuro-imaging point of view'', including functional magnetic resonance
+imaging (fMRI) and electroencephalography (EEG), for both short-term outcomes
+(altered states) and long-term outcomes (altered traits \cite{davidson1977role}) to
+improve future Vipassana research. \cite{chiesa2010vipassana} New developments into
+magnetoencephalography (MEG) and quantum MEG over the past decade may permit it as a
+third neuro-imaging technique.
 
@@
 Vipassana. \cite{patrik1994phenomenological} I have previously discussed this paradox
-in \textit{Vipassana for Hackers, Paper One: Curious Mechanics} \cite{deobald2017curious}
+in Appendix A of \textit{Vipassana for Hackers, Paper One: Curious Mechanics}.
+\cite{deobald2017curious}
 
-\subsection{Overview (?)}
+Last, statistical study of Vipassana, at the intermediate level, is currently not
+possible. Meditators who are in the process of transitioning from 10,000 hours of
+experience to 60,000 hours of experience are not readily available as research
+subjects in significant numbers. Because we have not studied such meditators, we are
+as of yet not exactly sure what we might be studying when we do study
+them. V.S. Ramachandran put this most succinctly:
+
+``I can't think of a single discovery of disease which had more than one initial
+sample. .... You can't do statistical analysis of an initial discovery.''
+\cite{ramachandran2019relevance}
+
+\subsubsection{Meditation as Research Tool}
+
+A lifelong study of a single, increasingly-experienced Vipassana meditator
+is, in essence, an exploratory process of discovering what researchers might be
+bothered to study in a (single) blind trial with active and inactive controls, over a
+much larger sample. This process could be described as xxx
+
+\todo{(Harari) / (Altered Traits) quotes}
+
+Research tool initially, and research tool + research subject later on.
+
+\subsubsection{Summary}
+
+Because Vipassana meditators are inherently self-selecting, double blinding is not
+possible, nor are randomized trials. Single blinding is possible, and should be
+employed in future studies where it is feasible. Combination active/inactive controls
+are also possible for all \textit{statistical} studies on Vipassana.
+
 \subsection{Hypothesis}

### 2019-12-03 paper-three/paper-three.tex
proposal: finally done the intro, maybe
proposal: cite OPM MEG research paper, move method to method section

@@
   \item digestion (``Make sure to pay attention to your poops!'' first course)
-  \item diet ^
+  \item diet
   \item schedule
@@
 \cite{hauke2018}
+
+\textbf{vedana-samosarana sabbe dhamma. ``Everything
+that arises in the mind starts flowing with a sensation
+on the body.''} \cite{goenka1999discourses}
 
@@
 
+\bibitem{goenka1999discourses}
+  Discourses on satipatthana sutta
+%  @article{goenka1999discourses,
+%  title={Discourses on satipatthana sutta},
+%  author={Goenka, Satya Narayan},
+%  journal={Igatpuri, Maharashtra, India: Vipassana Research Institute},
+%  year={1999}
+%}
+
 \bibitem{nagendra2012}
@@
   \url{https://doi.org/10.3389/fneur.2012.00054}
+
+\bibitem{maruthai2016senior}
+  Maruthai N., Nagendra R.P., Sasidharan A., Srikumar S., Datta K., Uchida S., Kutty B.M.
+  \textit{Senior Vipassana Meditation practitioners exhibit distinct REM sleep
+    organization from that of novice meditators and healthy controls.}
+  International Review of Psychiatry, Volume 28(3), Pages 279-287, April 2016.
+  \url{https://doi.org/10.3109/09540261.2016.1159949}
 

### 2019-12-03 the-proposal/the-proposal.tex
proposal: cite OPM MEG research paper, move method to method section

@@
 
+\subsection{Hypothesis}
+
+\section{Objectives and Aims}
+
+\subsection{Overall Objective}
+\subsection{Specific Aims}
+
+\section{Background and Significance}
+
+\subsection{Preliminary Research Review}
+- summary of related research
+- strengths and weaknesses
+- justification
+--- what hasn't been done by others?
+--- why is this research necessary?
+
+\subsection{Why Vipassana?}
+
+1. ``complete'', standardized, global, multilingual, 100\% free
+2. mundane (sleep) vs. supramundane (total eradication of suffering)
+
+3. who is steven deobald
+
+\subsection{Why now?}
+
+\section{Research Design and Methods}
+
+\subsection{Overview}
+
 As the structure of studies conducted on Vipassana is inherently very difficult, due to
-the nature of Vipassana meditation instruction, we must first examine which kinds of
+the strict nature of Vipassana meditation instruction, we must first examine which kinds of
 studies are possible, which are not, and what will make possible studies worthwhile.
@@
 the instruction and its target audience. On the lifelong scale of Vipassana
-practice, the spectrum spans from laypeople practicing two hours per day to
-renunciates (monks and nuns) who practice up to twenty-four hours per day once they
-reach the stage where deep sleep no longer occurs due to the practice itself. It is
+practice, the spectrum spans from a minimum of laypeople practicing two hours per day to
+a maximum exemplified by renunciates (monks and nuns) who practice up to twenty-four
+hours per day once they
+reach the advanced stage where deep sleep no longer occurs. It is
 also worth noting that a junior renunciate will still have less \textit{experience}
-than an senior lay meditator, which means that the only objective measure of
+than a senior lay meditator, which means that the only objective measure of
 experience is total number of hours meditated. \cite{goleman2017altered}
@@
 hours of experience are accumulated (hours per day). We must capture both, as accurately as
-possible. With a high hours-per-day rate and extended (lifelong) experience, the very
-broad category of \textit{expert} meditators quickly requires upward of 100,000 hours
+possible. Renunciates potentially have both a high hours-per-day rate and extended
+(lifelong) duration of experience. The existence of renunciates forces the very
+broad category of \textit{expert} meditators into the terrority of 100,000 hours
 of practice, even if we limit practice hours to formal, sitting meditation and
 estimate twelve (12) hours per day. If we take the much more modest rate of six (6)
-hours per day, assume an adult can learn meditation from approximately 20 years of
-age, and is of an age when they may participate in an extended study (say, 50-60
+hours per day, assume an adult renunciate can learn Vipassana, at the earliest, from
+approximately 20 years of
+age, and is now of an age when they may participate in an extended study (say, 50-60
 years of age), we arrive at roughly 60,000 hours of practice. If we use the oft-cited
@@
 Vipassana meditators are often predominantly observing beginners (7.9-8.6 mean years
-of experience with 2 hours of daily practice).
+of experience with 2 hours of daily practice). \cite{chiesa2010vipassana}
 
@@
 (altered states) and long-term outcomes (altered traits \cite{davidson1977role}) to
-improve future Vipassana research. \cite{chiesa2010vipassana} New developments into
-magnetoencephalography (MEG) and quantum MEG over the past decade may permit it as a
-third neuro-imaging technique.
+improve future Vipassana research. \cite{chiesa2010vipassana}
+Optically-pumped magnetometer (OPM)-based magnetoencephalography (MEG) performed using a
+mobile helmet capable of operating at room temperature \cite{boto2017new}, in development since Chiesa's
+paper was published, may permit it as a third neuro-imaging technique. Even older MEG
+technology based on superconducting quantum interference devices (SQUIDs) could
+potentially be used. The restriction of SQUID-based MEG is that subjects must remain
+extremely still, but this is all but a requirement of serious Vipassana meditation
+anyway.
 
@@
 through philosophical phenomenology becomes a hindrance to actually practicing
-Vipassana. \cite{patrik1994phenomenological} I have previously discussed this paradox
-in Appendix A of \textit{Vipassana for Hackers, Paper One: Curious Mechanics}.
+Vipassana. \cite{patrik1994phenomenological} The two are mututally exclusive, as is
+any attempt to contemplate Vipassana, the phenomenology of the mind, or the
+technique, during the actual practice of Vipassana. I have previously discussed this
+apparent paradox in Appendix A of \textit{Vipassana for Hackers, Paper One: Curious Mechanics}.
 \cite{deobald2017curious}
@@
 are also possible for all \textit{statistical} studies on Vipassana.
-
-\subsection{Hypothesis}
-
-\section{Objectives and Aims}
-
-\subsection{Overall Objective}
-\subsection{Specific Aims}
-
-\section{Background and Significance}
-
-\subsection{Preliminary Research Review}
-- summary of related research
-- strengths and weaknesses
-- justification
---- what hasn't been done by others?
---- why is this research necessary?
-
-\subsection{Why Vipassana?}
-
-1. ``complete'', standardized, global, multilingual, 100\% free
-2. mundane (sleep) vs. supramundane (total eradication of suffering)
-
-3. who is steven deobald
-
-\subsection{Why now?}
-
-\section{Research Design and Methods}
-\subsection{OVERVIEW}
 

### 2019-12-04 the-proposal/the-proposal.tex
proposal: study design

@@
 
+\begin{quotation}
   ``I can't think of a single discovery of disease which had more than one initial
-sample. .... You can't do statistical analysis of an initial discovery.''
+  sample. ... You can't do statistical analysis of an initial discovery.''
   \cite{ramachandran2019relevance}
+\end{quotation}
 
@@
 is, in essence, an exploratory process of discovering what researchers might be
-bothered to study in a (single) blind trial with active and inactive controls, over a
-much larger sample. This process could be described as xxx
+bothered to study in a controlled trial with both active and inactive controls and
+(single) blinding, over a much larger sample. This exploratory process has been
+described by Goleman and Davidson in \textit{Altered Traits}:
 
-\todo{(Harari) / (Altered Traits) quotes}
+\begin{quotation}
+  ``Perhaps one day an ultralong study will give us the equivalent of video on how
+  altered traits emerge. For now, as the Brewer group conjectured, meditation seems
+  to transform the resting state---the brain's default mode---to resemble the
+  meditative state.
 
-Research tool initially, and research tool + research subject later on.
+  Or, as we put it long ago, the after is the before for the next
+  during.'' \cite{goleman2017altered}
+\end{quotation}
+
+The demands of such an ``ultralong study'' were echoed by Harari in \textit{21
+  Lessons for the 21st Century}:
+
+\begin{quotation}
+  ``Some universities and laboratories have indeed begun using meditation as a
+  research tool rather than as a mere object for brain studies. Yet this process is
+  still in its infancy, partly because it requires an extraordinary investment on the
+  part of the researchers.'' \cite{harari201821}
+\end{quotation}
+
+The line between ``mere object for brain studies'' and ``research tool'' begins to
+blur when we consider using meditation for both simultaneously. However, it seems remiss to neglect
+gathering whatever objecting EEG/fMRI/MEG data from the long-term subject-researcher
+when that data is so readily available.
 
@@
 
-\subsection{Study Design / Research Method}
-- case study type?
+The study proposed is an ultralong (lifelong) study of an individual --- myself. As
+far as quantitative data gathered from brain scans is concerned, I am effectively
+making myself available as a human guinea pig. As far as qualitative data is
+concerned, research will take the form of a deep case study. As I will represent
+Ramachandran's ``initial sample'', statistical analysis will not be relevant. If the
+study proves productive, future studies targeting a specific hypothesis about
+Vipassana in a large sample can use learnings from this study in their statistical
+structure.
+
+\subsection{Study Design}
+
+\subsubsection{Qualitative}
+
+Qualitative research into the specific nature of human consciousness or the broader
+nature of consciousness as it applies to any organism or artificial intelligence can
+only be performed directly. In this case, a case study performed by an individual
+(myself) will explore the question of consciousness and the mind-body problem
+directly, through Vipassana meditation, for a prolonged period of time (the remainder
+of my life). Many theories and questions currently exist regarding
+consciousness, such as \textit{Theory of Mind}, \textit{Theory of Panpsychism}, and
+the \textit{Simulation Hypothesis}. This study will not address any one of them
+directly, nor will it absorb any specific hypothesis. The data recovered from
+practicing Vipassana meditation directly for prolonged periods of time will almost
+certainly overlap with many such ideas but themes and narratives are much more likely
+to be emergent, rather than conforming to an existing hypothesis of consciousness.
+
+Phenomenology, in the broadest sense, may be used to describe experiences as they
+pertain to a lifelong trajectory of altered traits. Phenomenological descriptions of
+deep meditative states, or even of altered traits, is not the goal,
+however. If other narrative tools emerge over the course of the study, I will augment
+phenomenological descriptions and imagery with those tools.
+
+Vipassana's instruction provides us with claims we can evaluate: ``[Vipassana] explores the entire field of
+mind and matter'' \cite{goenka1999discourses}, ``[Vipassana] is a technique that will
+eradicate suffering'' \cite{dhammaorg2014code}, and ``Vipassana aims at ... total
+liberation and full enlightenment'' \cite{dhammaorg2014code}.
+
+\subsubsection{Quantitative}
+
+Measurements performed during the lifelong qualitative study
 

### 2019-12-04 the-proposal/the-proposal.tex
proposal: quantitative case study description

@@
 of my life). Many theories and questions currently exist regarding
-consciousness, such as \textit{Theory of Mind}, \textit{Theory of Panpsychism}, and
-the \textit{Simulation Hypothesis}. This study will not address any one of them
+consciousness, such as \textit{Theory of Mind}, \textit{Theory of Panpsychism},
+\textit{Theory of the Entropic Brain},
+the \textit{Simulation Hypothesis}, etc. This study will not address any one of the
+many theories of consciousness
 directly, nor will it absorb any specific hypothesis. The data recovered from
 practicing Vipassana meditation directly for prolonged periods of time will almost
-certainly overlap with many such ideas but themes and narratives are much more likely
+certainly overlap with many such ideas but themes and narratives are likely
 to be emergent, rather than conforming to an existing hypothesis of consciousness.
@@
 eradicate suffering'' \cite{dhammaorg2014code}, and ``Vipassana aims at ... total
-liberation and full enlightenment'' \cite{dhammaorg2014code}.
+liberation and full enlightenment'' \cite{dhammaorg2014code}. Progress made in
+evaluation of these claims will be, by its very nature, emergent.
+
+The validity of these emergent observations made during the qualitative portion of
+the study can be asserted across most axes of validity: prolonged engagement, rich
+descriptions, external audits (from more experienced meditators), identification of
+researcher bias, peer debriefing (again from more experienced meditators), and
+searching for discrepant evidence are all possible. Respondent validation (member
+checking), although possible, may not carry much weight in terms of strengthening the
+validity of the study, given an initial sample of one. Triangulation will not be
+possible with a sample of one. If the study proves productive, in the future
+additional researcher-subjects intent on a lifelong Vipassana practice may strengthen
+respondent validation and triangulation for parallel studies. Triangulation of
+qualitative analysis of consciousness suffers from the paradox that any individual's
+consciousness is only directly observable by that individual and conclusions --- even
+those validated by triangulation --- will always be in the third-person.
+
+Data will be collected on a daily basis and themes and narratives regularly collected
+with the intention of
+describing mental phenomena and the evolution of traits as Vipassana practice
+progresses. Monthly or yearly schedules will be decided with a
+supervisor but an example daily schedule (subject to change before the study begins)
+is available in \textit{Appendix 1: Proposed Daily Schedule}.
 

### 2019-12-25 paper-four/paper-four.tex
one last note for p4
first commit on the proposal... here we go.
p4: abstract and target audience

@@
 \begin{abstract}
- \todo{Write the abstract.}
+ \todo{After someone has completed their first 10-day Vipassana course there often
+   remain a number of open questions regarding the practice. How, exactly, is the
+   meditation removing mental negativities? Is this really the be-all-end-all of
+   meditation techniques? Is the teaching methodology really systematic and global?
+   This paper attempts to answer these questions and others.}
 \end{abstract}
@@
 The target audience for this paper remains ``hackers'' for the broadest possible
-definition of that word. \todo{ref: from cherry?}
+definition of that word. \todo{add ref: Hackers book} If you are a curious and
+analytical person, you qualify. The new restriction placed on this paper is that you
+must have \textit{completed} at least one 10-day Vipassana course (as taught by
+S.N. Goenka). To ``complete'' the course means that you left the course centre on the
+eleventh day and no sooner. Although Sayagyi U Ba Khin authorized many other teachers
+in addition to S.N. Goenka, the portions of this paper which deal with the systematic
+and global approach to the teaching require taking a course from S.N. Goenka or one
+of his authorized Assistant Teachers. Sayagyi U Ba Khin's other students did not
+construct comparably systematic teaching methods.
 
-\begin{itemize}
-  \item Old students (Goenka; forward ref other methods of learning)
-  \item People looking to deepen practice
-  \item Meditators who haven't chosen a technique yet
-\end{itemize}
+This paper will discuss alternative methods of Vipassana instruction to S.N. Goenka's
+system but even those discussions will be within the context of S.N. Goenka's
+instruction as the global standard.
+
+Readers may have chosen to dedicate their lives to the technique of Vipassana or they
+may still be searching. Portions of this paper will be dedicated to analyzing reasons
+to dedicate oneself to a single technique --- as well as merits and drawbacks of
+choosing Vipassana as that technique.
+
+The paper is best suited to Vipassana meditators who are looking to deepen and expand
+their practice. It is important to remember, however, that this paper is authored by
+non-teaching meditators and is no replacement for expert advice from a trained and
+authorized Assistant Teacher.
+
+With these restrictions in mind, anyone is of course welcome to read the
+paper. However, much of what is being said will seem unnecessarily recondite to
+someone without the requisite practical meditation experience.
 
@@
 
-Anapana eliminates space.
+Anapana eliminates space. Continuity is the source of constantly-narrowing attention on space.
 
@@
 - Day 8: Meditate all the time.
+
+- structure: the requirement that meditation be taught for free (expand)
 

### 2019-12-26 paper-four/paper-four.tex
p4: comparison to other meditation techniques

@@
 
-\usepackage[english]{babel}
+\usepackage[english,hindi]{babel}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{hyperref}
-
-\setlength{\parindent}{1.3em}
-\setlength{\parskip}{0.7em}
 
@@
 
+From one's very first Vipassana course --- in fact, from the discourse on the second
+day --- a meditator is met with two messages: that of \textit{Eka Maggo} (the one and
+only path out of suffering) and the idea that the Noble Eightfold Path is
+``complete''. This section will address these two issues.
+
 \subsection{Learning Vipassana Elsewhere}
 
-\begin{itemize}
-  \item Ledi Sayadaw: 5 hours a day anapana, ``hairs of the head and the body''
-  \item Other students of U Ba Khin: ``The Quiet Mind'', ``Beyond the Breath''
-  \item ``Vipassana'': sensations matter more than terminology (3-day course, 10-day course), other forms of ``Vipassana'' not of interest, Zazen
-\end{itemize}
+For new meditators, Goenkaji addresses this issue during early
+courses. Although it isn't a point that is repeatedly addressed, there are two quotes
+worth highlighting in this regard:
+
+\begin{quote}
+  \textbf{You call it ``Vipassana'' --- you call it whatever you like.}
+\end{quote}
+
+\todo{add ref to quote}
+
+\begin{quote}
+  \textbf{Any technique which deals with sensations on the body is a technique of
+  liberation.}
+\end{quote}
+
+\todo{add ref to quote}
+
+Dissecting these two ideas quickly brings us to the conclusion that Goenka's
+instruction in no way implies that he is the only legitimate teacher of Vipassana ---
+just the most accessible. Anyone who has ever attempted a 10-day self course will
+attest to the sheer difficulty of attempting to recreate the environment of a
+Vipassana Centre in one's own home. But it can be done and is even recommended in the
+Day 11 discourse before the course ends. \todo{add reference to discourse}
+
+Between the 10-day introductory course, the 3-day course, and the Satipatthana Sutta
+course, the messaging becomes very clear: Vedana (sensations) are of utmost
+importance and terminology is of no importance. Many techniques refer to themselves
+as ``Vipassana'' but if they do not deal with sensation they are not relevant to the
+discussion, as we will see below in \textit{How is Vipassana ``Complete''?} Other
+techniques, such as Zazen and Kundalini, which deal with sensation at least in part
+are of interest to a meditator convinced of the value of sensation analysis. It is up
+to each meditator to decide whether they prefer one technique or the other for
+analyzing sensation and pursue that one technique vigorously. The reasons for
+choosing only a single technique will be discussed in \textit{Warnings: Do not mix
+  meditations}.
+
+If a meditator finds herself in the position of preferring Vipassana, with an
+emphasis on Vedana and Anicca, but preferring, for one reason or another, not to
+learn the technique from S.N. Goenka and his Assistant Teachers, there are other
+methods for learning within the same lineage.
+
+Going back one generation, it is possible to learn from the other students of Sayagyi
+U Ba Khin. Though most of them are getting quite old, some do still teach. John
+Coleman describes teaching his first 10-day course in the UK in \textit{The Quiet
+  Mind} \todo{add ref}, for example. Glickman, another student of Sayagyi U Ba Khin,
+attempts another strategy by laying out a progressive and systematic at-home training
+plan for readers in his book \textit{Beyond The Breath: Full Body Vipassana
+  Meditation} \todo{add ref}.
+
+We cannot go back two generations as Saya Thet Gyi did not leave any written
+instructional material, nor any known authorized teachers other than his student,
+Sayagyi U Ba Khin. However, we can go back three generations to Saya Thet's teacher:
+Ledi Sayadaw. Ledi Sayadaw was both a master meditator and accomplished scholar and
+wrote prolifically in Burmese and Pali. Although his work is often deeply arcane, in
+his \textit{Manuals of Dhamma} and \textit{Anapana Dipani} \todo{add ref} he
+specifies his instructions for practicing Anapana and Vipassana as a layperson:
+
+\begin{quote}
+  A person who wants to practise vipassanā, being an ordinary human being,
+    may not find it possible to put forth effort twenty-four hours a day. He must
+    therefore allocate three or four hours a day and put
+    forth effort punctually and regularly every day. When he starts to practise, he must
+    first overcome the wandering tendencies of the mind and establish mindfulness on the
+    breath. It is only after he has overcome the mind’s wandering tendencies that he can
+    direct the mind towards vipassanā.
+
+    ....
+
+    The way to rid oneself of [the belief that is firmly and deeply rooted in the minds
+      of worldlings (puthujjana)] is as follows: When the eye of wisdom penetrates to
+    these four primary elements and the ultimate reality is perceived, such things as
+    shape and form in the out-breath and in-breath disappear, and every time one
+    contemplates them, the deep and firm root of personality view disappears. One
+    perceives that there is in reality no shape and form --- no out-breath and
+    in-breath. One perceives that there exist only the four primary elements. Thus
+    purity of view (diṭṭhi-visuddhi) is achieved.
+
+    ....
+
+    It is the same with respect to the other parts of the body such as head-hairs,
+    body-hairs, etc. There exists, on the one hand, the deeply rooted habitual
+    perception of shape and form, such as, ``This is head-hair,'' and on the other, there
+    exist the four primary elements. When these four primary elements are penetrated
+    and clearly perceived with wisdom in the head-hairs, the deeply rooted wrong
+    perception of shape and form will disappear. It will be perceived that the
+    head-hairs do not exist in reality. When it is thus seen, purity of view in the
+    head-hairs is achieved. Proceed in the same way in the case of the other parts of
+    the body such as body-hairs, etc.
+\end{quote}
+
+Ledi Sayadaw's recommendations in \textit{The Manuals of Dhamma} are similar. To
+parapharse: Practice Anapanasati by observing the in-breath and the out-breath at the
+``point of touch'' above the upper lip for 4 to 5 hours a day for a few months. Then
+investigate the hairs of the head and the hairs of the body.
+
+One assumes subtle sensation would be fairly easy to detect if each individual hair
+is easily visible to the attention at the \textit{beginning} of one's Vipassana
+practice.
 
@@
 
-- 7 doors
-- TM says to use your imagination exclusively, with a mantra
-- Anapana variations say to use the breath (sensation) exclusively, with a limitation to a particular area
-- Void-based meditations (yoga, elsewhere) suggest you can divorce yourself of attention from any one of these phenomena… which is pretty unlikely for a beginner
-- “Sound of Silence” focuses on the ear sense door + tinitus
-- other forms of Vipassana focus entirely on thoughts
-- Zazen, generally, says open yourself to all 7 sense doors, all at once.
-... they all choose an object of meditation within these 7 doors.
+There is a finite set of possible destinations for our attention, a finite set of
+possible meditation objects. These can be
+enumerated in various ways but all enumerations tend to be equivalent to the six
+sense doors of traditional Buddhist theory (eyes, ears, tongue, nose, body skin, and mind) plus
+the seventh sense door of Vedana
+(sensation) inside the body. Some Westernized variations of this enumeration choose
+to divide the sixth sense door of \textit{mind} into two categories of
+\textit{thought} and \textit{emotion}, however, this division will provide a
+limitation in deeper meditative states where states of mind no longer resemble either
+of these two categories. Rather than devise new terminology, ever dividing mind
+states into narrower and more specific subcategories, it makes the most sense to
+simply leave the sixth sense door of \textit{mind} intact.
+
+\todo{add 8-sense diagram}
+
+The other common
+complication of categorization is that of \textit{body and body skin}, which is ultimately the sense
+door at which Vedana arise. Even within the scope of a 10-day Vipassana course,
+sensation is first addressed where it is most accessible: the body skin. Gross,
+solidified Vedana (say, internal sensation such as
+a stomach ache or external sensation such as a feather on the skin) can be thought of
+as the sense object and subtle Vedana (perhaps elsewhere on the body) can be thought of as the corresponding Vedana
+understood to be addressed in observation of impermanence (Anicca and Sampajañña
+\todo{add ref to importance of v and s}). This oversimplification will eventually prove insufficient, however,
+as the sense-door bodily sensations and Vedana of Sampajañña are recursively
+codependent --- you cannot have one without the other. Ultimately, they are ``two
+sides of the same coin'' \todo{add ref to discourse 10}.
+
+\todo{add sense reflection diagram}
+
+Having the experience of at least one 10-day Vipassana course, one understands ---
+experientially --- the relationship of sense door contact to sensation on the
+body. That is, when contact is made at any sense door there is a sensation on the
+body and it is possible to observe this sensation, objectively. In meditation, the
+most important sense door tends to be the mind. The meditator may go to a quiet,
+uninhabited place, sit down cross-legged, and proceed to close her eyes and
+mouth. It is easy to physically stopple the external sense doors. But the sense door
+of the mind may only become louder on such an occasion. The sense door of the mind is
+also reflective of the other five senses: it remembers and anticipates contact at the
+other five sense doors, in addition to its own unique processes of automatic
+verbalization and emotional habit. Regardless of which sense door is activated, mind
+or otherwise, a reflective sensation occurs within the body. This is not a point
+which can be explained in text or conveyed from person to person. It must absolutely
+be experienced by the individual.
+
+With both this reflective property of Vedana and the finite set of meditation
+objects in mind, let us examine some other meditation techniques.
+
+\textbf{Transcendental Meditation (TM)} is a popularization of mantra meditation,
+found in various forms of Yoga and elsewhere. It employs a system of bija
+mantras \todo{add reference to bija mantra list}, silently recited. In terms of sense
+doors, five are stoppled, the mind sense door is employed with a remembered object of
+the sound sense door. Bodily sensation is ignored.
+
+\textbf{Anapanasati} is the broad spectrum of meditations which employ the in-breath
+and out-breath as the object of meditation. The breath can only be \textit{felt} and
+therefore relies on either body-skin sensation or other Vedana. There are variations
+which place the attention at various points within the body: the tip of the nose, the
+upper lip, the passage of breath down to the diaphragm or navel, exclusively on
+the stomach, etc. These all demand the body as meditation object and thus rely on
+sensation as the door of perception for meditation but, in general, variations of
+Anapanasati do not extend to the \textit{entire} body. As some sensation on portions
+of the body irrelevant to the in-breath and out-breath are ignored, such meditation
+techniques can never be considered complete in the sense that they do not and cannot
+encompass the entirety of sensory experience. Exceptions do exist. Most notably,
+Anapanasati instruction by Webu Sayadaw, which directs attention to the ``point of
+touch'' above the upper lip but admits that very narrow Anapanasati, when continued
+for 24 hours per day at length, will eventually lead to awareness of sensation across
+the entire body.
+
+\textbf{Void (Emptiness) Meditations} found in some Yogic and Zen schools and
+elsewhere suggest that the meditator can
+divorce herself from attention entirely, removing attention from any one of these
+phenomena and placing it on nothing. Such a technique is unlikely to succeed for
+beginners and the path provided in Void Meditation teachings inherently begins with a
+mental ``visualization'' at any one of the six sense doors: an absence of sight
+object, sound object, feeling object, etc. Where an apparent void is possible to
+observe directly, a meditator may have attained access concentration \todo{add ref to
+  access concentration} or a jhana \todo{add ref to jhana}. However, attention
+directed toward a minimum (or zero, if that is possible) sense doors is inherently
+attention which misses something. By its very nature, it is attention which tries to
+miss as many inputs as possible. Such a meditation cannot be considered complete,
+given the vast surface area of sense contact which is intentionally missed.
+
+\textbf{Sound of Silence / Nada Yoga} is a void-adjacent meditation technique taught
+by Ajahn Amaro and others which, paradoxically, narrows attention specifically to the
+ear sense door. For most people ``the sound of silence'' is really the sound of
+tinitus \todo{add reference to tinitus} as the human ear is largely incapable of
+hearing nothing at all. As Anapanasati may eventually lead to full-body Sampajañña,
+so may the Sound of Silence. Sayagyi U Ba Khin describes access to full-body Vedana
+by means of the other five sense doors (other than the body skin) but insists that
+the body skin is by far the easiest and most accessible to lay meditators. \todo{add
+  ref to u ba khin discourse}
+
+\textbf{Vipassana (Cittanupassana)} \todo{add ref to cittanupassana} as taught by
+Mahasi Sayadaw and others is perhaps
+even less accessible than the Sound of Silence as an alternative sense door to
+internal Vedana. Cittanupassana, taught without Vedana as the meditation object,
+tends to embroil the meditator in the objects of gross, surface-level thought and
+emotional processes. Although it is theoretically possible for a meditator to work
+with sensation-based Sampajañña through this method, it would seem extremely unlikely
+for anyone but a monk or nun to do so.
+
+\textbf{Zazen} generally suggests that the meditator open herself to all seven
+(counting Vedana of the body and body-skin as separate) sense doors,
+simultaneously. In this sense, Zazen can be thought of as ``complete'' but from
+research of lay meditators it seems extremely difficult for the meditator to practice
+to a point where subtle sensation is felt on or inside the body.
+
+\textbf{Vipassana} asks the meditator, initially, to stopple the five external sense
+doors. Sit in a shunyagar (empty room or cell), in reduced light, in a quiet and uninhabited
+environment, with eyes and mouth closed. Since the mind is active, attention is
+repeatedly returned to bodily sensation: first at the narrowest possible location
+under the nostrils during the Anapana period and second at the widest possible set of
+locations --- across the entire body --- during the Vipassana period. As the six
+sense bases are fully reflected in bodily sensation, Vipassana on full-body sensation
+will (eventually) cover the entire field of sensory, mental, and supramundane
+experience. An explanation is provided in the context of Spacetime, below.
+
+\subsection{Mastering Spacetime}
 
@@
 
-- structure: the requirement that meditation be taught for free (expand)
+\section{structure: the requirement that meditation be taught for free (expand)}
 
@@
   \item Do not use meditation to get high
-  \item Do not mix meditations (1. dangers, 2. meditation ``doing itself'' - anatta)
 \end{itemize}
+
+\subsection{Do not mix meditations}
+
+(1. dangers, 2. meditation ``doing itself'' - anatta)
 

### 2019-12-26 paper-four/paper-four.tex
mastery over spacetime

@@
 
-- mind / body === brain / nervous system
+The completeness of a meditation can only be understood based on three criteria:
 
-- vipassana the largest superset
+\begin{itemize}
+  \item All time: It must cover all of time, for the smallest unit of time
+  \item All space: It must cover all of space, for both the smallest unit of space and the largest
+  \item Continuity: It must be practiced continuously, in every moment of life
+\end{itemize}
 
-- anapana covers time, vipassana covers space
+In this way, we have a meditation which covers all of time and all of space, for all
+available time. No ``largest unit of time'' is requested as time, directly perceived,
+can only be of the present moment --- and the present moment is infinitesimally small
+by its very nature. In place of the ``largest unit of time'', all moments of
+perceived time are to be captured: a total continuity.
 
-Anapana eliminates space. Continuity is the source of constantly-narrowing attention on space.
+In this section we will examine how, even within the 10-day Vipassana course, the
+teaching of Vipassana actually covers the largest possible superset of all other
+meditation techniques.
 
-It says “focus here, on this tiny area… make the area as tiny as possible, with no
-limit or exception. Now all you have to do is follow your breath in the tiniest
-slices possible, making time stretch out longer and longer with every sub-slice.”
+\subsubsection{Anapana}
 
-Anapana is mastery over time.
+Anapana, as taught in the 10-day course, Satipatthana Sutta course, and 20-day
+course, demands that the meditator make an attempt at \textit{eliminating space}: the
+attention placed below the nostrils is to become narrower and narrower as awareness
+and attention become more acute.
 
-Vipassana is harder.
-It stops ignoring space and actually adds it to the equation.
-You move your attention throughout your entire body (which is the only thing you have direct access to: sensation) with the narrow focus you trained yourself to achieve in the first 3 days of the course.
-As your apparent physical existence begins to break apart, you’ll open up your attention to the whole body, all at once (or as much as possible at once) but even after 9 Vipassana courses, I rarely ever do anything approximating that because I’m quite a beginner.
-🙂
+Item 3, continuity, already makes an appearance in the practice of Anapana. In
+addition to narrowing the area of awarenss, a secondary goal of Anapana Meditation is
+to fix the attention to this narrow area for as long a period as possible. Initially,
+this continuity is understood in terms of mundane time: 5 minutes, 15 minutes,
+etc. However, it is a natural by-product of unbroken attention that it makes apparent
+a new, subtler form of continuity. Between each breath, there is the time it takes
+for the length of the in-breath and out-breath themselves. If we break the in-breath,
+for example, into three pieces (beginning, middle, end) it is then natural to find
+that, after one has sustained continuous awareness of these subsections of the
+breath, that these three pieces can each be broken down further into smaller
+slices. And so on. It is in
+this way that Anapanasati teaches the meditator to meditate on \textbf{all of time}.
 
-Anyway.
+\todo{add anapana slice diagram}
 
-Vipassana is mastery over time + space.
+Anapana eliminates space. Continuity is the source of constantly-narrowing attention
+on space. Anapana is mastery over time.
 
-...once you have that, there’s nothing left.
+\subsubsection{Vipassana}
 
-- Day 8: Meditate all the time.
+Vipassana is harder. It stops ignoring space and actually adds it back into the
+equation. You move your attention throughout your entire body (which is the only
+meditation object you have direct access to) with the narrow focus you trained
+yourself to achieve using Anapana. As your apparent physical
+existence begins to break apart, you’ll open up your attention to the whole body, all
+at once (or as much as possible at once) but even after 10 Vipassana courses, I rarely
+ever do anything approximating that because I’m quite a beginner. 🙂
+
+Vedana provides a sort of backdoor to all contact at the other six sense
+doors. Direct observation of Vedana can only be of our own body. By observing the
+entire body and ``dividing and dissecting'' it with our attention, we eventually
+observe all Vedana and, therefore, all contact activity at the sense doors. This is
+all the space there is to meditate on. We are meditating on \textbf{all of space}.
+
+Vipassana maintains the mastery of time and continuity demanded in Anapana, adding
+all of space. Vipassana is mastery over time + space.
+
+\subsubsection{Continuity}
+
+On Day 8 of the introductory 10-day Vipassana course, a new meditator is introduced
+to the idea of complete continuity: meditating all the time. In the last days of the
+10-day course, the meditator is asked to meditate in all four postures (sitting,
+standing, walking, lying down) during all activities (walking, eating, using the
+toilet, bathing, washing, sleeping, etc.).
+
+Old student courses such as the 3-day course, Satipatthana Sutta course, and 20-day
+course emphasize this point more strictly, beginning in the Anapana period. The
+notion of continuity is expanded to Anapana and the idea of potentially meditating 24
+hours a day is introduced. As the meditator begins meditating in all positions, and
+particularly while lying down, sleep is naturally replaced with
+meditation. Eventually, the conscious/unconscious barrier is broken even during
+hours of dreaming and hours of deep sleep. One becomes conscious of bodily sensation,
+either through the breath or throughout the entire body, 24 hours a day --- a true
+continuity.
+
+Thus, the teaching fulfills our initial three criteria: Vipassana, practiced
+continuously, is a meditation on all of space, all of time, all the time.
+
+Vipassana is mastery over Spacetime.
 

### 2019-12-26 paper-four/paper-four.tex
p4: formatting

@@
 paper. However, much of what is being said will seem unnecessarily recondite to
-someone without the requisite practical meditation experience.
+someone without the requisite practical meditation experience. For meditators and
+non-meditators, both, it is recommended that one read the \textit{Warnings} section
+--- in particular, \textit{This document is a hindrance} --- before preceding
+further.
 
@@
 only path out of suffering) and the idea that the Noble Eightfold Path is
-``complete''. This section will address these two issues.
+``complete''. The following sections will address these two issues.
 
-\subsection{Learning Vipassana Elsewhere}
+\section{Learning Vipassana Elsewhere}
 
-For new meditators, Goenkaji addresses this issue during early
+For new meditators, S.N. Goenka addresses this issue during early
 courses. Although it isn't a point that is repeatedly addressed, there are two quotes
@@
 instruction in no way implies that he is the only legitimate teacher of Vipassana ---
-just the most accessible. Anyone who has ever attempted a 10-day self course will
+he is simply the most accessible. Anyone who has ever attempted a 10-day self course will
 attest to the sheer difficulty of attempting to recreate the environment of a
@@
 Between the 10-day introductory course, the 3-day course, and the Satipatthana Sutta
-course, the messaging becomes very clear: Vedana (sensations) are of utmost
+course, the messaging becomes very clear: Vedana \todo{add ref to vedana} (sensations) are of utmost
 importance and terminology is of no importance. Many techniques refer to themselves
@@
 
-\subsection{How is Vipassana ``Complete''?}
+\section{How is Vipassana ``Complete''?}
 
@@
 
+\textbf{Kundalini} is the practice of repeatedly examining the chakras, points
+largely located along the spinal cord. This is actually the final location of
+attention in Vipassana, once the body-skin and internal body are dissolved to the
+meditator's awareness. Kundalini can be thought of as a very specific subset of
+Vipassana.  \todo{add ref to q \& a}
+
 \textbf{Zazen} generally suggests that the meditator open herself to all seven
@@
 
-\subsection{Mastering Spacetime}
+\section{Mastering Spacetime}
 
@@
 
-\subsubsection{Anapana}
+\subsection{Anapana}
 
@@
 
-\subsubsection{Vipassana}
+\subsection{Vipassana}
 
@@
 
-\subsubsection{Continuity}
+\subsection{Continuity}
 
@@
 
-\section{structure: the requirement that meditation be taught for free (expand)}
+\section{Structure: Cost}
+
+I expect that many readers of this paper will be using it to triangulate their
+opinion on their personal choice of lifetime meditation technique. I hope that it is
+helpful in that regard and I would like to add one point, orthogonal to technique, to
+that matter: cost.
+
+Meditation, whatever technique you ultimately choose (and it is strongly advised that
+you choose only one: see \textit{Do not mix techniques} below), must be taught for
+free. The teacher, his or her organization, staff or servers, the facilities --- this
+must all be made available for free.
+
+Commercial meditations are always going to cause problems for the meditator for the
+simple reason that, while one may progress a great deal in one's own meditation while
+paying for it, the question of universality will arise in the mind of the meditator,
+sooner or later. \textit{Can everyone practice this technique? As I am learning it?}
+If not, the technique cannot be considered universal. If the question of universality
+has not yet arisen in your mind, trust that it will. A meditation capable of
+liberating only an individual, or only those who can afford it, is not much of a
+meditation at all. The teaching is unlikely to have longevity, to persist, and the
+value is unlikely to emerge as a global phenonmenon.
+
+The idea of \textit{Eka Maggo}, the one and only path, may be interpreted any number
+of ways. Some interpretations may even be paradoxical or inherently
+self-referential. However, the very concept of a single path implies that a technique
+is both \textit{complete} and \textit{universal}. It is strongly suggested you
+repeatedly apply this test to any technique you are considering for lifetime
+practice.
 
@@
 \begin{itemize}
-  \item This document itself is a hindrance
+  \item This document is a hindrance
   \item Don't get bogged down by ``perfect objectivity'' or ``perfect equanimity'' - yes muscles and blood flow follow your attention
@@
 
-\subsection{Do not mix meditations}
+\subsection{Do not mix techniques}
 

### 2020-03-29 paper-three/paper-three.tex
vipassana vs. blockchain
lucid dreams; service

@@
 
+The purpose of various forms of meditation should be understood clearly, as the
+intention of techniques can vary wildly from one to the next. Where the purpose of
+meditation is simply to relax, it is difficult to say that meditation is more
+valuable than a nap. Especially for our chronically sleep-deprived society, it's
+unlikely their effects would differ much.
+
+However, most meditation techniques are not intended to relax the meditator. Whatever
+the object of a meditation technique (one of the five sense doors, bodily sensation,
+or objects of mind) the very act of meditation is one of
+
 \todo{look through leather notebook}

### 2020-05-08 paper-three/paper-three.tex
p3: gratitude

@@
   \item emotion (i.e. anger)
+  \item gratitute (gratitude vs. fear; julie) - https://youtu.be/WPPPFqsECz0
   \item dealing with death
@@
   \item clarity: in thought, work, planning
+  \item federated, p2p, distributed systems like blockchain / holochain attempt to circumvent
+    structures of oppression and control -- what if corporations could be rebuilt from inside,
+    one employee/stakeholder at a time such that ethics were inherent to the company's very
+    existence?
   \item Die Standing Up

### 2020-05-14 paper-three/paper-three.tex
Neuron speed via Carmack

--- 2019-12-03 paper-three/paper-three.tex
@@
   \item Time: Nat Friedman's blog post?
+  \item Is it really possible? https://twitter.com/ID_AA_Carmack/status/1260698271422570497
+    The “speed of thought” (neuron conduction speed) is only 110m/s, which isn’t that fast. 9 ms of latency just in your arm!
+    https://www.youtube.com/watch?v=MeHoU-arXgY
   \item bible; ``a perfect set of rules''; nature itself <=> perception; ``perfect

### 2023-05-05 paper-one/paper-one.tex
Fixing epub builds
- fixing epub title and series in both epub scripts
- fixed references section order in paper-two.tex (`section` within bibilography was causing errors)
- fixed svg resize errors by putting SVGs in their own directory... I have no idea why `epub-convert` was looking at the SVGs since the .tex file doesn't even reference them, but whatever.
- added a refrences section title to paper one
- adding graphicsmagick to readme (may not be required but it can't hurt)
Complete Linux build - remove polyglossia, use fontspec directly
* fonts are included inline now
* debs listed in READ

@@
 
-\usepackage{polyglossia}
-\setdefaultlanguage{english}
-\setotherlanguage{sanskrit}
 \usepackage{fontspec}
-\newfontfamily\devanagarifont[Scale=MatchUppercase]{Kohinoor Devanagari}
+\newfontfamily\devanagarifontsans{NotoSansDevanagari-Regular.ttf}[
+  BoldFont = NotoSansDevanagari-Bold.ttf ]
+\newfontfamily\devanagarifontserif{NotoSerifDevanagari-Regular.ttf}[
+  BoldFont = NotoSerifDevanagari-Bold.ttf ]
 
@@
 The second, more common use of the term ``Sankhara'' refers to ``conditioned phenomena'' or ``psychological dispositions''.
-The equivalent Hindi/Sanskrit word, \foreignlanguage{sanskrit}{संस्कार} (sanskaar) roughly translates in this context to ``mental impression''.
-This second definition will apply to this paper.
+The equivalent Hindi/Sanskrit word, {\devanagarifontserif संस्कार} (sanskaar) roughly translates in this context to ``mental impression''. This second definition will apply to this paper.
 
@@
 
-\begin{thebibliography}{9}
+\section*{References}
+
+\begin{thebibliography}{99}
 \raggedright

### 2023-05-05 paper-one/build-pdf.sh

@@ -14,7 +14,9 @@ PATH=$PATH:$LIBPATH
 
 OUT=../output
 
-echo "Running pre-build. This takes a while..."
+echo "Running pre-build. This takes a while."
+echo "However, if it seems to hang, remove pipe to /dev/null"
+echo "..."
 xelatex -output-directory=$OUT paper-one.tex > /dev/null 2>&1
 echo "...done. Now running the build."
 xelatex -output-directory=$OUT paper-one.tex

### 2023-06-01 paper-three/paper-three.tex
FIX: get paper three rendering again (Linux)

--- 2020-05-08 paper-three/paper-three.tex
@@
-\documentclass[a4paper, amsfonts, amssymb, amsmath, reprint, showkeys, nofootinbib, twoside]{revtex4-1}
+\documentclass[a4paper, amsfonts, amssymb, amsmath, reprint, showkeys, nofootinbib, twoside]{revtex4-2}
 
@@
 
-\title{%
+\title{
   \large{Vipassana for Hackers} \\
@@
 principle as fact. Rather, a 10-day Vipassana course is a sort of laboratory where the
-principle can be tested and experienced for oneself.
+principle can be tested and experienced directly.
 
@@
   \item Time: Nat Friedman's blog post?
+  \item Is it really possible? https://twitter.com/ID\_AA\_Carmack/status/1260698271422570497
+    The “speed of thought” (neuron conduction speed) is only 110m/s, which isn’t that fast. 9 ms of latency just in your arm!
+    https://www.youtube.com/watch?v=MeHoU-arXgY
   \item bible; ``a perfect set of rules''; nature itself <=> perception; ``perfect

### 2023-06-01 letter-one/letter-one.tex
Adding Letter One (On Gurudom)

\documentclass{letter}
\usepackage{hyperref}
\usepackage{xcolor}
\signature{Steven Deobald}
\address{Lot 1 Beaver Harbour Rd. \\ Beaver Harbour \\ Canada B3H 4M1}

\NewDocumentCommand{\codeword}{v}{%
\texttt{\textcolor{blue}{#1}}%
}

\begin{document}
\begin{letter}{Adrianna Tan}
\opening{Dear Adrianna,}

Enclosed is a copy of Charlotte Joko Beck's ``Nothing Special''.
I do not endorse her teaching or her particular brand of Buddhism (nor any other) but, immediately following my second 10-day meditation course, I found her writing to be a great help in untangling the confusion which surfaced from my experience.
Perhaps you will also find it similarly useful.

In the past, I've written essays in lieu of letters (or emails) to individuals.
This time, I thought I might write and publish the letter instead.
Let me know what you think about that idea.

I write to you to speak, ostensibly, about gurus.
Starting with an essay at first, I penned a title of ``On Gurudom'' which caused the immediate \codeword{rm} of that file.
Such an essay telegraphs its conclusion loudly and obnoxiously: ``Gurus Bad!''
That's not my intention.
There's also a lot of periphery to the teachings and their respective teachers that don't have anything to do with gurudom, per se.

\end{letter}
\end{document}

### 2023-06-01 letter-one/build-pdf.sh

@@ -1,0 +1,23 @@
+#!/bin/sh
+
+# pdflatex was used originally but cannot render Sanskrit without a lot of hoop-jumping
+# pdflatex vipassana-for-hackers.tex
+
+# xelatex is apparently less capable than pdflatex at font-rendering in general, but
+# is capable of rendering non-Latin fonts (like Hindi/Sanskrit) by default
+
+# Compile twice to resolve references.
+
+SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
+LIBPATH="$SCRIPTPATH/../lib"
+PATH=$PATH:$LIBPATH
+
+OUT=../output
+
+echo "Running pre-build. This takes a while..."
+xelatex -output-directory=$OUT letter-one.tex # > /dev/null 2>&1
+echo "...done. Now running the build."
+xelatex -output-directory=$OUT letter-one.tex
+
+rm -f texput.log $OUT/*.aux $OUT/*.log $OUT/*.out $OUT/*.tdo $OUT/*.bib
+

### 2023-06-01 letter-one/letter-one.tex
On Gurudom

@@
 }
+\newcommand\hr{\par\vspace{-.5\ht\strutbox}\noindent\hrulefill\par}
 
@@
 Perhaps you will also find it similarly useful.
+It has little or nothing to do with the following letter.
 
@@
 
+It's impossible for me to write something for every economy, cultural context, linguistic background, and ambient religion.
+Instead, I'll try my best to relate my personal experiences in a way that might be appreciated globally -- as a story, if nothing else.
+
+\hr
+
+There were no gurus in my childhood.
+By the time I was old enough to appreciate what religion and spirituality meant to the adults in my community,
+
 \end{letter}

### 2023-06-01 letter-one/letter-one.tex
Done letter?

@@
 There were no gurus in my childhood.
-By the time I was old enough to appreciate what religion and spirituality meant to the adults in my community,
+By the time I was old enough to appreciate what religion and spirituality meant to the adults in my community, the various sects of Canadian Christianity (the only spiritual practice of note in a town with a population of 1000) had already begun to collapse.
+From most churches, gone were the severe priests and pastors of an immigrant generation.
+The death rattle of podunk churches sounded like soft rock on Friday nights amongst a congregation who now referred to their minister by her first name.
+At that age, I never thought of church leaders are \textit{teachers}, even to those they preached to.
+Certainly not to me.
+
+When I moved to India for the first time in 2006, I was a drunken wreck and no one in my life would mistake the year I spent in Pune for a trip of self-discovery.
+But when I moved to Bangalore in 2011, the question was surprisingly common: ``are you going to find a guru? to start a practice? to go on a spiritual quest?'' ... ``did you find yourself there?''
+The truth is, Bangalore was just another escape.
+I hated my job in high finance and, what the heck, maybe I could teach some kids Clojure for a year before finding another real job.
+
+I did take two meditation courses in India, but they were Burmese (Vipassana) and Japanese (Zazen), and bore little in common with meditation schools one would normally associate with India.
+At the time, my Indian friends found the long-bearded God Men of India as offensive as I might find the minister of a new age megachurch in the USA.
+I found them cartoonish.
+
+Leaving aside the post-Osho beards and robes (which many lower-and-middle-class Indians don't identify as red flags the way you and I do), I found it always only took a few seconds to search for ``XYZ net worth'' or ``XYZ controversy'' and justify my hesitations.
+These were clearly businessmen.
+Most didn't bother hiding it.
+
+Over the years, I began to navigate and solidify my own meditation practice.
+As I did this, I encountered a number of different schools, practices, ideas, philosophies, teachers, and students.
+Curiously, what I found was that the subtler schools and teachers were often the most dangerous.
+
+\hr
+
+The primary difficulty in finding a teacher, as you had mentioned in our conversation, is the power dynamic.
+This is unavoidable.
+And it is for this reason that I generally discourage my friends from seeking a teacher at all.
+Secondarily, not only is the power dynamic a problem, but any significant meditation practice will also severely alter the practitioner's mental state from time to time.
+This exacerbates the problem of power dynamics between teacher and student: it is when the student is at her most vulnerable that she needs the teacher's help the most.
+
+Paradoxically, I also discourage people from self-study.
+All but the lightest (MBSR and other common forms of mindfulness, for example) meditation practices are bound to place the practitioner into deeply troubled psychic states sooner or later.
+Alone in our living rooms, listening to pre-recorded instructions in an app without guidance or a safety net, is not the time to dig up our most penetrating unconscious psychological traumas.
+
+Instead, I have found that seeking out \textit{systems}, rather than gurus or my hubris-fuelled insistence on self-reliance, to be the most effective path.
+Along the way, I have encountered teachers, of course -- but these teachers tended to be litmus tests for selecting a system, rather than the other way around.
+
+\hr
+
+With that in mind, I'll return to the idea of subtle dangers for a moment.
+Outside of the more obvious warning signs (``this lady is teaching X to make money'' or ``this dude is in jail for rape; maybe his school isn't for me''), I have found gross-but-less-obvious warning signs over the years.
+
+The first is touching.
+It seems very strange to people accustomed to a lot of physical contact that even so much as a handshake should be forbidden between teacher and student (or between the students themselves).
+For some, a rule like ``no touching'' conjures up images of George Bluth Senior shouting in the jail in the first season of \textit{Arrested Development}.
+Still, the fact that ``no touching'' presents itself as a binary rule makes it very easy to adhere to, without wading into an extreme, such as forced chastity, which is known to have the opposite effect.
+There are any number of problematic events, such as the stripping of W.S. Merwin and Dana Naone by Chögyam Trungpa, which are obviated by such a rule.
+Though the Halloween party of 1975 is something we'd consider unforgiveable today, it was perhaps forgiven by all involved back in the 70s.
+There are new grey areas of abuse in this decade, but ``no touching'' (if it is adhered to), at least eliminates physical abuse.
+
+The second is alcohol.
+Abuses of power, both physical and psychological, are given unnecessary fuel by alcohol.
+For a long time, I was drawn to Zen.
+However, even with all my other hesitations set aside about Zen practice, I could never get over the use of alcohol -- although I have been sober for 8 years I still acutely remember the damage it does to the human decision-making process.
+Where there is alcohol, there is bound to be abuse of power.
+
+The last is money.
+There are a number of East-Meets-West schools and practices which intentionally adopt a commercial position.
+Much like cartoonish God Men, meditation schools and apps which make no bones about their for-profit nature often aren't as dangerous as those which sit somewhere in the middle.
+It is unlikely that anyone will learn meditation on the floor of an American yoga studio, much less harm themselves psychically while doing so.
+The danger comes when a school presents itself as a spiritual outlet -- a temple, a monastery, or a meditation centre.
+In these locations, if participation is transactional, the lines become blurred and there is a risk that students will be exploited.
+Many meditation schools teach for free, on a donation basis.
+If you choose to explore schools which require payment for their teaching, I strongly suggest extra attention while exploring them.
+I have personally attended meditation courses which cost money; often they were useful and informative, but ultimately they were not for me.
+
+The subtle dangers are hinted at by the danger of money.
+From the God Men of India to the spandex spirituality of the USA to stoic philosophies of Europe to African Cosmology to the Buddhism of Southeast Asia... there's a lot out there.
+For better or worse, it's all getting a bit blended up now.
+The subtlety comes in situations where a path, school, or teacher is not just \textit{not obviously dangerous} but perhaps seems particularly safe or inviting.
+As we become increasingly invested in any particular path, the dangers present on that path (and they're on every path, obviously) are amplified ...
+we always find it harder to divest ourselves of something we've sacrificed for.
+
+Although it would be presumptuous of me to assume you're specifically and narrowly interested in the teachings of the Buddha, I do think the foundational advice he provides is a safe starting position (for me, at least).
+To paraphrase, in one sutra he suggests to students: ``Inspect my teaching like a golden trinket you might buy in the market. Try to find flaws in what I am teaching you.''
+In others, he insists on relying on ``yathā-bhūta'' or ``seeing things as they are, with direct experience, devoid of any concepts''.
+As soon as I have found myself relying on someone -- or some belief -- I have experienced the aforementioned sacrifice, which is a dangerous commitment even in small doses.
+
+\hr
+
+And so I wish you the best of luck in your pursuit of a system (and, ultimately, a teacher) which you find satisfactory.
+I hope you try many systems and practices.
+Begin with a breadth-first search, if you can.
+Don't commit yourself to an avenue until you are certain.
+There are a lot of blind alleys.
+Tread carefully.
+
+\closing{Sincerely,}
 

### 2023-06-02 letter-one/letter-one.tex
Some revisions to Gurudom letter

@@
 \signature{Steven Deobald}
-\address{Lot 1 Beaver Harbour Rd. \\ Beaver Harbour \\ Canada B3H 4M1}
+\address{Lot 1 Beaver Harbour Rd. \\ Beaver Harbour, NS \\ Canada B3H 4M1}
 
@@
 
-Enclosed is a copy of Charlotte Joko Beck's ``Nothing Special''.
+Enclosed is a copy of Charlotte Joko Beck's \textit{Nothing Special}.
 I do not endorse her teaching or her particular brand of Buddhism (nor any other) but, immediately following my second 10-day meditation course, I found her writing to be a great help in untangling the confusion which surfaced from my experience.
@@
 In the past, I've written essays in lieu of letters (or emails) to individuals.
-This time, I thought I might write and publish the letter instead.
+This time, I thought I might do the opposite and write (then publish) the letter instead.
 Let me know what you think about that idea.
@@
 It's impossible for me to write something for every economy, cultural context, linguistic background, and ambient religion.
-Instead, I'll try my best to relate my personal experiences in a way that might be appreciated globally -- as a story, if nothing else.
+Instead, I'll try my best to relate my personal experiences in a way that might be appreciated betwixt our two homelands -- as a story, if nothing else.
 
@@
 The death rattle of podunk churches sounded like soft rock on Friday nights amongst a congregation who now referred to their minister by her first name.
-At that age, I never thought of church leaders are \textit{teachers}, even to those they preached to.
+At that age, I never thought of church leaders as \textit{teachers}, even to those they preached to.
 Certainly not to me.
@@
 I hated my job in high finance and, what the heck, maybe I could teach some kids Clojure for a year before finding another real job.
+Had I known in advance that I'd spend nearly a decade in India, I probably never would have gone.
 
@@
 
-Leaving aside the post-Osho beards and robes (which many lower-and-middle-class Indians don't identify as red flags the way you and I do), I found it always only took a few seconds to search for ``XYZ net worth'' or ``XYZ controversy'' and justify my hesitations.
+Leaving aside the post-Osho beards and robes (which many lower-and-middle-class Indians don't identify as red flags the way you and I might), I found it only took a few seconds to search for ``XYZ net worth'' or ``XYZ controversy'' and justify my hesitations.
 These were clearly businessmen.
@@
 Secondarily, not only is the power dynamic a problem, but any significant meditation practice will also severely alter the practitioner's mental state from time to time.
-This exacerbates the problem of power dynamics between teacher and student: it is when the student is at her most vulnerable that she needs the teacher's help the most.
+This exacerbates the problem of power dynamics between teacher and student: it is when the student is vulnerable that she needs the teacher's help the most.
 
-Paradoxically, I also discourage people from self-study.
+Without contradiction, I also discourage people from self-study.
 All but the lightest (MBSR and other common forms of mindfulness, for example) meditation practices are bound to place the practitioner into deeply troubled psychic states sooner or later.
 Alone in our living rooms, listening to pre-recorded instructions in an app without guidance or a safety net, is not the time to dig up our most penetrating unconscious psychological traumas.
+I've heard too many horror stories of dabbling students painting themselves into a corner to believe that exclusive self-study is a safe approach for anyone.
 
@@
 Along the way, I have encountered teachers, of course -- but these teachers tended to be litmus tests for selecting a system, rather than the other way around.
+A teacher is a representative of her chosen practice.
 
@@
 With that in mind, I'll return to the idea of subtle dangers for a moment.
-Outside of the more obvious warning signs (``this lady is teaching X to make money'' or ``this dude is in jail for rape; maybe his school isn't for me''), I have found gross-but-less-obvious warning signs over the years.
+Outside of the more obvious warning signs (``this lady is teaching X to make money'' or ``this dude is in jail for rape -- maybe his school isn't for me''), I have found gross-but-less-obvious warning signs over the years.
 
 The first is touching.
-It seems very strange to people accustomed to a lot of physical contact that even so much as a handshake should be forbidden between teacher and student (or between the students themselves).
+It seems very strange to people accustomed to a lot of physical contact that so much as a handshake should be forbidden between teacher and student (or between the students themselves, in a practice environment).
 For some, a rule like ``no touching'' conjures up images of George Bluth Senior shouting in the jail in the first season of \textit{Arrested Development}.
@@
 There are any number of problematic events, such as the stripping of W.S. Merwin and Dana Naone by Chögyam Trungpa, which are obviated by such a rule.
-Though the Halloween party of 1975 is something we'd consider unforgiveable today, it was perhaps forgiven by all involved back in the 70s.
-There are new grey areas of abuse in this decade, but ``no touching'' (if it is adhered to), at least eliminates physical abuse.
+Though the Halloween party of 1975 is something we'd consider unforgiveable today, it may have been forgiven by all involved back in the 70s.
+There are new grey areas of abuse in this decade, but ``no touching'' (if it is adhered to), at least eliminates direct physical abuse.
+I attended a week-long meditation course once which encouraged the students to ``greet each other'' (with hugs) at the point of course completion.
+It was innocent, but I still found that much physical contact in a meditation environment a rather shocking experience.
 
@@
 However, even with all my other hesitations set aside about Zen practice, I could never get over the use of alcohol -- although I have been sober for 8 years I still acutely remember the damage it does to the human decision-making process.
-Where there is alcohol, there is bound to be abuse of power.
+Where there is alcohol, there is bound to be abuse of power, eventually.
 
@@
 There are a number of East-Meets-West schools and practices which intentionally adopt a commercial position.
-Much like cartoonish God Men, meditation schools and apps which make no bones about their for-profit nature often aren't as dangerous as those which sit somewhere in the middle.
+Much like cartoonish God Men, meditation schools and apps which make no bones about their for-profit nature don't seem as dangerous to me as those which sit somewhere in the middle.
 It is unlikely that anyone will learn meditation on the floor of an American yoga studio, much less harm themselves psychically while doing so.
@@
 Many meditation schools teach for free, on a donation basis.
-If you choose to explore schools which require payment for their teaching, I strongly suggest extra attention while exploring them.
-I have personally attended meditation courses which cost money; often they were useful and informative, but ultimately they were not for me.
+Although I have personally attended meditation courses which cost money, and although they were often useful and informative, ultimately none of them were for me.
 
 The subtle dangers are hinted at by the danger of money.
-From the God Men of India to the spandex spirituality of the USA to stoic philosophies of Europe to African Cosmology to the Buddhism of Southeast Asia... there's a lot out there.
+From the God Men of India to the spandex spirituality of the USA to stoic philosophies of Europe to African Cosmology to Buddhism in Southeast Asia... there's a lot out there.
 For better or worse, it's all getting a bit blended up now.
@@
 As soon as I have found myself relying on someone -- or some belief -- I have experienced the aforementioned sacrifice, which is a dangerous commitment even in small doses.
+On the occasions when I place my trust in myself, my intuition, and my skepticism, I have managed to avoid these dangers.
 
@@
 
-And so I wish you the best of luck in your pursuit of a system (and, ultimately, a teacher) which you find satisfactory.
+Regarding intuition, and the teachers I have studied under, the litmus test I spoke of earlier has always materialized only by chance.
+The conditions for such a test cannot be created at will; we have to be patient.
+
+In one such example, a minor irritant (a malfunctioning microphone) managed to push the teacher to a point of visible frustration with her co-teacher and her partner (who was volunteering).
+It was surprising to me how quickly my opinion of that teacher, and the practice she taught, was shattered by observing her reaction to such an insignificant stimulus.
+
+In another pair of examples, one teacher was verbally assaulted by a student during a Q\&A period and another had to deal with a massive snow storm and power outage.
+In the first instance, the student accused the teacher of being a fraud, essentially, and then demanded the teacher (who was not Japanese) translate some Japanese words for him.
+The teacher was clearly shocked by the interaction... but didn't respond with irritation or frustration.
+Instead, he smiled and responded to the student patiently.
+During the snow storm, the teachers not only maintained their composure under difficult circumstances, but continued to support the course volunteers and coordinate safety measures, day after day, for students.
+
+No single event (or response to an event) could convince me of the benefits a particular practice has conveyed (or not) onto a teacher.
+Still, every observation you make about your teachers will paint pictures of them for you -- and I think our own intuition can tell us a lot about the people we interact with in this world.
+
+\hr
+
+And so I wish you the best of luck in your pursuit of a system and teacher.
 I hope you try many systems and practices.
-Begin with a breadth-first search, if you can.
-Don't commit yourself to an avenue until you are certain.
-There are a lot of blind alleys.
-Tread carefully.
+I hope you begin with a breadth-first search, if you can.
+I hope you can avoid commitment until you are certain.
+
+When it comes to meditation, there are many blind alleys.
+Still, there is no other aspect of my life as precious as meditation.
+It is the greatest thing I know.
+The search is all made worth it when you find a practice that really changes you.
 

### 2024-06-26 the-proposal/the-proposal.tex
long proposal: citations for intro

@@
 \begin{abstract}
-Vipassana meditation (as taught by S.N. Goenka in the tradition of Sayagyi U Ba Khin)
-is unlike other meditation techniques insofar as its claims to completeness and
-outcomes are concerned. Vipassana claims to ultimately explore the entire field of
-mind and matter, with the goals of total liberation and full enlightenment. Implicit
+Vipassana meditation claims to ultimately explore the entire field of
+mind and matter, with the goal of total liberation from suffering. Implicit
 within these claims is a complete understanding of human consciousness. These are
-difficult claims to prove because the time commitment required to research the
-technique orders on multiple decades --- if not multiple generations. This is complicated by the
+difficult claims to prove or disprove because the time commitment required to research the
+technique orders on multiple decades --- if not multiple generations.
+This is complicated by the
 fact that the time commitment is demanded of both the researcher \textit{and the
 subject}. Until now, research on meditation of all kinds has covered only one of
-two fields: (1) controlled experiments which must inherently rely on superficial data
-gathered from beginners and (2) observational exploratory research of monks --- expert
-meditators who have dedicated their lives to the practice. I propose bridging this
+two fields:
+(1) controlled experiments which must inherently rely on superficial data
+gathered from beginners \cite{https://doi.org/10.1002/jclp.20761,
+  https://doi.org/10.1037/a0018441,
+  https://doi.org/10.1093/scan/nss054,
+  https://doi.org/10.31231/osf.io/hzv65}
+and (2) observational exploratory research of monks --- expert
+meditators who have dedicated their lives to the practice.
+\cite{https://doi.org/10.1073/pnas.0407401101,
+https://doi.org/10.1371/journal.pone.0073417}
+
+I propose bridging this
 gap by submitting myself to the middle ground. While remaining a layperson, I will
-commit to a high ratio of waking medition hours for the rest of my life. Individually
+commit to a high ratio of waking medition hours. Individually
 and internally, I will conduct qualitative research into the consequences of
@@
 studies with a broader community of researchers.
-
-\todo{add references to (1) and (2)}
 

### 2024-06-26 the-proposal/the-proposal.tex
review/update abstract

@@
 mind and matter, with the goal of total liberation from suffering. Implicit
-within these claims is a complete understanding of human consciousness. These are
-difficult claims to prove or disprove because the time commitment required to research the
-technique orders on multiple decades --- if not multiple generations.
-This is complicated by the
-fact that the time commitment is demanded of both the researcher \textit{and the
-subject}. Until now, research on meditation of all kinds has covered only one of
-two fields:
+within this claim is a complete understanding of human consciousness.
+This is a difficult claim to prove or disprove because the time commitment
+required to research the technique orders on multiple decades
+--- if not multiple generations.
+Only a longitudinal study has any meaning.
+This is complicated by the fact that the time commitment is demanded of
+both the researcher \textit{and the subject}.
+Until now, research on meditation of all kinds has covered only one of two fields:
 (1) controlled experiments which must inherently rely on superficial data
@@
 gap by submitting myself to the middle ground. While remaining a layperson, I will
-commit to a high ratio of waking medition hours. Individually
-and internally, I will conduct qualitative research into the consequences of
-Vipassana meditation and the nature of consciousness while collectively and
-externally pledging myself as a subject for long-term quantitative
+commit to a high ratio of waking medition hours.
+Individually and internally, I will conduct qualitative research into the consequences of
+Vipassana meditation and the nature of consciousness.
+Collectively and externally I will pledge myself as a subject for long-term quantitative
 studies with a broader community of researchers.
@@
   sampajāno satimā} (continuous piercing awareness of constantly
-changing bodily sensation \cite{goenka1999discourses}) is not only a
+changing bodily sensation \cite{goenka1999discourses, analayo2003satipatthana})
+is not only a
 requirement of serious Vipassana practice but could very well act as a surrogate

### 2024-06-26 the-proposal/the-proposal.tex
reviewing previous methodology written

@@
 
+There is a dearth of high-quality scientific meditation research.
+Due to decades of misunderstanding the subject, it's estimated that
+--- of the 50,000+ papers published on the broad topic of ``meditation'' ---
+only a few dozen conform to scientific rigour.
+\cite{goleman2017altered,semanticscholar2024meditation}
+
+There are exceptions. Notably, Richard Davidson's 2002 study on Minyur Rinpoche and
+the subsequent EEG and fMRI studies of serious Tibetan monks.
+\cite{https://doi.org/10.1073/pnas.0407401101}
+There have also been effective double-blind trials of householders (the laity),
+predominantly studying Mindfulness-Based Stress Reduction (MBSR) and
+Mindfulness-based Cognitive Therapy (MBCT), which represent two very mild
+techniques of meditation.
+
+Sampling from the extremes is a natural consequence of the accessibility of subjects.
+Contemplative monks, serious in their meditation practice, are not easily accessed.
+Modern householders tend to peak at 2 or 3 hours of meditation per day,
+due to other responsibilities.
+This is the dichotomic choice in meditation research:
+either sporadically studies monks, who can have tremendous meditative capacity
+--- or regularly studies householders, who have little to none.
+
+Bridging this monastic / householder divide requires individuals willing to
+(a) dedicate themselves to long periods (years or decades) of continuous meditation
+practice with a single technique and
+(b) submit themselves to continuous scientific inquiry.
+
 \subsection{Overview}
@@
 approximately 20 years of
-age, and is now of an age when they may participate in an extended study (say, 50-60
-years of age), we arrive at roughly 60,000 hours of practice. If we use the oft-cited
+age, and is now of an age when they may participate in an extended study (say, 40-60
+years of age), we arrive at roughly 60,000 hours of practice.
+We can reduce this number to 50,000 to match the upper-bound of groups
+categorized in studies by Lumma and Brewer.
+\cite{https://doi.org/10.1016/j.ijpsycho.2015.04.017,https://doi.org/10.1073/pnas.1112029108}
+If we use the oft-cited
 ``10,000 hours'' measurement for competence in a subject, we might reasonably
@@
 	\item Beginner: 0-10,000 hours
-	\item Intermediate: 10,000-60,000 hours
-	\item Expert: 60,000+ hours
+	\item Intermediate: 10,000-50,000 hours
+	\item Expert: 50,000+ hours
 \end{itemize}
@@
 methodological issues with studies of Vipassana, as described by Alberto Chiesa in
-\textit{Vipassana meditation: systematic review of current evidence},
+\textit{Vipassana Meditation: Systematic Review of Current Evidence},
 2010. \cite{chiesa2010vipassana} These include the lack of: study replication,
@@
 Optically-pumped magnetometer (OPM)-based magnetoencephalography (MEG) performed using a
-mobile helmet capable of operating at room temperature \cite{boto2017new}, in development since Chiesa's
+mobile helmet capable of operating at room temperature \cite{boto2017new},
+in development since Chiesa's
 paper was published, may permit it as a third neuro-imaging technique. Even older MEG
@@
 technique, during the actual practice of Vipassana. I have previously discussed this
-apparent paradox in Appendix A of \textit{Vipassana for Hackers, Paper One: Curious Mechanics}.
+apparent paradox in Appendix A of
+\textit{Vipassana for Hackers, Paper One: Curious Mechanics}.
 \cite{deobald2017curious}
@@
 possible. Meditators who are in the process of transitioning from 10,000 hours of
-experience to 60,000 hours of experience are not readily available as research
+experience to 50,000 hours of experience are not readily available as research
 subjects in significant numbers. Because we have not studied such meditators, we are
@@
 The line between ``mere object for brain studies'' and ``research tool'' begins to
-blur when we consider using meditation for both simultaneously. However, it seems remiss to neglect
-gathering whatever objecting EEG/fMRI/MEG data from the long-term subject-researcher
+blur when we consider using meditation for both simultaneously. However, we would be
+remiss not to
+gather whatever objective EEG/fMRI/MEG data from the long-term subject-researcher
 when that data is so readily available.
@@
 
-Because Vipassana meditators are inherently self-selecting, double blinding is not
+Because Vipassana meditators are inherently self-selecting, double blind trials are not
 possible, nor are randomized trials. Single blinding is possible, and should be
@@
 
-The study proposed is an ultralong (lifelong) study of an individual --- myself. As
-far as quantitative data gathered from brain scans is concerned, I am effectively
+The study proposed is an ultralong (lifelong) study of an individual --- myself.
+There is potential for others to join this ultralong study, should it show promise.
+As far as quantitative data gathered from brain scans is concerned, I am effectively
 making myself available as a human guinea pig. As far as qualitative data is
 concerned, research will take the form of a deep case study. As I will represent
-Ramachandran's ``initial sample'', statistical analysis will not be relevant. If the
-study proves productive, future studies targeting a specific hypothesis about
+Ramachandran's ``initial sample'', statistical analysis will not be relevant.
+If the study proves productive, future studies targeting a specific hypothesis about
 Vipassana in a large sample can use learnings from this study in their statistical
@@
 (myself) will explore the question of consciousness and the mind-body problem
-directly, through Vipassana meditation, for a prolonged period of time (the remainder
-of my life). Many theories and questions currently exist regarding
+directly, through Vipassana meditation, for a prolonged period of time.
+Many theories and questions currently exist regarding
 consciousness, such as \textit{Theory of Mind}, \textit{Theory of Panpsychism},
@@
 
-Vipassana's instruction provides us with claims we can evaluate: ``[Vipassana] explores the entire field of
-mind and matter'' \cite{goenka1999discourses}, ``[Vipassana] is a technique that will
-eradicate suffering'' \cite{dhammaorg2014code}, and ``Vipassana aims at ... total
-liberation and full enlightenment'' \cite{dhammaorg2014code}. Progress made in
-evaluation of these claims will be, by its very nature, emergent.
+Vipassana's instruction provides us with claims we can evaluate:
+``[Vipassana] explores the entire field of mind and matter'' \cite{goenka1999discourses},
+``[Vipassana] is a technique that will eradicate suffering'' \cite{dhammaorg2014code},
+and ``Vipassana aims at ... total liberation and full enlightenment''
+\cite{dhammaorg2014code}.
+Progress made in evaluation of these claims will be, by its very nature, emergent.
 
@@
 researcher bias, peer debriefing (again from more experienced meditators), and
-searching for discrepant evidence are all possible. Respondent validation (member
+searching for discrepancies in evidence are all possible. Respondent validation (member
 checking), although possible, may not carry much weight in terms of strengthening the
@@
 progresses. Monthly or yearly schedules will be decided with a
-supervisor but an example daily schedule (subject to change before the study begins)
-is available in \textit{Appendix 1: Proposed Daily Schedule}.
+study supervisor but an example daily schedule
+(subject to change) is available in \textit{Appendix 1: Example Daily Schedule}.
 
@@
 \section{APPENDICES}
-\subsection{Appendix 1: Proposed Daily Schedule}
+
+\subsection{Appendix 1: Example Daily Schedule}
+
+EXAMPLE: \\
+05:00 - Wake \\
+05:30 - 1 hour meditation \\
+06:30 - Breakfast \\
+07:30 - Exercise \\
+09:00 - 3 hours meditation \\
+12:00 - Lunch \\
+13:00 - 3 hours meditation \\
+16:00 - Writing period \\
+18:00 - Evening meal \\
+19:00 - 1 hour meditation \\
+21:00 - Sleep
+
+It is worth noting that this schedule generally persists across weekends and contains
+flexibility for longer or shorter writing periods.
+It is also worth noting that this is a sample schedule for the beginning of the study.
+As the study progresses, it is expected that the duration of sleep will lessen.
 

### 2024-06-26 the-proposal/the-proposal.tex
include up to "specific aims"

@@
 
+\subsection{Overview}
+
 There is a dearth of high-quality scientific meditation research.
@@
 
-\subsection{Overview}
+\subsection{Hypothesis}
 
-\subsection{Hypothesis}
+As far as quantitative measures, such as future EEG, MEG, and fMRI readings, are concerned,
+this study hypothesizes that readings taken from subjects will progress toward readings
+found in ``expert'' groups of previous studies.
+\cite{https://doi.org/10.1073/pnas.0407401101,https://doi.org/10.1371/journal.pone.0073417}
+See \textit{RESEARCH DESIGN AND METHODS, Expertise}.
+
+However, both quantitative and qualitative measures will be used primarily for
+initial discovery and exploration.
+Since no other study of this kind has ever been run, it is not possible to meaningfully
+predict outputs.
+See \textit{RESEARCH DESIGN AND METHODS, Methodological Issues}.
 
@@
 \subsection{Overall Objective}
+
+As an ``ultralong'' longitudinal study of continuous practice,
+the primary goal is open discovery.
+
+Most pre-existing studies on meditation focus on narrow outputs of the
+practitioners' meditation activity.
+Qualitiative examples include effects on emotional regulation, negative self-belief,
+depression, anxiety, and amygdala response outside of meditation.
+\cite{https://doi.org/10.1093/scan/nss054,
+  https://doi.org/10.1037/a0018441,
+  https://doi.org/10.31231/osf.io/hzv65}
+Quantiative examples include discovery of high-amplitude gamma synchrony during meditation,
+higher Parietal-Occipital EEG gamma activity during NREM sleep,
+Heart Rate (HR) and High-Frequency Heart Rate Variability (HF-HRV) as useful proxies for effort,
+and meditation experience correlated to reduced Default Mode Network (DMN) activity.
+\cite{https://doi.org/10.1073/pnas.0407401101,
+  https://doi.org/10.1371/journal.pone.0073417,
+  https://doi.org/10.1016/j.ijpsycho.2015.04.017,
+  https://doi.org/10.1073/pnas.1112029108}
+
+This ultralong study will, instead, examine all aspects of prolonged practice of
+Vipassana meditation.
+Qualitative outputs will begin after the first quarter of the study.
+Quantitative outputs will emerge as long-term access to research equipment becomes available.
+
 \subsection{Specific Aims}
+
+The study will produce qualitative writing describing the research-subject's experience.
+
+The study will produce quantitative measures, based on equipment availability.
+
+The study will also act as a template study methodology for others to follow,
+or for other researcher-subjects to join if the study proves successful in
+uncovering new discoveries.
+
+There has been limited, unprofessional writing on the topic of the dangers of meditation.
+\cite{ingram2020mastering,https://doi.org/10.1080/03060497.1983.11084578}
+Even though these writings focus on Christian ideals and meditative concepts
+(such as ``The Dark Night of The Soul''),
+which have no relationship to Vipassana meditation whatsoever,
+they do arise in discussions of Vipassana meditation from time to time.
+As a result, it is a specific aim of this study to determine if there are
+psychological or physical dangers presented by prolonged meditation practice.
 
@@
 19:00 - 1 hour meditation \\
-21:00 - Sleep
+21:00 - Sleep \\
 

### 2024-06-27 the-proposal/the-proposal.tex
up to design & methods

@@
 \cite{ingram2020mastering,https://doi.org/10.1080/03060497.1983.11084578}
-Even though these writings focus on Christian ideals and meditative concepts
-(such as ``The Dark Night of The Soul''),
-which have no relationship to Vipassana meditation whatsoever,
-they do arise in discussions of Vipassana meditation from time to time.
+Even though these writings focus on Christian ideals, struggles of faith,
+and meditative concepts (such as ``The Dark Night of The Soul''),
+which have no relationship to meditation whatsoever,
+they do arise in discussions of meditation from time to time.
 As a result, it is a specific aim of this study to determine if there are
@@
 \subsection{Preliminary Research Review}
-- summary of related research
-- strengths and weaknesses
-- justification
---- what hasn't been done by others?
---- why is this research necessary?
+
+Related research, as mentioned previously, falls on the spectrum of exploratory
+research conducted on ``expert'' (50,000 hours of practice or more) meditators
+in various traditions or acute, quantitative research conducted on novice
+(0 - 10,000 hours) or intermediate (10,000 - 50,000 hours) meditators.
+
+The strength of each form of study is demonstrated in its results.
+Exploratory research has given us a clear understanding that the brains of
+very experienced meditators function in a manner very different from the
+societal mean.
+Quantitative research with a narrow hypothesis has provided repeated
+statistical evidence that meditation practices do change the minds and bodies
+of the participant, and that those changes are measurable.
+Statistical research has also taught us that the effects of many
+meditation practices become more pronounced with a prolonged practice period.
+
+Longitudinal studies have largely not been conducted in the various fields
+of meditation research.
+Where they have been conducted, ``longitudinal'' may only represent a period
+of nine months.
+Generally, these longitudinal studies focus on larger groups of novice meditators.
+While useful in its own right, this is equivalent to studying hobby pianists
+for long periods of time: none of the subjects will ever master the piano.
+
+True longitudinal study of an individual (or individuals) will allow for a
+new form of exploratory research, which will itself uncover new questions
+we might ask of meditation.
+We can then probe with further statistical research once we know what
+questions we want to ask of the technique or techniques under study.
 
@@
 
-1. ``complete'', standardized, global, multilingual, 100\% free
-2. mundane (sleep) vs. supramundane (total eradication of suffering)
+Vipassana is a ``complete'' system of meditation, based on two premises:
 
-3. who is steven deobald
+1. that every mental and emotional formation in the consciousness of an individual
+corresponds to an equivalent physical formation in the bodily sensations
+of that individual and
+
+2. that the totality of physical, bodily sensations are open to exploration by
+the method of self-directed human awareness
+
+These premises can be experientially validated, and are thus known to be true.
+As with all experience-oriented factual information found in the fields of
+consciousness studies, we must operate based on the assumption that if a
+third party wishes to validate these premises, they can and will do so
+experientially.
+This presents a difficulty, but not an insurmountable one, as we'll see below.
+
+\subsubsection{Standardization and Access}
+
+Vipassana meditation is a globally standardized meditation practice.
+It is taught based on the same instructions across the globe,
+and has been translated into dozens of languages, covering the vast
+majority of global society.
+Those instructions are taught verbally, via audio and video recording,
+and do not require the participant to be literate in the language they speak.
+Introductory courses in Vipassana meditation are all identical, covering
+a 10-day period in which meditators participate in complete silence,
+devoid of external input and confounding variables.
+
+These introductory courses (as with all Vipassana meditation courses)
+are taught completely free of charge, making access universal.
+
+The only limitation to access is the mental and physical fitness of
+individuals applying to take Vipassana meditation courses.
+Course management will, at their discretion, request someone not to
+take a course if they suffer from mental illness, for example.
+
+\subsubsection{Purpose of Practice}
+
+The intended purpose of Vipassana meditation is inherently supramundane:
+``the total eradication of suffering.''
+While this purpose may appear unreasonable or unattainable to skeptics,
+it is not scientific to judge the potential outcomes of such a meditation
+based on our assumptions.
+We must instead test the methodology of Vipassana meditation to see
+how it holds up to these claims, and to what degree.
+
+Conversely, the purpose of many contemporary meditation techniques
+(particularly those studied in existing literature) is inherently mundane:
+to improve sleep, to reduce anxiety, to reduce PTSD, and so on.
+While some of these effects may be visible in the study of Vipassana
+meditation, they are not the end goal of the technique under study.
 
 \subsection{Why now?}
+
+Instruction in Vipassana meditation has been available outside of
+Burma / Myanmar since 1969.
+However, it has taken five decades of proliferation at the hands of
+volunteers to make Vipassana meditation accessible to large portions of
+the globe.
+Even though there are permanent Vipassana Centres (teaching schools) in
+many countries worldwide, there is still extremely limited presence in
+Africa and the Middle East, though there is widespread interest there.
+
+10-Day introductory courses in Vipassana meditation have long waitlists
+in all of these Centres.
+Demand for instruction in Vipassana meditation far outstrips supply.
+
+Vipassana instruction is clearly on a path of growth, at present.
+With a globally-accessible technique of meditation which is increasingly
+discussed within the scope of popular culture, it is an appropriate
+time to study its effects for the benefit of the global population.
+\cite{marriage2024theretreat,muraskin2024perils,sacchet2024advanced}
 
@@
 
+\subsubsection{Why Steven Deobald?}
+
 \subsection{ETHICS AND HUMAN SUBJECTS ISSUES}

### 2024-06-27 the-proposal/the-proposal.tex
add ethics, up to 'timeframes'

@@
 
+Outside the scope of Vipassana meditation, the broader topic of
+``meditation'' has received a great deal of confused analysis in
+recent years.
+Many authors will make the mistake of confusing their own subjective
+analysis of their personal experiences as somehow objective, or
+present it as such.
+\cite{ingram2020mastering,https://doi.org/10.1007/s12671-020-01389-4}
+It is also an appropriate time to clarify the conclusions of such
+confused analysis.
+
 \section{Research Design and Methods}
@@
 
-Chiesa also notes that it would be beneficial to capture data ``both from a clinical
-and from a neuro-imaging point of view'', including functional magnetic resonance
-imaging (fMRI) and electroencephalography (EEG), for both short-term outcomes
+Chiesa also notes that it would be beneficial to capture data
+``both from a clinical and from a neuro-imaging point of view'',
+including functional magnetic resonance imaging (fMRI)
+and electroencephalography (EEG), for both short-term outcomes
 (altered states) and long-term outcomes (altered traits \cite{davidson1977role}) to
@@
 
-\subsubsection{Qualitative}
+\subsubsection{Qualitative Measures}
 
@@
 progresses. Monthly or yearly schedules will be decided with a
-study supervisor but an example daily schedule
-(subject to change) is available in \textit{Appendix 1: Example Daily Schedule}.
+study supervisor but an example daily schedule is available in
+\textit{Appendix 1: Example Daily Schedule}.
 
-\subsubsection{Quantitative}
+\subsubsection{Quantitative Measures}
 
-Measurements performed during the lifelong qualitative study
+Measurements performed during the qualitative study will be used to create a
+dataset upon which later inferences will be based.
+Measurements taken throughout the study will be subject to equipment availability,
+and as such will only be taken based on the future availability of the same
+equipment.
 
-\subsection{POPULATION AND STUDY SAMPLE}
-\subsection{SAMPLE SIZE AND SELECTION OF SAMPLE}
-\subsection{SOURCES OF DATA}
-\subsection{COLLECTION OF DATA}
-\subsection{EXPOSURE ASSESSMENT}
-\subsection{DATA MANAGEMENT}
-\subsection{DATA ANALYSIS STRATEGIES}
+In general, quantitative measurements will be taken by additional researchers,
+as the study progresses.
+This study is, in and of itself, not intended to interpret quantitative measures
+of brain activity or otherwise.
+Interpretation will be left to researchers familiar with such data, but raw
+data will be publicized.
+
+The simplest data available is from off-the-shelf EEG machines, available to
+the general public with open-source hardware and open APIs, such as those
+sold by BlueBCI \footnote{https://bluebci.com/}, making raw data
+easily and permanently accessible to the study.
+More advanced EEG machines will require partnership with a university or
+research group in the future.
+
+ECG data producing HR and HF-HRV data can be collected from off-the-shelf
+systems such as the Zephyr Bioharness 3 \footnote{https://www.zephyranywhere.com/},
+previously validated against physiology monitor systems used in laboratories.
+\cite{johnstone2012bioharness}
+
+fMRI scans are available at commercial
+diagnostic laboratories in India for as little as \$100 USD per scan.
+MEG, as a newer technology, will likely require a university partnership.
+
+\subsection{Population and Study Sample}
+
+Initially, only one researcher-subject will be used in this study.
+
+Additional researcher-subjects will be added to the study in the future
+if such persons are found capable of dedicating themselves to the task.
+
+\subsection{Collection of Data}
+
+Initial data collection will be done in the form of experiential reports.
+
+\subsection{Data Management}
+
+Data will be the intellectual property of Atapi Research Inc.
+
+Data will be stored and backed up in raw formats produced by diagnostic
+equipment under a version controlled data store.
 
 \subsection{Participants}
-- people and roles
+
+Initially, the only participant in this study will by me (Steven Deobald).
+In the future, additional researchers or research-subjects may be added
+to the study.
 
@@
 
-\subsection{ETHICS AND HUMAN SUBJECTS ISSUES}
-\subsection{TIMEFRAMES}
+My willingness to submit myself to this study, for the length of time required,
+is based on a number of factors.
+
+First, I have been exclusively practicing Vipassana meditation for over 10 years.
+I have taken month-long courses, and will commit myself to taking courses of
+one or two months every year for the duration of the study.
+
+Second, I am in the peculiar position of having severe eye damage due to a failed
+surgery in 2014.
+Although I can still use computers (and will need to do so to write experience
+reports), it causes me quite a lot of physical harm to use computers for
+extended periods of time.
+A severe light sensitivity also caused by the failed surgery means that, peculiarly
+enough, my eyes are most comfortable when they are closed in a dark room.
+
+Third, I have been considering this form of research since 2016
+--- nearly 8 years at the time of this writing.
+I have also seriously considered monastic life, but ordaining as a monk is limiting
+in terms of scientific output and there is no monastic order under the Vipassana
+lineage yet.
+
+Last, after 10+ years living in India, I am literate in the Devanagari script and
+familiar with the Pāli language, which all original Buddhist texts are written in.
+While researching meditation directly, it will be important to check findings
+against the original teachings to validate my understanding before publishing it.
+
+\subsection{Ethics and Human Subjects Issues}
+
+The ethics of undertaking such a study are personal until and unless another
+researcher-subject gets involved.
+At that point, those individuals will need to assert for themselves that this
+is definitely a process they want to undertake.
+
+That said, there are some personal ethical constraints to address.
+
+The study will leave little room for any other activity.
+It is likely that other past-times will need to be nearly eliminated
+to pursue the study in earnest.
+``Weekends'', as a concept, won't really exist in a study which is intended
+to operate continuously.
+
+Other factors which are declared monastic rules, but which seem natural
+while practicing deeper meditative states, may come into play.
+Monastics are celibate and do not listen to music.
+They are also not permitted to run.
+While I do intend to become a celibate, eliminate physical exercise,
+or stop listening to music, I am cognizant that these are risks of taking
+a meditation practice serious to this degree.
+
+\subsubsection{``Getting Paid to Meditate''}
+
+On the subject of \textit{societal} ethics, the topic of
+``getting paid to meditate'' has been one I'm acutely aware of, and one
+that other commentators on the study have mentioned as well.
+
+If the intention of the study were only to engage in the act of full-time
+meditation, I would ordain as a monk.
+At present, the system of rules laid out by most monasteries still reflect
+those of the Vinaya: a minimum period of 5 years is required to live and
+practice independently.
+\footnote{https://www.watpahnanachat.org/joining}
+Still, if I did not care to publish my investigations into Vipassana
+meditation, five years of life is a small price to pay for the
+independence to practice continuously.
+\cite{nanatusita2014patimokkha}
+
+However, the purpose of financing this study by sponsorship is to pay for
+its outputs, not its inputs.
+Sponsors who participate are justifiably expecting that the study will
+produce readable outcomes, of which they will be the first recipients.
+
+\subsection{Timeframes}
 
 \section{STRENGTHS AND WEAKNESSES OF THE STUDY}
+
+\section{Risks}
+
+* devolves into performance art
+* line between science / art is blurry
+* forced to take robes
+* financially unsustainable
+
 \section{PUBLIC HEALTH SIGNIFICANCE}

### 2024-06-27 the-proposal/clean.sh

@@ -1,0 +1,3 @@
+#!/usr/bin/env bash
+
+rm *.aux *.bbl *.lof *.out *.toc

### 2024-06-27 the-proposal/the-proposal.tex
timeframes, risks, etc... first draft done

@@
 
-\section{STRENGTHS AND WEAKNESSES OF THE STUDY}
+The nominal timeframe for the study will be throughout the course of the
+life of the initial researcher-subject (me), and any further subjects
+added to the study over time.
+
+Results of the study and experience reports will be sent to sponsors
+on a quarterly basis.
+Less frequent aggregate results will be published in the form of
+papers, articles, and books, with no fixed timeline.
+
+\section{Strengths and Weaknesses of the Study}
+
+The study will provide an in-depth examination of Vipassana meditation
+as no other prior study has been able to formulate of any meditation
+technique.
+The sponsorship model will prevent the study from getting mired in
+the examination of narrow outcomes.
+
+As mentioned previously, the study is not structured to provide
+significant statistical data out of which other researchers may draw
+data-oriented conclusions.
 
@@
 
-* devolves into performance art
-* line between science / art is blurry
-* forced to take robes
-* financially unsustainable
+There is a significant risk that such a broad study will devolve into
+something of a performative art piece.
+Since the writing produced by the study will be inherently subjective,
+it will represent a sort of literary art form.
+The line between Science and Art under these circumstances is blurry.
+Avoiding this degredation of outputs is a primary concern.
 
-\section{PUBLIC HEALTH SIGNIFICANCE}
-\section{BUDGET AND MOTIVATION}
-- resources required
+There is a more acute risk that, after years of heavy meditation practice,
+it will be natural to choose to renounce the householder life and take on
+the role of a monk entirely, regardless of what other difficulities that
+presents.
 
-\section{Conflict of Interest Statement?}
+There is a final acute risk of financial unsustainability.
+If the sponsorship model for the study cannot sustain even one person
+financially, it will not be possible to continue with the study
+indefinitely.
 
-\section{REFERENCES}
-\section{APPENDICES}
+\section{Public Health Significance}
+
+The significance of the study's outcomes for public health are parallel
+to the current accelerating adoption of Vipassana meditation as the
+primary meditation technique for many thousands, if not millions, of
+householder and monastic practitioners.
+
+If the outcomes of the study are resoundingly positive, it strongly
+implies that Vipassana meditation could be an effective tool in
+improving public health.
+However, if negative effects are detected throughout the course of the
+study, it could imply the opposite --- that we should be cautious in
+employing Vipassana meditation as a tool for public health improvements.
+
+\section{Budget}
+
+Initially, the resources required to see this study through to completion
+will be measured entirely by an individual's annual income.
+To reduce the temptation of returning to a career in software development
+(which I engaged in previously), that annual income would need to be
+comparable to that of an intermediate software developer.
+In Canada (where I currently reside), that figure is approximately
+\$150,000 - \$200,000 CAD per annum.
+
+As the study progresses, other resources will become a constraint, such as
+access to commercial EEG, ECG, and fMRI equipment.
+As such, the study will be run through Atapi Research Inc. to separate
+study finances from my personal finances.
+
+\section{Conflict of Interest Statement}
+
+There is an inherent conflict of interest produced by the fact that I have
+chosen Vipassana meditation as the single meditation technique I will use
+for the rest of my life.
+However, there is no way around this if the study is to focus on an
+individual who has made precisely such a choice.
+
+\section{Appendices}
 
@@
 As the study progresses, it is expected that the duration of sleep will lessen.
+Monthly and yearly schedules will include periods of total and partial seclusion,
+such as meditation courses (24 hours of meditation per day, less deep sleep cycles)
+and meditation retreats (16+ hours of meditation per day).
 
@@
 
-\section*{References}
+\newpage
+
+% \section*{References}
 

### 2024-06-28 the-proposal/build-pdf.sh
removing proposal from v4h, adding to atapi

@@ -1,26 +1,0 @@
-#!/bin/sh
-
-# pdflatex was used originally but cannot render Sanskrit without a lot of hoop-jumping
-# pdflatex vipassana-for-hackers.tex
-
-# xelatex is apparently less capable than pdflatex at font-rendering in general, but
-# is capable of rendering non-Latin fonts (like Hindi/Sanskrit) by default
-
-# Compile twice to resolve references.
-
-SCRIPTPATH="$( cd "$(dirname "$0")" ; pwd -P )"
-LIBPATH="$SCRIPTPATH/../lib"
-PATH=$PATH:$LIBPATH
-
-echo "Running bibliography pre-build. This takes a while..."
-xelatex the-proposal.tex > /dev/null 2>&1
-echo "Generating bibliography..."
-bibtex the-proposal.aux
-read -p "Press enter to continue"
-
-echo "Running PDF pre-build. This takes a while..."
-xelatex the-proposal.tex > /dev/null 2>&1
-echo "...done. Now running the PDF build."
-xelatex the-proposal.tex
-
-mv the-proposal.pdf ../output/the-proposal.pdf

### 2024-06-28 the-proposal/clean.sh

@@ -1,3 +1,0 @@
-#!/usr/bin/env bash
-
-rm *.aux *.bbl *.lof *.out *.toc
